### 2024-11-10 doc/sysdesc.tex
more

\documentclass{llncs}
\usepackage[show]{ed}
\usepackage{cleveref}
\usepackage{wrapfig}
\usepackage{xspace}
\usepackage{stex-logo}

\usepackage[style=alphabetic,hyperref=auto,defernumbers=true,backend=bibtex,firstinits=true,maxbibnames=9,maxcitenames=3,isbn=false]{biblatex}
\addbibresource{kwarcpubs.bib}
\addbibresource{extpubs.bib}
\addbibresource{kwarccrossrefs.bib}
\addbibresource{extcrossrefs.bib}

% I do not want to annotate just yet.

\newcommand\ALeA{\textsf{ALeA}\xspace}

\title{Bulk Semantic Annotation with an Partially Known Knowledge Base}
\author{Michael Kohlhase, Jan Frederik Schaefer}
\institute{Computer Science, FAU Erlangen N\"urnberg, Germany}
\begin{document}
\maketitle
\begin{abstract}
tbw
\end{abstract}

\section{Introduction}
Arguably, dealing with large document collections is one of the key factors in the
knowledge-driven society and economy. There are currently two main contenders for machine
support in this area. Symbolic/logic-based technologies and subsymbolic,
machine-learning-based AI, e.g. via LLMs or chatbots; they have complementary strengths
and challenges: Symbolic technologies offer precision and explainability out of the box,
but face scalability challenges, because the prerequisite background knowledge has to be
(manually) formalized. ML-based approaches can be trained on all data of the Internet, but
face challenges in precision and explanations. Aarne Ranta matches these profiles to the
notion of \textbf{producer tasks} -- i.e. tasks where precision is key, but limited
coverage ($10^{3\pm1}$ concepts) is ok; e.g. for multi-language/variant manuals for very
expensive tool machines -- and \textbf{consumer tasks}, where coverage is key and
precision secondary; e.g. for consumer-grade machine translation like \textsf{Google
  translate} in \cite{Ranta:atcp17}.

In this paper, we address tools that help attack the coverage problem in symbolic/semantic
approaches in practice. The producer task we use as a case study is that of adaptive
learning assistants for MINT subject, concretely the \ALeA system
\cite{BerBetChu:lssmkm23}, which uses \sTeX \cite{MueKo:sdstex22,URL:sTeX:github} -- a
variant of {\LaTeX} that allows to embed semantic annotations -- for a knowledge
representation and generates learner-adaptive learning objects instrumented with
learning-support interactions from that. The \ALeA ecosystem

\section{Conclusion}
\printbibliography
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2024-11-10 doc/sysdesc.tex
more

@@
 \newcommand\ALeA{\textsf{ALeA}\xspace}
+\newcommand\srf{\textsf{srify}\xspace}
 
@@
 knowledge-driven society and economy. There are currently two main contenders for machine
-support in this area. Symbolic/logic-based technologies and subsymbolic,
+support in this area. Symbolic/logic-based technologies and sub-symbolic,
 machine-learning-based AI, e.g. via LLMs or chatbots; they have complementary strengths
@@
 notion of \textbf{producer tasks} -- i.e. tasks where precision is key, but limited
-coverage ($10^{3\pm1}$ concepts) is ok; e.g. for multi-language/variant manuals for very
+coverage ($10^{3\pm1}$ concepts) is OK; e.g. for multi-language/variant manuals for very
 expensive tool machines -- and \textbf{consumer tasks}, where coverage is key and
@@
 approaches in practice. The producer task we use as a case study is that of adaptive
-learning assistants for MINT subject, concretely the \ALeA system
-\cite{BerBetChu:lssmkm23}, which uses \sTeX \cite{MueKo:sdstex22,URL:sTeX:github} -- a
+learning assistants for MINT subjects, concretely the \ALeA system
+\cite{BerBetChu:lssmkm23}, which uses \sTeX \cite{MueKo:sdstex22,sTeX:github:on} -- a
 variant of {\LaTeX} that allows to embed semantic annotations -- for a knowledge
 representation and generates learner-adaptive learning objects instrumented with
-learning-support interactions from that. The \ALeA ecosystem
+learning-support interactions from that. The \textbf{\sTeX/\ALeA content commons}
+\ednote{MK@MK: continue to describe the size and distribution, and authorship}
+
+We contend that while education -- especially higher education -- as a whole is a
+wide-coverage task for society education of individual learners in specific domains --
+especially, if we want to tailor the educational offerings to said individual/domain -- is
+a producer task, at least until ML-based methods reach the precision and explainability
+required by the ethics of teaching.
+
+\section{Semantic Authoring in a Content Common}
+
+The main practical problem of annotating presentational {\LaTeX} course materials into
+semantic documents that carry enough information to support meaningful learning support
+services is to reference
+
+\sTeX authoring is already supported by an VSCode IDE plugin \cite{sTeX-IDE:git} that
+analyzes annotations on the fly, displays the underlying reference semantics, reports
+errors and redundancies, and even offers a concept search interface.
 
 \section{Conclusion}
+
+In this system description we have presented the \srf tool, a command-line tool
+
 \printbibliography
@@
 %%% End:
+
+% LocalWords:  tbw Aarne Ranta

### 2024-11-10 doc/sysdesc.tex
more

@@
 \usepackage{xspace}
+\usepackage{graphicx}
 \usepackage{stex-logo}
+\usepackage{listings}
+\definecolor{codegray}{rgb}{0.9,0.9,0.9}
+\lstset{basicstyle=\sf,columns=fullflexible,backgroundcolor = \color{codegray}}
 
@@
 \newcommand\srf{\textsf{srify}\xspace}
+\def\llangle{\langle\kern-.2em\langle}
+\def\rrangle{\rangle\kern-.2em\rangle}
 
@@
 learning-support interactions from that. The \textbf{\sTeX/\ALeA content commons}
-\ednote{MK@MK: continue to describe the size and distribution, and authorship}
+\ednote{MK@MK: continue to describe the size and distribution, and
+  authorship}\ednote{explain the \textbf{domain model} (the SMGloM) and the
+  \textbf{formulation model} (cf. \cite{BerBetChu:lssmkm23}) in the commons and give their
+  releative sizes.}
 
@@
 semantic documents that carry enough information to support meaningful learning support
-services is to reference
+services is to add semantic references to technical terms from the domain jargon. Take for
+instance a learning object like the paragraph in \cref{fig:lo}, where we want to annotate
+the the term ``state space'' with the (semantic) concept of a ``state space of a search
+problem''. The latter can either be defined earlier in the course or in the domain mode
+(or both).
+
+\begin{figure}[ht]\centering
+  \includegraphics[width=8cm]{problem}
+  \caption{An annotated Learning Object from an AI Lecture}\label{fig:lo}
+\end{figure}
+
+Concretely this is about converting the {\LaTeX} string
+\begin{lstlisting}[language={[LaTeX]TeX}]
+\item \blue{Problem:} The state space has local minima, e.g. the board on the
+  right has $h=1$ but every successor has $h>1$. 
+\end{lstlisting}
+into the (partially annotated item:  
+\begin{lstlisting}[language={[LaTeX]TeX},morekeywords={sn,importmodule}]
+\importmodule[smglom/search]{mod?state-space}
+[...]  
+\item \blue{Problem:} The \sn{state space} has local minima, e.g. the
+  board on the right has $h=1$ but every successor has $h>1$. 
+\end{lstlisting}
+Concretely, the author has to remember the module for the term ``state space'' from
+\cref{fig:state-space}, the symbol name \lstinline|state space|, annotate it with the
+\sTeX macro \lstinline|\sn{...}|, and add the module import in the first line (unless it
+is already imported). In this case the symbol name happens to be the same as its English
+verbalization, otherwise we would have to use
+\lstinline[mathescape]|\sr{$\llangle symbol name\rrangle$}{state space}|.
+
+\begin{figure}[ht]\centering
+  \includegraphics[width=12cm]{state-space}
+  \caption{The definition of the term ``state space'' from the domain model}\label{fig:state-space}
+\end{figure}
+
+For this, the author has to be aware that the module from \cref{fig:state-space} exists,
+the symbol name, know the URL and manage reduncancy of the imports. At ca 150.000 words in
+a semester's words of lecture notes and ca. 10-15\% of technical terms a rather daunting
+task.
 

### 2024-11-10 doc/sysdesc.tex
more

@@
 
-For this, the author has to be aware that the module from \cref{fig:state-space} exists,
-the symbol name, know the URL and manage reduncancy of the imports. At ca 150.000 words in
-a semester's words of lecture notes and ca. 10-15\% of technical terms a rather daunting
-task.
+For this, the author has to be aware ``state space'' is a technical term, that the module
+from \cref{fig:state-space} exists, know the symbol name and its URL, and manage
+reduncancy of the imports. At ca 150.000 words in a semester's words of lecture notes and
+ca. 10-15\% of technical terms a rather daunting task.
 
@@
 analyzes annotations on the fly, displays the underlying reference semantics, reports
-errors and redundancies, and even offers a concept search interface.
+errors and redundancies, and even offers a concept search interface. But these only offer
+support already annotated terms, so they do not solve the annotation problem above.
+
+\section{The \srf System}
 

### 2024-11-11 doc/cicm25-sysdesc/sysdesc.tex
renaming

--- 2024-11-10 doc/sysdesc.tex
@@
 \begin{figure}[ht]\centering
-  \includegraphics[width=8cm]{problem}
+  \fbox{\includegraphics[width=8cm]{../img/problem}}
   \caption{An annotated Learning Object from an AI Lecture}\label{fig:lo}
@@
 \begin{figure}[ht]\centering
-  \includegraphics[width=12cm]{state-space}
+  \fbox{\includegraphics[width=12cm]{../img/state-space}}
   \caption{The definition of the term ``state space'' from the domain model}\label{fig:state-space}
@@
 
+The \srf system is a simple command-line tool that creates a database of
+symbol-verbalization pairs\ednote{explain that first}, analyzes \sTeX source files and
+then steps through all word occurrences in the document that share the stem with a
+verbalization in the database. For each such word, the user is presented with an
+annotation choice and interactions that allow to fine-tune the wordwise annotation
+workflow.
+
+As \srf addresses mostly power-users, the UI of \srf is command-line based, much like the
+early \textsf{ispell} system, see
+
 \section{Conclusion}

### 2024-11-11 stextools/srify/annotate_command.py
fix

@@ -6,6 +6,7 @@ import click
 from pylatexenc.latexwalker import LatexMacroNode, LatexSpecialsNode, LatexMathNode, LatexGroupNode, \
     LatexEnvironmentNode, LatexCommentNode, LatexCharsNode, LatexWalker
 
+from stextools.core.config import get_config
 from stextools.core.linker import Linker
 from stextools.core.macros import STEX_CONTEXT_DB
 from stextools.core.mathhub import make_filter_fun
@@ -297,7 +298,8 @@ class LookupCommand(Command, AnnotateMixin):
         AnnotateMixin.__init__(self, state, linker)
 
     def execute(self, *, state: State, call: str) -> list[CommandOutcome]:
-        fzf_path = shutil.which('fzfx')
+        fzf_path = shutil.which('fzf')
+        get_config().get('stextools', 'fzf_path', fallback=None)
         if fzf_path is None:
             print(click.style('fzf not found', fg='red'))
             print('Please install fzf to use this feature.')

### 2024-11-11 stextools/srify/controller.py
fix #46

@@ -261,9 +261,10 @@ class Controller:
         return command_collection.apply(state=self.state)
 
     def _get_current_command_collection(self) -> CommandCollection:
-        candidate_symbols = self.get_verb_trie(self.get_current_lang()).find_first_match(
+        match_info = self.get_verb_trie(self.get_current_lang()).find_first_match(
             string_to_stemmed_word_sequence_simplified(self.state.get_selected_text(), self.get_current_lang())
-        )[2]
+        )
+        candidate_symbols = match_info[2] if match_info is not None else []
         filter_fun = make_filter_fun(self.state.filter_pattern, self.state.ignore_pattern)
         candidate_symbols = [s for s in candidate_symbols if filter_fun(s.declaring_file.archive.name)]
         annotate_command = AnnotateCommand(
@@ -363,10 +364,12 @@ class Controller:
 def srify(files: list[str], filter: str, ignore: str):
     session_storage = SessionStorage()
     state = session_storage.get_session_dialog()
+    is_new = False
     if state is None:
         state = State(files=[Path(file) for file in files], filter_pattern=filter, ignore_pattern=ignore,
                       cursor=PositionCursor(file_index=0, offset=0))
-    controller = Controller(state, is_new=True)
+        is_new = True
+    controller = Controller(state, is_new=is_new)
     unfinished = controller.run()
     if unfinished:
         session_storage.store_session_dialog(state)

### 2024-11-12 doc/cicm25-sysdesc/sysdesc.tex
more

@@
 
-Concretely this is about converting the {\LaTeX} string
+Concretely this is about converting the {\LaTeX} string\ednote{add numbers, start with 3
+  so that corresponding lines with the next listing.}
 \begin{lstlisting}[language={[LaTeX]TeX}]
@@
 
-The \srf system is a simple command-line tool that creates a database of
-symbol-verbalization pairs\ednote{explain that first}, analyzes \sTeX source files and
-then steps through all word occurrences in the document that share the stem with a
-verbalization in the database. For each such word, the user is presented with an
+The \srf system (see \cite{stextools:git} is a simple command-line tool that creates a
+database of symbol-verbalization pairs\ednote{explain that first}, analyzes \sTeX source
+files and then steps through all word occurrences in the document that share the stem with
+a verbalization in the database. For each such word, the user is presented with an
 annotation choice and interactions that allow to fine-tune the wordwise annotation
@@
 In this system description we have presented the \srf tool, a command-line tool
+\ednote{continue}
+
+In an earlier attempt to support semantic annotation in IDEs, we tried named-entity
+recognition (NER) for classification of ``likely annotation candidate words''
+\cite{hutterer:msc23}, however this classification was not precise enough in the
+distinction in ``technical terms'' and ordinary english noun phrases and named entities --
+the relevant task for annotation, and so made it impractical, especially, since it could
+not enter the actual annotations and import directives automatically. But maybe using the
+NER-based approach, after the \srf annotation process might change the tradeoffs
+involved. In any case, the overall workflow suggested by the NER-based approach --
+especially after the verbalizations covered by the existing content commons\ednote{better
+  to speak of the domain model here?} -- is geared towards adding to the content commons.
+
+This is complementary to the \srf approach,
 

### 2024-11-13 experiment.py
first support for structures #38

@@ -1,34 +1,16 @@
-from pathlib import Path
+from stextools.core.cache import Cache
+import logging
 
-from lxml import etree
-from pylatexenc.latexwalker import LatexWalker
+from stextools.core.linker import Linker
 
-from stextools.macros import STEX_CONTEXT_DB
-from stextools.translation import stex_to_html, documents_to_html, html_to_stex, bergamot_html_translate
+logging.getLogger('pylatexenc.latexwalker').setLevel(logging.WARNING)
+# logging.getLogger('stextools.core.linker').setLevel(logging.FATAL)
+logging.basicConfig(level=logging.INFO)
 
-walker = LatexWalker('\\definiendum[x]{a}{b}', latex_context=STEX_CONTEXT_DB)
-print(walker.get_latex_nodes()[0])
-walker = LatexWalker('\\definiendum{a}{b}', latex_context=STEX_CONTEXT_DB)
-print(walker.get_latex_nodes()[0])
+Cache.clear = lambda: None  # type: ignore
 
+mh = Cache.get_mathhub(update_all=True)
 
-print('------------------')
-
-
-element = stex_to_html(Path('/home/jfs/git/gl.mathhub.info/smglom/cs/source/mod/cryptology.en.tex'))
-print(etree.tostring(element, pretty_print=True).decode())
+linker1 = Linker(mh)
 
 
-print('------------------')
-
-html = documents_to_html([Path('/home/jfs/git/gl.mathhub.info/smglom/cs/source/mod/cryptology.en.tex')])
-print(html)
-
-print(html_to_stex(html)[0][1])
-
-print('------------------')
-
-html2 = html.replace('\n', '<img data-replace="&#10;"/>')
-print(bergamot_html_translate(html2))
-
-print(html_to_stex(bergamot_html_translate(html2))[0][1])

### 2024-11-13 stextools/core/linker.py

@@ -1,13 +1,36 @@
 import logging
 from collections import defaultdict
-from collections.abc import Iterable
-from typing import Optional
+from collections.abc import Iterable, Callable
+from typing import Optional, Union
 
 from stextools.core.mathhub import MathHub
-from stextools.core.stexdoc import STeXDocument, Symbol, Verbalization
+from stextools.core.stexdoc import STeXDocument, Symbol, Verbalization, DocInfo, ModuleInfo
 from stextools.utils.intifier import Intifier
 
 logger = logging.getLogger(__name__)
+
+
+def top_sort(doc_deps: dict[int, Iterable[int]], error_doc_info: Callable[[int], str]) -> Iterable[int]:
+    full_processed: set[int] = set()
+    already_under_consideration: set[int] = set()
+    result: list[int] = []
+
+    def visit(doc: int):
+        if doc in full_processed:
+            return
+        if doc in already_under_consideration:
+            logger.error(f'Circular dependency detected (involving {error_doc_info(doc)})')
+            return
+        already_under_consideration.add(doc)
+        for dep in doc_deps[doc]:
+            visit(dep)
+        full_processed.add(doc)
+        result.append(doc)
+
+    for doc in list(doc_deps):
+        visit(doc)
+
+    return result
 
 
 class Linker:
@@ -33,6 +56,10 @@ class Linker:
     module_to_symbs: dict[int, set[int]]        # module -> set[symbol]
     symbol_to_module: dict[int, int]            # symbol -> module
     symbs_by_name: dict[str, set[tuple[int, int]]]   # name -> set[(module, symbol)]
+    structures_by_name: dict[str, set[int]]      # name -> set[structure (module)]
+    structure_info: dict[int, tuple[str, int, set[str], int]]  # structure -> (name, containing module, dependencies, start_pos)
+    structures_by_file: dict[int, list[int]]      # file -> set[structure (module)]
+    use_math_structs: dict[int, list[tuple[str, int, int]]]  # file -> set[(struct name, start, end)]
 
     # data from _transitive_imports
     transitive_imports: dict[int, set[int]]    # module -> set[module] (transitive closure of module imports, including self)
@@ -55,12 +82,15 @@ class Linker:
         b = time.time()
         self._compute_transitive_imports()
         c = time.time()
-        self._link_symbols()
+        self._link_structures()
         d = time.time()
+        self._link_symbols()
+        e = time.time()
[…]

### 2024-11-13 stextools/core/macros.py

@@ -59,6 +59,8 @@ STEX_ENV_SPECS: list = [
     std_environment('sparagraph', '['),
     std_environment('assignment', '['),
     std_environment('nparagraph', '['),
+    std_environment('mathstructure', '{['),
+    std_environment('extstructure', '*{[{'),
     std_environment('scb', '['),
     std_environment('smcb', '['),
 ]

### 2024-11-13 stextools/core/simple_api.py

@@ -63,6 +63,19 @@ class SimpleModule:
         self._linker = linker
 
     @property
+    def is_structure(self) -> bool:
+        return self._module.is_structure
+
+    @property
+    def struct_name(self) -> Optional[str]:
+        return self._module.struct_name
+
+    def get_structures_containing_module(self) -> Optional['SimpleModule']:
+        if self._module_int not in self._linker.structure_info:
+            return None
+        return SimpleModule(self._linker.structure_info[self._module_int][1], self._linker)
+
+    @property
     def path_rel_to_archive(self) -> str:
         file_path = self.file.get_relative_path(drop_tex_extension=True, drop_lang_tag=True)
         module_path = self.path_in_doc

### 2024-11-13 stextools/core/stexdoc.py

@@ -25,6 +25,7 @@ class DependencyPropFlag:  # note: enum.IntFlag is really slow -- maybe this is 
     IS_USE = 2
     TARGET_NO_TEX = 4
     IS_INPUT = 8
+    IS_USE_STRUCT = 16
 
 
 # NOTE: According to my benchmark, using `slots=True` noticably slows down pickling/unpickling
@@ -48,6 +49,11 @@ class Dependency:
     def is_use(self) -> bool:
         """Symbols from the dependency are not exported"""
         return bool(self.flags & DependencyPropFlag.IS_USE)
+
+    @property
+    def is_use_struct(self) -> bool:
+        """ \\usestructure """
+        return bool(self.flags & DependencyPropFlag.IS_USE_STRUCT)
 
     @property
     def target_no_tex(self) -> bool:
@@ -138,10 +144,13 @@ class ModuleInfo:
     """Basic information about a document (dependencies, created symbols, verbalizations, etc.)"""
     name: str
     valid_range: tuple[int, int]
+    struct_name: Optional[str] = None
     dependencies: list[Dependency] = dataclasses.field(default_factory=list)
     symbols: list[Symbol] = dataclasses.field(default_factory=list)
     # submodules
     modules: list[ModuleInfo] = dataclasses.field(default_factory=list)
+    is_structure: bool = False
+    struct_deps: Optional[list[str]] = None
 
     def flattened_dependencies(self) -> Iterator[Dependency]:
         yield from self.dependencies
@@ -194,6 +203,7 @@ class DependencyProducer:
     is_use: bool = False
     target_no_tex: bool = False
     is_input: bool = False
+    is_use_struct: bool = False
 
     def produce(self, node: LatexMacroNode, from_archive: str, from_subdir: str, mh: MathHub,
                 valid_range: tuple[int, int], lang: str = '*') -> Optional[Dependency]:
@@ -206,6 +216,8 @@ class DependencyProducer:
             flag |= DependencyPropFlag.TARGET_NO_TEX
         if self.is_input:
             flag |= DependencyPropFlag.IS_INPUT
+        if self.is_use_struct:
+            flag |= DependencyPropFlag.IS_USE_STRUCT
 
         # STEP 1: Determine the target archive
         target_archive: Optional[str] = None
@@ -279,6 +291,8 @@ DEPENDENCY_PRODUCERS = [
     DependencyProducer('requiremodule', references_module=True, opt_param_is_archive=True, is_use=True),
[…]

### 2024-11-13 stextools/srify/annotate_command.py

@@ -65,7 +65,13 @@ class AnnotateMixin:
         args = ''
         if symbol.declaring_file.archive != file.archive:
             args += f'[{symbol.declaring_file.archive.name}]'
-        args += f'{{{symbol.declaring_module.path_rel_to_archive}}}'
+        structure_use: Optional[str] = None
+        if cont_module := symbol.declaring_module.get_structures_containing_module():
+            args += f'{{{cont_module.path_rel_to_archive}}}'
+            structure_use = f'\\usestructure{{{symbol.declaring_module.struct_name}}}'
+        else:
+            args += f'{{{symbol.declaring_module.path_rel_to_archive}}}'
+
 
         file_text = self.state.get_current_file_text()
 
@@ -81,12 +87,17 @@ class AnnotateMixin:
         explain_loc = lambda loc: ' after ' + latex_format(
             '\\begin{' + loc + '}') if loc else ' at the beginning of the file'
 
+        def _get_use_struct(pos: int) -> str:
+            if structure_use:
+                return _get_indentation(pos) + structure_use
+            return ''
+
         commands = [
             ImportCommand(
                 'u', 'semodule' + explain_loc(import_locations[3]),
                      'Inserts \\usemodule' + explain_loc(import_locations[3]),
                 ImportInsertionOutcome(
-                    _get_indentation(import_locations[0]) + f'\\usemodule{args}',
+                    _get_indentation(import_locations[0]) + f'\\usemodule{args}' + _get_use_struct(import_locations[0]),
                     import_locations[0]
                 )
             ),
@@ -99,7 +110,7 @@ class AnnotateMixin:
                 if import_locations[4] else
                 'Inserts \\usemodule at the top of the document (in this case same as [u])',
                 ImportInsertionOutcome(
-                    _get_indentation(import_locations[1]) + f'\\usemodule{args}',
+                    _get_indentation(import_locations[1]) + f'\\usemodule{args}' + _get_use_struct(import_locations[1]),
                     import_locations[1]
                 )
             )
@@ -110,7 +121,7 @@ class AnnotateMixin:
                 'i', 'mportmodule',
                 'Inserts \\importmodule',
                 ImportInsertionOutcome(
-                    _get_indentation(import_locations[2]) + f'\\importmodule{args}',
+                    _get_indentation(import_locations[2]) + f'\\importmodule{args}' + _get_use_struct(import_locations[2]),
                     import_locations[2]
                 )
             ))

### 2024-11-13 experiment.py
more debug info #48

@@ -12,5 +12,3 @@ Cache.clear = lambda: None  # type: ignore
 mh = Cache.get_mathhub(update_all=True)
 
 linker1 = Linker(mh)
-
-

### 2024-11-13 stextools/srify/selection.py

@@ -13,6 +13,7 @@ from stextools.srify.state import PositionCursor
 from stextools.srify.state import SelectionCursor, State
 from stextools.srify.stemming import string_to_stemmed_word_sequence, string_to_stemmed_word_sequence_simplified
 from stextools.utils.linked_str import LinkedStr, string_to_lstr
+from stextools.utils.ui import standard_header
 
 # By default, macros are not searched for potential annotations.
 # This is a list of exceptions to this rule.
@@ -49,6 +50,16 @@ def get_linked_strings(latex_text: str) -> list[LinkedStr]:
             if node.nodeType() == LatexMacroNode:
                 if node.macroname in MACRO_RECURSION_RULES:
                     for arg_idx in MACRO_RECURSION_RULES[node.macroname]:
+                        if arg_idx >= len(node.nodeargs):
+                            click.clear()
+                            standard_header('Error', bg='red')
+                            print(f"Macro {node.macroname} does not have argument {arg_idx}")
+                            print('Context:')
+                            print(latex_text[max(node.pos - 100, 0):min(len(latex_text) - 1, node.pos + 300]))
+                            print()
+                            print('Please report this error')
+                            click.pause()
+                            continue
                         _recurse([node.nodeargs[arg_idx]])
             elif node.nodeType() == LatexEnvironmentNode:
                 if node.envname in ENVIRONMENT_RECURSION_RULES:

### 2024-11-13 stextools/srify/selection.py
typo

@@ -55,7 +55,7 @@ def get_linked_strings(latex_text: str) -> list[LinkedStr]:
                             standard_header('Error', bg='red')
                             print(f"Macro {node.macroname} does not have argument {arg_idx}")
                             print('Context:')
-                            print(latex_text[max(node.pos - 100, 0):min(len(latex_text) - 1, node.pos + 300]))
+                            print(latex_text[max(node.pos - 100, 0):min(len(latex_text) - 1, node.pos + 300)])
                             print()
                             print('Please report this error')
                             click.pause()

### 2024-11-13 doc/cicm25-sysdesc/sysdesc.tex
draining

@@
 
+\section{Practical Evaluation}
+
+The step-through
+
 \section{Conclusion}
@@
 involved. In any case, the overall workflow suggested by the NER-based approach --
-especially after the verbalizations covered by the existing content commons\ednote{better
-  to speak of the domain model here?} -- is geared towards adding to the content commons.
-
-This is complementary to the \srf approach,
+especially after the verbalizations covered by the existing domain model\ednote{introduce
+  in the definition above that the domain model is all definitions, not only in the
+  smglom?} -- is geared towards adding modules and symbols -- the ones discovered by NER
+-- to the domain model.
 

### 2024-11-13 doc/cicm25-sysdesc/sysdesc.tex
more

@@
 As \srf addresses mostly power-users, the UI of \srf is command-line based, much like the
-early \textsf{ispell} system, see
+early \textsf{ispell} system, see\ednote{continue}
 
@@
 
-The step-through
+In our experience, the step-through workflow for annotating term references for symbols
+from the domain model is almost an order of magnitude more efficient than writing the
+annotations and imports by hand, even for annotators who are familiar with the domain
+model. For annotators unfamiliar with the domain model, the unassisted annotation task is
+almost infeasible, and the average unfamiliarity naturally grows with the domain
+model. The symbol disambiguation process -- on average a word induces $n$\ednote{MK@FS: we
+  should try to measure this: for every verbalization (i.e. symbol/word pair) we should
+  compute the ratio of those words with the same stem.} choices is still manageable,
+requires considerable concentration and domain knowledge, but little knowledge of the
+domain model flexiformalization.
+
+The command-line interface is simple and responsive and gives all the necessary
+information in a single glance if underlying shell area exceeds ca. $80\times 35$
+characters. It is very much geared towards annotating existing documents with respect to a
+relatively complete -- pre-existing -- domain model. Otherwise we have to skip too many
+terms that should ultimately be annotated and annotation efficiency suffers. This is
+currently the case for all non-English languages in the \sTeX commons. Coverage in German
+is only about half of that of English and we can already see the practical effects. We
+have also experimented with a Slovene introductory math book, and it seems clear that
+apart from having to introduce a Slovene stemmer\ednote{MK@FS: there is one for python at
+  \url{https://repo.ijs.si/pboskoski/slo_stemmer}}, we would need to manually annotate all
+definienda in the book before we can harvest the symbol/verbalization pairs which are a
+prerequisite for annotation.
+
+The current interface is not well-suited for on-the-fly annotation while editing while
+authoring. For that, the underlying information (the symbol/verbalization pairs harvested
+from the domain model) can be integrated into any IDE. In fact we plan to do this for the
+next version of the sTeX plugin for VSCode \cite{sTeX-IDE:git}.
 

### 2024-11-13 doc/cicm25-sysdesc/sysdesc.tex
more

@@
 characters. It is very much geared towards annotating existing documents with respect to a
-relatively complete -- pre-existing -- domain model. Otherwise we have to skip too many
-terms that should ultimately be annotated and annotation efficiency suffers. This is
-currently the case for all non-English languages in the \sTeX commons. Coverage in German
-is only about half of that of English and we can already see the practical effects. We
-have also experimented with a Slovene introductory math book, and it seems clear that
-apart from having to introduce a Slovene stemmer\ednote{MK@FS: there is one for python at
+relatively complete -- pre-existing -- domain model, and it seems unlikely that a more
+sophisticated UI would add value for this use-case.
+
+For less complete domain models we have to skip too many terms that should ultimately be
+annotated and annotation efficiency suffers. This is currently the case for all
+non-English languages in the \sTeX corpus we work with. Coverage in German is only about
+half of that of English and we can already see the practical effects. We have also
+experimented with a Slovene introductory math book, and it seems clear that apart from
+having to introduce a Slovene stemmer\ednote{MK@FS: there is one for python at
   \url{https://repo.ijs.si/pboskoski/slo_stemmer}}, we would need to manually annotate all

### 2024-11-14 doc/cicm25-sysdesc/sysdesc.tex
more text

@@
 \usepackage[show]{ed}
-\usepackage{cleveref}
 \usepackage{wrapfig}
@@
 \usepackage{stex-logo}
+%\usepackage{lststex} debug that the gray is not solid
 \usepackage{listings}
@@
 \lstset{basicstyle=\sf,columns=fullflexible,backgroundcolor = \color{codegray}}
-
+\lstset{numberstyle=\tiny}
+\lstset{language={[LaTeX]TeX}}
 \usepackage[style=alphabetic,hyperref=auto,defernumbers=true,backend=bibtex,firstinits=true,maxbibnames=9,maxcitenames=3,isbn=false]{biblatex}
@@
 \addbibresource{extcrossrefs.bib}
+\usepackage[noabbrev]{cleveref}
 
@@
   \textbf{formulation model} (cf. \cite{BerBetChu:lssmkm23}) in the commons and give their
-  releative sizes.}
+  relative sizes.}
 
@@
   so that corresponding lines with the next listing.}
-\begin{lstlisting}[language={[LaTeX]TeX}]
+\begin{lstlisting}[numbers=left,firstnumber=3,
+caption=The {\LaTeX} sources of \cref{fig:lo},label=lst:los]
 \item \blue{Problem:} The state space has local minima, e.g. the board on the
@@
 \end{lstlisting}
-into the (partially annotated item:
-\begin{lstlisting}[language={[LaTeX]TeX},morekeywords={sn,importmodule}]
+into the (partially) annotated item:
+\begin{lstlisting}[morekeywords={sn},numbers=left,
+caption=Annotating ``state space'' in \cref{lst:los},label=lst:losa]
 \importmodule[smglom/search]{mod?state-space}
@@
 
+The annotation is legitimized by a definition in the domain model module shown in
+\cref{fig:state-space}. The \lstinline|\definame| in line 3 introduces a
+\textbf{verbalization} ``state space'' for the symbol \lstinline|state-space?state space|
+introduced be \lstinline|\symdecl*| declaration in line 3.
+
 \begin{figure}[ht]\centering
   \fbox{\includegraphics[width=12cm]{../img/state-space}}
+\begin{lstlisting}[morekeywords={definame,symdecl},numbers=left]
+\begin{smodule}[title=State Space]{state-space}
+[... some imports ...]
+\symdecl*{state space}
+\begin{sdefinition}
+  [...] we call the \sn{graph} [...] the \definame{state space} of $\Pi$.
+\end{sdefinition}
+\end{smodule}
+\end{lstlisting}
   \caption{The definition of the term ``state space'' from the domain model}\label{fig:state-space}
@@
 
-For this, the author has to be aware ``state space'' is a technical term, that the module
-from \cref{fig:state-space} exists, know the symbol name and its URL, and manage
-reduncancy of the imports. At ca 150.000 words in a semester's words of lecture notes and
-ca. 10-15\% of technical terms a rather daunting task.
+For the annotation \lstinline|\sn{state-space}|, the author has to be aware ``state
+space'' is a technical term, that the module from \cref{fig:state-space} exists, know the
+symbol name and its URL, and manage redundancy of the imports -- i.e. only adding the
+\lstinline|\importmodule| directive if it is not already (recursively) implied and
+possibly removing directives that become redundant by the new one. At ca 150.000 words in
+a semester's words of lecture notes and ca. 10-15\% of technical terms a rather daunting
+task.
 
@@
 The \srf system (see \cite{stextools:git} is a simple command-line tool that creates a
-database of symbol-verbalization pairs\ednote{explain that first}, analyzes \sTeX source
-files and then steps through all word occurrences in the document that share the stem with
-a verbalization in the database. For each such word, the user is presented with an
-annotation choice and interactions that allow to fine-tune the wordwise annotation
-workflow.
+database of symbol-verbalization pairs, analyzes \sTeX source files and then steps through
+all word occurrences in the document that share the stem with a verbalization in the
+database. For each such word, the user is presented with an annotation choice and
+interactions that allow to fine-tune the wordwise annotation workflow. \Cref{fig:srify}
+shows a typical situation.\ednote{MK: I like the example less and less! We need an example
+  that has at least one homostem (another verbalization that has the same stem; introduce
+  the concept her and use it throughout) so that we can show the choice mechanism, and
+  where the verbalization is different from the symbol name. Having a two-word compound is
+  good though.}
 
-As \srf addresses mostly power-users, the UI of \srf is command-line based, much like the
-early \textsf{ispell} system, see\ednote{continue}
+\begin{figure}[ht]
+  \fbox{\includegraphics[width=12cm]{../img/srify}}
+  \caption{\srf in Action: Annotating the term ``state space''.}\label{fig:srify}
+\end{figure}
+
+Here the annotator can choose an annotation by typing the corresponding choice
+number. Here, the module \lstinline|state-space| is already imported (see the little green
+check-mark); otherwise \srf would offer the annotator the choice to import it -- if we are
+in a module context -- or to add a \lstinline|\usemodule| directive and where (in the
+local environment or top-level). Alternatively, typing \lstinline|s| skips this word,
+\lstinline|S| skips and adds a comment to always skip this word, and -- if \srf first
+suggests the term `state` \lstinline|n| extends it to the compound ``state space'' by
+including the next word.
+
+\srf is usually called on a set of files -- e.g. in a directory or math archive -- which
+together form a \textbf{session}, which can be interrupted and resumed without having to
+re-do all the \lstinline|s| skips. Alternatively, \srf can be called e.g. on the top-level
+lecture notes; then the session consists of all included files. Other productivity
+features include a focus mode that can be used to propagate a particular choice to the
+rest of the file, session, or the whole \sTeX/\ALeA corpus before resuming regular
+annotation This reduces the cognitive effort from choosing a symbol from a list to
+deciding whether a particular choice is applicable in the current document context. Of
+course the include management still needs to be done locally.
 
@@
 In this system description we have presented the \srf tool, a command-line tool
-\ednote{continue}
+\ednote{continue} \srf is open source\ednote{MK@FS: do we have an explicit license yet in
+  the repository.}; the source and documentation are available from \cite{stextools:git}.
 
@@
 \cite{hutterer:msc23}, however this classification was not precise enough in the
-distinction in ``technical terms'' and ordinary english noun phrases and named entities --
+distinction in ``technical terms'' and ordinary English noun phrases and named entities --
 the relevant task for annotation, and so made it impractical, especially, since it could
 not enter the actual annotations and import directives automatically. But maybe using the
-NER-based approach, after the \srf annotation process might change the tradeoffs
+NER-based approach, after the \srf annotation process might change the trade-offs
 involved. In any case, the overall workflow suggested by the NER-based approach --
@@
 
-% LocalWords:  tbw Aarne Ranta
+% LocalWords:  tbw Aarne Ranta homostem pre

### 2024-11-14 stextools/srify/selection.py
improve skipping

@@ -206,7 +206,7 @@ class VerbTrie:
                         original_word = original_string[word_lstrs[j].get_start_ref()-shift:word_lstrs[j].get_end_ref()-shift]
                         if IgnoreList.contains(lang=self.lang, word=original_word):
                             skip = True
-                        if srskipped is not None and original_word in srskipped.skipped_literal:
+                        if srskipped is not None and srskipped.should_skip_literal(original_word):
                             skip = True
                     if srskipped and ' '.join(words_stemmed[match_start:j + 1]) in srskipped.skipped_stems:
                         skip = True

### 2024-11-14 stextools/srify/skip_and_ignore.py

@@ -1,5 +1,6 @@
 import dataclasses
 import itertools
+import re
 from pathlib import Path
 
 from stextools.core.linker import Linker
@@ -162,9 +163,13 @@ class SrSkipped:
                 for e in line[len('% srskip '):].split(','):
                     e = e.strip()
                     if e.startswith('s:'):
+                        if e[2:] in self.skipped_stems or not e[2:]:
+                            continue
                         self.skipped_stems_ordered.append(e[2:])
                         self.skipped_stems.add(e[2:])
                     elif e.startswith('l:'):
+                        if e[2:] in self.skipped_literal or not e[2:]:
+                            continue
                         self.skipped_literal_ordered.append(e[2:])
                         self.skipped_literal.add(e[2:])
                     else:   # legacy
@@ -176,8 +181,12 @@ class SrSkipped:
         self.skipped_stems.add(stem)
 
     def add_literal(self, literal: str):
+        literal = re.sub(r'\s+', ' ', literal)
         self.skipped_literal_ordered.append(literal)
         self.skipped_literal.add(literal)
+
+    def should_skip_literal(self, literal: str) -> bool:
+        return re.sub(r'\s+', ' ', literal) in self.skipped_literal
 
     def to_new_text(self) -> str:
         have_added = False

### 2024-11-14 stextools/srify/skip_and_ignore.py
fix #50

@@ -173,6 +173,8 @@ class SrSkipped:
                         self.skipped_literal_ordered.append(e[2:])
                         self.skipped_literal.add(e[2:])
                     else:   # legacy
+                        if e in self.skipped_stems or not e:
+                            continue
                         self.skipped_stems_ordered.append(e)
                         self.skipped_stems.add(e)
 

### 2024-11-14 stextools/srify/annotate_command.py
implemented [s!]

@@ -72,7 +72,6 @@ class AnnotateMixin:
         else:
             args += f'{{{symbol.declaring_module.path_rel_to_archive}}}'
 
-
         file_text = self.state.get_current_file_text()
 
         def _get_indentation(pos: int) -> str:
@@ -121,7 +120,8 @@ class AnnotateMixin:
                 'i', 'mportmodule',
                 'Inserts \\importmodule',
                 ImportInsertionOutcome(
-                    _get_indentation(import_locations[2]) + f'\\importmodule{args}' + _get_use_struct(import_locations[2]),
+                    _get_indentation(import_locations[2]) + f'\\importmodule{args}' + _get_use_struct(
+                        import_locations[2]),
                     import_locations[2]
                 )
             ))
@@ -238,7 +238,6 @@ class AnnotateMixin:
         return outcomes
 
 
-
 class AnnotateCommand(Command, AnnotateMixin):
     """ Has to be re-instantiated for each state! """
 
@@ -289,8 +288,8 @@ def symbol_display(file: SimpleFile, symbol: SimpleSymbol, state: State, style: 
             symb_path[0] + '?' +
             (
                 (
-                    click.style(symb_path[1], bg=color('bright_cyan', (180, 180, 255))) + '?' +
-                    click.style(symb_path[2], bg=color('bright_green', (180, 255, 180)))
+                        click.style(symb_path[1], bg=color('bright_cyan', (180, 180, 255))) + '?' +
+                        click.style(symb_path[2], bg=color('bright_green', (180, 255, 180)))
                 )
                 if style else (symb_path[1] + '?' + symb_path[2])
             )
@@ -305,7 +304,7 @@ class LookupCommand(Command, AnnotateMixin):
             pattern_regex='^l$',
             description_short='ookup a symbol',
             description_long='Look up a symbol for annotation')
-        )
+                         )
         AnnotateMixin.__init__(self, state, linker)
 
     def execute(self, *, state: State, call: str) -> list[CommandOutcome]:

### 2024-11-14 stextools/srify/selection.py

@@ -1,3 +1,5 @@
+import re
+from pathlib import Path
 from typing import Optional
 
 import click
@@ -6,7 +8,7 @@ from pylatexenc.latexwalker import LatexWalker, LatexMacroNode, LatexMathNode, L
 
 from stextools.core.linker import Linker
 from stextools.core.macros import STEX_CONTEXT_DB
-from stextools.core.simple_api import get_symbols, SimpleSymbol
+from stextools.core.simple_api import get_symbols, SimpleSymbol, SimpleFile
 from stextools.srify.commands import Command, CommandInfo, CommandOutcome, SetNewCursor
 from stextools.srify.skip_and_ignore import IgnoreList, SrSkipped
 from stextools.srify.state import PositionCursor
@@ -184,9 +186,12 @@ class VerbTrie:
     def find_first_match(
             self,
             words_stemmed: list[str],
-            # words are only ignored if the next two arguments are provided
+            # words are only ignored if the next two arguments are provided (necessary for skipping literal (i.e. unstemmed) words)
             word_lstrs: Optional[list[LinkedStr]] = None,
             original_string: Optional[str] = None,
+            # the temporarily skipped words are only considered if the next two arguments are provided
+            state: Optional[State] = None,
+            file: Optional[Path] = None,
             shift: Optional[int] = None,
             srskipped: Optional[SrSkipped] = None,
     ) -> Optional[tuple[int, int, list[SimpleSymbol]]]:
@@ -204,12 +209,28 @@ class VerbTrie:
                     skip = False
                     if word_lstrs is not None and original_string is not None:
                         original_word = original_string[word_lstrs[j].get_start_ref()-shift:word_lstrs[j].get_end_ref()-shift]
+                        original_word = re.sub(r'\s+', ' ', original_word)
                         if IgnoreList.contains(lang=self.lang, word=original_word):
                             skip = True
                         if srskipped is not None and srskipped.should_skip_literal(original_word):
                             skip = True
-                    if srskipped and ' '.join(words_stemmed[match_start:j + 1]) in srskipped.skipped_stems:
+                        if state is not None:
+                            if file is not None:
+                                if original_word in state.skip_literal_by_file.get(file, set()):
+                                    skip = True
+                            if original_word in state.skip_literal_all_session.get(self.lang, set()):
+                                skip = True
+
+                    stemmed_word = ' '.join(words_stemmed[match_start:j + 1])
+                    if srskipped and stemmed_word in srskipped.skipped_stems:
                         skip = True
[…]

### 2024-11-14 stextools/srify/skip_and_ignore.py

@@ -4,7 +4,8 @@ import re
 from pathlib import Path
 
 from stextools.core.linker import Linker
-from stextools.srify.commands import Command, CommandInfo, CommandOutcome, SetNewCursor, TextRewriteOutcome
+from stextools.srify.commands import Command, CommandInfo, CommandOutcome, SetNewCursor, TextRewriteOutcome, \
+    StateSkipOutcome
 from stextools.srify.state import State, SelectionCursor, PositionCursor
 from stextools.srify.stemming import string_to_stemmed_word_sequence_simplified
 
@@ -23,6 +24,27 @@ class SkipOnceCommand(Command):
         return [SetNewCursor(PositionCursor(state.cursor.file_index, state.cursor.selection_start + 1))]
 
 
+class SkipUntilFileEnd(Command):
+    def __init__(self):
+        super().__init__(CommandInfo(
+            pattern_presentation='s!',
+            pattern_regex='^s!$',
+            description_short='kip until file end',
+            description_long='Do not propose any occurrences of the selected phrase in this file anymore (until end of session).')
+        )
+
+    def execute(self, *, state: State, call: str) -> list[CommandOutcome]:
+        assert isinstance(state.cursor, SelectionCursor)
+        return [
+            StateSkipOutcome(
+                word=state.get_current_file_text()[state.cursor.selection_start:state.cursor.selection_end],
+                is_stem=False,
+                session_wide=False
+            ),
+            SetNewCursor(PositionCursor(state.cursor.file_index, state.cursor.selection_start)),
+        ]
+
+
 class AddWordToSrSkip(Command):
     def __init__(self):
         super().__init__(CommandInfo(
@@ -38,7 +60,7 @@ class AddWordToSrSkip(Command):
         srskipped = SrSkipped(state.files[state.cursor.file_index].read_text())
         srskipped.add_literal(state.get_selected_text())
         return [
-            TextRewriteOutcome(srskipped.to_new_text()),
+            TextRewriteOutcome(srskipped.to_new_text(), requires_reparse=False),
             # note: this is risky (position would change if there are, for weird reasons, early % srskip comments)
             SetNewCursor(PositionCursor(state.cursor.file_index, state.cursor.selection_start + 1))
         ]
@@ -191,7 +213,6 @@ class SrSkipped:
         return re.sub(r'\s+', ' ', literal) in self.skipped_literal
 
     def to_new_text(self) -> str:
-        have_added = False
[…]

### 2024-11-28 stextools/__main__.py
remove deprecated srify (aliasing did not actually work as intended)

@@ -80,6 +80,7 @@ def translate(path):
     print(translate(Path(path)))
 
 
+@cli.command(name='snify', help='\\sn-ify sTeX documents')
 @click.argument('files', nargs=-1)
 @click.option('--filter',
               default=lambda: get_config().get('stextools.snify', 'filter', fallback=None),
@@ -91,9 +92,6 @@ def snify_actual(files, filter, ignore):
     from stextools.snify.controller import snify
     snify(files, filter, ignore)
 
-
-cli.command(name='srify', help='\\sr-ify sTeX documents (deprecated, use snify instead)')(snify_actual)
-cli.command(name='snify', help='\\sn-ify sTeX documents')(snify_actual)
 
 if __name__ == '__main__':
     cli()

### 2024-11-28 stextools/core/config.py
refactoring

@@ -1,5 +1,4 @@
 import configparser
-import dataclasses
 import functools
 import logging
 from pathlib import Path

### 2024-11-28 stextools/core/linker.py

@@ -130,19 +130,19 @@ class Linker:
                 dep_doc_int = _doc_intify(dep_doc)
                 if dep.is_use and dep_mod is not None:
                     _available_module_ranges[int_doc].add((_mod_intify((dep_doc_int, dep_mod.name)),
-                                                                dep.valid_range[0], dep.valid_range[1]))
+                                                           dep.scope[0], dep.scope[1]))
                 else:
                     _file_import_graph[int_doc].add(dep_doc_int)
                     if int_source_mod is not None and dep_mod is not None:
                         mod_int = _mod_intify((dep_doc_int, dep_mod.name))
                         _module_import_graph[int_source_mod].add(mod_int)
-                        _available_module_ranges[int_doc].add((mod_int, dep.valid_range[0], dep.valid_range[1]))
+                        _available_module_ranges[int_doc].add((mod_int, dep.scope[0], dep.scope[1]))
 
             for dep in doc_info.dependencies:
                 if dep.is_use_struct:
                     # will be processed later in _link_structures as it requires the import graph
                     assert dep.module_name is not None
-                    self.use_math_structs[int_doc].append((dep.module_name, dep.valid_range[0], dep.valid_range[1]))
+                    self.use_math_structs[int_doc].append((dep.module_name, dep.scope[0], dep.scope[1]))
                 else:
                     process_dep(dep, doc)
 

### 2024-11-28 stextools/core/stexdoc.py

@@ -38,7 +38,7 @@ class Dependency:
     file_hint: Optional[str]
     module_name: Optional[str]
     flags: int     # DependencyPropFlag
-    valid_range: tuple[int, int]  # range of the document where the dependency is valid
+    scope: tuple[int, int]  # range of the document where the dependency is valid
     intro_range: Optional[tuple[int, int]] = None  # range of the document where the dependency is introduced
     file_may_be_relative: bool = False
 
@@ -236,7 +236,7 @@ class DependencyProducer:
     is_use_struct: bool = False
 
     def produce(self, node: LatexMacroNode, from_archive: str, from_subdir: str, mh: MathHub,
-                valid_range: tuple[int, int], lang: str = '*') -> Optional[Dependency]:
+                scope: tuple[int, int], lang: str = '*') -> Optional[Dependency]:
         flag = 0   # DependencyPropFlag(0)
         if self.is_lib:
             flag |= DependencyPropFlag.IS_LIB
@@ -272,7 +272,7 @@ class DependencyProducer:
         if archive is None:    # not locally installed, but we still want to store a dependency
             return Dependency(archive=target_archive or from_archive, file_hint=None,
                               file_may_be_relative=True, module_name=None,
-                              flags=int(flag), valid_range=valid_range, intro_range=intro_range)
+                              flags=int(flag), scope=scope, intro_range=intro_range)
 
         if self.references_module:
             if '?' in main_arg:
@@ -282,16 +282,16 @@ class DependencyProducer:
                 module_name = main_arg.split('/')[-1]
 
             return Dependency(archive.get_archive_name(), file_hint=path, file_may_be_relative=True,
-                              module_name=module_name, flags=int(flag), valid_range=valid_range,
+                              module_name=module_name, flags=int(flag), scope=scope,
                               intro_range=intro_range)
         else:
             if self.target_no_tex:
                 # TODO: determine file (though we don't really care about it, to be honest)
                 return Dependency(archive.get_archive_name(), None, module_name=None,
-                                  flags=int(flag), valid_range=valid_range, intro_range=intro_range)
+                                  flags=int(flag), scope=scope, intro_range=intro_range)
             else:
                 return Dependency(archive.get_archive_name(), file_hint=main_arg, file_may_be_relative=True,
-                                  module_name=None, flags=int(flag), valid_range=valid_range, intro_range=intro_range)
[…]

### 2024-11-28 stextools/snify/annotate_command.py
fix #35

@@ -1,6 +1,7 @@
+import dataclasses
 import shutil
 import subprocess
-from typing import Optional, Union
+from typing import Optional, Union, Any
 
 import click
 from pylatexenc.latexwalker import LatexMacroNode, LatexSpecialsNode, LatexMathNode, LatexGroupNode, \
@@ -10,9 +11,9 @@ from stextools.core.config import get_config
 from stextools.core.linker import Linker
 from stextools.core.macros import STEX_CONTEXT_DB
 from stextools.core.mathhub import make_filter_fun
-from stextools.core.simple_api import SimpleSymbol, get_symbols, SimpleFile
+from stextools.core.simple_api import SimpleSymbol, get_symbols, SimpleFile, SimpleModule
 from stextools.snify.commands import Command, CommandInfo, CommandOutcome, SubstitutionOutcome, SetNewCursor, \
-    ImportInsertionOutcome, ImportCommand, CommandCollection, show_current_selection, StatisticUpdateOutcome, \
+    ImportCommand, CommandCollection, show_current_selection, StatisticUpdateOutcome, \
     QuitSubdialogCommand, Exit
 from stextools.snify.state import State, SelectionCursor, PositionCursor
 from stextools.utils.ui import standard_header, pale_color, option_string, color, latex_format
@@ -32,6 +33,54 @@ def symbol_to_sorting_key(symbol: SimpleSymbol) -> tuple:
     return primary, secondary
 
 
+@dataclasses.dataclass
+class ImportPositions:
+    use_pos: int
+    top_use_pos: int
+    import_pos: Optional[int]
+    use_env: Optional[str]
+    top_use_env: Optional[str]
+    # scopes
+    top_use_scope: tuple[int, int]
+    use_scope: tuple[int, int]
+    import_scope: Optional[tuple[int, int]]
+
+
+def get_redundant_import_removals(
+        file: SimpleFile, text: str, imported_module: SimpleModule, import_scope: tuple[int, int], is_use: bool,
+) -> list[SubstitutionOutcome]:
+    # TODO: do this in simple_api (requires extending the API)
+    linker = file._linker
+    result: list[SubstitutionOutcome] = []
+    for dep in file._stex_doc.get_doc_info(linker.mh).dependencies:
+        if dep.is_input or dep.is_use_struct or dep.is_lib or dep.intro_range is None:
+            continue
+        if is_use and not dep.is_use:
+            continue
+        target_doc, target_mod = dep.get_target(linker.mh, file._stex_doc)
[…]

### 2024-11-28 stextools/snify/commands.py

@@ -2,7 +2,7 @@ import abc
 import dataclasses
 import re
 from pathlib import Path
-from typing import Optional, Any
+from typing import Optional, Any, Sequence
 
 import click
 
@@ -53,13 +53,6 @@ class StatisticUpdateOutcome(CommandOutcome):
         self.value = value
 
 
-class ImportInsertionOutcome(CommandOutcome):
-    """Note: command is responsible for ensuring that the index is correct *after* the previous file modification outcomes."""
-    def __init__(self, inserted_str: str, insert_pos: int):
-        self.inserted_str = inserted_str
-        self.insert_pos = insert_pos
-
-
 class SubstitutionOutcome(CommandOutcome):
     """Note: command is responsible for ensuring that the index is correct *after* the previous file modification outcomes."""
     def __init__(self, new_str: str, start_pos: int, end_pos: int):
@@ -108,7 +101,7 @@ class Command(abc.ABC):
         self.command_info = command_info
 
     @abc.abstractmethod
-    def execute(self, *, state: State, call: str) -> list[CommandOutcome]:
+    def execute(self, *, state: State, call: str) -> Sequence[CommandOutcome]:
         ...
 
     def standard_display(self, *, state: State) -> str:
@@ -136,7 +129,7 @@ class QuitSubdialogCommand(Command):
             description_long='Quits current subdialog')
         )
 
-    def execute(self, *, state: State, call: str) -> list[CommandOutcome]:
+    def execute(self, *, state: State, call: str) -> Sequence[CommandOutcome]:
         return [Exit()]
 
 
@@ -149,7 +142,7 @@ class QuitProgramCommand(Command):
             description_long='Quits the program')
         )
 
-    def execute(self, *, state: State, call: str) -> list[CommandOutcome]:
+    def execute(self, *, state: State, call: str) -> Sequence[CommandOutcome]:
         return [Exit()]
 
 
@@ -167,12 +160,13 @@ class RescanCommand(Command):
             description_long='Rescans all of MathHub')
         )
 
-    def execute(self, *, state: State, call: str) -> list[CommandOutcome]:
+    def execute(self, *, state: State, call: str) -> Sequence[CommandOutcome]:
         return [RescanOutcome()]
 
 
 class ImportCommand(Command):
-    def __init__(self, letter: str, description_short: str, description_long: str, outcome: ImportInsertionOutcome):
+    def __init__(self, letter: str, description_short: str, description_long: str, outcome: SubstitutionOutcome,
[…]

### 2024-11-28 stextools/snify/controller.py

@@ -1,6 +1,6 @@
 import abc
 from pathlib import Path
-from typing import Optional, Any
+from typing import Optional, Any, Sequence
 
 import click
 
@@ -11,7 +11,7 @@ from stextools.core.simple_api import file_from_path
 from stextools.core.stexdoc import Dependency
 from stextools.snify.annotate_command import AnnotateCommand, LookupCommand
 from stextools.snify.commands import CommandCollection, QuitProgramCommand, Exit, CommandOutcome, \
-    show_current_selection, ImportInsertionOutcome, SubstitutionOutcome, SetNewCursor, \
+    show_current_selection, SubstitutionOutcome, SetNewCursor, \
     ExitFileCommand, UndoOutcome, RedoOutcome, UndoCommand, RedoCommand, ViewCommand, View_i_Command, \
     TextRewriteOutcome, StatisticUpdateOutcome, ReplaceCommand, RescanCommand, RescanOutcome, StateSkipOutcome, \
     FocusOutcome, StemFocusCommand, StemFocusCommandPlus, StemFocusCommandPlusPlus
@@ -247,13 +247,6 @@ class Controller:
 
                 if isinstance(outcome, Exit):
                     return True
-                elif isinstance(outcome, ImportInsertionOutcome):
-                    text = self.state.get_current_file_text()
-                    modification = FileModification(
-                        file=self.state.get_current_file(),
-                        old_text=text,
-                        new_text=text[:outcome.insert_pos] + outcome.inserted_str + text[outcome.insert_pos:]
-                    )
                 elif isinstance(outcome, SubstitutionOutcome):
                     text = self.state.get_current_file_text()
                     modification = FileModification(
@@ -328,7 +321,7 @@ class Controller:
                 assert doc is not None
                 doc.delete_doc_info_if_outdated()
 
-    def _get_and_run_command(self) -> list[CommandOutcome]:
+    def _get_and_run_command(self) -> Sequence[CommandOutcome]:
         command_collection = self._get_current_command_collection()
         print()
         return command_collection.apply(state=self.state)

### 2024-11-28 doc/cicm25-sysdesc/sysdesc.tex
changes

@@
 \newcommand\ALeA{\textsf{ALeA}\xspace}
-\newcommand\srf{\textsf{srify}\xspace}
+\newcommand\snify{\textsf{snify}\xspace}
 \def\llangle{\langle\kern-.2em\langle}
@@
 
-\title{Bulk Semantic Annotation with an Partially Known Knowledge Base}
+\title{Bulk Semantic Annotation with a Partially Known Knowledge Base}
 \author{Michael Kohlhase, Jan Frederik Schaefer}
@@
 support in this area. Symbolic/logic-based technologies and sub-symbolic,
-machine-learning-based AI, e.g. via LLMs or chatbots; they have complementary strengths
+machine-learning-based AI, e.g.\ via LLMs or chatbots. They have complementary strengths
 and challenges: Symbolic technologies offer precision and explainability out of the box,
-but face scalability challenges, because the prerequisite background knowledge has to be
+but face scalability challenges because the prerequisite background knowledge has to be
 (manually) formalized. ML-based approaches can be trained on all data of the Internet, but
-face challenges in precision and explanations. Aarne Ranta matches these profiles to the
-notion of \textbf{producer tasks} -- i.e. tasks where precision is key, but limited
-coverage ($10^{3\pm1}$ concepts) is OK; e.g. for multi-language/variant manuals for very
-expensive tool machines -- and \textbf{consumer tasks}, where coverage is key and
-precision secondary; e.g. for consumer-grade machine translation like \textsf{Google
-  translate} in \cite{Ranta:atcp17}.
+face challenges in precision and explainability. In \cite{Ranta:atcp17},
+Aarne Ranta matches these profiles to the notion of \textbf{producer tasks} -- i.e.\ tasks
+where precision is key, but limited coverage ($10^{3\pm1}$ concepts) is OK;
+e.g.\ for multi-language/variant manuals for very
+expensive machines -- and \textbf{consumer tasks}, where coverage is key and
+precision secondary; e.g.\ for consumer-grade machine translation like \textsf{Google translate}.
 
@@
 approaches in practice. The producer task we use as a case study is that of adaptive
-learning assistants for MINT subjects, concretely the \ALeA system
-\cite{BerBetChu:lssmkm23}, which uses \sTeX \cite{MueKo:sdstex22,sTeX:github:on} -- a
-variant of {\LaTeX} that allows to embed semantic annotations -- for a knowledge
-representation and generates learner-adaptive learning objects instrumented with
+learning assistants for MINT subjects, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.
+\ALeA uses \sTeX \cite{MueKo:sdstex22,sTeX:github:on} -- a
+variant of {\LaTeX} that allows to embed semantic annotations -- for knowledge
+representation, and generates learner-adaptive learning objects instrumented with
 learning-support interactions from that. The \textbf{\sTeX/\ALeA content commons}
@@
 \section{Semantic Authoring in a Content Common}
-
 The main practical problem of annotating presentational {\LaTeX} course materials into
@@
 instance a learning object like the paragraph in \cref{fig:lo}, where we want to annotate
-the the term ``state space'' with the (semantic) concept of a ``state space of a search
-problem''. The latter can either be defined earlier in the course or in the domain mode
+the the term ``terminal'' with the (semantic) concept of ``a goal state of a search problem''. The latter can either be defined earlier in the course or in the domain model
 (or both).
@@
 \begin{figure}[ht]\centering
-  \fbox{\includegraphics[width=8cm]{../img/problem}}
+  \fbox{\includegraphics[width=10cm]{../example2.pdf}}
   \caption{An annotated Learning Object from an AI Lecture}\label{fig:lo}
@@
 
-Concretely this is about converting the {\LaTeX} string\ednote{add numbers, start with 3
-  so that corresponding lines with the next listing.}
+Concretely this is about converting the {\LaTeX} string
 \begin{lstlisting}[numbers=left,firstnumber=3,
 caption=The {\LaTeX} sources of \cref{fig:lo},label=lst:los]
-\item \blue{Problem:} The state space has local minima, e.g. the board on the
-  right has $h=1$ but every successor has $h>1$. 
+\item \blue{Note:} Depth-limited minimax requires an evaluation for every
+cut-off state $s$. If $s$ is terminal, we use its utility, and otherwise
+an estimate.
 \end{lstlisting}
 into the (partially) annotated item:
-\begin{lstlisting}[morekeywords={sn},numbers=left,
-caption=Annotating ``state space'' in \cref{lst:los},label=lst:losa]
-\importmodule[smglom/search]{mod?state-space}
+\begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
+caption=Annotating ``terminal'' in \cref{lst:los},label=lst:losa]
+\importmodule[smglom/search]{mod?search-problem}
 [...]
-\item \blue{Problem:} The \sn{state space} has local minima, e.g. the
-  board on the right has $h=1$ but every successor has $h>1$. 
+\item \blue{Note:} Depth-limited minimax requires an evaluation for every
+cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its utility,
+and otherwise an estimate.
 \end{lstlisting}
-Concretely, the author has to remember the module for the term ``state space'' from
-\cref{fig:state-space}, the symbol name \lstinline|state space|, annotate it with the
-\sTeX macro \lstinline|\sn{...}|, and add the module import in the first line (unless it
-is already imported). In this case the symbol name happens to be the same as its English
-verbalization, otherwise we would have to use
-\lstinline[mathescape]|\sr{$\llangle symbol name\rrangle$}{state space}|.
+% Concretely, the author has to remember the module that introduces the concept
+% of terminal states (\cref{fig:state-space}),
+% the symbol name \lstinline|goal state|, annotate it with the
+% \sTeX macro \lstinline[mathescape]|\sr{$\llangle symbol name \rrangle$}{$\llangle verbalization\rrangle$}|, and add the module import in the first line (unless it
+% is already imported).
+% If the symbol name and the desired verbalization coincide,
+% the short-hand macro \lstinline|\sn| can be used (e.g. \lstinline|\sn{utility}|).
 
-The annotation is legitimized by a definition in the domain model module shown in
-\cref{fig:state-space}. The \lstinline|\definame| in line 3 introduces a
-\textbf{verbalization} ``state space'' for the symbol \lstinline|state-space?state space|
+% In this case the symbol name happens to be the same as its English
+% verbalization, otherwise we would have to use
+% \lstinline[mathescape]|\sr{$\llangle symbol name\rrangle$}{state space}|.
+The annotation is legitimized\ednote{grounded?} by a definition in the domain model module shown in
+\cref{fig:state-space}. The \lstinline|\definame| in line 3 introduces the
+\textbf{verbalizations} ``goal state'' and ``terminal state''
+for the symbol \lstinline|search-problem?goal state|
 introduced be \lstinline|\symdecl*| declaration in line 3.
@@
 \begin{figure}[ht]\centering
-  \fbox{\includegraphics[width=12cm]{../img/state-space}}
+  \fbox{\includegraphics[width=12cm]{../search-problem.en.pdf}}
 \begin{lstlisting}[morekeywords={definame,symdecl},numbers=left]
-\begin{smodule}[title=State Space]{state-space}
+\begin{smodule}[title=Search Problem]{search-problem}
 [... some imports ...]
-\symdecl*{state space}
+\symdecl*{goal state}
 \begin{sdefinition}
-  [...] we call the \sn{graph} [...] the \definame{state space} of $\Pi$.
+  [...] Certain \sns{state} are [...] \definame[post=s]{goal state} [...]
+  (also called \definiendum{goal state}{terminal states}).
 \end{sdefinition}
@@
 \end{lstlisting}
-  \caption{The definition of the term ``state space'' from the domain model}\label{fig:state-space}
+  \caption{Simplified definition of ``goal state'' from the domain model}\label{fig:state-space}
 \end{figure}
 
-For the annotation \lstinline|\sn{state-space}|, the author has to be aware ``state
-space'' is a technical term, that the module from \cref{fig:state-space} exists, know the
-symbol name and its URL, and manage redundancy of the imports -- i.e. only adding the
+To annotate the string ``terminal'' in the learning object,
+the author has to be aware that it is a technical term,
+that the module from \cref{fig:state-space} exists, know the
+symbol name and its URL\ednote{path?}, and manage redundancy of the imports -- i.e. only adding the
 \lstinline|\importmodule| directive if it is not already (recursively) implied and
-possibly removing directives that become redundant by the new one. At ca 150.000 words in
-a semester's words of lecture notes and ca. 10-15\% of technical terms a rather daunting
+possibly removing directives that become redundant by the new one.
+At ca 150,000 words in
+a semester's words of lecture notes and ca. 10--15\%\ednote{can we actually measure this for AI?} of technical terms a rather daunting
 task.
@@
 errors and redundancies, and even offers a concept search interface. But these only offer
-support already annotated terms, so they do not solve the annotation problem above.
+support already for annotated terms, so they do not solve the annotation problem above.
 
-\section{The \srf System}
+\section{The \snify System}
 
-The \srf system (see \cite{stextools:git} is a simple command-line tool that creates a
+The \snify system (see \cite{stextools:git}) is a simple command-line tool that creates a
 database of symbol-verbalization pairs, analyzes \sTeX source files and then steps through
@@
 database. For each such word, the user is presented with an annotation choice and
-interactions that allow to fine-tune the wordwise annotation workflow. \Cref{fig:srify}
+interactions that allow to fine-tune the wordwise annotation workflow. \Cref{fig:snify}
 shows a typical situation.\ednote{MK: I like the example less and less! We need an example
@@
   \fbox{\includegraphics[width=12cm]{../img/srify}}
-  \caption{\srf in Action: Annotating the term ``state space''.}\label{fig:srify}
+  \caption{\snify in Action: Annotating the term ``state space''.}\label{fig:snify}
 \end{figure}
@@
 number. Here, the module \lstinline|state-space| is already imported (see the little green
-check-mark); otherwise \srf would offer the annotator the choice to import it -- if we are
+check-mark); otherwise \snify would offer the annotator the choice to import it -- if we are
 in a module context -- or to add a \lstinline|\usemodule| directive and where (in the
 local environment or top-level). Alternatively, typing \lstinline|s| skips this word,
-\lstinline|S| skips and adds a comment to always skip this word, and -- if \srf first
+\lstinline|S| skips and adds a comment to always skip this word, and -- if \snify first
 suggests the term `state` \lstinline|n| extends it to the compound ``state space'' by
@@
 
-\srf is usually called on a set of files -- e.g. in a directory or math archive -- which
+\snify is usually called on a set of files -- e.g. in a directory or math archive -- which
 together form a \textbf{session}, which can be interrupted and resumed without having to
-re-do all the \lstinline|s| skips. Alternatively, \srf can be called e.g. on the top-level
+re-do all the \lstinline|s| skips. Alternatively, \snify can be called e.g. on the top-level
 lecture notes; then the session consists of all included files. Other productivity
@@
 
-In this system description we have presented the \srf tool, a command-line tool
-\ednote{continue} \srf is open source\ednote{MK@FS: do we have an explicit license yet in
+In this system description we have presented the \snify tool, a command-line tool
+\ednote{continue} \snify is open source\ednote{MK@FS: do we have an explicit license yet in
   the repository.}; the source and documentation are available from \cite{stextools:git}.
@@
 not enter the actual annotations and import directives automatically. But maybe using the
-NER-based approach, after the \srf annotation process might change the trade-offs
+NER-based approach, after the \snify annotation process might change the trade-offs
 involved. In any case, the overall workflow suggested by the NER-based approach --

### 2024-11-28 stextools/snify/annotate_command.py
minor improvements

@@ -364,7 +364,6 @@ class AnnotateCommand(Command, AnnotateMixin):
             line = option_string(
                 str(i),
                 ' ' + symbol_display(file, symbol, state)
-                # + '\n      ' + click.style(symbol.declaring_file.path, italic=True, fg=pale_color())
             )
             lines.append(line)
         return '\n'.join(lines)

### 2024-11-28 stextools/snify/commands.py

@@ -6,6 +6,7 @@ from typing import Optional, Any, Sequence
 
 import click
 
+from stextools.core.config import get_config
 from stextools.core.linker import Linker
 from stextools.core.mathhub import make_filter_fun
 from stextools.core.simple_api import SimpleSymbol, get_files
@@ -14,14 +15,14 @@ from stextools.utils.ui import option_string, standard_header, pale_color, get_l
     standard_header_str
 
 
-def show_current_selection(state, with_header: bool = True):
+def show_current_selection(state: State, with_header: bool = True):
     if with_header:
         status = [
             f'File {state.cursor.file_index + 1}/{len(state.files)}'.ljust(15),
             f'Annotations added: {state.statistic_annotations_added}'.ljust(25)
         ]
         print(' | ' + ' | '.join(status) + ' |')
-        standard_header(str(state.get_current_file()), bg='bright_green')
+        standard_header(str(state.get_current_file().relative_to(Path.cwd())), bg='bright_green')
 
     cursor = state.cursor
     assert isinstance(cursor, SelectionCursor)
@@ -29,7 +30,8 @@ def show_current_selection(state, with_header: bool = True):
     a, b, c, line_no_start = get_lines_around(
         state.get_current_file_text(),
         cursor.selection_start,
-        cursor.selection_end
+        cursor.selection_end,
+        n_lines=int(get_config().get('stextools.snify', 'context-lines', fallback='7'))
     )
     doc = latex_format(a) + (
         '\n'.join(click.style(p, bg='bright_yellow', bold=True) for p in b.split('\n'))

### 2024-11-28 stextools/snify/controller.py

@@ -19,7 +19,7 @@ from stextools.snify.selection import VerbTrie, PreviousWordShouldBeIncluded, Ne
     get_linked_strings, FirstWordShouldntBeIncluded, LastWordShouldntBeIncluded
 from stextools.snify.session_storage import SessionStorage
 from stextools.snify.skip_and_ignore import SkipOnceCommand, IgnoreWordOutcome, IgnoreCommand, IgnoreList, \
-    AddWordToSrSkip, AddStemToSrSkip, SrSkipped, SkipUntilFileEnd
+    AddWordToSrSkip, AddStemToSrSkip, SrSkipped, SkipUntilFileEnd, SkipForRestOfSession
 from stextools.snify.state import PositionCursor, Cursor, SelectionCursor
 from stextools.snify.state import State
 from stextools.snify.stemming import string_to_stemmed_word_sequence_simplified, string_to_stemmed_word_sequence
@@ -361,6 +361,7 @@ class Controller:
                 LastWordShouldntBeIncluded(self.get_current_lang()),
                 SkipOnceCommand(),
                 SkipUntilFileEnd(),
+                SkipForRestOfSession(),
                 IgnoreCommand(self.get_current_lang()),
                 AddWordToSrSkip(),
                 AddStemToSrSkip(self.get_current_lang()),

### 2024-11-28 stextools/snify/skip_and_ignore.py

@@ -46,6 +46,28 @@ class SkipUntilFileEnd(Command):
         ]
 
 
+class SkipForRestOfSession(Command):
+    def __init__(self):
+        super().__init__(CommandInfo(
+            show=False,
+            pattern_presentation='s!!',
+            pattern_regex='^s!!$',
+            description_short='kip until end of session',
+            description_long='Do not propose any occurrences of the selected phrase anymore (until end of session).')
+        )
+
+    def execute(self, *, state: State, call: str) -> list[CommandOutcome]:
+        assert isinstance(state.cursor, SelectionCursor)
+        return [
+            StateSkipOutcome(
+                word=state.get_current_file_text()[state.cursor.selection_start:state.cursor.selection_end],
+                is_stem=False,
+                session_wide=True
+            ),
+            SetNewCursor(PositionCursor(state.cursor.file_index, state.cursor.selection_start)),
+        ]
+
+
 class AddWordToSrSkip(Command):
     def __init__(self):
         super().__init__(CommandInfo(

### 2024-11-28 doc/cicm25-sysdesc/sysdesc.tex
some work on sysdesc

@@
 \begin{figure}[ht]\centering
-  \fbox{\includegraphics[width=10cm]{../example2.pdf}}
+  \fbox{\includegraphics[width=10cm]{../img/minimax-remarks.pdf}}
   \caption{An annotated Learning Object from an AI Lecture}\label{fig:lo}
@@
 \begin{figure}[ht]\centering
-  \fbox{\includegraphics[width=12cm]{../search-problem.en.pdf}}
+  \fbox{\includegraphics[width=12cm, trim={1cm 0 0 0}, clip]{../img/search-problem.en.pdf}}
 \begin{lstlisting}[morekeywords={definame,symdecl},numbers=left]
@@
 
-The \snify system (see \cite{stextools:git}) is a simple command-line tool that creates a
-database of symbol-verbalization pairs, analyzes \sTeX source files and then steps through
-all word occurrences in the document that share the stem with a verbalization in the
-database. For each such word, the user is presented with an annotation choice and
+The \snify system (see \cite{stextools:git}) is a simple command-line tool.
+It creates a database of symbol-verbalization pairs, analyzes \sTeX source files,
+and then steps through all word occurrences in the document that match a
+verbalization in the database.
+Matching is based on word stems using off-the-shelf stemmers, which allows us to
+e.g.\ match ``utilities'' with ``utility''.
+For each matched word, the user is presented with an annotation choice and
 interactions that allow to fine-tune the wordwise annotation workflow. \Cref{fig:snify}
-shows a typical situation.\ednote{MK: I like the example less and less! We need an example
-  that has at least one homostem (another verbalization that has the same stem; introduce
-  the concept her and use it throughout) so that we can show the choice mechanism, and
-  where the verbalization is different from the symbol name. Having a two-word compound is
-  good though.}
+shows a typical situation.
 
 \begin{figure}[ht]
-  \fbox{\includegraphics[width=12cm]{../img/srify}}
-  \caption{\snify in Action: Annotating the term ``state space''.}\label{fig:snify}
+  \setlength{\fboxsep}{0pt}
+  \fbox{\includegraphics[width=12cm,trim={0 3cm 0 0},clip]{../img/snify}}
+  \caption{\snify in Action: Annotating the word ``terminal''.}\label{fig:snify}
 \end{figure}
 
-Here the annotator can choose an annotation by typing the corresponding choice
-number. Here, the module \lstinline|state-space| is already imported (see the little green
-check-mark); otherwise \snify would offer the annotator the choice to import it -- if we are
-in a module context -- or to add a \lstinline|\usemodule| directive and where (in the
-local environment or top-level). Alternatively, typing \lstinline|s| skips this word,
-\lstinline|S| skips and adds a comment to always skip this word, and -- if \snify first
-suggests the term `state` \lstinline|n| extends it to the compound ``state space'' by
-including the next word.
+The annotator can choose an annotation by typing the corresponding choice number,
+in this case \lstinline|0|.
+The little green check-mark indicates that the relevant module is already imported.
+Otherwise \snify would offer the annotator the choice to import it -- if we are
+in a module context -- or to add a \lstinline|\usemodule| directive, either in the
+local environment or at the top-level.
+
+To skip the word, the user can type \lstinline|s|,
+\lstinline|s!| skips it until the end of the file, and \lstinline|S|
+will skip it in future runs as well by appending a comment to the file.
+There are numerous other commands, e.g.\ to adjust the selection,
+search for alternative annotation targets,
+or view the file introducing one of the choices.
 
@@
 lecture notes; then the session consists of all included files. Other productivity
-features include a focus mode that can be used to propagate a particular choice to the
+features include a focus mode that can be used to only annotate a particular word in the
 rest of the file, session, or the whole \sTeX/\ALeA corpus before resuming regular
-annotation This reduces the cognitive effort from choosing a symbol from a list to
-deciding whether a particular choice is applicable in the current document context. Of
-course the include management still needs to be done locally.
+annotation. This reduces the cognitive overhead from switching between different 
+words and symbol lists.
+% Of course the include management still needs to be done locally.
 
@@
 
-The current interface is not well-suited for on-the-fly annotation while editing while
-authoring. For that, the underlying information (the symbol/verbalization pairs harvested
+The current interface is not well-suited for on-the-fly annotation while authoring.
+For that, the underlying information (the symbol/verbalization pairs harvested
 from the domain model) can be integrated into any IDE. In fact we plan to do this for the
@@
 
-In this system description we have presented the \snify tool, a command-line tool
-\ednote{continue} \snify is open source\ednote{MK@FS: do we have an explicit license yet in
-  the repository.}; the source and documentation are available from \cite{stextools:git}.
+In this system description we have presented \snify, a command-line tool
+\ednote{continue} \snify is open source;
+the source and documentation are available from \cite{stextools:git}.
 

### 2024-11-28 experiment.py

@@ -1,10 +1,11 @@
 from stextools.core.cache import Cache
 import logging
+from pathlib import Path
 
 from stextools.core.linker import Linker
 
 logging.getLogger('pylatexenc.latexwalker').setLevel(logging.WARNING)
-# logging.getLogger('stextools.core.linker').setLevel(logging.FATAL)
+logging.getLogger('stextools.core.linker').setLevel(logging.FATAL)
 logging.basicConfig(level=logging.INFO)
 
 Cache.clear = lambda: None  # type: ignore
@@ -12,3 +13,13 @@ Cache.clear = lambda: None  # type: ignore
 mh = Cache.get_mathhub(update_all=True)
 
 linker1 = Linker(mh)
+
+import stextools.core.simple_api as sa
+
+# sa.file_from_path(Path('/home/jfs/MMT/MMT-content/smglom/computing/source/mod/termination.en.tex'), linker1)
+symbs = list(sa.get_symbols(linker1, name='termination'))
+
+for doc in sa.get_files(linker1):
+    for verb in doc._stex_doc.get_doc_info(linker1.mh).verbalizations:
+        if verb.symbol_name == 'termination':
+            print(doc.path)

### 2024-12-04 stextools/cycle_finder.py
no heuristic for ignoring stop words

@@ -5,7 +5,7 @@ import click
 
 from stextools.core.cache import Cache
 from stextools.core.linker import Linker
-from stextools.utils.ui import get_lines_around, latex_format, pale_color, print_highlight_selection
+from stextools.utils.ui import print_highlight_selection
 
 
 class CycleFound(Exception):

### 2024-12-04 stextools/snify/selection.py

@@ -229,8 +229,8 @@ class VerbTrie:
             if not verbalizations:
                 verbalizations = [symbol.name]
             for verb in verbalizations:
-                if len(verb) < 3:
-                    continue
+                # if len(verb) < 3 and sum(1 for c in verb if c.isupper()) < 2:
+                #     continue
 
                 words = [str(w) for w in string_to_stemmed_word_sequence_simplified(verb, lang)]
                 if repr(words) in covered_verbs:

### 2024-12-04 doc/cicm25-sysdesc/sysdesc.tex
drain

@@
 
-The \snify system (see \cite{stextools:git}) is a simple command-line tool.
-It creates a database of symbol-verbalization pairs, analyzes \sTeX source files,
+The \snify\ednote{explain name? write as sn-ify?}
+system (see \cite{stextools:git}) is a simple command-line tool
+that creates a database of symbol-verbalization pairs, analyzes \sTeX source files,
 and then steps through all word occurrences in the document that match a

### 2024-12-04 scripts/annocounthistory.py

@@ -1,0 +1,96 @@
+import datetime
+import itertools
+import json
+import re
+import subprocess
+import sys
+from pathlib import Path
+
+paths: list[Path] = [Path(p) for p in sys.argv[1:]]
+
+repo_paths: list[Path] = []
+
+for path in paths:
+    for line in subprocess.run(['sh', '-c', 'find -name .git'], cwd=path, capture_output=True, text=True).stdout.splitlines():
+        print(line)
+        line = line.strip()
+        if line:
+            repo_paths.append((path / line).parent)
+
+
+def checkout_date(repo_path: Path, date: datetime.date):
+    result = subprocess.run(['sh', '-c', f'git rev-list -n1 --before={date.isoformat()} main | xargs git checkout'], cwd=repo_path)
+    if result.returncode != 0:
+        print(f'Error checking out {date.isoformat()} in {repo_path}')
+
+
+# MacroSpec('sn', '[{'),
+# MacroSpec('sns', '[{'),
+# MacroSpec('Sn', '[{'),
+# MacroSpec('Sns', '[{'),
+# MacroSpec('sr', '[{{'),
+
+current_names = ['sn', 'sns', 'Sn', 'Sns', 'sr']
+intermediate_names = ['symname', 'Symname', 'symref']
+old_names = [prefix + core + suff + suffsuff for core in ['tref', 'Tref'] for suff in ['i', 'ii', 'iii', 'iv'] for suffsuff in ['', 's'] for prefix in ['', 'm']]
+
+
+def anno_count(repo_path: Path):
+    regex = re.compile(r'\\(' + '|'.join(current_names + intermediate_names + old_names) + r')[{[]')
+    count = 0
+    for texfile in repo_path.glob('**/*.tex'):
+        if any(x in texfile.name for x in ['all.', 'glossary', 'dictionary']):   # those are generated files
+            continue
+        try:
+            count += len(regex.findall(texfile.read_text()))
+        except Exception as e:
+            print(e)
+            print()
+    return count
+
+end_date = datetime.date.today()
+
+HISTORY = {}
+HISTORY_BY_REPO = {}
+
+for i in range(100):
+    print(i)
+    cur_date = end_date - i * datetime.timedelta(days=14)
+
+    HISTORY[cur_date.isoformat()] = 0
+
+    for repo_path in repo_paths:
[…]

### 2024-12-04 stextools/snify/selection.py
bugfix

@@ -230,7 +230,8 @@ class VerbTrie:
                 verbalizations = [symbol.name]
             for verb in verbalizations:
                 # if len(verb) < 3 and sum(1 for c in verb if c.isupper()) < 2:
-                #     continue
+                if len(verb) < 1:
+                    continue
 
                 words = [str(w) for w in string_to_stemmed_word_sequence_simplified(verb, lang)]
                 if repr(words) in covered_verbs:

### 2024-12-05 stextools/core/simple_api.py
scripts for annotation statistics

@@ -193,6 +193,10 @@ class SimpleArchive:
     def __hash__(self) -> int:
         return hash((self._repo, self._linker))
 
+    @property
+    def path(self) -> Path:
+        return self._repo.path
+
 
 @dataclasses.dataclass
 class SimpleVerbalization:
@@ -231,6 +235,11 @@ def file_from_path(path: Path, linker: Linker) -> Optional[SimpleFile]:
 def get_files(linker: Linker) -> Iterable[SimpleFile]:
     for doc_int in linker.document_ints.int_iter():
         yield SimpleFile(doc_int, linker)
+
+
+def get_repos(linker: Linker) -> Iterable[SimpleArchive]:
+    for repo in linker.mh.iter_stex_archives():
+        yield SimpleArchive(repo, linker)
 
 
 def get_linker() -> Linker:

### 2024-12-05 doc/cicm25-sysdesc/sysdesc.tex
more

--- 2024-11-28 doc/cicm25-sysdesc/sysdesc.tex
@@
 
-\section{Semantic Authoring in a Content Common}
+\section{Semantic Authoring in a Content Commons}
+
+Before we discuss semantic authoring in general, let us introduce a concrete example from
+the \sTeX/\ALeA world. We will use it to sharpen our intuitions of the problems an issues
+involved and to ground the solution we propose.
+
+\subsection{Running Example: Annotating  ``Terminal''}
+
 The main practical problem of annotating presentational {\LaTeX} course materials into
@@
 instance a learning object like the paragraph in \cref{fig:lo}, where we want to annotate
-the the term ``terminal'' with the (semantic) concept of ``a goal state of a search problem''. The latter can either be defined earlier in the course or in the domain model
+the the term ``terminal'' with the (semantic) concept of ``a goal state of a search
+problem''. The latter can either be defined earlier in the course or in the domain model
 (or both).
@@
 % \lstinline[mathescape]|\sr{$\llangle symbol name\rrangle$}{state space}|.
-The annotation is legitimized\ednote{grounded?} by a definition in the domain model module shown in
+The annotation rerfers to a definition in the domain model module shown in
 \cref{fig:state-space}. The \lstinline|\definame| in line 3 introduces the
-\textbf{verbalizations} ``goal state'' and ``terminal state''
-for the symbol \lstinline|search-problem?goal state|
-introduced be \lstinline|\symdecl*| declaration in line 3.
+\textbf{verbalizations} ``goal state'' and ``terminal state'' for the symbol
+\lstinline|search-problem?goal state| introduced be \lstinline|\symdecl*| declaration in
+line 3.
 
 \begin{figure}[ht]\centering
-  \fbox{\includegraphics[width=12cm, trim={1cm 0 0 0}, clip]{../img/search-problem.en.pdf}}
+  \fbox{\includegraphics[width=12cm]{../img/search-problem.en.pdf}}
 \begin{lstlisting}[morekeywords={definame,symdecl},numbers=left]
@@
 
-To annotate the string ``terminal'' in the learning object,
-the author has to be aware that it is a technical term,
-that the module from \cref{fig:state-space} exists, know the
-symbol name and its URL\ednote{path?}, and manage redundancy of the imports -- i.e. only adding the
-\lstinline|\importmodule| directive if it is not already (recursively) implied and
-possibly removing directives that become redundant by the new one.
-At ca 150,000 words in
-a semester's words of lecture notes and ca. 10--15\%\ednote{can we actually measure this for AI?} of technical terms a rather daunting
-task.
+To annotate the string ``terminal'' in the learning object, the author has to be aware
+that it is a technical term, that the module from \cref{fig:state-space} exists, know the
+symbol name and its URL\ednote{path?}, and manage redundancy of the imports -- i.e. only
+adding the \lstinline|\importmodule| directive if it is not already (recursively) implied
+and possibly removing directives that become redundant by the new one.  At ca 150,000
+words in a semester's words of lecture notes and ca. 10--15\%\ednote{can we actually
+  measure this for AI?} of technical terms a rather daunting task.
 
@@
 support already for annotated terms, so they do not solve the annotation problem above.
+
+\subsection{Semantic Authoring as a Distinct Problem}
+
+\begin{newpart}{MK: looking at the problem in general and compare it to the informal
+    authoring and formal authoring problems e.g. programming in an IDE.}
+We contend that semantic authoring is a problem that is distinct from traditional
+(informal) authoring and authoring fully formal corpora (e.g. programs or formalizations),
+while sharing some of the aspects of either.
+
+In traditional authoring e.g. of a scientific article or a textbook, the domain model -- a
+highly structured model of the domain of discourse -- is in the authors' heads, and is
+partially verbalized e.g. in the preliminaries section or the main sections of the
+respective article. The content is informal -- usually a natural language like english
+augmented with technical jargon, formulae, tables, and diagrams -- and designed for
+processing via the brain of a colleague or student -- in any case, a human who shares a
+canonical part of the domain model with the authors.\ednote{MK: this shared domain model
+  could be thought of as the informal content commons; hmmm, but maybe the literature is
+  the content commons} Consequently a large part of the authoring problem lies in
+predicting the state and extent of the domain model of the reader and creating text that
+bridges the gap between the reader's domain model and the payload content of the article
+-- the knowledge the article or textbook intends to impart -- ideally using the technical
+vocabulary the reader is familiar with to reduce this gap. Tool support for traditional
+authoring is minimal and usually restricted to spell/grammar-checkers and online thesauri
+if we disregard LLM writing/formulation support for now.
+
+In fully formal authoring e.g. for programming, the content commons is given formally in
+the form of the libraries of the respective programming language, correspondingly the
+authored content is also formal -- a program -- and intended for processing by a computer,
+for which the formality is a prerequisite.
+
+\ednote{talk about formalization in logic -- e.g. as theorem prover libraries as well}
+\end{newpart}
 

### 2024-12-05 stextools/snify/controller.py
better colors for dark terminals

@@ -49,8 +49,8 @@ class FileModification(Modification):
     def apply(self, state: State):
         current_text = self.file.read_text()
         if current_text != self.old_text:
-            print(click.style(f"File {self.file} has been modified since the last time it was read", bg='bright_yellow'))
-            print(click.style(f"I will not change the file", bg='bright_yellow'))
+            print(click.style(f"File {self.file} has been modified since the last time it was read", fg='black', bg='bright_yellow'))
+            print(click.style(f"I will not change the file", fg='black', bg='bright_yellow'))
             click.pause()
             return
 
@@ -59,8 +59,8 @@ class FileModification(Modification):
     def unapply(self, state: State):
         current_text = self.file.read_text()
         if current_text != self.new_text:
-            print(click.style(f"File {self.file} has been modified since the last time it was written", bg='bright_yellow'))
-            print(click.style(f"I will not change the file", bg='bright_yellow'))
+            print(click.style(f"File {self.file} has been modified since the last time it was written", fg='black', bg='bright_yellow'))
+            print(click.style(f"I will not change the file", fg='black', bg='bright_yellow'))
             click.pause()
             return
         self.file.write_text(self.old_text)
@@ -494,7 +494,14 @@ def snify(files: list[str], filter: str, ignore: str):
     if state is None:
         state = State(files=[], filter_pattern=filter,
                       ignore_pattern=ignore, cursor=PositionCursor(file_index=0, offset=0))
-        controller = Controller(state, new_files=[Path(file).absolute().resolve() for file in files])
+        paths: list[Path] = []
+        for file in files:
+            path = Path(file).absolute().resolve()
+            if path.is_dir():
+                paths.extend(path.rglob('*.tex'))
+            else:
+                paths.append(path)
+        controller = Controller(state, new_files=paths)
     else:
         controller = Controller(state)
     unfinished = controller.run()

### 2024-12-05 stextools/utils/ui.py

@@ -34,7 +34,7 @@ def standard_header(title, bg: str = 'bright_green'):
 
 
 def standard_header_str(title, bg: str = 'bright_green') -> str:
-    return click.style(f'{title:^{width()}}', bold=True, bg=bg)
+    return click.style(f'{title:^{width()}}', bold=True, fg='black', bg=bg)
 
 
 def pale_color():
@@ -52,7 +52,7 @@ def simple_choice_prompt(options: list[str]):
 def print_highlight_selection(doc_text: str, start: int, end: int, n_lines: int = 7, *, bold: bool = True):
     a, b, c, line_no_start = get_lines_around(doc_text, start, end, n_lines)
     doc = latex_format(a) + (
-        '\n'.join(click.style(p, bg='bright_yellow', bold=bold) for p in b.split('\n'))
+        '\n'.join(click.style(p, fg='black', bg='bright_yellow', bold=bold) for p in b.split('\n'))
     ) + latex_format(c)
 
     for i, line in enumerate(doc.split('\n'), line_no_start):

### 2024-12-05 stextools/__main__.py
better colors for dark terminals

@@ -87,7 +87,7 @@ def translate(path):
 
 
 @cli.command(name='snify', help='\\sn-ify sTeX documents')
-@click.argument('files', nargs=-1)
+@click.argument('files', nargs=-1)  # Path to files or directories to snify
 @click.option('--filter',
               default=lambda: get_config().get('stextools.snify', 'filter', fallback=None),
               help='Filter pattern to only include some archives (e.g. \'smglom/*,courses/*\')')

### 2024-12-05 stextools/snify/annotate_command.py

@@ -385,8 +385,8 @@ def symbol_display(file: SimpleFile, symbol: SimpleSymbol, state: State, style: 
             symb_path[0] + '?' +
             (
                 (
-                        click.style(symb_path[1], bg=color('bright_cyan', (180, 180, 255))) + '?' +
-                        click.style(symb_path[2], bg=color('bright_green', (180, 255, 180)))
+                        click.style(symb_path[1], fg='black', bg=color('bright_cyan', (180, 180, 255))) + '?' +
+                        click.style(symb_path[2], fg='black', bg=color('bright_green', (180, 255, 180)))
                 )
                 if style else (symb_path[1] + '?' + symb_path[2])
             )

### 2024-12-08 doc/cicm25-sysdesc/sysdesc.tex
draining

@@
 authored content is also formal -- a program -- and intended for processing by a computer,
-for which the formality is a prerequisite.
+for which the formality is a prerequisite. Formal authoring is usually supported by an
+integraded development environment (IDE) that makes use of the fully formal content and
+offers services like semantic tab completion
 
@@
 
-The current interface is not well-suited for on-the-fly annotation while authoring.
-For that, the underlying information (the symbol/verbalization pairs harvested
+One practical problem with \snify that remains is that the system only harvests
+symbol/verbalization pairs from \sTeX files on the local file system. This means that for
+effective annotation users (have to) download the whole content commons and keep it
+updated. Unfortunately, the current MMT-based MathHub knowledge management
+API\ednote{intro this above} does not supply verbalization information. The iMMT-based
+API~\cite{iMMT:on} currently under development will, so \snify will be able to request the
+pairs for the remaining (non-local) content commons. If a pair is chosen in an annotation,
+\snify can then automatically download the corresponding archive so that the annotated
+file can be formattted with pdflatex or rusTeX locally.
+
+Finally, the current interface is not well-suited for on-the-fly annotation while
+authoring.  For that, the underlying information (the symbol/verbalization pairs harvested
 from the domain model) can be integrated into any IDE. In fact we plan to do this for the
@@
 
-In this system description we have presented \snify, a command-line tool
-\ednote{continue} \snify is open source;
-the source and documentation are available from \cite{stextools:git}.
+In this paper we have presented the semantic authoring problem using the \sTeX corpus and
+the adaptive learning assistant \ALeA as a concrete example. We have shown that it differs
+from both classical authoring and formal authoring in terms of the necesssary system
+support. As an example of specialized authoring support, we have presented \snify, a
+command-line tool that bridges the gap between formal symbols in \sTeX and
+
+\snify is open source; the source and documentation are available from
+\cite{stextools:git}.
 

### 2024-12-09 doc/cicm25-sysdesc/sysdesc.tex
minor edits

@@
 \subsection{Running Example: Annotating  ``Terminal''}
-
 The main practical problem of annotating presentational {\LaTeX} course materials into
 semantic documents that carry enough information to support meaningful learning support
-services is to add semantic references to technical terms from the domain jargon. Take for
-instance a learning object like the paragraph in \cref{fig:lo}, where we want to annotate
-the the term ``terminal'' with the (semantic) concept of ``a goal state of a search
+services is to add semantic references to technical terms from the domain jargon.
+% TODO: This the following better?
+% With \sTeX, authors can add semantic references to technical terms,
+% which enables straight-forward learning support services like
+% showing the definition of a term, but also informs the system
+% about the content/prerequisites of a learning object.
+% Adding these references is a key practical problem when converting presentational {\LaTeX}
+% course materials into semantic documents.
+Take for instance a learning object like the paragraph in \cref{fig:lo}, where we want to annotate
+the term ``terminal'' with the (semantic) concept of ``a goal state of a search
 problem''. The latter can either be defined earlier in the course or in the domain model
@@
   \fbox{\includegraphics[width=10cm]{../img/minimax-remarks.pdf}}
-  \caption{An annotated Learning Object from an AI Lecture}\label{fig:lo}
+  \caption{An annotated learning object from an AI lecture}\label{fig:lo}
 \end{figure}
@@
 \begin{lstlisting}[numbers=left,firstnumber=3,
-caption=The {\LaTeX} sources of \cref{fig:lo},label=lst:los]
+caption=The (unannotated) {\LaTeX} sources of \cref{fig:lo},label=lst:los]
 \item \blue{Note:} Depth-limited minimax requires an evaluation for every
@@
 % \lstinline[mathescape]|\sr{$\llangle symbol name\rrangle$}{state space}|.
-The annotation rerfers to a definition in the domain model module shown in
-\cref{fig:state-space}. The \lstinline|\definame| in line 3 introduces the
-\textbf{verbalizations} ``goal state'' and ``terminal state'' for the symbol
-\lstinline|search-problem?goal state| introduced be \lstinline|\symdecl*| declaration in
-line 3.
+The annotation refers to a definition in the domain model module shown in \cref{fig:state-space}.
+The \lstinline|\definame| and \lstinline|\definiendum| in lines 5--6 introduce the \textbf{verbalizations} ``goal state'' and ``terminal state'' for the symbol \lstinline|search-problem?goal state| introduced by \lstinline|\symdecl*| declaration in line 3.
 
@@
   \fbox{\includegraphics[width=12cm]{../img/search-problem.en.pdf}}
-\begin{lstlisting}[morekeywords={definame,symdecl},numbers=left]
+\begin{lstlisting}[morekeywords={definame,symdecl,definiendum},numbers=left]
 \begin{smodule}[title=Search Problem]{search-problem}
@@
 analyzes annotations on the fly, displays the underlying reference semantics, reports
-errors and redundancies, and even offers a concept search interface. But these only offer
-support already for annotated terms, so they do not solve the annotation problem above.
+errors and redundancies, and even offers a concept search interface.
+But these features only support already annotated terms, so they do not solve the annotation problem above.
 

### 2024-12-11 doc/cicm25-sysdesc/sysdesc.tex
more

@@
 
-\begin{newpart}{MK: looking at the problem in general and compare it to the informal
-    authoring and formal authoring problems e.g. programming in an IDE.}
 We contend that semantic authoring is a problem that is distinct from traditional
@@
 integraded development environment (IDE) that makes use of the fully formal content and
-offers services like semantic tab completion
+offers services like semantic highlighting, tab completion, jumping to variable
+declarations, function definitions, and interface definitions, as well as refactoring
+support. All of these rely on a syntactic/semantic analysis of the formal content. The
+situation is very similar for authoring of formal knowledge, e.g. in program verification
+of the formalization of mathematics.
 
-\ednote{talk about formalization in logic -- e.g. as theorem prover libraries as well}
-\end{newpart}
+In semantic authoring, e.g. the \sTeX sources of educational content for the \ALeA system
+presented above. The base of the text is classical narrative text to which most of the
+constraints of classical authoring apply. The exception is the authoring-time tailoring to
+the competencies of the assumed reader, which can (partially) be automated by systems like
+\ALeA. Note that it is not a restriction that \ALeA is initially targeted to tertiary
+education; the underlying mechanisms generalize to all scientific communication and
+technical documents. After all, the purpose is to transfer information (instruct), even if
+the roles of instructor and instructee are not predetermined by academic seniority.
+
+But in addition to the base text, semantic authoring also has to provide the semantic
+annotations that drive the personalization and interaction. Techniques from formal
+authoring could help here, but \ednote{cite Luka's CICM paper \cite{VreWelKam:tsmmdui24}
+  for the formulae; introduce the \sTeX way of writing content formulae above.}
 
@@
 
-The \snify\ednote{explain name? write as sn-ify?}
-system (see \cite{stextools:git}) is a simple command-line tool
-that creates a database of symbol-verbalization pairs, analyzes \sTeX source files,
-and then steps through all word occurrences in the document that match a
-verbalization in the database.
-Matching is based on word stems using off-the-shelf stemmers, which allows us to
-e.g.\ match ``utilities'' with ``utility''.
-For each matched word, the user is presented with an annotation choice and
-interactions that allow to fine-tune the wordwise annotation workflow. \Cref{fig:snify}
-shows a typical situation.
+The \snify utility (from ``\texttt{\textbackslash sn}-ify'' for adding term references;
+see \cite{stextools:git}) is a simple command-line tool that creates a database of
+symbol-verbalization pairs, analyzes \sTeX source files, and then steps through all word
+occurrences in the document that match a verbalization in the database.  Matching is based
+on word stems using off-the-shelf stemmers, which allows us to e.g.\ match ``utilities''
+with ``utility''.  For each matched word, the user is presented with an annotation choice
+and interactions that allow to fine-tune the wordwise annotation
+workflow. \Cref{fig:snify} shows a typical situation.
 

### 2024-12-11 doc/cicm25-sysdesc/sysdesc.tex
more

@@
 
+\paragraph{Traditional, informal authoring}
 In traditional authoring e.g. of a scientific article or a textbook, the domain model -- a
@@
 
+\paragraph{Formal Authoring}
 In fully formal authoring e.g. for programming, the content commons is given formally in
@@
 
+\paragraph{Semantic Authoring}
 In semantic authoring, e.g. the \sTeX sources of educational content for the \ALeA system
@@
 authoring could help here, but \ednote{cite Luka's CICM paper \cite{VreWelKam:tsmmdui24}
-  for the formulae; introduce the \sTeX way of writing content formulae above.}
+  for the formulae; introduce the \sTeX way of writing content formulae above.} But in
+contrast to formal authoring, we cannot just trigger the IDE functionality by the formal
+syntax, as the authored content is -- so far -- informal and the content commons (the
+target of the annotations) is itself only flexiformal. In a way, we need a local
+flexiformalization workflow that helps authors bridge the informal-formal gap.
+
+\paragraph{Semantic Authoring in \sTeX}
+How this can concretely work depends on the particular semantic framework, the specific
+annotations we want to support, and the semantics embedded in the content commons.
 

### 2024-12-11 doc/cicm25-sysdesc/sysdesc.tex
more

@@
 partially verbalized e.g. in the preliminaries section or the main sections of the
-respective article. The content is informal -- usually a natural language like english
+respective article. The content is informal -- usually a natural language like English
 augmented with technical jargon, formulae, tables, and diagrams -- and designed for
@@
 authoring is minimal and usually restricted to spell/grammar-checkers and online thesauri
-if we disregard LLM writing/formulation support for now.
+if we disregard LLM-based writing/formulation support for now.
 
@@
 for which the formality is a prerequisite. Formal authoring is usually supported by an
-integraded development environment (IDE) that makes use of the fully formal content and
+integrated development environment (IDE) that makes use of the fully formal content and
 offers services like semantic highlighting, tab completion, jumping to variable
@@
 annotations that drive the personalization and interaction. Techniques from formal
-authoring could help here, but \ednote{cite Luka's CICM paper \cite{VreWelKam:tsmmdui24}
-  for the formulae; introduce the \sTeX way of writing content formulae above.} But in
-contrast to formal authoring, we cannot just trigger the IDE functionality by the formal
-syntax, as the authored content is -- so far -- informal and the content commons (the
-target of the annotations) is itself only flexiformal. In a way, we need a local
-flexiformalization workflow that helps authors bridge the informal-formal gap.
+authoring could help here, but in contrast to formal authoring, we cannot just trigger the
+IDE functionality by the formal syntax, as the authored content is -- so far -- informal
+and the content commons (the target of the annotations) is itself only flexiformal. In a
+way, we need a local flexiformalization workflow that helps authors bridge the
+informal-formal gap.
 
@@
 How this can concretely work depends on the particular semantic framework, the specific
-annotations we want to support, and the semantics embedded in the content commons.
+annotations we want to support, and the semantics embedded in the content commons. In the
+\sTeX framework, where we semantically annotate formulae and technical terms with pointers
+to their definienda, we can bridge the presentation/semantics gap -- we annotate the words
+in sentences and glyphs in formulae with the verbalizations and notations in the content
+commons.\ednote{MK: maybe explain them again}. A tool for supporting semantic authoring of
+formulae via notations has been presented by Vre{\v{c}}ar et al. in
+\cite{VreWelKam:tsmmdui24}. In this paper we present a tool that supports semantic
+annotation of technical terms via verbalizations.
 
@@
 \snify can then automatically download the corresponding archive so that the annotated
-file can be formattted with pdflatex or rusTeX locally.
+file can be formatted with pdflatex or rusTeX locally.
 
@@
 the adaptive learning assistant \ALeA as a concrete example. We have shown that it differs
-from both classical authoring and formal authoring in terms of the necesssary system
+from both classical authoring and formal authoring in terms of the necessary system
 support. As an example of specialized authoring support, we have presented \snify, a
@@
 
-% LocalWords:  tbw Aarne Ranta homostem pre
+% LocalWords:  tbw Aarne Ranta homostem pre hmmm Vre ar al sn ify

### 2024-12-11 doc/cicm25-sysdesc/sysdesc.tex
abstracct

@@
 \begin{abstract}
-tbw
+  Semantic Annotations can be utilized for learning support services in technical
+  documents. The resulting active documents support more direct and personalized
+  interaction with the knowledge conveyed, but require a large investment in the authoring
+  process.
+
+  In this paper we discuss the nature of the semantic authoring problem and argue that
+  while sharing aspects of classical, informal authoring and formal authoring, it is a
+  problem of its own that needs special editing support facilities. We present a simple
+  utility that supports the process of annotating technical terms with semantic references
+  to a flexiformal content commons that enhances annotation productivity by almost an
+  order of magnitude.
 \end{abstract}

### 2024-12-11 doc/cicm25-sysdesc/sysdesc.tex
typos

@@
 approaches in practice. The producer task we use as a case study is that of adaptive
-learning assistants for MINT subjects, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.
+learning assistants for STEM subjects, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.
 \ALeA uses \sTeX \cite{MueKo:sdstex22,sTeX:github:on} -- a
@@
 Before we discuss semantic authoring in general, let us introduce a concrete example from
-the \sTeX/\ALeA world. We will use it to sharpen our intuitions of the problems an issues
+the \sTeX/\ALeA world. We will use it to sharpen our intuitions of the problems and issues
 involved and to ground the solution we propose.
@@
 \begin{lstlisting}[numbers=left,firstnumber=3,
-caption=The (unannotated) {\LaTeX} sources of \cref{fig:lo},label=lst:los]
+caption=The unannotated {\LaTeX} sources of \cref{fig:lo},label=lst:los]
 \item \blue{Note:} Depth-limited minimax requires an evaluation for every
@@
 \end{lstlisting}
-into the (partially) annotated item:
+into the partially annotated item:
 \begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
@@
 that it is a technical term, that the module from \cref{fig:state-space} exists, know the
-symbol name and its URL\ednote{path?}, and manage redundancy of the imports -- i.e. only
+symbol name and its URL\ednote{path/URI/identifier?}, and manage redundancy of the imports -- i.e. only
 adding the \lstinline|\importmodule| directive if it is not already (recursively) implied
@@
 
-\sTeX authoring is already supported by an VSCode IDE plugin \cite{sTeX-IDE:git} that
+\sTeX authoring is already supported by a VSCode IDE plugin \cite{sTeX-IDE:git} that
 analyzes annotations on the fly, displays the underlying reference semantics, reports
@@
 situation is very similar for authoring of formal knowledge, e.g. in program verification
-of the formalization of mathematics.
+or the formalization of mathematics.
 
@@
 In semantic authoring, e.g. the \sTeX sources of educational content for the \ALeA system
-presented above. The base of the text is classical narrative text to which most of the
+presented above, the base of the text is classical narrative text to which most of the
 constraints of classical authoring apply. The exception is the authoring-time tailoring to
@@
 
-In our experience, the step-through workflow for annotating term references for symbols
+In our experience, \snify's step-through workflow for annotating term references for symbols
 from the domain model is almost an order of magnitude more efficient than writing the
@@
 The command-line interface is simple and responsive and gives all the necessary
-information in a single glance if underlying shell area exceeds ca. $80\times 35$
+information in a single glance if the underlying shell area exceeds ca. $80\times 35$
 characters. It is very much geared towards annotating existing documents with respect to a

### 2024-12-11 doc/cicm25-sysdesc/sysdesc.tex
stats

@@
 search for alternative annotation targets,
-or view the file introducing one of the choices.
+or view/edit the file introducing one of the choices.
 
@@
 almost infeasible, and the average unfamiliarity naturally grows with the domain
-model. The symbol disambiguation process -- on average a word induces $n$\ednote{MK@FS: we
-  should try to measure this: for every verbalization (i.e. symbol/word pair) we should
-  compute the ratio of those words with the same stem.} choices is still manageable,
+model. The symbol disambiguation process -- on average a word induces 4.1 choices\ednote{
+    more details? median is 3, for existing annotations, the average is 2.6 and the median 2...
+} -- is still manageable,
 requires considerable concentration and domain knowledge, but little knowledge of the

### 2024-12-11 scripts/symbolstats.py

@@ -1,0 +1,68 @@
+import dataclasses
+import logging
+from collections import defaultdict
+
+from stextools.core.simple_api import get_linker, get_repos, get_symbols, SimpleFile
+from stextools.core.cache import Cache
+
+
+def prepare():
+    Cache.clear = lambda: None  # type: ignore
+    logging.getLogger('pylatexenc.latexwalker').setLevel(logging.WARNING)
+    # TODO: the linker indicates both real sTeX issues and missing features – we should not suppress them in general
+    logging.getLogger('stextools.core.linker').setLevel(logging.FATAL)
+    logging.basicConfig(level=logging.INFO)
+
+
+prepare()
+linker = get_linker()
+
+
+@dataclasses.dataclass
+class Stat:
+    symbols: int = 0   # number of declared symbols
+    defined_by_lang: dict[str, int] = dataclasses.field(default_factory=lambda: defaultdict(int))
+    annotated_refs: int = 0
+    alt_annotated_refs: int = 0
+
+
+def repo_to_category(repo):
+    if repo.name.startswith('smglom'):
+        return 'smglom'
+    if repo.name.endswith('course') or repo.name.endswith('ComSem') or repo.name.endswith('ComLog') or repo.name.endswith('TheoCS') or repo.name.endswith('eida') or repo.name.endswith('WebEng') or repo.name.endswith('lecture-notes'):
+        return 'courses'
+    if repo.name.endswith('problems'):
+        return 'problems'
+    elif 'talks' in repo.name:
+        return 'talks'
+    return 'other'
+
+
+stats = defaultdict(Stat)
+
+for repo in get_repos(linker):
+    category = repo_to_category(repo)
+    if category == 'other':
+        print('Unknown category:', repo.name)
+
+    for file in repo.files:
+        for symbol in file.declared_symbols:
+            stats[category].symbols += 1
+            for verb in symbol.get_verbalizations():
+                if verb.is_defining:
+                    stats[category].defined_by_lang[verb.lang] += 1
+        for verb in file.iter_verbalizations():
+            if verb.is_defining:
+                continue
+            stats[category].alt_annotated_refs += 1
+
+for symbol in get_symbols(linker):
+    for verb in symbol.get_verbalizations():
+        if verb.is_defining:
+            continue
+        stats[repo_to_category(verb.declaring_file.archive)].annotated_refs += 1
+
+
+
+
+print(stats)

### 2025-01-06 doc/cicm25-sysdesc/sysdesc.tex
paper

@@
 \newcommand\ALeA{\textsf{ALeA}\xspace}
+\newcommand\SMGloM{\textsf{SMGloM}\xspace}
 \newcommand\snify{\textsf{snify}\xspace}
@@
 
-\title{Bulk Semantic Annotation with a Partially Known Knowledge Base}
+\title{Bulk Semantic Annotation with a Partially Known Knowledge Base\ednote{
+    Too \snify specific...
+}}
 \author{Michael Kohlhase, Jan Frederik Schaefer}
@@
 \begin{abstract}
-  Semantic Annotations can be utilized for learning support services in technical
+  Semantic annotations can be utilized for learning support services in technical
   documents. The resulting active documents support more direct and personalized
@@
 \section{Introduction}
-Arguably, dealing with large document collections is one of the key factors in the
-knowledge-driven society and economy. There are currently two main contenders for machine
-support in this area. Symbolic/logic-based technologies and sub-symbolic,
-machine-learning-based AI, e.g.\ via LLMs or chatbots. They have complementary strengths
-and challenges: Symbolic technologies offer precision and explainability out of the box,
-but face scalability challenges because the prerequisite background knowledge has to be
-(manually) formalized. ML-based approaches can be trained on all data of the Internet, but
-face challenges in precision and explainability. In \cite{Ranta:atcp17},
-Aarne Ranta matches these profiles to the notion of \textbf{producer tasks} -- i.e.\ tasks
-where precision is key, but limited coverage ($10^{3\pm1}$ concepts) is OK;
-e.g.\ for multi-language/variant manuals for very
-expensive machines -- and \textbf{consumer tasks}, where coverage is key and
-precision secondary; e.g.\ for consumer-grade machine translation like \textsf{Google translate}.
+Arguably, dealing with large document collections is one of the key factors in the knowledge-driven society and economy.
+There are currently two main contenders for machine support in this area.
+Symbolic/logic-based technologies and sub-symbolic/machine-learning-based AI, e.g.\ via LLMs or chatbots.
+They have complementary strengths and challenges:
+ML-based approaches can be trained on all data of the Internet, but face challenges in precision and explainability. 
+Symbolic technologies, on the other hand, offer precision and explainability out of the box, but face scalability challenges
+-- in particular because of the cost of semantic authoring\ednote{
+    \emph{semantic authoring} already used in context of semantic web (RDFa).
+    Tools exist and it doesn't have exactly the same challenges as we face, though there is some overlap.
+    We have a module system and need more dependency management, something else?
+    This has to be described somewhere.
 
-In this paper, we address tools that help attack the coverage problem in symbolic/semantic
-approaches in practice. The producer task we use as a case study is that of adaptive
-learning assistants for STEM subjects, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.
-\ALeA uses \sTeX \cite{MueKo:sdstex22,sTeX:github:on} -- a
-variant of {\LaTeX} that allows to embed semantic annotations -- for knowledge
-representation, and generates learner-adaptive learning objects instrumented with
-learning-support interactions from that. The \textbf{\sTeX/\ALeA content commons}
-\ednote{MK@MK: continue to describe the size and distribution, and
-  authorship}\ednote{explain the \textbf{domain model} (the SMGloM) and the
-  \textbf{formulation model} (cf. \cite{BerBetChu:lssmkm23}) in the commons and give their
-  relative sizes.}
+    Related work: ``User Interfaces for Semantic Authoring of Textual Content: A Systematic Literature Review.''
+}, which we discuss in this paper.
 
-We contend that while education -- especially higher education -- as a whole is a
-wide-coverage task for society education of individual learners in specific domains --
-especially, if we want to tailor the educational offerings to said individual/domain -- is
-a producer task, at least until ML-based methods reach the precision and explainability
-required by the ethics of teaching.
+As a case study, we will use adaptive learning assistants for STEM subjects in higher education, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.
+We contend that machine-learning based technologies are largely unsuitable for this task -- at least until they reach the
+precision and explainability required by the ethics of teaching --
+while the scope of specific domains is just narrow enough that symbolic approaches are viable.
+Conventional (\LaTeX-based) lecture materials can essentially be directly imported into \ALeA,
+but its ability to generate learner-adaptive learning objects -- instrumented with learning support interactions --
+depends on semantic annotations.
+These can be embedded \LaTeX\ documents using the \sTeX package~\cite{MueKo:sdstex22,sTeX:github:on}.
+In this paper, we will mostly focus on annotating terms with semantic concept they refer to,
+but \sTeX supports a much wider range of annotations, such as annotating problems with the learning objectives
+or annotating the defining phrase (definiens) in concept definitions,
+and can, in principle, even be used for full formalization.\ednote{True?}
+As \sTeX supports the full spectrum from informal (unannotated) to fully formalized content,
+we also call \sTeX content \textbf{flexiformal}.
 
-\section{Semantic Authoring in a Content Commons}
+The \textbf{\sTeX/\ALeA content commons} consists of \SMGloM~\cite{GinIanJuc:spsttom16},
+the domain model for semantic annotations,
+as well as re-usable lecture materials such as lecture slides and problems -- mostly from the area of computer science.
+The sheer size of the commons is one of the key challenges for authors:
+\SMGloM contains definitions of $X$\ednote{TODO} STEM concepts in multiple languages,
+created and maintained by over 30 contributors.
+Additionally, there are roughly 4900\ednote{
+    Includes 440 from UPS, which seems unannotated... Other issues?
+    Command (tex comment): % grep -R --include='*.tex' -F -e '\begin{frame}' | wc -l
+}
+semantically annotated lecture slides and 4000 problems.
+
+\paragraph{Overview}
+We argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
+but is nevertheless a distinct problem requiring special tool support (\cref{sec:semauth}).
+Afterwards, we discuss (potential) tools in various stages of development (\cref{sec:tools}) and then
+describe and evaluate one concrete tool, \snify, that supports the annotation
+of terms with the semantic concepts they refer to (\cref{sec:snify,sec:eval}).
+The paper ends with a brief conclusion (\cref{sec:conclusion}).
+
+% % OLD INTRO
+% Arguably, dealing with large document collections is one of the key factors in the
+% knowledge-driven society and economy. There are currently two main contenders for machine
+% support in this area. Symbolic/logic-based technologies and sub-symbolic,
+% machine-learning-based AI, e.g.\ via LLMs or chatbots. They have complementary strengths
+% and challenges: Symbolic technologies offer precision and explainability out of the box,
+% but face scalability challenges because the prerequisite background knowledge has to be
+% (manually) formalized. ML-based approaches can be trained on all data of the Internet, but
+% face challenges in precision and explainability. In \cite{Ranta:atcp17},
+% Aarne Ranta matches these profiles to the notion of \textbf{producer tasks} -- i.e.\ tasks
+% where precision is key, but limited coverage ($10^{3\pm1}$ concepts) is OK;
+% e.g.\ for multi-language/variant manuals for very
+% expensive machines -- and \textbf{consumer tasks}, where coverage is key and
+% precision secondary; e.g.\ for consumer-grade machine translation like \textsf{Google translate}.
+% 
+% In this paper, we address tools that help attack the coverage problem in symbolic/semantic
+% approaches in practice. The producer task we use as a case study is that of adaptive
+% learning assistants for STEM subjects, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.
+% \ALeA uses \sTeX \cite{MueKo:sdstex22,sTeX:github:on} -- a
+% variant of {\LaTeX} that allows to embed semantic annotations -- for knowledge
+% representation, and generates learner-adaptive learning objects instrumented with
+% learning-support interactions from that. The \textbf{\sTeX/\ALeA content commons}
+% \ednote{MK@MK: continue to describe the size and distribution, and
+%   authorship}\ednote{explain the \textbf{domain model} (the SMGloM) and the
+%   \textbf{formulation model} (cf. \cite{BerBetChu:lssmkm23}) in the commons and give their
+%   relative sizes.}
+% 
+% We contend that while education -- especially higher education -- as a whole is a
+% wide-coverage task for society education of individual learners in specific domains --
+% especially, if we want to tailor the educational offerings to said individual/domain -- is
+% a producer task, at least until ML-based methods reach the precision and explainability
+% required by the ethics of teaching.
+
+\section{Semantic Authoring in a Content Commons}\label{sec:semauth}
 
@@
 \subsection{Running Example: Annotating  ``Terminal''}
-The main practical problem of annotating presentational {\LaTeX} course materials into
-semantic documents that carry enough information to support meaningful learning support
-services is to add semantic references to technical terms from the domain jargon.
-% TODO: This the following better?
-% With \sTeX, authors can add semantic references to technical terms,
-% which enables straight-forward learning support services like
-% showing the definition of a term, but also informs the system
-% about the content/prerequisites of a learning object.
-% Adding these references is a key practical problem when converting presentational {\LaTeX}
-% course materials into semantic documents.
+With \sTeX, authors can add semantic references to technical terms,
+which enables straight-forward learning support services like
+showing the definition of a term, but also informs the system
+about the content/prerequisites of a learning object.
+Adding these references is a key practical problem when converting presentational {\LaTeX}
+course materials into semantic documents.
+
+\begin{figure}[ht]\centering
+  \fbox{\includegraphics[width=10cm]{../img/minimax-remarks.pdf}}
+  \caption{An annotated learning object from an AI lecture}\label{fig:lo}
+\end{figure}
+
 Take for instance a learning object like the paragraph in \cref{fig:lo}, where we want to annotate
@@
 
-\begin{figure}[ht]\centering
-  \fbox{\includegraphics[width=10cm]{../img/minimax-remarks.pdf}}
-  \caption{An annotated learning object from an AI lecture}\label{fig:lo}
-\end{figure}
+% The main practical problem of annotating presentational {\LaTeX} course materials into
+% semantic documents that carry enough information to support meaningful learning support
+% services is to add semantic references to technical terms from the domain jargon.
 
@@
 and possibly removing directives that become redundant by the new one.  At ca 150,000
-words in a semester's words of lecture notes and ca. 10--15\%\ednote{can we actually
+words in a semester's worth of lecture notes and ca. 10--15\%\ednote{can we actually
   measure this for AI?} of technical terms a rather daunting task.
@@
 errors and redundancies, and even offers a concept search interface.
-But these features only support already annotated terms, so they do not solve the annotation problem above.
+But these features only support already annotated terms, so they do not solve the annotation problem above.\ednote{
+    last paragraph doesn't fit here anymore, as it only motivates \snify
+}
 
 \subsection{Semantic Authoring as a Distinct Problem}
-
 We contend that semantic authoring is a problem that is distinct from traditional
@@
 presented above, the base of the text is classical narrative text to which most of the
-constraints of classical authoring apply. The exception is the authoring-time tailoring to
-the competencies of the assumed reader, which can (partially) be automated by systems like
+constraints of classical authoring apply.
+The exception is the tailoring to the competencies of the assumed reader
+that traditionally happens during authoring, but which can (partially) be automated by systems like
 \ALeA. Note that it is not a restriction that \ALeA is initially targeted to tertiary
@@
 \cite{VreWelKam:tsmmdui24}. In this paper we present a tool that supports semantic
-annotation of technical terms via verbalizations.
+annotation of technical terms via verbalizations.\ednote{move this in the next section}
 
-\section{The \snify System}
+\section{Existing and Envisioned Authoring Support for \sTeX/\ALeA}\label{sec:tools}
+
+\section{The \snify System}\label{sec:snify}
 
@@
 
-\section{Practical Evaluation}
+\section{Practical Evaluation}\label{sec:eval}
 
@@
 
-\section{Conclusion}
+\section{Conclusion}\label{sec:conclusion}
 

### 2025-01-09 doc/cicm25-sysdesc/sysdesc.tex
more writing

@@
 The paper ends with a brief conclusion (\cref{sec:conclusion}).
+\ednote{rework this paragraph a bit}
 
@@
 
-\sTeX authoring is already supported by a VSCode IDE plugin \cite{sTeX-IDE:git} that
-analyzes annotations on the fly, displays the underlying reference semantics, reports
-errors and redundancies, and even offers a concept search interface.
-But these features only support already annotated terms, so they do not solve the annotation problem above.\ednote{
-    last paragraph doesn't fit here anymore, as it only motivates \snify
-}
+% \sTeX authoring is already supported by a VSCode IDE plugin \cite{sTeX-IDE:git} that
+% analyzes annotations on the fly, displays the underlying reference semantics, reports
+% errors and redundancies, and even offers a concept search interface.
+% But these features only support already annotated terms, so they do not solve the annotation problem above.\ednote{
+%     last paragraph doesn't fit here anymore, as it only motivates \snify
+% }
 
@@
 
-\paragraph{Semantic Authoring in \sTeX}
-How this can concretely work depends on the particular semantic framework, the specific
-annotations we want to support, and the semantics embedded in the content commons. In the
-\sTeX framework, where we semantically annotate formulae and technical terms with pointers
-to their definienda, we can bridge the presentation/semantics gap -- we annotate the words
-in sentences and glyphs in formulae with the verbalizations and notations in the content
-commons.\ednote{MK: maybe explain them again}. A tool for supporting semantic authoring of
-formulae via notations has been presented by Vre{\v{c}}ar et al. in
-\cite{VreWelKam:tsmmdui24}. In this paper we present a tool that supports semantic
-annotation of technical terms via verbalizations.\ednote{move this in the next section}
+% \paragraph{Semantic Authoring in \sTeX}
+% How this can concretely work depends on the particular semantic framework, the specific
+% annotations we want to support, and the semantics embedded in the content commons. In the
+% \sTeX framework, where we semantically annotate formulae and technical terms with pointers
+% to their definienda, we can bridge the presentation/semantics gap -- we annotate the words
+% in sentences and glyphs in formulae with the verbalizations and notations in the content
+% commons.\ednote{MK: maybe explain them again}. A tool for supporting semantic authoring of
+% formulae via notations has been presented by Vre{\v{c}}ar et al. in
+% \cite{VreWelKam:tsmmdui24}. In this paper we present a tool that supports semantic
+% annotation of technical terms via verbalizations.\ednote{move this in the next section}
 
-\section{Existing and Envisioned Authoring Support for \sTeX/\ALeA}\label{sec:tools}
+\section{Tool Support for Semantic Authoring in \sTeX}\label{sec:tools}
+Semantic authoring can ``inherit'' tool support from both informal
+and formal authoring.
+In the case of \sTeX, we can re-use spell checkers for traditional informal authoring.
+Additionally, there is a custom VSCode IDE plugin \cite{sTeX-IDE:git} that
+analyzes annotations on the fly, displays the underlying reference semantics, reports
+errors and redundancies, and even offers a concept search interface.
+But such features only support existing annotations and are, therefore,
+of limited use when creating new annotations, which arguably is the
+most time consuming aspects of semantic authoring.
+
+How the annotation process can be supported depends on the particular semantic framework,
+the specific annotations we want to support, and the semantics embedded in the content commons.
+In \sTeX/\ALeA, we primarily annotate formulae and technical terms with pointers to their definienda.
+A tool for supporting semantic authoring of formulae via notations has been presented by Vre{\v{c}}ar et al.\ in~\cite{VreWelKam:tsmmdui24}.
+It uses a grammar to parse informal, presentation-focused formulae (e.g.\ \lstinline|A\times B|) into abstract syntax trees based on
+semantic macros defined in \sTeX, which can then replace the original formula
+(e.g.\ \lstinline|\cart{A,B}| for the cartesian product of sets).
+In the case of ambiguity, authors can pick the correct reading in a
+graphical user interface that visualizes the different abstract syntax trees.
+
+Efficient support for the annotation of technical terms in natural language
+has been elusive so far.
+\ednote{
+    Move Hutter from conclusion here?
+    Also: Plivelic}
 

### 2025-01-10 doc/cicm25-sysdesc/sysdesc.tex
more writing

@@
+% KEY TODOs:
+%  * better title
+%  * turn overview into combined overview and contribution
+%  * Finish tool support section
+%  *  -> reduce future work paragraph in conclusion accordingly
+
 \documentclass{llncs}
@@
 Additionally, there are roughly 4900\ednote{
-    Includes 440 from UPS, which seems unannotated... Other issues?
+    Includes 440 from UPS, which seems unannotated...
+    Excludes talks as well...
+    Other issues?
     Command (tex comment): % grep -R --include='*.tex' -F -e '\begin{frame}' | wc -l
@@
 The paper ends with a brief conclusion (\cref{sec:conclusion}).
-\ednote{rework this paragraph a bit}
+\ednote{rework this paragraph a bit -- make it contribution and overview at the same time?}
 
@@
 most time consuming aspects of semantic authoring.
+\ednote{also: SPARQL API}
 
@@
 A tool for supporting semantic authoring of formulae via notations has been presented by Vre{\v{c}}ar et al.\ in~\cite{VreWelKam:tsmmdui24}.
-It uses a grammar to parse informal, presentation-focused formulae (e.g.\ \lstinline|A\times B|) into abstract syntax trees based on
+It uses a grammar to parse informal, presentation-focused formulae (e.g.\ \lstinline[keywordstyle={}]|A\times B|) into abstract syntax trees based on
 semantic macros defined in \sTeX, which can then replace the original formula
@@
 re-do all the \lstinline|s| skips. Alternatively, \snify can be called e.g. on the top-level
-lecture notes; then the session consists of all included files. Other productivity
+lecture notes; then the session consists of all (transitively) included files. Other productivity
 features include a focus mode that can be used to only annotate a particular word in the
@@
 One practical problem with \snify that remains is that the system only harvests
-symbol/verbalization pairs from \sTeX files on the local file system. This means that for
-effective annotation users (have to) download the whole content commons and keep it
+symbol/verbalization pairs from \sTeX files on the local file system. This means that, for
+effective annotation, users (have to) download the whole content commons and keep it
 updated. Unfortunately, the current MMT-based MathHub knowledge management
@@
 Finally, the current interface is not well-suited for on-the-fly annotation while
-authoring.  For that, the underlying information (the symbol/verbalization pairs harvested
+authoring.
+For that, the underlying information (the symbol/verbalization pairs harvested
 from the domain model) can be integrated into any IDE. In fact we plan to do this for the
 next version of the sTeX plugin for VSCode \cite{sTeX-IDE:git}.
+For now, the typical workflow of early adopters is to first write the document (e.g.\ a set of quiz questions) without annotations, and then run \snify to annotate the new content.
 
@@
 support. As an example of specialized authoring support, we have presented \snify, a
-command-line tool that bridges the gap between formal symbols in \sTeX and
+command-line tool that bridges the gap between informal text and formal symbols in \sTeX.
 

### 2025-01-27 doc/cicm25-sysdesc/sysdesc.tex

@@
 
-\title{Bulk Semantic Annotation with a Partially Known Knowledge Base\ednote{
-    Too \snify specific...
+\title{
+    Semantic Authoring in the \sTeX/\ALeA World
+    \ednote{Find better title. Was ``Bulk Semantic Annotation with a Partially Known Knowledge Base''
 }}
@@
 
-\paragraph{Overview}
-We argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
+% \paragraph{Overview \& Contribution}
+In this paper, we argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
 but is nevertheless a distinct problem requiring special tool support (\cref{sec:semauth}).
-Afterwards, we discuss (potential) tools in various stages of development (\cref{sec:tools}) and then
-describe and evaluate one concrete tool, \snify, that supports the annotation
+Using \sTeX/\ALeA as an exemplary case study,
+we will provide an overview of the authoring support (\cref{sec:tools})
+and then describe and evaluate one new, concrete tool, \snify, that supports the annotation
 of terms with the semantic concepts they refer to (\cref{sec:snify,sec:eval}).
 The paper ends with a brief conclusion (\cref{sec:conclusion}).
-\ednote{rework this paragraph a bit -- make it contribution and overview at the same time?}
 

### 2025-01-27 stextools/symnamer.py
--focus support in snify #76

@@ -1,0 +1,50 @@
+from pathlib import Path
+
+from pylatexenc.latexwalker import LatexWalker, LatexMathNode, LatexCommentNode, LatexSpecialsNode, LatexMacroNode, \
+    LatexEnvironmentNode, LatexCharsNode, LatexGroupNode
+
+from stextools.core.macros import STEX_CONTEXT_DB
+
+
+def quickcheck(filecontent: str) -> bool:
+    """ returns true iff the file may have to be processed, i.e. if it introduces symbols """
+    if 'symdecl' in filecontent or 'symdef' in filecontent:
+        return True
+    return False
+
+
+def recurse(nodes):
+    for node in nodes:
+        if node is None or node.nodeType() in {LatexMathNode, LatexCommentNode, LatexSpecialsNode, LatexCharsNode}:
+            continue
+        if node.nodeType() == LatexMacroNode:
+            if node.macroname in MACRO_RECURSION_RULES:
+                for arg_idx in MACRO_RECURSION_RULES[node.macroname]:
+                    if arg_idx >= len(node.nodeargd.argnlist):
+                        click.clear()
+                        standard_header('Error', bg='red')
+                        print(f"Macro {node.macroname} does not have argument {arg_idx}")
+                        print('Context:')
+                        print(latex_text[max(node.pos - 100, 0):min(len(latex_text) - 1, node.pos + 300)])
+                        print()
+                        print('Please report this error')
+                        click.pause()
+                        continue
+                    _recurse([node.nodeargd.argnlist[arg_idx]])
+        elif node.nodeType() in {LatexEnvironmentNode, LatexGroupNode}:
+            recurse(node.nodelist)
+        else:
+            raise RuntimeError(f"Unexpected node type: {node.nodeType()}")
+
+
+
+def symname(file: Path):
+    with open(file) as fp:
+        filecontent = fp.read()
+
+    if not quickcheck(filecontent):
+        return
+
+    walker = LatexWalker(filecontent, latex_context=STEX_CONTEXT_DB)
+    nodes = walker.get_latex_nodes()[0]
+

### 2025-01-27 stextools/__main__.py
--focus support in snify #76

@@ -96,9 +96,11 @@ def translate(path):
 @click.option('--ignore',
               default=lambda: get_config().get('stextools.snify', 'ignore', fallback=None),
               help='Pattern to exclude some archives (e.g. \'Papers/*,smglom/mv\')')
-def snify_actual(files, filter, ignore):
+@click.option('--focus',
+              help='Immediately focus on a specific word')
+def snify_actual(files, filter, ignore, focus):
     from stextools.snify.controller import snify
-    snify(files, filter, ignore)
+    snify(files, filter, ignore, focus)
 
 
 @cli.command(name='version', help='Print the version of stextools.')

### 2025-01-27 stextools/snify/controller.py

@@ -161,7 +161,7 @@ class StateSkipModification(Modification):
 
 
 class Controller:
-    def __init__(self, state: State, new_files: Optional[list[Path]] = None):
+    def __init__(self, state: State, new_files: Optional[list[Path]] = None, stem_focus: Optional[str] = None):
         self.state: State = state
         self.mh = Cache.get_mathhub(update_all=True)
         self._linker: Optional[Linker] = None
@@ -170,9 +170,10 @@ class Controller:
         self._modification_future: list[list[Modification]] = []   # for re-doing
 
         if new_files:
-            self.load_files_dialog(new_files)
+            self.load_files_dialog(new_files, stem_focus)
 
-    def load_files_dialog(self, new_files: list[Path]):
+    def load_files_dialog(self, new_files: list[Path], stem_focus: Optional[str]):
+        # note: better design would be to move this out of the controller, I think
         have_inputs = False
         for file in new_files:
             stexdoc = self.mh.get_stex_doc(file)
@@ -215,9 +216,9 @@ class Controller:
                 else:
                     print(f'File {path} is not loaded')
 
-            self.state.push_focus(new_files=all_files)
+            self.state.push_focus(new_files=all_files, select_only_stem=stem_focus)
         else:
-            self.state.push_focus(new_files=new_files)
+            self.state.push_focus(new_files=new_files, select_only_stem=stem_focus)
 
     @property
     def linker(self) -> Linker:
@@ -499,9 +500,11 @@ class Controller:
         return self.state.get_current_lang(self.linker)
 
 
-def snify(files: list[str], filter: str, ignore: str):
+def snify(files: list[str], filter: str, ignore: str, focus: Optional[str]):
     session_storage = SessionStorage()
-    state = session_storage.get_session_dialog()
+    state: Optional[State] = None
+    if state is None and focus is None:
+        state = session_storage.get_session_dialog()
     if state is None:
         state = State(files=[], filter_pattern=filter,
                       ignore_pattern=ignore, cursor=PositionCursor(file_index=0, offset=0))
@@ -512,7 +515,7 @@ def snify(files: list[str], filter: str, ignore: str):
                 paths.extend(path.rglob('*.tex'))
             else:
                 paths.append(path)
-        controller = Controller(state, new_files=paths)
+        controller = Controller(state, new_files=paths, stem_focus=focus)
     else:
         controller = Controller(state)
     unfinished = controller.run()

### 2025-02-02 doc/cicm25-sysdesc/sysdesc.tex
more

@@
 \usepackage{stex-logo}
+\usepackage{paralist}
 %\usepackage{lststex} debug that the gray is not solid
@@
 and can, in principle, even be used for full formalization.\ednote{True?}
-As \sTeX supports the full spectrum from informal (unannotated) to fully formalized content,
+As \sTeX supports the full spectrum from informal (un-annotated) to fully formalized content,
 we also call \sTeX content \textbf{flexiformal}.
@@
 The \textbf{\sTeX/\ALeA content commons} consists of \SMGloM~\cite{GinIanJuc:spsttom16},
-the domain model for semantic annotations,
-as well as re-usable lecture materials such as lecture slides and problems -- mostly from the area of computer science.
-The sheer size of the commons is one of the key challenges for authors:
-\SMGloM contains definitions of $X$\ednote{TODO} STEM concepts in multiple languages,
-created and maintained by over 30 contributors.
-Additionally, there are roughly 4900\ednote{
-    Includes 440 from UPS, which seems unannotated...
-    Excludes talks as well...
-    Other issues?
-    Command (tex comment): % grep -R --include='*.tex' -F -e '\begin{frame}' | wc -l
-}
-semantically annotated lecture slides and 4000 problems.
+the domain model for semantic annotations, as well as re-usable lecture materials such as
+lecture slides and problems -- mostly from the area of computer science.  The sheer size
+of the commons is one of the key challenges for authors: \SMGloM contains definitions of
+$X$\ednote{TODO} STEM concepts in multiple languages, created and maintained by over 30
+contributors.  Additionally, there are roughly 4900\ednote{ Includes 440 from UPS, which
+  seems un-annotated...  Excludes talks as well...  Other issues?  Command (tex
+  comment): % grep -R --include='*.tex' -F -e '\begin{frame}' | wc -l
+} semantically annotated lecture slides and 4000 problems.
 
@@
 
-Concretely this is about converting the {\LaTeX} string
+Concretely this is about converting the {\LaTeX} string\ednote{MK: There is/was a discrepancy
+  in the last line of the example; I tried to fix it, but failed}
 \begin{lstlisting}[numbers=left,firstnumber=3,
@@
 \begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
-caption=Annotating ``terminal'' in \cref{lst:los},label=lst:losa]
+caption=Annotating ``terminal'' and ``utility'' in \cref{lst:los},label=lst:losa]
 \importmodule[smglom/search]{mod?search-problem}
@@
 \item \blue{Note:} Depth-limited minimax requires an evaluation for every
-cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its utility,
+cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its \sn{utility},
 and otherwise an estimate.
 \end{lstlisting}
-% Concretely, the author has to remember the module that introduces the concept
-% of terminal states (\cref{fig:state-space}),
-% the symbol name \lstinline|goal state|, annotate it with the
-% \sTeX macro \lstinline[mathescape]|\sr{$\llangle symbol name \rrangle$}{$\llangle verbalization\rrangle$}|, and add the module import in the first line (unless it
-% is already imported).
-% If the symbol name and the desired verbalization coincide,
-% the short-hand macro \lstinline|\sn| can be used (e.g. \lstinline|\sn{utility}|).
+In the latter, the author has the word ``terminal'' with the concept it refers to with 
+via the \sTeX macro
+\lstinline[mathescape]|\sr{$\llangle URI \rrangle$}{$\llangle verbalization\rrangle$}|.
 
-% In this case the symbol name happens to be the same as its English
-% verbalization, otherwise we would have to use
-% \lstinline[mathescape]|\sr{$\llangle symbol name\rrangle$}{state space}|.
-The annotation refers to a definition in the domain model module shown in \cref{fig:state-space}.
-The \lstinline|\definame| and \lstinline|\definiendum| in lines 5--6 introduce the \textbf{verbalizations} ``goal state'' and ``terminal state'' for the symbol \lstinline|search-problem?goal state| introduced by \lstinline|\symdecl*| declaration in line 3.
+\begin{newpart}{MK: more definitions}
+  In \sTeX, objects and concepts are represented by \textbf{symbols}, which are identified
+globally and unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module
+  name}, and a \textbf{symbol name}. Every symbol can have one or more
+\textbf{verbalizations}: words that of the domain jargon used to refer to the concept or
+object. Verbalizations can vary between between subdomains and communities of practice. In
+our example, the verbalization ``terminal [state]'' was annotated with the symbol with
+name \lstinline|goal state|. It comes from the module with name \lstinline|search-problem|
+in the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
+macro in line 1. As the symbol name \lstinline|goal state| is unique in the imports here
+it suffices as a relative URI in the \lstinline|\sr| annotation.
+
+If the symbol name and the desired verbalization coincide (as they often do in English),
+the short-hand macro \lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| in
+\cref{lst:losa}).
+\end{newpart}
+
+The annotation refers to a definition in the domain model module shown in
+\cref{fig:state-space}.  The \lstinline|\definame| and \lstinline|\definiendum| in lines
+5--6 introduce the verbalizations ``goal state'' and ``terminal state'' for the symbol
+\lstinline|search-problem?goal state| introduced by \lstinline|\symdecl*| declaration in
+line 3.
 
@@
 
-To annotate the string ``terminal'' in the learning object, the author has to be aware
-that it is a technical term, that the module from \cref{fig:state-space} exists, know the
-symbol name and its URL\ednote{path/URI/identifier?}, and manage redundancy of the imports -- i.e. only
-adding the \lstinline|\importmodule| directive if it is not already (recursively) implied
-and possibly removing directives that become redundant by the new one.  At ca 150,000
-words in a semester's worth of lecture notes and ca. 10--15\%\ednote{can we actually
-  measure this for AI?} of technical terms a rather daunting task.
+\begin{newpart}{MK: trying to make some general remark about annotation before we go into
+    semantic authoring}
+\subsection{Annotation Workflows and Requirements}
 
-% \sTeX authoring is already supported by a VSCode IDE plugin \cite{sTeX-IDE:git} that
-% analyzes annotations on the fly, displays the underlying reference semantics, reports
-% errors and redundancies, and even offers a concept search interface.
-% But these features only support already annotated terms, so they do not solve the annotation problem above.\ednote{
-%     last paragraph doesn't fit here anymore, as it only motivates \snify
-% }
+To annotate the word ``terminal'' in the document, the author has to be aware that it is a
+technical term, that the module from \cref{fig:state-space} exists, know the symbol URL,
+and manage redundancy of the imports -- i.e. only adding the \lstinline|\importmodule|
+directive if it is not already (recursively) implied and possibly removing directives that
+become redundant by the new one.  At ca 150,000 words in a semester's worth of lecture
+notes and ca. 10--15\% of technical terms a rather daunting task.
+
+While annotation needs to happen during authoring and curation of documents, the majority
+of annotation tasks in practice concern the semantic augmentation of existing documents
+(\textbf{bulk annotation}) with respect to an existing semantic domain model. In any case,
+it is probably useful to separate the roles of author and semantic annotator conceptually,
+even though they may well be the same person in practice.
+
+The most important resource for annotators is a verbalization to symbol URI mapping -- we
+call it the \textbf{annotation catalog} -- either implicitly in the mind of the annotator
+or explicitly as a technical artifact, which allows to
+\begin{inparaenum}[\em i\rm)]
+\item identify the known technical terms,
+\item to map them to the corresponding symbols, and 
+\item manage the necessary import directives.
+\end{inparaenum}
+This catalog can be harvested from the \textbf{formal declarations}
+(\lstinline|\definiendum|, \lstinline|\begin|/\lstinline|\end{smodule}| and
+\lstinline|\importmodule|, etc.) of the domain model and existing documents in the
+semantic corpus.
+
+In $L$-language documents, where $L$ is not English the annotation practices are a bit
+more complex still, in particular because the annotation catalog is usually incomplete.
+
+Note that for $L$-annotation, we need a $L$-specific annotation catalog. 
+\sTeX already supports multilinguality:
+\begin{itemize}
+\item Documents have explicit language annotations (usually in form of the file extension
+  \lstinline[mathescape]|.$\llangle lang\rrangle$.tex| where $\llangle lang\rrangle$ is the
+  ISO-639 language identifier.)
+\item Modules in the domain model allow $L$-translations that share the formal
+  declarations with another module. In particular all definienda in $L$-modules are
+  $L$-specific.
+\end{itemize}
+
+But still, $L$-annotation catalogs are usually inadequate for $L$-annotation tasks in
+practice, as
+\begin{enumerate}[\em i\rm)]
+\item $L$-verbalizations and symbol names do not often coincide and
+\item the domain model may not even have $L$-modules -- especially for the ``smaller''
+  languages
+\end{enumerate}
+ In this situation, we usually have to invest in
+establishing a suitable catalog e.g. by providing $L$-translations (or at least dictionary
+information) in the domain model or delineating modules and annotating definienda that the
+domain model does not cover in the documents to be annotated.
+\end{newpart}
 
@@
 \section{Tool Support for Semantic Authoring in \sTeX}\label{sec:tools}
-Semantic authoring can ``inherit'' tool support from both informal
-and formal authoring.
+Semantic authoring can ``inherit'' tool support from both informal and formal authoring.
 In the case of \sTeX, we can re-use spell checkers for traditional informal authoring.
-Additionally, there is a custom VSCode IDE plugin \cite{sTeX-IDE:git} that
-analyzes annotations on the fly, displays the underlying reference semantics, reports
-errors and redundancies, and even offers a concept search interface.
-But such features only support existing annotations and are, therefore,
-of limited use when creating new annotations, which arguably is the
-most time consuming aspects of semantic authoring.
-\ednote{also: SPARQL API}
+Additionally, there is a custom VSCode IDE plugin \cite{sTeX-IDE:git} that analyzes
+annotations on the fly, displays the underlying reference semantics, reports errors and
+redundancies, and even offers a concept search interface.  But such features only support
+existing annotations and are, therefore, of limited use when creating new annotations,
+which arguably is the most time consuming aspects of semantic authoring.  \ednote{also:
+  SPARQL API}
 
 How the annotation process can be supported depends on the particular semantic framework,
-the specific annotations we want to support, and the semantics embedded in the content commons.
-In \sTeX/\ALeA, we primarily annotate formulae and technical terms with pointers to their definienda.
-A tool for supporting semantic authoring of formulae via notations has been presented by Vre{\v{c}}ar et al.\ in~\cite{VreWelKam:tsmmdui24}.
-It uses a grammar to parse informal, presentation-focused formulae (e.g.\ \lstinline[keywordstyle={}]|A\times B|) into abstract syntax trees based on
-semantic macros defined in \sTeX, which can then replace the original formula
-(e.g.\ \lstinline|\cart{A,B}| for the cartesian product of sets).
-In the case of ambiguity, authors can pick the correct reading in a
-graphical user interface that visualizes the different abstract syntax trees.
+the specific annotations we want to support, and the semantics embedded in the content
+commons.  In \sTeX/\ALeA, we primarily annotate formulae and technical terms with pointers
+to their definienda.  A tool for supporting semantic authoring of formulae via notations
+has been presented by Vre{\v{c}}ar et al.\ in~\cite{VreWelKam:tsmmdui24}.  It uses a
+grammar to parse informal, presentation-focused formulae (e.g.\
+\lstinline[keywordstyle={}]|A\times B|) into abstract syntax trees based on semantic
+macros defined in \sTeX, which can then replace the original formula (e.g.\
+\lstinline|\cart{A,B}| for the Cartesian product of sets).  In the case of ambiguity,
+authors can pick the correct reading in a graphical user interface that visualizes the
+different abstract syntax trees.
 
@@
 
-% LocalWords:  tbw Aarne Ranta homostem pre hmmm Vre ar al sn ify
+% LocalWords:  tbw Aarne Ranta homostem pre hmmm Vre ar al sn ify TODO et Hutter Plivelic
+% LocalWords:  lang

### 2025-05-04 doc/cicm25-sysdesc/sysdesc.tex
better title
work on paper

@@
 \usepackage{paralist}
+\usepackage{wrapfig}
 %\usepackage{lststex} debug that the gray is not solid
@@
 \newcommand\snify{\textsf{snify}\xspace}
+\newcommand\defianno{\textsf{defianno}\xspace}
 \def\llangle{\langle\kern-.2em\langle}
@@
 \title{
-    Semantic Authoring in the \sTeX/\ALeA World
-    \ednote{Find better title. Was ``Bulk Semantic Annotation with a Partially Known Knowledge Base''
-}}
+    % Semantic Authoring in the \sTeX/\ALeA World
+    % \ednote{Find better title. Was ``Bulk Semantic Annotation with a Partially Known Knowledge Base''}
+Semantic authoring in a flexiformal context --- Bulk annotation of rigorous documents
+}
 \author{Michael Kohlhase, Jan Frederik Schaefer}
@@
 \begin{abstract}
-  Semantic annotations can be utilized for learning support services in technical
-  documents. The resulting active documents support more direct and personalized
-  interaction with the knowledge conveyed, but require a large investment in the authoring
-  process.
+%   Semantic annotations can be utilized for learning support services in technical
+%   documents. The resulting active documents support more direct and personalized
+%   interaction with the knowledge conveyed, but require a large investment in the authoring
+%   process.
+% 
+%   In this paper we discuss the nature of the semantic authoring problem and argue that
+%   while sharing aspects of classical, informal authoring and formal authoring, it is a
+%   problem of its own that needs special editing support facilities. We present a simple
+%   utility that supports the process of annotating technical terms with semantic references
+%   to a flexiformal content commons that enhances annotation productivity by almost an
+%   order of magnitude.
 
-  In this paper we discuss the nature of the semantic authoring problem and argue that
-  while sharing aspects of classical, informal authoring and formal authoring, it is a
-  problem of its own that needs special editing support facilities. We present a simple
-  utility that supports the process of annotating technical terms with semantic references
-  to a flexiformal content commons that enhances annotation productivity by almost an
-  order of magnitude.
+    Semantic annotations enable services that make the knowledge
+    in documents more accessible.
+    In the case of \ALeA, a learning assistant we use as a case study,
+    they enable, for example, guided tours and practice problems
+    tailored to the learner's competencies.
+    However, the cost of semantic authoring is substantial
+    and tool support is limited.
+
+    In this paper we discuss the nature of the semantic authoring
+    problem and argue that, while sharing aspects of classical
+    (informal) authoring and formal authoring,
+    it is a problem of its own that needs special editing support facilities.
+    We present a simple utility that supports the process of annotating
+    technical terms with semantic references to a flexiformal content commons
+    that enhances annotation productivity by almost an order of magnitude.
 \end{abstract}
@@
 Symbolic technologies, on the other hand, offer precision and explainability out of the box, but face scalability challenges
--- in particular because of the cost of semantic authoring\ednote{
-    \emph{semantic authoring} already used in context of semantic web (RDFa).
-    Tools exist and it doesn't have exactly the same challenges as we face, though there is some overlap.
-    We have a module system and need more dependency management, something else?
-    This has to be described somewhere.
-
-    Related work: ``User Interfaces for Semantic Authoring of Textual Content: A Systematic Literature Review.''
-}, which we discuss in this paper.
+-- in particular because of the cost of semantic authoring, which we discuss in this paper.
 
@@
 These can be embedded \LaTeX\ documents using the \sTeX package~\cite{MueKo:sdstex22,sTeX:github:on}.
-In this paper, we will mostly focus on annotating terms with semantic concept they refer to,
+In this paper, we will mostly focus on annotating terms with the semantic concepts they refer to,
 but \sTeX supports a much wider range of annotations, such as annotating problems with the learning objectives
@@
 As \sTeX supports the full spectrum from informal (un-annotated) to fully formalized content,
-we also call \sTeX content \textbf{flexiformal}.
+we say that \sTeX content is \textbf{flexiformal}.
 
 The \textbf{\sTeX/\ALeA content commons} consists of \SMGloM~\cite{GinIanJuc:spsttom16},
-the domain model for semantic annotations, as well as re-usable lecture materials such as
-lecture slides and problems -- mostly from the area of computer science.  The sheer size
-of the commons is one of the key challenges for authors: \SMGloM contains definitions of
-$X$\ednote{TODO} STEM concepts in multiple languages, created and maintained by over 30
-contributors.  Additionally, there are roughly 4900\ednote{ Includes 440 from UPS, which
-  seems un-annotated...  Excludes talks as well...  Other issues?  Command (tex
-  comment): % grep -R --include='*.tex' -F -e '\begin{frame}' | wc -l
-} semantically annotated lecture slides and 4000 problems.
+a glossary acting as the domain model for semantic annotations,
+as well as re-usable lecture materials such as lecture slides and problems.
+% -- mostly from the area of computer science.
+The sheer size
+of the commons is one of the key challenges for authors: \SMGloM contains definitions of over
+5400 concepts in multiple languages, created and maintained by over 30
+contributors.  Additionally, there are roughly 4900
+semantically annotated lecture slides and 4000 problems
+-- mostly from the area of computer science.
+% grep -R --include='*.tex' -F -e '\begin{frame}' | wc -l
+
+\ednote{
+    \textbf{Related work: semantic web authoring}
+    \emph{semantic authoring} already used in context of semantic web (RDFa).
+    Tools exist and it doesn't have exactly the same challenges as we face, though there is some overlap.
+    We have a module system and need more dependency management, something else?
+    This has to be described somewhere.
+
+    Related work: ``User Interfaces for Semantic Authoring of Textual Content: A Systematic Literature Review.''
+}
 
@@
 
-% % OLD INTRO
-% Arguably, dealing with large document collections is one of the key factors in the
-% knowledge-driven society and economy. There are currently two main contenders for machine
-% support in this area. Symbolic/logic-based technologies and sub-symbolic,
-% machine-learning-based AI, e.g.\ via LLMs or chatbots. They have complementary strengths
-% and challenges: Symbolic technologies offer precision and explainability out of the box,
-% but face scalability challenges because the prerequisite background knowledge has to be
-% (manually) formalized. ML-based approaches can be trained on all data of the Internet, but
-% face challenges in precision and explainability. In \cite{Ranta:atcp17},
-% Aarne Ranta matches these profiles to the notion of \textbf{producer tasks} -- i.e.\ tasks
-% where precision is key, but limited coverage ($10^{3\pm1}$ concepts) is OK;
-% e.g.\ for multi-language/variant manuals for very
-% expensive machines -- and \textbf{consumer tasks}, where coverage is key and
-% precision secondary; e.g.\ for consumer-grade machine translation like \textsf{Google translate}.
-% 
-% In this paper, we address tools that help attack the coverage problem in symbolic/semantic
-% approaches in practice. The producer task we use as a case study is that of adaptive
-% learning assistants for STEM subjects, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.
-% \ALeA uses \sTeX \cite{MueKo:sdstex22,sTeX:github:on} -- a
-% variant of {\LaTeX} that allows to embed semantic annotations -- for knowledge
-% representation, and generates learner-adaptive learning objects instrumented with
-% learning-support interactions from that. The \textbf{\sTeX/\ALeA content commons}
-% \ednote{MK@MK: continue to describe the size and distribution, and
-%   authorship}\ednote{explain the \textbf{domain model} (the SMGloM) and the
-%   \textbf{formulation model} (cf. \cite{BerBetChu:lssmkm23}) in the commons and give their
-%   relative sizes.}
-% 
-% We contend that while education -- especially higher education -- as a whole is a
-% wide-coverage task for society education of individual learners in specific domains --
-% especially, if we want to tailor the educational offerings to said individual/domain -- is
-% a producer task, at least until ML-based methods reach the precision and explainability
-% required by the ethics of teaching.
-
 \section{Semantic Authoring in a Content Commons}\label{sec:semauth}
@@
 Before we discuss semantic authoring in general, let us introduce a concrete example from
-the \sTeX/\ALeA world. We will use it to sharpen our intuitions of the problems and issues
+the \sTeX/\ALeA world. We will use it to sharpen our intuitions of the challenges
 involved and to ground the solution we propose.
@@
 
-% The main practical problem of annotating presentational {\LaTeX} course materials into
-% semantic documents that carry enough information to support meaningful learning support
-% services is to add semantic references to technical terms from the domain jargon.
-
-Concretely this is about converting the {\LaTeX} string\ednote{MK: There is/was a discrepancy
-  in the last line of the example; I tried to fix it, but failed}
+Concretely this is about converting the {\LaTeX} string
+    (for conciseness completely unannotated)
+    \ednote{MK: There is/was a discrepancy
+  in the last line of the example; I tried to fix it, but failed\\
+    FS: Is it fixed now? I don't see any discrepancy.}
 \begin{lstlisting}[numbers=left,firstnumber=3,
@@
 \item \blue{Note:} Depth-limited minimax requires an evaluation for every
-cut-off state $s$. If $s$ is terminal, we use its utility, and otherwise
-an estimate.
+cut-off state $s$. If $s$ is terminal, we use its utility,
+and otherwise an estimate.
 \end{lstlisting}
@@
 \begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
-caption=Annotating ``terminal'' and ``utility'' in \cref{lst:los},label=lst:losa]
+caption=Annotating ``terminal'' in \cref{lst:los},label=lst:losa]
 \importmodule[smglom/search]{mod?search-problem}
@@
 \item \blue{Note:} Depth-limited minimax requires an evaluation for every
-cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its \sn{utility},
+cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its utility,
 and otherwise an estimate.
@@
   name}, and a \textbf{symbol name}. Every symbol can have one or more
-\textbf{verbalizations}: words that of the domain jargon used to refer to the concept or
-object. Verbalizations can vary between between subdomains and communities of practice. In
+\textbf{verbalizations}:
+words of the domain jargon used to refer to the concept or object.
+Verbalizations can vary between between subdomains and communities of practice. In
 our example, the verbalization ``terminal [state]'' was annotated with the symbol with
@@
 If the symbol name and the desired verbalization coincide (as they often do in English),
-the short-hand macro \lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| in
-\cref{lst:losa}).
+the short-hand macro \lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| for ``utility'' in \cref{lst:losa}).\ednote{
+    FS: Do we need the last paragraph?
+        I've removed the ``sn utility'' annotation from the listing
+        because I thought it breaks the flow a bit.
+        But if we want to show the sn macro, I can add it back.
+}
 \end{newpart}
@@
   \fbox{\includegraphics[width=12cm]{../img/search-problem.en.pdf}}
-\begin{lstlisting}[morekeywords={definame,symdecl,definiendum},numbers=left]
+\begin{lstlisting}[morekeywords={definame,symdecl,definiendum},numbers=left,
+escapechar=!]
 \begin{smodule}[title=Search Problem]{search-problem}
@@
     semantic authoring}
-\subsection{Annotation Workflows and Requirements}
+    \subsection{Annotation Workflows and Requirements}\label{sec:workflows}
 
 To annotate the word ``terminal'' in the document, the author has to be aware that it is a
-technical term, that the module from \cref{fig:state-space} exists, know the symbol URL,
+technical term, that the module from \cref{fig:state-space} exists, know the symbol URI,
 and manage redundancy of the imports -- i.e. only adding the \lstinline|\importmodule|
@@
 This catalog can be harvested from the \textbf{formal declarations}
-(\lstinline|\definiendum|, \lstinline|\begin|/\lstinline|\end{smodule}| and
+(\lstinline|\definiendum|, \lstinline|\sr|, \lstinline[language={}]|\begin|/\lstinline[language={}]|\end{smodule}| and
 \lstinline|\importmodule|, etc.) of the domain model and existing documents in the
@@
 \end{newpart}
+\ednote{FS: Todo: verbalizations can also be
+inferred from existing annotations (without any $L$-modules).}
 
@@
 
-\section{Tool Support for Semantic Authoring in \sTeX}\label{sec:tools}
+\section{Tool Support for Semantic Authoring in \sTeX/\ALeA}\label{sec:tools}
 Semantic authoring can ``inherit'' tool support from both informal and formal authoring.
 In the case of \sTeX, we can re-use spell checkers for traditional informal authoring.
-Additionally, there is a custom VSCode IDE plugin \cite{sTeX-IDE:git} that analyzes
+There also is prototypical support for authoring \ALeA content in Microsoft Word
+with a custom plugin (see \cite{KohKoh:woide24}),
+which allows authors to use Word's spell checker and grammar checker.
+
+On the formal side, there is a custom VSCode IDE plugin for \sTeX \cite{sTeX-IDE:git} that analyzes
 annotations on the fly, displays the underlying reference semantics, reports errors and
-redundancies, and even offers a concept search interface.  But such features only support
+redundancies, and even offers a concept search interface.
+But such features only support
 existing annotations and are, therefore, of limited use when creating new annotations,
-which arguably is the most time consuming aspects of semantic authoring.  \ednote{also:
-  SPARQL API}
+which arguably is the most time consuming aspects of semantic authoring.
+% \ednote{also: SPARQL API}
 
@@
 has been elusive so far.
-\ednote{
-    Move Hutter from conclusion here?
-    Also: Plivelic}
 
 \section{The \snify System}\label{sec:snify}
-
-The \snify utility (from ``\texttt{\textbackslash sn}-ify'' for adding term references;
-see \cite{stextools:git}) is a simple command-line tool that creates a database of
+The \snify\footnote{
+From ``\texttt{\textbackslash sn}-ify''; ``\texttt{\textbackslash sn}'' is
+a frequently used \sTeX short-hand macro for term annotations.}
+utility \cite{stextools:git}
+is a simple command-line tool that creates a catalog of
 symbol-verbalization pairs, analyzes \sTeX source files, and then steps through all word
-occurrences in the document that match a verbalization in the database.  Matching is based
+occurrences in the document that match a verbalization in the catalog.  Matching is based
 on word stems using off-the-shelf stemmers, which allows us to e.g.\ match ``utilities''
@@
 and interactions that allow to fine-tune the wordwise annotation
-workflow. \Cref{fig:snify} shows a typical situation.
+workflow.
+The interface is inspired by traditional spell checkers like \lstinline|ispell|.
+\Cref{fig:snify} shows a typical situation.
 
@@
 
-\section{Practical Evaluation}\label{sec:eval}
+As a companion to \snify,
+we have \defianno, a prototype tool for annotating definienda in documents
+that were originally authored without semantic annotations.
+It can be configured to use different \LaTeX\ macros
+to identify definienda (e.g.\ \lstinline|\emph| or \lstinline|\textbf|),
+and then steps through them similarly to \snify.
+For each potential definiendum, the user can then use fuzzy search to find
+the corresponding symbol in the domain model.
+This populates the catalog with verbalizations relevant to the document,
+making subsequent \snify runs more effective,
+especially for documents in languages that are not well covered by the
+domain model (i.e.\ not English).
 
+\section{Practical Evaluation and Future Work}\label{sec:eval}
+\begin{wrapfigure}{r}{0.4\textwidth}
+  \centering
+  \includegraphics[width=0.4\textwidth]{../img/annocounts.pdf}
+  \caption{Number of annotated terms in the \sTeX/\ALeA commons over time.}\label{fig:annocounts}
+\end{wrapfigure}
 In our experience, \snify's step-through workflow for annotating term references for symbols
@@
 annotations and imports by hand, even for annotators who are familiar with the domain
-model. For annotators unfamiliar with the domain model, the unassisted annotation task is
+model.
+Given the sheer number of term annotations (see \cref{fig:annocounts}),
+this results in a considerable productivity gain.
+For annotators unfamiliar with the domain model, the unassisted annotation task is
 almost infeasible, and the average unfamiliarity naturally grows with the domain
-model. The symbol disambiguation process -- on average a word induces 4.1 choices\ednote{
-    more details? median is 3, for existing annotations, the average is 2.6 and the median 2...
-} -- is still manageable,
+model. The symbol disambiguation process -- on average a word induces 4.1 choices
+% \ednote{
+%     more details? median is 3, for existing annotations, the average is 2.6 and the median 2...
+% }
+-- is still manageable,
 requires considerable concentration and domain knowledge, but little knowledge of the
@@
 For less complete domain models we have to skip too many terms that should ultimately be
-annotated and annotation efficiency suffers. This is currently the case for all
+annotated and annotation efficiency suffers.\ednote{
+    FS: Isn't this wrong? We don't have to skip them, the problem is that \snify skips them.
+    Efficiency does not suffer, but rather effectiveness.
+}
+This is currently the case for all
 non-English languages in the \sTeX corpus we work with. Coverage in German is only about
-half of that of English and we can already see the practical effects. We have also
-experimented with a Slovene introductory math book, and it seems clear that apart from
-having to introduce a Slovene stemmer\ednote{MK@FS: there is one for python at
-  \url{https://repo.ijs.si/pboskoski/slo_stemmer}}, we would need to manually annotate all
-definienda in the book before we can harvest the symbol/verbalization pairs which are a
-prerequisite for annotation.
+half of that of English and we can already see the practical effects.
+The \defianno tool (see end of \cref{sec:snify}) can mitigate this partially by
+providing verbalizations for the definienda in the document,
+but we also want to annotate terms that are not explicitly defined in the document.
+A potential solution would be to extend \snify with a machine learning-based
+tool for identifying unannotated technical terms that are not in the catalog.
+The user could then pick a symbol for annotation, which would add a verbalization to the catalog.
+In an earlier (pre-\snify) attempt, we used named-entity
+recognition (NER) for classification of ``likely annotation candidate words''
+\cite{hutterer:msc23}. However, this classification was not precise enough in the
+distinction in ``technical terms'' and ordinary English noun phrases
+to be practical on its own.
+A combination with \snify's catalog-based approach might change the trade-offs involved.
+
+% For less complete domain models we have to skip too many terms that should ultimately be
+% annotated and annotation efficiency suffers. This is currently the case for all
+% non-English languages in the \sTeX corpus we work with. Coverage in German is only about
+% half of that of English and we can already see the practical effects. We have also
+% experimented with a Slovene introductory math book, and it seems clear that apart from
+% having to introduce a Slovene stemmer\ednote{MK@FS: there is one for python at
+%   \url{https://repo.ijs.si/pboskoski/slo_stemmer}}, we would need to manually annotate all
+% definienda in the book before we can harvest the symbol/verbalization pairs which are a
+% prerequisite for annotation.\ednote{this is in principle possible with \defianno}
+% \ednote{don't even need definienda - sn annos are enough to populate the catalog}
+% \ednote{TODO: I've commented out Hutter in conclusion - we can integrate it here as a potential workflow}
 
@@
 API~\cite{iMMT:on} currently under development will, so \snify will be able to request the
-pairs for the remaining (non-local) content commons. If a pair is chosen in an annotation,
+pairs for the remaining (non-local) content commons.\ednote{
+    This is already outdated.
+    What is the current API status?
+    How is it related to FLAMS?
+}
+If a pair is chosen in an annotation,
 \snify can then automatically download the corresponding archive so that the annotated
@@
 
-In an earlier attempt to support semantic annotation in IDEs, we tried named-entity
-recognition (NER) for classification of ``likely annotation candidate words''
-\cite{hutterer:msc23}, however this classification was not precise enough in the
-distinction in ``technical terms'' and ordinary English noun phrases and named entities --
-the relevant task for annotation, and so made it impractical, especially, since it could
-not enter the actual annotations and import directives automatically. But maybe using the
-NER-based approach, after the \snify annotation process might change the trade-offs
-involved. In any case, the overall workflow suggested by the NER-based approach --
-especially after the verbalizations covered by the existing domain model\ednote{introduce
-  in the definition above that the domain model is all definitions, not only in the
-  smglom?} -- is geared towards adding modules and symbols -- the ones discovered by NER
--- to the domain model.
+% In an earlier attempt to support semantic annotation in IDEs, we tried named-entity
+% recognition (NER) for classification of ``likely annotation candidate words''
+% \cite{hutterer:msc23}, however this classification was not precise enough in the
+% distinction in ``technical terms'' and ordinary English noun phrases and named entities --
+% the relevant task for annotation, and so made it impractical, especially, since it could
+% not enter the actual annotations and import directives automatically. But maybe using the
+% NER-based approach, after the \snify annotation process might change the trade-offs
+% involved. In any case, the overall workflow suggested by the NER-based approach --
+% especially after the verbalizations covered by the existing domain model\ednote{introduce
+%   in the definition above that the domain model is all definitions, not only in the
+%   smglom?} -- is geared towards adding modules and symbols -- the ones discovered by NER
+% -- to the domain model.
 

### 2025-05-08 doc/cicm25-sysdesc/sysdesc.tex
more

--- 2025-02-02 doc/cicm25-sysdesc/sysdesc.tex
@@
 \usepackage[noabbrev]{cleveref}
+\usepackage{tikz}
+\usetikzlibrary{shapes}
+\usetikzlibrary{arrows}
+\usetikzlibrary{docicon}
 
@@
 
+\begin{wrapfigure}r{5.5cm}
+  \begin{tikzpicture}[scale=.7]
+    \tikzstyle{doc}=[draw,thick,align=center,color=black,
+                  shape=document,minimum width=10mm,minimum height=15mm]
+    \tikzstyle{database}=[cylinder,shape border rotate=90,aspect=0.25,draw, 
+     cylinder uses custom fill,cylinder body fill=yellow!30,cylinder end fill=yellow!30]
+     \tikzstyle{include}=[right hook-angle 45,thick]
+     \tikzstyle{includeleft}=[left hook-angle 45,thick]
+    \node[doc] (cc) at (0,0) {\shortstack{Content\\ Commons}};
+    \node[doc] (pl) at (5,0) {\shortstack{Private/\\Local}};
+    \node[database] (db) at (3,2.5) {\shortstack{\tiny Verbalization\\Cache}};
+    \node (sn) at (0,4.5) {\texttt{snify}};
+    \node (id) at (2.5,5) {IDE};
+    \node (wo) at (5,4.5) {WOIDE};
+    \draw[include]  (cc) -- (db);
+    \draw[includeleft] (pl) -- (db);
+    \draw[->] (db) -- (sn);
+    \draw[->] (db) -- (id);
+    \draw[->] (db) -- (wo);
+    \draw[dashed,thick] (3,1.6)  -- (3,-1.4);
+  \end{tikzpicture}
+  \caption{Architecture}\label{fig:arch}\vspace*{-2em}
+\end{wrapfigure}
 One practical problem with \snify that remains is that the system only harvests

### 2025-05-08 doc/cicm25-sysdesc/sysdesc.tex
writing

--- 2025-05-04 doc/cicm25-sysdesc/sysdesc.tex
@@
 
-\documentclass{llncs}
+\documentclass[runningheads]{llncs}
 \usepackage[show]{ed}
@@
 \usepackage{wrapfig}
+\usepackage[hidelinks]{hyperref}
+\usepackage{orcidlink}
 %\usepackage{lststex} debug that the gray is not solid
@@
 }
-\author{Michael Kohlhase, Jan Frederik Schaefer}
+\author{Michael Kohlhase\orcidlink{0000-0002-9859-6337} \and Jan Frederik Schaefer\orcidlink{0000-0003-2545-4626}}
+
+% \keywords{semantic authoring, \snify, \sTeX, \ALeA, flexiformal content commons}
 \institute{Computer Science, FAU Erlangen N\"urnberg, Germany}
@@
     it is a problem of its own that needs special editing support facilities.
-    We present a simple utility that supports the process of annotating
+    We present and evaluate a simple utility that supports the process of annotating
     technical terms with semantic references to a flexiformal content commons
@@
 \end{abstract}
+% \setcounter{tocdepth}{3}
+% \tableofcontents
 
@@
 or annotating the defining phrase (definiens) in concept definitions,
-and can, in principle, even be used for full formalization.\ednote{True?}
-As \sTeX supports the full spectrum from informal (un-annotated) to fully formalized content,
+and could, in principle, even be used for full formalization.
+As \sTeX supports the spectrum from informal (un-annotated) to fully formalized content,
 we say that \sTeX content is \textbf{flexiformal}.
@@
 
-\ednote{
-    \textbf{Related work: semantic web authoring}
-    \emph{semantic authoring} already used in context of semantic web (RDFa).
-    Tools exist and it doesn't have exactly the same challenges as we face, though there is some overlap.
-    We have a module system and need more dependency management, something else?
-    This has to be described somewhere.
+% \ednote{
+%     \textbf{Related work: semantic web authoring}
+%     \emph{semantic authoring} already used in context of semantic web (RDFa).
+%     Tools exist and it doesn't have exactly the same challenges as we face, though there is some overlap.
+%     We have a module system and need more dependency management, something else?
+%     This has to be described somewhere.
+% 
+%     Related work: ``User Interfaces for Semantic Authoring of Textual Content: A Systematic Literature Review.''
+% }
 
-    Related work: ``User Interfaces for Semantic Authoring of Textual Content: A Systematic Literature Review.''
-}
-
-% \paragraph{Overview \& Contribution}
+\paragraph{Overview \& Contribution}
 In this paper, we argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
@@
 
+\paragraph{Acknowledgments}
+The work reported in this article was conducted as part of the VoLL-KI project (see
+\url{https://voll-ki.de}) funded by the German Research/Education Ministry under grant
+16DHBKI089.
+
 \section{Semantic Authoring in a Content Commons}\label{sec:semauth}
@@
 
-\subsection{Running Example: Annotating  ``Terminal''}
+\subsection{Running Example: Annotating ``terminal''}
 With \sTeX, authors can add semantic references to technical terms,
@@
 
-\begin{newpart}{MK: trying to make some general remark about annotation before we go into
-    semantic authoring}
 \subsection{Annotation Workflows and Requirements}\label{sec:workflows}
@@
 domain model does not cover in the documents to be annotated.
-\end{newpart}
 \ednote{FS: Todo: verbalizations can also be
@@
 
-% \paragraph{Semantic Authoring in \sTeX}
-% How this can concretely work depends on the particular semantic framework, the specific
-% annotations we want to support, and the semantics embedded in the content commons. In the
-% \sTeX framework, where we semantically annotate formulae and technical terms with pointers
-% to their definienda, we can bridge the presentation/semantics gap -- we annotate the words
-% in sentences and glyphs in formulae with the verbalizations and notations in the content
-% commons.\ednote{MK: maybe explain them again}. A tool for supporting semantic authoring of
-% formulae via notations has been presented by Vre{\v{c}}ar et al. in
-% \cite{VreWelKam:tsmmdui24}. In this paper we present a tool that supports semantic
-% annotation of technical terms via verbalizations.\ednote{move this in the next section}
+\paragraph{Semantic Authoring and the Semantic Web}
+The phrase ``semantic authoring'' has also been used in the context of the semantic web with varying meanings.
+This includes ontology authoring, which we would classify as formal authoring,
+and metadata annotation (e.g. the title, author, and creation date of a document),
+which is a somewhat different concern.
+A closer match is the annotation of informal content with
+references to a domain ontology,
+but there are some differences to our situation with \sTeX/\ALeA:
+The domain ontology would be a fully formal target
+-- possibly containing verbalizations --
+while in our case the target is flexiformal.
+Furthermore, \sTeX has a module system, which requires
+dependency management during authoring.
 

### 2025-05-08 scripts/annocounthistory2.py

@@ -52,7 +52,11 @@ def process_repo(path: Path):
         branch = 'master'
     if 'courses/FAU/GDP/problems' in path.as_uri():
         branch = 'main'
-    _, code = run_sh(f'git checkout {branch}', path)
+    _, code = run_sh(f'git checkout stex4', path)
+    if code == 0:
+        branch = 'stex4'
+    else:
+        _, code = run_sh(f'git checkout {branch}', path)
     assert code == 0, f'Error checking out {branch} in {path}'
     history, code = run_sh("git log --pretty='format:%H!%ci!%cn'", path)
     assert code == 0, f'Error getting history in {path}'

### 2025-05-08 scripts/annocountplot.py

@@ -30,9 +30,9 @@ def accumulate(data: list[CommitInfo]) -> tuple[list[float], list[int]]:
     return dates, counts
 
 
-def plot_cumulative(data: list[CommitInfo], label: str):
+def plot_cumulative(data: list[CommitInfo], label: str, lines: str):
     dates, counts = accumulate(data)
-    plt.plot(dates, counts, label=label)
+    plt.plot(dates, counts, label=label, linestyle=lines)
 
 
 def filtered(*pattern: str) -> list[CommitInfo]:
@@ -64,9 +64,9 @@ def plot_diff(data: list[CommitInfo], label: str):
 
 
 def cumulative_plots():
-    plot_cumulative(filtered('smglom/*'), 'smglom')
-    plot_cumulative(filtered('courses/*/course'), 'course')
-    plot_cumulative(filtered('*/problems'), 'problems')
+    plot_cumulative(filtered('smglom/*'), 'smglom', lines='dotted')
+    plot_cumulative(filtered('courses/*/course'), 'course', lines='dashed')
+    plot_cumulative(filtered('*/problems'), 'problems', lines='dashdot')
     # plot_cumulative(filtered('smglom/*', 'courses/*/course', '*/problems'), 'annotations')
 
 
@@ -86,7 +86,7 @@ def main():
     labels = []
     now = datetime.now()
     for year in range(s.year, now.year + 1):
-        for month in range(1, 13):
+        for month in range(1, 13, 3):
             if year == s.year and month < s.month:
                 continue
             if year == now.year and month > now.month:

### 2025-05-09 doc/cicm25-paper/paper.tex
paragraph on scaling snify

--- 2025-05-08 doc/cicm25-sysdesc/sysdesc.tex
@@
 
-    Semantic annotations enable services that make the knowledge
-    in documents more accessible.
-    In the case of \ALeA, a learning assistant we use as a case study,
-    they enable, for example, guided tours and practice problems
-    tailored to the learner's competencies.
-    However, the cost of semantic authoring is substantial
-    and tool support is limited.
+  Semantic annotations enable services that make the knowledge in documents more
+  accessible.  In the case of \ALeA, a learning assistant we use as a case study, they
+  enable, for example, guided tours and practice problems tailored to the learner's
+  competencies.  However, the cost of semantic authoring is substantial and tool support
+  is limited.
 
-    In this paper we discuss the nature of the semantic authoring
-    problem and argue that, while sharing aspects of classical
-    (informal) authoring and formal authoring,
-    it is a problem of its own that needs special editing support facilities.
-    We present and evaluate a simple utility that supports the process of annotating
-    technical terms with semantic references to a flexiformal content commons
-    that enhances annotation productivity by almost an order of magnitude.
+  In this paper we discuss the nature of the semantic authoring problem and argue that,
+  while sharing aspects of classical (informal) authoring and formal authoring, it is a
+  problem of its own that needs special editing support facilities.  We present and
+  evaluate a simple utility that supports the process of annotating technical terms with
+  semantic references to a flexiformal content commons that enhances annotation
+  productivity by almost an order of magnitude.
 \end{abstract}
@@
 \section{Tool Support for Semantic Authoring in \sTeX/\ALeA}\label{sec:tools}
+
 Semantic authoring can ``inherit'' tool support from both informal and formal authoring.
 In the case of \sTeX, we can re-use spell checkers for traditional informal authoring.
-There also is prototypical support for authoring \ALeA content in Microsoft Word
-with a custom plugin (see \cite{KohKoh:woide24}),
-which allows authors to use Word's spell checker and grammar checker.
+There also is prototypical support for authoring \ALeA content in Microsoft Word with a
+custom plugin (see \cite{KohKoh:woide24}), which allows authors to use Word's spell
+checker and grammar checker.\ednote{MK: Mention CPoint and SMD as well? Probably yes.}
 
-On the formal side, there is a custom VSCode IDE plugin for \sTeX \cite{sTeX-IDE:git} that analyzes
-annotations on the fly, displays the underlying reference semantics, reports errors and
-redundancies, and even offers a concept search interface.
-But such features only support
-existing annotations and are, therefore, of limited use when creating new annotations,
-which arguably is the most time consuming aspects of semantic authoring.
+On the formal side, there is a custom VSCode IDE plugin for \sTeX \cite{sTeX-IDE:git} that
+analyzes annotations on the fly, displays the underlying reference semantics, reports
+errors and redundancies, and even offers a concept search interface.  But such features
+only support existing annotations and are, therefore, of limited use when creating new
+annotations, which arguably is the most time consuming aspects of semantic authoring.
 % \ednote{also: SPARQL API}
@@
 \section{The \snify System}\label{sec:snify}
-The \snify\footnote{
-From ``\texttt{\textbackslash sn}-ify''; ``\texttt{\textbackslash sn}'' is
-a frequently used \sTeX short-hand macro for term annotations.}
-utility \cite{stextools:git}
-is a simple command-line tool that creates a catalog of
+
+The \snify\footnote{ From ``\texttt{\textbackslash sn}-ify''; ``\texttt{\textbackslash
+    sn}'' is a frequently used \sTeX short-hand macro for term annotations.}  utility
+\cite{stextools:git} is a simple command-line tool that creates a catalog of
 symbol-verbalization pairs, analyzes \sTeX source files, and then steps through all word
@@
 with ``utility''.  For each matched word, the user is presented with an annotation choice
-and interactions that allow to fine-tune the wordwise annotation
-workflow.
-The interface is inspired by traditional spell checkers like \lstinline|ispell|.
-\Cref{fig:snify} shows a typical situation.
+and interactions that allow to fine-tune the wordwise annotation workflow.  The interface
+is inspired by traditional spell checkers like \lstinline|ispell|.  \Cref{fig:snify} shows
+a typical situation.
 
@@
 together form a \textbf{session}, which can be interrupted and resumed without having to
-re-do all the \lstinline|s| skips. Alternatively, \snify can be called e.g. on the top-level
-lecture notes; then the session consists of all (transitively) included files. Other productivity
-features include a focus mode that can be used to only annotate a particular word in the
-rest of the file, session, or the whole \sTeX/\ALeA corpus before resuming regular
-annotation. This reduces the cognitive overhead from switching between different 
-words and symbol lists.
+re-do all the \lstinline|s| skips. Alternatively, \snify can be called e.g. on the
+top-level lecture notes; then the session consists of all (transitively) included
+files. Other productivity features include a focus mode that can be used to only annotate
+a particular word in the rest of the file, session, or the whole \sTeX/\ALeA corpus before
+resuming regular annotation. This reduces the cognitive overhead from switching between
+different words and symbol lists.
 % Of course the include management still needs to be done locally.
 
-As a companion to \snify,
-we have \defianno, a prototype tool for annotating definienda in documents
-that were originally authored without semantic annotations.
-It can be configured to use different \LaTeX\ macros
-to identify definienda (e.g.\ \lstinline|\emph| or \lstinline|\textbf|),
-and then steps through them similarly to \snify.
-For each potential definiendum, the user can then use fuzzy search to find
-the corresponding symbol in the domain model.
-This populates the catalog with verbalizations relevant to the document,
-making subsequent \snify runs more effective,
-especially for documents in languages that are not well covered by the
-domain model (i.e.\ not English).
+\paragraph{Definition Handling as a Pre-process for Multilingual Documents}
+As a companion to \snify, we have \defianno, a prototype tool for annotating definienda in
+documents that were originally authored without semantic annotations.  It can be
+configured to use different \LaTeX\ macros to identify definienda (e.g.\ \lstinline|\emph|
+or \lstinline|\textbf|), and then steps through them similarly to \snify.  For each
+potential definiendum, the user can then use fuzzy search to find the corresponding symbol
+in the domain model.  This populates the catalog with verbalizations relevant to the
+document, making subsequent \snify runs more effective, especially for documents in
+languages that are not well covered by the domain model (i.e.\ not English).
+
+\begin{newpart}{MK: rewritten as existing but experimental. This needs to be placed
+    somewhere else.}
+  \paragraph{Scaling \snify to the Flexiformal Content Commons }
+One practical problem with \snify system as presented so far is that it only harvests
+symbol/verbalization pairs from \sTeX files on the local file system. This means that, for
+effective annotation, users (have to) download the whole content commons and keep it
+updated -- an assumption/requirement that becomes increasingly impractical as the \sTeX
+ecosystem grows.
+
+The general situation is that on the one hand there is a constantly growing, shared
+\textbf{content commons} of public math archives -- currently 200 with about 14.000 \sTeX
+files -- that are regularly converted to FTML (\underline{F}lexiformal \underline{T}ext
+\underline{M}arkup \underline{L}anguage) and served on \url{https://MathHub.info}. On the
+other hand authors are working on
+\begin{compactenum}[\em i\rm)]
+\item a set of local working copies of the archives -- with the intention of publishing
+  them on MathHub.info and their dependencies (they need to be local, since pdflatex
+  cannot deal with remote files) and
+\item possibly a set of private archive (think papers under development or exam
+  problems/solutions) 
+\end{compactenum}
+This situation is familiar from software development and also from e.g. the \LaTeX
+ecosystem. Semantic authoring -- and thus the \snify utility -- inherits the requirements
+from both.
+
+\begin{wrapfigure}r{5.5cm}\vspace*{-2em}
+  \begin{tikzpicture}[scale=.7]
+    \tikzstyle{doc}=[draw,thick,align=center,color=black,
+                  shape=document,minimum width=10mm,minimum height=15mm]
+    \tikzstyle{database}=[cylinder,shape border rotate=90,aspect=0.25,draw, 
+     cylinder uses custom fill,cylinder body fill=yellow!30,cylinder end fill=yellow!30]
+     \tikzstyle{include}=[right hook-angle 45,thick]
+     \tikzstyle{includeleft}=[left hook-angle 45,thick]
+    \node[doc] (cc) at (0,-1) {\shortstack{Content\\ Commons}};
+    \node[doc] (pl) at (5,-1) {\shortstack{Private/\\Local}};
+    \node[database] (db) at (3,2.5) {\shortstack{\tiny Verbalization\\Cache}};
+    \node[draw] (sn) at (0,4.5) {\snify};
+    \node[draw] (id) at (2.5,5) {IDE};
+    \node[draw] (wo) at (5,4.5) {WOIDE};
+    \draw[draw,fill=black!10] (-1.3,.5) rectangle ++ (6.4,1); 
+    \node (f) at (-.5,.9) {\flams};
+    \draw[include]  (cc) -- node[left] {\tiny F2RDF} (db);
+    \draw[includeleft] (pl) -- node[right] {\tiny linter} (db);
+    \draw[->] (db) -- (sn);
+    \draw[->] (db) -- (id);
+    \draw[->] (db) -- (wo);
+    \draw[dashed,thick] (3,1.6)  -- (3,-2);
+    \draw[dashed,thick] (3,1.6)  -- (3,-2);
+    \draw[includeleft] (pl) to[bend right=10] (wo);
+  \end{tikzpicture}
+  \caption{Architecture}\label{fig:arch}\vspace*{-1em}
+\end{wrapfigure}
+To cope with this situation, We have extended the brand-new \flams system
+\cite{flams:on}\footnote{A paper on \flams is submitted to CICM 2025 and will be cited
+  here if accepted.} with functionality for handling definienda in the FTML2RDF harvester,
+so that all symbol/verbalizations are available in the \flams triple-store. Thus a harvest
+from the flexiformal content commons is just a simple SPARQL query and some postprocessing
+of the result (de-duplication, stemming, and pairing) away. Note that \flams works
+directly on FTML, which is generated from \sTeX input by \textsf{rus\TeX} and exported
+from annoated MS Word files via WOIDE.
+
+In local/private archives, the \snify harvester needs to work directly on the \sTeX
+sources. Here we have extended the \flams \textbf{linter} -- a simple \sTeX parser that
+drives the LSP functionality the VSCode and \texttt{emacs} \sTeX IDE plugins use to deal
+with definienda. This is much more efficient (and robust) than the initial regex-based
+python parser). The analogon for this on the WOIDE side is still under development it
+could be based on FTML generated by WOIDE, or WOIDE could export symbol/verbalization
+pairs into the \snify verbalization cache directly.
+
+In this situation, the main functionality of \snify is to maintain the verbalization cache
+and answer front-end queries for verbalizations efficiently so that we can power various
+front-ends from that. These include
+\begin{compactenum}[\em i\rm)]
+\item The traditional text/terminal-based \snify UI for bulk annotation,
+\item the \sTeX IDEs in VSCode and \texttt{emacs} that can supply while-editing annotation
+  support, and 
+\item WOIDE, which can be seen as a semantic authoring IDE based on MS Word.\ednote{MK:
+    CPoint and semantic wikis.}
+\end{compactenum}
+
+Note that implemented this broadly \snify can provide additionall logistic services: If a
+pair is chosen in an annotation, \snify can then automatically download the corresponding
+archive so that the annotated file can be formatted with \texttt{pdflatex} or
+\textsf{rus\TeX} locally.
+\end{newpart}
 
@@
 
-\begin{wrapfigure}r{5.5cm}
-  \begin{tikzpicture}[scale=.7]
-    \tikzstyle{doc}=[draw,thick,align=center,color=black,
-                  shape=document,minimum width=10mm,minimum height=15mm]
-    \tikzstyle{database}=[cylinder,shape border rotate=90,aspect=0.25,draw, 
-     cylinder uses custom fill,cylinder body fill=yellow!30,cylinder end fill=yellow!30]
-     \tikzstyle{include}=[right hook-angle 45,thick]
-     \tikzstyle{includeleft}=[left hook-angle 45,thick]
-    \node[doc] (cc) at (0,0) {\shortstack{Content\\ Commons}};
-    \node[doc] (pl) at (5,0) {\shortstack{Private/\\Local}};
-    \node[database] (db) at (3,2.5) {\shortstack{\tiny Verbalization\\Cache}};
-    \node (sn) at (0,4.5) {\texttt{snify}};
-    \node (id) at (2.5,5) {IDE};
-    \node (wo) at (5,4.5) {WOIDE};
-    \draw[include]  (cc) -- (db);
-    \draw[includeleft] (pl) -- (db);
-    \draw[->] (db) -- (sn);
-    \draw[->] (db) -- (id);
-    \draw[->] (db) -- (wo);
-    \draw[dashed,thick] (3,1.6)  -- (3,-1.4);
-  \end{tikzpicture}
-  \caption{Architecture}\label{fig:arch}\vspace*{-2em}
-\end{wrapfigure}
-One practical problem with \snify that remains is that the system only harvests
-symbol/verbalization pairs from \sTeX files on the local file system. This means that, for
-effective annotation, users (have to) download the whole content commons and keep it
-updated. Unfortunately, the current MMT-based MathHub knowledge management
-API\ednote{intro this above} does not supply verbalization information. The iMMT-based
-API~\cite{iMMT:on} currently under development will, so \snify will be able to request the
-pairs for the remaining (non-local) content commons.\ednote{
-    This is already outdated.
-    What is the current API status?
-    How is it related to FLAMS?
-}
-If a pair is chosen in an annotation,
-\snify can then automatically download the corresponding archive so that the annotated
-file can be formatted with pdflatex or rusTeX locally.
-
 Finally, the current interface is not well-suited for on-the-fly annotation while
@@
 % LocalWords:  tbw Aarne Ranta homostem pre hmmm Vre ar al sn ify TODO et Hutter Plivelic
-% LocalWords:  lang
+% LocalWords:  lang DHBKI FS Todo FTML lexiformal arkup anguage CICM RDF postprocessing
+% LocalWords:  de

### 2025-05-13 doc/cicm25-paper/paper.tex
ednots

@@
 
-Take for instance a learning object like the paragraph in \cref{fig:lo}, where we want to annotate
-the term ``terminal'' with the (semantic) concept of ``a goal state of a search
+Take for instance a learning object like the paragraph in \cref{fig:lo}, where we want to
+annotate the term ``terminal'' with the (semantic) concept of ``a goal state of a search
 problem''. The latter can either be defined earlier in the course or in the domain model
-(or both).
+(or both). Concretely this is about converting the {\LaTeX} string
 
-Concretely this is about converting the {\LaTeX} string
-    (for conciseness completely unannotated)
-    \ednote{MK: There is/was a discrepancy
-  in the last line of the example; I tried to fix it, but failed\\
-    FS: Is it fixed now? I don't see any discrepancy.}
 \begin{lstlisting}[numbers=left,firstnumber=3,
@@
 \end{lstlisting}
-In the latter, the author has the word ``terminal'' with the concept it refers to with 
-via the \sTeX macro
+In the latter, the author has annotated the word ``terminal'' with the concept it refers
+to with via the \sTeX macro
 \lstinline[mathescape]|\sr{$\llangle URI \rrangle$}{$\llangle verbalization\rrangle$}|.
 
-\begin{newpart}{MK: more definitions}
 In \sTeX, objects and concepts are represented by \textbf{symbols}, which are identified
@@
   name}, and a \textbf{symbol name}. Every symbol can have one or more
-\textbf{verbalizations}:
-words of the domain jargon used to refer to the concept or object.
-Verbalizations can vary between between subdomains and communities of practice. In
-our example, the verbalization ``terminal [state]'' was annotated with the symbol with
-name \lstinline|goal state|. It comes from the module with name \lstinline|search-problem|
-in the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
-macro in line 1. As the symbol name \lstinline|goal state| is unique in the imports here
-it suffices as a relative URI in the \lstinline|\sr| annotation.
+\textbf{verbalizations}: words of the domain jargon used to refer to the concept or
+object.  Verbalizations can vary between between subdomains and communities of
+practice. In our example, the verbalization ``terminal [state]'' was annotated with the
+symbol with name \lstinline|goal state|. It comes from the module with name
+\lstinline|search-problem| in the namespace \lstinline|smglom/search/mod| imported via the
+\lstinline|\importmodule| macro in line 1. As the symbol name \lstinline|goal state| is
+unique in the imports here it suffices as a relative URI in the \lstinline|\sr|
+annotation. If the symbol name and the desired verbalization coincide (as they often do in
+English), the short-hand macro \lstinline|\sn| can be used (e.g. as
+\lstinline|\sn{utility}| for ``utility'' in \cref{lst:losa}).
 
-If the symbol name and the desired verbalization coincide (as they often do in English),
-the short-hand macro \lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| for ``utility'' in \cref{lst:losa}).\ednote{
-    FS: Do we need the last paragraph?
-        I've removed the ``sn utility'' annotation from the listing
-        because I thought it breaks the flow a bit.
-        But if we want to show the sn macro, I can add it back.
-}
-\end{newpart}
-
-The annotation refers to a definition in the domain model module shown in
+Both variants of the annotation refer to a definition in the domain model module shown in
 \cref{fig:state-space}.  The \lstinline|\definame| and \lstinline|\definiendum| in lines

### 2025-05-13 doc/cicm25-paper/paper.tex
more

@@
 
+\begin{newpart}{MK: please re-read}
+  A distinctive feature of semantic authoring that is not apparent at the first glance is
+  that the added investment in annotation moves the sweet spot in the re-use/customize
+  trade-off towards sharing and interoperability. In our experience sufficiently annotated
+  documents can be automatically customized, if the semantics has multiple
+  notations/verbalizations/presentations. Consider for instance course materials on
+  complex arithmetic, which become re-usable across the electrical
+  engineering/mathematics border, if the imaginary unit in formulae is represented e.g. by
+  a macro \lstinline|\imaginaryunit|, which can alternatively expand to $i$ (for uses in
+  maths) and $j$ (for EE uses). As a consequence semantic authoring is more geared towards
+  creating reusable/shareable document fragments organized in central or federated
+  repositories and hubs. This is quite similar to the formal authoring system, where we
+  see a tendency towards massively shared math libraries in formal systems and package
+  hubs for programming libraries. As a consequence, semantic authoring support systems
+  should facilitate the logistics of using and curating shared libraries. 
+\end{newpart}
+
 \paragraph{Semantic Authoring and the Semantic Web}
@@
 directly on FTML, which is generated from \sTeX input by \textsf{rus\TeX} and exported
-from annoated MS Word files via WOIDE.
+from annotated MS Word files via WOIDE.
 
@@
 
-Note that implemented this broadly \snify can provide additionall logistic services: If a
+Note that implemented this broadly \snify can provide additional logistic services: If a
 pair is chosen in an annotation, \snify can then automatically download the corresponding
@@
 % LocalWords:  lang DHBKI FS Todo FTML lexiformal arkup anguage CICM RDF postprocessing
-% LocalWords:  de
+% LocalWords:  de maths EE CPoint SMD rus WOIDE linter LSP emacs

### 2025-05-13 doc/cicm25-paper/paper.tex
more

@@
   notations/verbalizations/presentations. Consider for instance course materials on
-  complex arithmetic, which become re-usable across the electrical
-  engineering/mathematics border, if the imaginary unit in formulae is represented e.g. by
-  a macro \lstinline|\imaginaryunit|, which can alternatively expand to $i$ (for uses in
-  maths) and $j$ (for EE uses). As a consequence semantic authoring is more geared towards
+  complex arithmetic, which become re-usable across the electrical engineering/mathematics
+  border, if the imaginary unit in formulae is represented e.g. by a semantic macro
+  \lstinline|\imaginaryunit|, which can alternatively expand to $i$ (for uses in maths)
+  and $j$ (for EE uses). As a consequence semantic authoring is more geared towards
   creating reusable/shareable document fragments organized in central or federated
@@
 \paragraph{Semantic Authoring and the Semantic Web}
-The phrase ``semantic authoring'' has also been used in the context of the semantic web with varying meanings.
-This includes ontology authoring, which we would classify as formal authoring,
-and metadata annotation (e.g. the title, author, and creation date of a document),
-which is a somewhat different concern.
-A closer match is the annotation of informal content with
-references to a domain ontology,
-but there are some differences to our situation with \sTeX/\ALeA:
-The domain ontology would be a fully formal target
--- possibly containing verbalizations --
-while in our case the target is flexiformal.
-Furthermore, \sTeX has a module system, which requires
-dependency management during authoring.
+The phrase ``semantic authoring'' has also been used in the context of the semantic web
+with varying meanings.  This includes ontology authoring, which we would classify as
+formal authoring, and metadata annotation (e.g. the title, author, and creation date of a
+document), which is a somewhat different concern.  A closer match is the annotation of
+informal content with references to a domain ontology, but there are some differences to
+our situation with \sTeX/\ALeA: The domain ontology would be a fully formal target --
+possibly containing verbalizations -- while in our case the target is flexiformal.
+Furthermore, \sTeX has a module system, which requires dependency management during
+authoring.\ednote{MK: We need more citations here}
 
@@
 
-\begin{newpart}{MK: rewritten as existing but experimental. This needs to be placed
-    somewhere else.}
   \paragraph{Scaling \snify to the Flexiformal Content Commons }
@@
 \textsf{rus\TeX} locally.
-\end{newpart}
 
@@
 For less complete domain models we have to skip too many terms that should ultimately be
-annotated and annotation efficiency suffers.\ednote{
-    FS: Isn't this wrong? We don't have to skip them, the problem is that \snify skips them.
-    Efficiency does not suffer, but rather effectiveness.
-}
-This is currently the case for all
+annotated and annotation effectiveness suffers. This is currently the case for all
 non-English languages in the \sTeX corpus we work with. Coverage in German is only about
-half of that of English and we can already see the practical effects.
-The \defianno tool (see end of \cref{sec:snify}) can mitigate this partially by
-providing verbalizations for the definienda in the document,
-but we also want to annotate terms that are not explicitly defined in the document.
-A potential solution would be to extend \snify with a machine learning-based
-tool for identifying unannotated technical terms that are not in the catalog.
-The user could then pick a symbol for annotation, which would add a verbalization to the catalog.
-In an earlier (pre-\snify) attempt, we used named-entity
+half of that of English and we can already see the practical effects.  The \defianno tool
+(see end of \cref{sec:snify}) can mitigate this partially by providing verbalizations for
+the definienda in the document, but we also want to annotate terms that are not explicitly
+defined in the document.  A potential solution would be to extend \snify with a machine
+learning-based tool for identifying unannotated technical terms that are not in the
+catalog.  The user could then pick a symbol for annotation, which would add a
+verbalization to the catalog.  In an earlier (pre-\snify) attempt, we used named-entity
 recognition (NER) for classification of ``likely annotation candidate words''
 \cite{hutterer:msc23}. However, this classification was not precise enough in the
-distinction in ``technical terms'' and ordinary English noun phrases
-to be practical on its own.
-A combination with \snify's catalog-based approach might change the trade-offs involved.
+distinction in ``technical terms'' and ordinary English noun phrases to be practical on
+its own.  A combination with \snify's catalog-based approach might change the trade-offs
+involved.
 

### 2025-05-21 doc/cicm25-paper/paper.tex
polishing

@@
 \newcommand\defianno{\textsf{defianno}\xspace}
+\newcommand\FTML{\textsf{FTML}\xspace}
+\newcommand\WOIDE{\textsf{WOIDE}\xspace}
 \def\llangle{\langle\kern-.2em\langle}
@@
 }
+\titlerunning{
+    Semantic authoring in a flexiformal context \textellipsis
+}
+
 \author{Michael Kohlhase\orcidlink{0000-0002-9859-6337} \and Jan Frederik Schaefer\orcidlink{0000-0003-2545-4626}}
@@
 depends on semantic annotations.
-These can be embedded \LaTeX\ documents using the \sTeX package~\cite{MueKo:sdstex22,sTeX:github:on}.
+These can be embedded in \LaTeX\ documents using the \sTeX package~\cite{MueKo:sdstex22,sTeX:github:on}.
 In this paper, we will mostly focus on annotating terms with the semantic concepts they refer to,
@@
 In this paper, we argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
-but is nevertheless a distinct problem requiring special tool support (\cref{sec:semauth}).
+but is nevertheless a distinct problem requiring special tool support (\Cref{sec:semauth}).
 Using \sTeX/\ALeA as an exemplary case study,
-we will provide an overview of the authoring support (\cref{sec:tools})
+we will provide an overview of the authoring support (\Cref{sec:tools})
 and then describe and evaluate one new, concrete tool, \snify, that supports the annotation
-of terms with the semantic concepts they refer to (\cref{sec:snify,sec:eval}).
-The paper ends with a brief conclusion (\cref{sec:conclusion}).
+of terms with the semantic concepts they refer to (\Cref{sec:snify}).
+Afterweards, we discuss on-going work to integrate \snify better
+into the \sTeX/\ALeA ecosystem and content commons (\Cref{sec:scaling}).
+The paper ends with a brief discussion of related work (\Cref{sec:relwork}) and the conclusion (\Cref{sec:conclusion}).
 
@@
 
-Take for instance a learning object like the paragraph in \cref{fig:lo}, where we want to
+Take for instance a learning object like the paragraph in \Cref{fig:lo}, where we want to
 annotate the term ``terminal'' with the (semantic) concept of ``a goal state of a search
@@
 \begin{lstlisting}[numbers=left,firstnumber=3,
-caption=The unannotated {\LaTeX} sources of \cref{fig:lo},label=lst:los]
+caption=The unannotated {\LaTeX} sources of \Cref{fig:lo},label=lst:los]
 \item \blue{Note:} Depth-limited minimax requires an evaluation for every
@@
 \begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
-caption=Annotating ``terminal'' in \cref{lst:los},label=lst:losa]
+caption=Annotating ``terminal'' in \Cref{lst:los},label=lst:losa]
 \importmodule[smglom/search]{mod?search-problem}
@@
 English), the short-hand macro \lstinline|\sn| can be used (e.g. as
-\lstinline|\sn{utility}| for ``utility'' in \cref{lst:losa}).
+\lstinline|\sn{utility}| for ``utility'' in \Cref{lst:losa}).
 
-Both variants of the annotation refer to a definition in the domain model module shown in
-\cref{fig:state-space}.  The \lstinline|\definame| and \lstinline|\definiendum| in lines
+The annotation refers to a definition in the domain model module shown in
+\Cref{fig:state-space}.  The \lstinline|\definame| and \lstinline|\definiendum| in lines
 5--6 introduce the verbalizations ``goal state'' and ``terminal state'' for the symbol
@@
 To annotate the word ``terminal'' in the document, the author has to be aware that it is a
-technical term, that the module from \cref{fig:state-space} exists, know the symbol URI,
+technical term, that the module from \Cref{fig:state-space} exists, know the symbol URI,
 and manage redundancy of the imports -- i.e. only adding the \lstinline|\importmodule|
@@
 
-In $L$-language documents, where $L$ is not English the annotation practices are a bit
+For documents in a language $L$,
+where $L$ is not English, the annotation practices are a bit
 more complex still, in particular because the annotation catalog is usually incomplete.
 
-Note that for $L$-annotation, we need a $L$-specific annotation catalog. 
+Note that for $L$-annotation, we need an $L$-specific annotation catalog. 
 \sTeX already supports multilinguality:
@@
   \lstinline[mathescape]|.$\llangle lang\rrangle$.tex| where $\llangle lang\rrangle$ is the
-  ISO-639 language identifier.)
+  ISO-639 language identifier).
 \item Modules in the domain model allow $L$-translations that share the formal
@@
 \item the domain model may not even have $L$-modules -- especially for the ``smaller''
-  languages
+  languages.
 \end{enumerate}
@@
 domain model does not cover in the documents to be annotated.
-\ednote{FS: Todo: verbalizations can also be
-inferred from existing annotations (without any $L$-modules).}
 
@@
 \paragraph{Traditional, informal authoring}
-In traditional authoring e.g. of a scientific article or a textbook, the domain model -- a
+In traditional authoring, e.g. of a scientific article or a textbook, the domain model -- a
 highly structured model of the domain of discourse -- is in the authors' heads, and is
 partially verbalized e.g. in the preliminaries section or the main sections of the
-respective article. The content is informal -- usually a natural language like English
-augmented with technical jargon, formulae, tables, and diagrams -- and designed for
+respective article. The content is informal (usually a natural language like English
+augmented with technical jargon, formulae, tables, and diagrams) and designed for
 processing via the brain of a colleague or student -- in any case, a human who shares a
@@
 \paragraph{Formal Authoring}
-In fully formal authoring e.g. for programming, the content commons is given formally in
+In fully formal authoring, e.g. programming, the content commons is given formally in
 the form of the libraries of the respective programming language, correspondingly the
-authored content is also formal -- a program -- and intended for processing by a computer,
+authored content -- a program -- is also formal and intended for processing by a computer,
 for which the formality is a prerequisite. Formal authoring is usually supported by an
@@
 
-But in addition to the base text, semantic authoring also has to provide the semantic
+In addition to the base text, semantic authoring also has to provide the semantic
 annotations that drive the personalization and interaction. Techniques from formal
@@
   trade-off towards sharing and interoperability. In our experience sufficiently annotated
-  documents can be automatically customized, if the semantics has multiple
+  documents can be automatically customized if the semantics has multiple
   notations/verbalizations/presentations. Consider for instance course materials on
   complex arithmetic, which become re-usable across the electrical engineering/mathematics
-  border, if the imaginary unit in formulae is represented e.g. by a semantic macro
+  border if the imaginary unit in formulae is represented e.g. by a semantic macro
   \lstinline|\imaginaryunit|, which can alternatively expand to $i$ (for uses in maths)
@@
 
-\paragraph{Semantic Authoring and the Semantic Web}
-The phrase ``semantic authoring'' has also been used in the context of the semantic web
-with varying meanings.  This includes ontology authoring, which we would classify as
-formal authoring, and metadata annotation (e.g. the title, author, and creation date of a
-document), which is a somewhat different concern.  A closer match is the annotation of
-informal content with references to a domain ontology, but there are some differences to
-our situation with \sTeX/\ALeA: The domain ontology would be a fully formal target --
-possibly containing verbalizations -- while in our case the target is flexiformal.
-Furthermore, \sTeX has a module system, which requires dependency management during
-authoring.\ednote{MK: We need more citations here}
-
 \section{Tool Support for Semantic Authoring in \sTeX/\ALeA}\label{sec:tools}
@@
 In the case of \sTeX, we can re-use spell checkers for traditional informal authoring.
-There also is prototypical support for authoring \ALeA content in Microsoft Word with a
-custom plugin (see \cite{KohKoh:woide24}), which allows authors to use Word's spell
+There also is prototypical support for authoring \ALeA content in Microsoft Word with
+the \WOIDE plugin \cite{KohKoh:woide24}, which allows authors to use Word's spell
 checker and grammar checker.\ednote{MK: Mention CPoint and SMD as well? Probably yes.}
@@
 and interactions that allow to fine-tune the wordwise annotation workflow.  The interface
-is inspired by traditional spell checkers like \lstinline|ispell|.  \Cref{fig:snify} shows
+is inspired by traditional spell checkers like \lstinline|ispell|~\cite{ispellman}.  \Cref{fig:snify} shows
 a typical situation.
@@
 
-To skip the word, the user can type \lstinline|s|,
+To skip the word, the user can type \lstinline|s|;
 \lstinline|s!| skips it until the end of the file, and \lstinline|S|
@@
 
-\paragraph{Definition Handling as a Pre-process for Multilingual Documents}
+\paragraph{\defianno: Definition Handling as a Pre-process for Multilingual Documents}
 As a companion to \snify, we have \defianno, a prototype tool for annotating definienda in
 documents that were originally authored without semantic annotations.  It can be
-configured to use different \LaTeX\ macros to identify definienda (e.g.\ \lstinline|\emph|
-or \lstinline|\textbf|), and then steps through them similarly to \snify.  For each
+configured to use different \LaTeX\ macros to identify definienda (e.g.\ \lstinline[language={}]|\emph|
+or \lstinline[language={}]|\textbf|), and then steps through them similarly to \snify.  For each
 potential definiendum, the user can then use fuzzy search to find the corresponding symbol
@@
 
-  \paragraph{Scaling \snify to the Flexiformal Content Commons }
-One practical problem with \snify system as presented so far is that it only harvests
-symbol/verbalization pairs from \sTeX files on the local file system. This means that, for
-effective annotation, users (have to) download the whole content commons and keep it
-updated -- an assumption/requirement that becomes increasingly impractical as the \sTeX
-ecosystem grows.
-
-The general situation is that on the one hand there is a constantly growing, shared
-\textbf{content commons} of public math archives -- currently 200 with about 14.000 \sTeX
-files -- that are regularly converted to FTML (\underline{F}lexiformal \underline{T}ext
-\underline{M}arkup \underline{L}anguage) and served on \url{https://MathHub.info}. On the
-other hand authors are working on
-\begin{compactenum}[\em i\rm)]
-\item a set of local working copies of the archives -- with the intention of publishing
-  them on MathHub.info and their dependencies (they need to be local, since pdflatex
-  cannot deal with remote files) and
-\item possibly a set of private archive (think papers under development or exam
-  problems/solutions) 
-\end{compactenum}
-This situation is familiar from software development and also from e.g. the \LaTeX
-ecosystem. Semantic authoring -- and thus the \snify utility -- inherits the requirements
-from both.
-
-\begin{wrapfigure}r{5.5cm}\vspace*{-2em}
-  \begin{tikzpicture}[scale=.7]
-    \tikzstyle{doc}=[draw,thick,align=center,color=black,
-                  shape=document,minimum width=10mm,minimum height=15mm]
-    \tikzstyle{database}=[cylinder,shape border rotate=90,aspect=0.25,draw, 
-     cylinder uses custom fill,cylinder body fill=yellow!30,cylinder end fill=yellow!30]
-     \tikzstyle{include}=[right hook-angle 45,thick]
-     \tikzstyle{includeleft}=[left hook-angle 45,thick]
-    \node[doc] (cc) at (0,-1) {\shortstack{Content\\ Commons}};
-    \node[doc] (pl) at (5,-1) {\shortstack{Private/\\Local}};
-    \node[database] (db) at (3,2.5) {\shortstack{\tiny Verbalization\\Cache}};
-    \node[draw] (sn) at (0,4.5) {\snify};
-    \node[draw] (id) at (2.5,5) {IDE};
-    \node[draw] (wo) at (5,4.5) {WOIDE};
-    \draw[draw,fill=black!10] (-1.3,.5) rectangle ++ (6.4,1); 
-    \node (f) at (-.5,.9) {\flams};
-    \draw[include]  (cc) -- node[left] {\tiny F2RDF} (db);
-    \draw[includeleft] (pl) -- node[right] {\tiny linter} (db);
-    \draw[->] (db) -- (sn);
-    \draw[->] (db) -- (id);
-    \draw[->] (db) -- (wo);
-    \draw[dashed,thick] (3,1.6)  -- (3,-2);
-    \draw[dashed,thick] (3,1.6)  -- (3,-2);
-    \draw[includeleft] (pl) to[bend right=10] (wo);
-  \end{tikzpicture}
-  \caption{Architecture}\label{fig:arch}\vspace*{-1em}
-\end{wrapfigure}
-To cope with this situation, We have extended the brand-new \flams system
-\cite{flams:on}\footnote{A paper on \flams is submitted to CICM 2025 and will be cited
-  here if accepted.} with functionality for handling definienda in the FTML2RDF harvester,
-so that all symbol/verbalizations are available in the \flams triple-store. Thus a harvest
-from the flexiformal content commons is just a simple SPARQL query and some postprocessing
-of the result (de-duplication, stemming, and pairing) away. Note that \flams works
-directly on FTML, which is generated from \sTeX input by \textsf{rus\TeX} and exported
-from annotated MS Word files via WOIDE.
-
-In local/private archives, the \snify harvester needs to work directly on the \sTeX
-sources. Here we have extended the \flams \textbf{linter} -- a simple \sTeX parser that
-drives the LSP functionality the VSCode and \texttt{emacs} \sTeX IDE plugins use to deal
-with definienda. This is much more efficient (and robust) than the initial regex-based
-python parser). The analogon for this on the WOIDE side is still under development it
-could be based on FTML generated by WOIDE, or WOIDE could export symbol/verbalization
-pairs into the \snify verbalization cache directly.
-
-In this situation, the main functionality of \snify is to maintain the verbalization cache
-and answer front-end queries for verbalizations efficiently so that we can power various
-front-ends from that. These include
-\begin{compactenum}[\em i\rm)]
-\item The traditional text/terminal-based \snify UI for bulk annotation,
-\item the \sTeX IDEs in VSCode and \texttt{emacs} that can supply while-editing annotation
-  support, and 
-\item WOIDE, which can be seen as a semantic authoring IDE based on MS Word.\ednote{MK:
-    CPoint and semantic wikis.}
-\end{compactenum}
-
-Note that implemented this broadly \snify can provide additional logistic services: If a
-pair is chosen in an annotation, \snify can then automatically download the corresponding
-archive so that the annotated file can be formatted with \texttt{pdflatex} or
-\textsf{rus\TeX} locally.
-
-\section{Practical Evaluation and Future Work}\label{sec:eval}
+\paragraph{Evaluation}
+% \section{Practical Evaluation and Future Work}\label{sec:eval}
 \begin{wrapfigure}{r}{0.4\textwidth}
@@
 model.
-Given the sheer number of term annotations (see \cref{fig:annocounts}),
+Given the sheer number of term annotations (see \Cref{fig:annocounts}),
 this results in a considerable productivity gain.
@@
 half of that of English and we can already see the practical effects.  The \defianno tool
-(see end of \cref{sec:snify}) can mitigate this partially by providing verbalizations for
-the definienda in the document, but we also want to annotate terms that are not explicitly
-defined in the document.  A potential solution would be to extend \snify with a machine
-learning-based tool for identifying unannotated technical terms that are not in the
-catalog.  The user could then pick a symbol for annotation, which would add a
-verbalization to the catalog.  In an earlier (pre-\snify) attempt, we used named-entity
-recognition (NER) for classification of ``likely annotation candidate words''
-\cite{hutterer:msc23}. However, this classification was not precise enough in the
-distinction in ``technical terms'' and ordinary English noun phrases to be practical on
-its own.  A combination with \snify's catalog-based approach might change the trade-offs
-involved.
-
-% For less complete domain models we have to skip too many terms that should ultimately be
-% annotated and annotation efficiency suffers. This is currently the case for all
-% non-English languages in the \sTeX corpus we work with. Coverage in German is only about
-% half of that of English and we can already see the practical effects. We have also
-% experimented with a Slovene introductory math book, and it seems clear that apart from
-% having to introduce a Slovene stemmer\ednote{MK@FS: there is one for python at
-%   \url{https://repo.ijs.si/pboskoski/slo_stemmer}}, we would need to manually annotate all
-% definienda in the book before we can harvest the symbol/verbalization pairs which are a
-% prerequisite for annotation.\ednote{this is in principle possible with \defianno}
-% \ednote{don't even need definienda - sn annos are enough to populate the catalog}
-% \ednote{TODO: I've commented out Hutter in conclusion - we can integrate it here as a potential workflow}
+can mitigate this partially by providing verbalizations for
+the definienda in the document, but it does not help with terms that are not explicitly
+defined in the document.
 
@@
 
+\section{Scaling \snify to the Flexiformal Content Commons }\label{sec:scaling}
+One practical problem with \snify system as presented so far is that it only harvests
+symbol/verbalization pairs from \sTeX files on the local file system. This means that, for
+effective annotation, users (have to) download the whole content commons and keep it
+updated -- an assumption/requirement that becomes increasingly impractical as the \sTeX
+ecosystem grows.
+
+The general situation is that on the one hand there is a constantly growing, shared
+\textbf{content commons} of public math archives -- currently 200 with about 14.000 \sTeX
+files -- that are regularly converted to \FTML (\underline{F}lexiformal \underline{T}ext
+\underline{M}arkup \underline{L}anguage) and served on \url{https://MathHub.info}. On the
+other hand, authors are working on
+\begin{compactenum}[\em i\rm)]
+\item a set of local working copies of the archives -- with the intention of publishing
+  them on MathHub.info -- and their dependencies (they need to be local since pdflatex
+  cannot deal with remote files) and
+\item possibly a set of private archives (e.g.\ papers under development or exam
+  problems/solutions).
+\end{compactenum}
+This situation is familiar from software development and also from e.g. the \LaTeX{}
+ecosystem. Semantic authoring -- and thus the \snify utility -- inherits the requirements
+from both.
+
+\begin{wrapfigure}r{5.5cm}\vspace*{-2em}
+  \begin{tikzpicture}[scale=.7]
+    \tikzstyle{doc}=[draw,thick,align=center,color=black,
+                  shape=document,minimum width=10mm,minimum height=15mm]
+    \tikzstyle{database}=[cylinder,shape border rotate=90,aspect=0.25,draw, 
+     cylinder uses custom fill,cylinder body fill=yellow!30,cylinder end fill=yellow!30]
+     \tikzstyle{include}=[right hook-angle 45,thick]
+     \tikzstyle{includeleft}=[left hook-angle 45,thick]
+    \node[doc] (cc) at (0,-1) {\shortstack{Content\\ Commons}};
+    \node[doc] (pl) at (5,-1) {\shortstack{Private/\\Local}};
+    \node[database] (db) at (3,2.5) {\shortstack{\tiny Verbalization\\Cache}};
+    \node[draw] (sn) at (0,4.5) {\snify};
+    \node[draw] (id) at (2.5,5) {IDE};
+    \node[draw] (wo) at (5,4.5) {\WOIDE};
+    \draw[draw,fill=black!10] (-1.3,.5) rectangle ++ (6.4,1); 
+    \node (f) at (-.5,.9) {\flams};
+    \draw[include]  (cc) -- node[left] {\tiny F2RDF} (db);
+    \draw[includeleft] (pl) -- node[right] {\tiny linter} (db);
+    \draw[->] (db) -- (sn);
+    \draw[->] (db) -- (id);
+    \draw[->] (db) -- (wo);
+    \draw[dashed,thick] (3,1.6)  -- (3,-2);
+    \draw[dashed,thick] (3,1.6)  -- (3,-2);
+    \draw[includeleft] (pl) to[bend right=10] (wo);
+  \end{tikzpicture}
+  \caption{Architecture}\label{fig:arch}\vspace*{-1em}
+\end{wrapfigure}
+To cope with this situation, we are working on extending the brand-new \flams system
+\cite{flams:on}\footnote{A paper on \flams is submitted to CICM 2025 and will be cited
+  here if accepted.} with functionality for handling definienda in the FTML2RDF harvester,
+so that all symbol/verbalizations are available in the \flams triple-store. Thus a harvest
+from the flexiformal content commons would be just a simple SPARQL query and some postprocessing
+of the result (de-duplication, stemming, and pairing) away. Note that \flams works
+directly on \FTML, which is generated from \sTeX input by \textsf{rus\TeX} and exported
+from annotated MS Word files via \WOIDE~\cite{KohKoh:woide24}.
+
+In local/private archives, the \snify harvester needs to work directly on the \sTeX
+sources.
+Currently, \snify uses a custom Python implementation for harvesting the verbalization cache,
+but we are working on a implementation based on the new \flams \textbf{linter} -- a simple \sTeX parser that
+drives the LSP functionality \sTeX IDE plugins for \textsf{VSCode} and \textsf{emacs}.
+% Here we have extended the \flams \textbf{linter} -- a simple \sTeX parser that
+% drives the LSP functionality the VSCode and \texttt{emacs} \sTeX IDE plugins use to deal
+% with definienda. This is much more efficient (and robust) than the initial regex-based
+% python parser).
+The analogon for this on the \WOIDE side is still under development it
+could be based on \FTML generated by \WOIDE, or \WOIDE could directly export symbol/verbalization
+pairs into the \snify verbalization cache.
+
+In this situation, the main functionality of \snify would be to maintain the verbalization cache
+and answer front-end queries for verbalizations efficiently so that we can power various
+front-ends from that. These include
+\begin{compactenum}[\em i\rm)]
+\item The traditional text/terminal-based \snify UI for bulk annotation,
+\item the \sTeX IDEs in \textsf{VSCode} and \textsf{emacs} that could supply annotation
+  support during editing, and 
+\item \WOIDE, which can be seen as a semantic authoring IDE based on MS Word.\ednote{MK:
+    CPoint and semantic wikis.}
+\end{compactenum}
+
+Note that, implemented this broadly, \snify could provide additional logistic services: If a
+pair is chosen in an annotation, \snify can then automatically download the corresponding
+archive and its dependencies so that the annotated file can be formatted with \texttt{pdflatex} or
+\textsf{rus\TeX} locally.
+
+\section{Related Work}\label{sec:relwork}
+The phrase ``semantic authoring'' has been used in the
+context of the semantic web with varying meanings
+(\cite{khalili2013user} provides an overview of the tool support).
+This includes ontology authoring (e.g.\ with \textsf{Prot{\'e}g{\'e}}
+\cite{musen2015protege}), which we would classify as
+formal authoring, and metadata annotation
+(e.g. the title, author, and creation date of a document with the Dublin Core vocabulary~\cite{DublinCore:on}),
+which is a somewhat different concern.
+A closer match is the annotation of
+informal content with references to a domain ontology (e.g.\ \cite{goerz2010adaptation}),
+but there are some differences to
+our situation with \sTeX/\ALeA: The domain ontology would be a fully formal target --
+possibly containing verbalizations -- while in our case the target is flexiformal.
+Furthermore, \sTeX has a module system, which requires dependency management during
+authoring.
+
+Another example of formal/informal authoring is code documentation.
+Often, the documentation is embedded in the code itself,
+e.g.\ in Python docstrings, which may follow a specific
+machine-readable format.
+IDEs, which are primarily used for (formal) code authoring,
+typically provide some support, e.g.\ by generating stubs
+and by extending refactoring support
+to (the machine-readable parts of) the documentation
+like parameter names and types.
+
+\snify's interface is inspired traditional spell checkers
+like \lstinline|ispell|~\cite{ispellman},
+which step through documents and suggest corrections for misspelled words.
+
+The bulk annotation workflow for existing informal text
+is also relevant when creating machine learning datasets,
+e.g.\ for named entity recognition.
+There are many (often browser-based) tools for this task;
+a prominent example is \textsf{brat}~\cite{brat:on}.
+
 \section{Conclusion}\label{sec:conclusion}
-
-In this paper we have presented the semantic authoring problem using the \sTeX corpus and
+In this paper we have presented the semantic authoring problem in
+a flexiformal context using the \sTeX corpus and
 the adaptive learning assistant \ALeA as a concrete example. We have shown that it differs
@@
 command-line tool that bridges the gap between informal text and formal symbols in \sTeX.
+Given the impact of \snify on the authoring process,
+we are currently working on a deeper integration of \snify into the \sTeX ecosystem and content commons
+as described in \Cref{sec:scaling}.
 
@@
 \cite{stextools:git}.
+
+\paragraph{Future Work}
+\snify depends on the existence of a comprehensive annotation catalog,
+which is a severe limitation for languages other than English.
+A potential solution would be to extend \snify with a machine
+learning-based tool for identifying unannotated technical terms that are not in the
+catalog.  The user could then pick a symbol for the annotation, which would add a
+verbalization to the catalog.  In an earlier (pre-\snify) attempt, we used named-entity
+recognition (NER) for classification of ``likely annotation candidate words''
+\cite{hutterer:msc23}. However, this classification was not precise enough in the
+distinction between technical terms and ordinary noun phrases to be practical on
+its own.  A combination with \snify's catalog-based approach might change the trade-offs
+involved.
 

### 2025-05-21 doc/cicm25-paper/paper.tex
more

@@
 of terms with the semantic concepts they refer to (\Cref{sec:snify}).
-Afterweards, we discuss on-going work to integrate \snify better
+Afterwards, we discuss on-going work to integrate \snify better
 into the \sTeX/\ALeA ecosystem and content commons (\Cref{sec:scaling}).
@@
 The work reported in this article was conducted as part of the VoLL-KI project (see
-\url{https://voll-ki.de}) funded by the German Research/Education Ministry under grant
-16DHBKI089.
+\url{https://voll-ki.de}) funded by the German Research/Education Ministry (BMBF) under
+grant 16DHBKI089.
 
@@
 object.  Verbalizations can vary between between subdomains and communities of
-practice. In our example, the verbalization ``terminal [state]'' was annotated with the
-symbol with name \lstinline|goal state|. It comes from the module with name
-\lstinline|search-problem| in the namespace \lstinline|smglom/search/mod| imported via the
-\lstinline|\importmodule| macro in line 1. As the symbol name \lstinline|goal state| is
-unique in the imports here it suffices as a relative URI in the \lstinline|\sr|
-annotation. If the symbol name and the desired verbalization coincide (as they often do in
-English), the short-hand macro \lstinline|\sn| can be used (e.g. as
-\lstinline|\sn{utility}| for ``utility'' in \Cref{lst:losa}).
+practice. In our example, the verbalization -- i.e. the natural language phrase used to
+express the symbol in English -- ``terminal [state]'' was annotated with the symbol with
+name \lstinline|goal state|. It comes from the module with name \lstinline|search-problem|
+in the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
+macro in line 1. As the symbol name \lstinline|goal state| is unique in the imports here
+it suffices as a relative URI in the \lstinline|\sr| annotation. If the symbol name and
+the desired verbalization coincide (as they often do in English), the short-hand macro
+\lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| for ``utility'' in
+\Cref{lst:losa}).
 
@@
 
-To annotate the word ``terminal'' in the document, the author has to be aware that it is a
-technical term, that the module from \Cref{fig:state-space} exists, know the symbol URI,
-and manage redundancy of the imports -- i.e. only adding the \lstinline|\importmodule|
-directive if it is not already (recursively) implied and possibly removing directives that
-become redundant by the new one.  At ca 150,000 words in a semester's worth of lecture
-notes and ca. 10--15\% of technical terms a rather daunting task.
+To annotate the word ``terminal'' in the document, the author has to 
+\begin{inparaenum}[\em i\rm)]
+\item be aware that it is a
+technical term, 
+\item that the module from \Cref{fig:state-space} exists,
+\item know the symbol URI, 
+\item manage redundancy of the imports -- i.e. only adding the \lstinline|\importmodule|
+  directive if it is not already (recursively) implied and possibly removing directives
+  that become redundant by the new one.
+\end{inparaenum}
+At ca 150,000 words in a semester's worth of lecture notes and ca. 10--15\% of technical
+terms a rather daunting task.
 
@@
 \paragraph{Traditional, informal authoring}
-In traditional authoring, e.g. of a scientific article or a textbook, the domain model -- a
-highly structured model of the domain of discourse -- is in the authors' heads, and is
+In traditional authoring, e.g. of a scientific article or a textbook, the domain model --
+a highly structured model of the domain of discourse -- is in the authors' heads, and is
 partially verbalized e.g. in the preliminaries section or the main sections of the
@@
 processing via the brain of a colleague or student -- in any case, a human who shares a
-canonical part of the domain model with the authors.\ednote{MK: this shared domain model
-  could be thought of as the informal content commons; hmmm, but maybe the literature is
-  the content commons} Consequently a large part of the authoring problem lies in
-predicting the state and extent of the domain model of the reader and creating text that
-bridges the gap between the reader's domain model and the payload content of the article
--- the knowledge the article or textbook intends to impart -- ideally using the technical
-vocabulary the reader is familiar with to reduce this gap. Tool support for traditional
-authoring is minimal and usually restricted to spell/grammar-checkers and online thesauri
-if we disregard LLM-based writing/formulation support for now.
+canonical part of the domain model with the authors. Consequently a large part of the
+authoring problem lies in predicting the state and extent of the domain model of the
+reader and creating text that bridges the gap between the reader's domain model and the
+payload content of the article -- the knowledge the article or textbook intends to impart
+-- ideally using the technical vocabulary the reader is familiar with to reduce this
+gap. Tool support for traditional authoring is minimal and usually restricted to
+spell/grammar-checkers and online thesauri if we disregard LLM-based writing/formulation
+support for now.
 
 \paragraph{Formal Authoring}
-In fully formal authoring, e.g. programming, the content commons is given formally in
-the form of the libraries of the respective programming language, correspondingly the
-authored content -- a program -- is also formal and intended for processing by a computer,
-for which the formality is a prerequisite. Formal authoring is usually supported by an
-integrated development environment (IDE) that makes use of the fully formal content and
-offers services like semantic highlighting, tab completion, jumping to variable
-declarations, function definitions, and interface definitions, as well as refactoring
-support. All of these rely on a syntactic/semantic analysis of the formal content. The
-situation is very similar for authoring of formal knowledge, e.g. in program verification
-or the formalization of mathematics.
+In programming, the content commons is given formally in the form of the libraries of the
+respective programming language, correspondingly the authored content -- a program -- is
+also formal and intended for processing by a computer, for which the formality is a
+prerequisite. Formal authoring is usually supported by an integrated development
+environment (IDE) that makes use of the fully formal content and offers services like
+semantic highlighting, tab completion, jumping to variable declarations, function
+definitions, and interface definitions, as well as refactoring support. All of these rely
+on a syntactic/semantic analysis of the formal content. The situation is very similar for
+authoring of formal knowledge, e.g. in program verification or the formalization of
+mathematics.
 
@@
 
-\begin{newpart}{MK: please re-read}
 A distinctive feature of semantic authoring that is not apparent at the first glance is
@@
 documents can be automatically customized if the semantics has multiple
-  notations/verbalizations/presentations. Consider for instance course materials on
-  complex arithmetic, which become re-usable across the electrical engineering/mathematics
-  border if the imaginary unit in formulae is represented e.g. by a semantic macro
-  \lstinline|\imaginaryunit|, which can alternatively expand to $i$ (for uses in maths)
-  and $j$ (for EE uses). As a consequence semantic authoring is more geared towards
-  creating reusable/shareable document fragments organized in central or federated
-  repositories and hubs. This is quite similar to the formal authoring system, where we
-  see a tendency towards massively shared math libraries in formal systems and package
-  hubs for programming libraries. As a consequence, semantic authoring support systems
-  should facilitate the logistics of using and curating shared libraries.
-\end{newpart}
+notations/verbalizations/presentations. Consider for instance course materials on complex
+arithmetic, which become re-usable across the electrical engineering/mathematics border if
+the imaginary unit in formulae is represented e.g. by a semantic macro
+\lstinline|\imaginaryunit|, which can alternatively expand to $i$ (for uses in maths) and
+$j$ (for EE uses). As a consequence semantic authoring is more geared towards creating
+reusable/shareable document fragments organized in central or federated repositories and
+hubs. This is quite similar to the formal authoring arena, where we see a tendency towards
+massively shared math libraries in formal systems and package/library hubs for programming
+libraries. As a consequence, semantic authoring support systems should facilitate the
+logistics of using and curating shared libraries.
 
@@
 the \WOIDE plugin \cite{KohKoh:woide24}, which allows authors to use Word's spell
-checker and grammar checker.\ednote{MK: Mention CPoint and SMD as well? Probably yes.}
+checker and grammar checker.
 
@@
 documents that were originally authored without semantic annotations.  It can be
-configured to use different \LaTeX\ macros to identify definienda (e.g.\ \lstinline[language={}]|\emph|
-or \lstinline[language={}]|\textbf|), and then steps through them similarly to \snify.  For each
-potential definiendum, the user can then use fuzzy search to find the corresponding symbol
-in the domain model.  This populates the catalog with verbalizations relevant to the
-document, making subsequent \snify runs more effective, especially for documents in
-languages that are not well covered by the domain model (i.e.\ not English).
+configured to consider different \LaTeX\ macros to identify definienda (e.g.\
+\lstinline[language={}]|\emph| or \lstinline[language={}]|\textbf|), and then steps
+through them as \snify does.  For each potential definiendum, the user can then use fuzzy
+search to find the corresponding symbol in the domain model.  This populates the catalog
+with verbalizations relevant to the document, making subsequent \snify runs more
+effective, especially for documents in languages that are not well covered by the domain
+model (i.e.\ not English).
 
@@
 \item a set of local working copies of the archives -- with the intention of publishing
-  them on MathHub.info -- and their dependencies (they need to be local since pdflatex
-  cannot deal with remote files) and
+  them on MathHub.info -- and their dependencies (they need to be local since
+  \textsf{pdflatex} cannot deal with remote files) and
 \item possibly a set of private archives (e.g.\ papers under development or exam
@@
 so that all symbol/verbalizations are available in the \flams triple-store. Thus a harvest
-from the flexiformal content commons would be just a simple SPARQL query and some postprocessing
-of the result (de-duplication, stemming, and pairing) away. Note that \flams works
-directly on \FTML, which is generated from \sTeX input by \textsf{rus\TeX} and exported
-from annotated MS Word files via \WOIDE~\cite{KohKoh:woide24}.
+from the flexiformal content commons would be just a simple SPARQL query and some
+postprocessing of the result (de-duplication, stemming, and pairing) away. Note that
+\flams works directly on \FTML, which is generated from \sTeX input by \textsf{rus\TeX}
+and exported from annotated MS Word files via \WOIDE~\cite{KohKoh:woide24}. There are two
+more systems that could profit from this: The CPoint system that allows annotating MS
+PowerPoint presentations with the \sTeX ontology and can already generate a precursor of
+\FTML \cite{Kohlhase:SemanticInteractionDesignDiss:biblatex}, and a semantic extension of
+Markdown we are developing as a lightweight alternative to \sTeX; it also transforms into
+FTML instead of plain HTML.
 
@@
   support during editing, and 
-\item \WOIDE, which can be seen as a semantic authoring IDE based on MS Word.\ednote{MK:
-    CPoint and semantic wikis.}
+\item \WOIDE, which can be seen as a semantic authoring IDE based on MS Word.
 \end{compactenum}
@@
 
-The bulk annotation workflow for existing informal text
-is also relevant when creating machine learning datasets,
-e.g.\ for named entity recognition.
-There are many (often browser-based) tools for this task;
-a prominent example is \textsf{brat}~\cite{brat:on}.
+The bulk annotation workflow for existing informal text is also relevant when creating
+machine learning datasets, e.g.\ for named entity recognition.  There are many (often
+browser-based) tools for this task; a prominent example is
+\textsf{brat}~\cite{brat:on}.\ednote{MK@FS: discuss semantic wikis}
+
+In all cases, the setting, solution, and resulting workflows differ from the one presented
+in this paper. The WissKI system \cite{goerz2010adaptation} has a rudimentary text edition
+facility that allows annotating document with ontology -- mostly ABox -- references; this
+is the nearest match to our \snify system.
 
 \section{Conclusion}\label{sec:conclusion}
-In this paper we have presented the semantic authoring problem in
-a flexiformal context using the \sTeX corpus and
-the adaptive learning assistant \ALeA as a concrete example. We have shown that it differs
-from both classical authoring and formal authoring in terms of the necessary system
-support. As an example of specialized authoring support, we have presented \snify, a
-command-line tool that bridges the gap between informal text and formal symbols in \sTeX.
-Given the impact of \snify on the authoring process,
-we are currently working on a deeper integration of \snify into the \sTeX ecosystem and content commons
-as described in \Cref{sec:scaling}.
+In this paper we have presented the semantic authoring problem in a flexiformal context
+using the \sTeX corpus and the adaptive learning assistant \ALeA as a concrete example. We
+have shown that it differs from both classical authoring and formal authoring in terms of
+the necessary system support. As an example of specialized authoring support, we have
+presented \snify, a command-line tool that bridges the gap between informal text and
+formal symbols in \sTeX.  Given the impact of \snify on the authoring process, we are
+currently working on a deeper integration of \snify into the \sTeX ecosystem and content
+commons as described in \Cref{sec:scaling}.
 
@@
 % LocalWords:  lang DHBKI FS Todo FTML lexiformal arkup anguage CICM RDF postprocessing
-% LocalWords:  de maths EE CPoint SMD rus WOIDE linter LSP emacs
+% LocalWords:  de maths EE CPoint SMD rus WOIDE linter LSP emacs Prot BMBF

### 2025-05-22 doc/cicm25-paper/paper.tex
final touches

@@
 \usepackage{paralist}
-\usepackage{wrapfig}
 \usepackage[hidelinks]{hyperref}
@@
 
+To cope with this situation, we are working on extending the brand-new \flams system
+\cite{flams:on}\footnote{A paper on \flams is submitted to CICM 2025 and will be cited
+  here if accepted.} with functionality for handling definienda in the FTML2RDF harvester,
+so that all symbol/verbalizations are available in the \flams triple-store. Thus a harvest
+from the flexiformal content commons would be just a simple SPARQL query and some
+postprocessing of the result (de-duplication, stemming, and pairing) away. Note that
+\flams works directly on \FTML, which is generated from \sTeX input by \textsf{rus\TeX}
+and exported from annotated MS Word files via \WOIDE~\cite{KohKoh:woide24}. There are two
+more systems that could profit from this: The CPoint system that allows annotating MS
+PowerPoint presentations with the \sTeX ontology and can already generate a precursor of
+\FTML \cite{Kohlhase:SemanticInteractionDesignDiss:biblatex}, and a semantic extension of
+Markdown we are developing as a lightweight alternative to \sTeX; it also transforms into
+FTML instead of plain HTML.
+
 \begin{wrapfigure}r{5.5cm}\vspace*{-2em}
@@
 \end{wrapfigure}
-To cope with this situation, we are working on extending the brand-new \flams system
-\cite{flams:on}\footnote{A paper on \flams is submitted to CICM 2025 and will be cited
-  here if accepted.} with functionality for handling definienda in the FTML2RDF harvester,
-so that all symbol/verbalizations are available in the \flams triple-store. Thus a harvest
-from the flexiformal content commons would be just a simple SPARQL query and some
-postprocessing of the result (de-duplication, stemming, and pairing) away. Note that
-\flams works directly on \FTML, which is generated from \sTeX input by \textsf{rus\TeX}
-and exported from annotated MS Word files via \WOIDE~\cite{KohKoh:woide24}. There are two
-more systems that could profit from this: The CPoint system that allows annotating MS
-PowerPoint presentations with the \sTeX ontology and can already generate a precursor of
-\FTML \cite{Kohlhase:SemanticInteractionDesignDiss:biblatex}, and a semantic extension of
-Markdown we are developing as a lightweight alternative to \sTeX; it also transforms into
-FTML instead of plain HTML.
 
@@
 authoring.
+A similar application requiring semantic authoring
+are semantic wikis, which enrich regular/syntactic
+wikis with semantic information
+(see e.g.\ \cite{semmediawiki}).
 
@@
 browser-based) tools for this task; a prominent example is
-\textsf{brat}~\cite{brat:on}.\ednote{MK@FS: discuss semantic wikis}
+\textsf{brat}~\cite{brat:on}.
 

### 2025-05-22 doc/cicm25-paper/paper.tex
some more fixes

@@
 With \sTeX, authors can add semantic references to technical terms,
-which enables straight-forward learning support services like
+which enables straightforward learning support services like
 showing the definition of a term, but also informs the system
@@
   \fbox{\includegraphics[width=12cm,trim={0 3cm 0 0},clip]{../img/snify}}
-  \caption{\snify in Action: Annotating the word ``terminal''.}\label{fig:snify}
+  \caption{\snify in Action: Annotating the word ``terminal''}\label{fig:snify}
 \end{figure}
@@
 % python parser).
-The analogon for this on the \WOIDE side is still under development it
+The analogon for this on the \WOIDE side is still under development; it
 could be based on \FTML generated by \WOIDE, or \WOIDE could directly export symbol/verbalization
@@
 in this paper. The WissKI system \cite{goerz2010adaptation} has a rudimentary text edition
-facility that allows annotating document with ontology -- mostly ABox -- references; this
+facility that allows annotating documents with ontology -- mostly ABox -- references; this
 is the nearest match to our \snify system.

### 2025-07-03 doc/cicm25-paper/rebuttal.txt
preamble

We thank the reviewers for their thorough and thoughtful comments and suggestions.

We concur with the reviewers that the paper is somewhere between a research paper and a
system description, and we have struggled with this in the paper writing process. Indeed
we had started out as the former, and then realized that there is more to the topic than a
system description could convey. We did not try to bloat the text to make it more suitable
to the category, but tried to layout the task of semantic authoring -- which we see as a
novel category -- as a general problem to be addressed by the CICM community.

### 2025-07-03 doc/cicm25-paper/rebuttal.txt
rebuttal work-in-progress

@@
 system description, and we have struggled with this in the paper writing process. Indeed
-we had started out as the former, and then realized that there is more to the topic than a
-system description could convey. We did not try to bloat the text to make it more suitable
+we had started out with the former, and then realized that there is more to the topic than a
+system description could convey.
+We did not try to bloat the text to make it more suitable
 to the category, but tried to layout the task of semantic authoring -- which we see as a
 novel category -- as a general problem to be addressed by the CICM community.
+Given reviewer 1's feedback, we will take hard look at the paper
+and see where we can shorten/streamline it.
+While writing, we assumed that not all readers would find the ideas as easy as reviewer 1 expressed, considering the broad range of backgrounds at CICM.
+
+We will, of course, incorporate the minor corrections suggested by the reviewers.
+
+To address a few questions/concerns:
+
+> While the whole project is a significant effort, the special tools described here  'snify' and 'defianno' seem to this reviewer not particularly able to scale up in the way desired.
+The measurement proposed for the increase of productivity seems very much subjective and the 'STEX/ALeA content commons' also seem specialized and not very `common', i.e.
+not shared by many.
+The problem with ambiguity of definitions and their contexts also seems not very easy to address with the tools proposed.
+
+TODO: This summarizes the criticism of rev. 2... Not sure what to do with the first and last sentence.
+
+> 'STEX/ALeA content commons' also seem specialized and not very `common', i.e. not shared by many.
+
+True.
+Most content is about mathematics and computer science,
+but some other disciplines are represented to a lesser extent (e.g. law and linguistics).
+Calling it a 'commons' is somewhat aspirational.
+
+> be more precise and quantitative when you write "one order of magnitude more efficient"
+
+We agree.
+The efficiency gain cannot easily be quantified.
+It strongly depends on the familiarity of the author with the domain model.
+Also, authors tend to skip more technical terms during annotation
+if they are not using snify.
+We measured values 
+ranging from a factor of ?? (for an author who knows the domain model by heart),
+to a factor of ?? (for an author who has never annotated before).
+We will elaborate on this in the paper.
+
+> Why do symbolic technologies offer explainability "out of the box"?
+
+> where does [the 150,000 words] estimate come from?
+
+This is an estimate based on our own lecture materials (a combination of slides and
+notes), specifically two lectures on artificial intelligence.
+The concrete numbers will of course vary by course and by instructor,
+but we were trying to give an idea of the scale.
+We will clarify this in the paper.
+
+> Is there more related work?

### 2025-07-03 doc/cicm25-paper/rebuttal.txt
more

@@
 system description could convey.
-We did not try to bloat the text to make it more suitable
-to the category, but tried to layout the task of semantic authoring -- which we see as a
-novel category -- as a general problem to be addressed by the CICM community.
-Given reviewer 1's feedback, we will take hard look at the paper
-and see where we can shorten/streamline it.
-While writing, we assumed that not all readers would find the ideas as easy as reviewer 1 expressed, considering the broad range of backgrounds at CICM.
 
-We will, of course, incorporate the minor corrections suggested by the reviewers.
+Howver, we did not try to bloat the text to make it more suitable to the category, but
+tried to layout the task of semantic authoring -- which we see as a novel task category --
+as a general problem to be addressed by the CICM community.  Given reviewer 1's feedback,
+we will take hard look at the paper and see where we can shorten/streamline it.  While
+writing, we assumed that not all readers would find the ideas as easy as reviewer 1
+expressed, considering the broad range of backgrounds at CICM.
 
-To address a few questions/concerns:
+We will, of course, incorporate the minor corrections suggested by the reviewers; these
+are very helpful to us.
+
+To address a few other questions/concerns:
 
@@
 We agree.
-The efficiency gain cannot easily be quantified.
-It strongly depends on the familiarity of the author with the domain model.
-Also, authors tend to skip more technical terms during annotation
-if they are not using snify.
+The efficiency gain cannot easily be quantified and we have therefore kept the assertions
+about that intentionally vague.
+It strongly depends on the author's familiarity with the domain model (having to search
+for the system URI of a concept that "should be in the domain model" is so time-consuming
+that we get the efficiency gain of "one order of magnitude in this part alone") .
+Also, authors tend to skip more technical terms during annotation if they are not using
+snify; so they retain efficiency at the cost of lowering the quality of annotation.
 We measured values 
-ranging from a factor of ?? (for an author who knows the domain model by heart),
-to a factor of ?? (for an author who has never annotated before).
-We will elaborate on this in the paper.
+ranging from a factor of 3 (for an author who knows the domain model by heart),
+to a factor of 20 (for an author who has never annotated before).
+Given the reviewer questions we will elaborate on this in the paper.
 
 > Why do symbolic technologies offer explainability "out of the box"?
+
+This statement has to be read in the context of comparing with machine-learning or neural
+AI technologies, where the "context and inference" only exist in the form of a
+weight/parameter distribution over the network. In symbolic technolgies, the inferences
+(and/or transformation steps) can be traced and turned into argumentations or
+justifications. While this may not be technically be "out of the box" at least the
+necessary resources are all there.
 
@@
 > Is there more related work?
+
+XXXX not to our knowledge, otherwise we would have discussed it XXXX

### 2025-07-04 doc/cicm25-paper/rebuttal.txt
a few more changes

--- 2025-07-03 doc/cicm25-paper/rebuttal.txt
@@
-We thank the reviewers for their thorough and thoughtful comments and suggestions.
+We thank the reviewers for their thorough and thoughtful comments and
+suggestions.
 
-We concur with the reviewers that the paper is somewhere between a research paper and a
-system description, and we have struggled with this in the paper writing process. Indeed
-we had started out as the former, and then realized that there is more to the topic than a
-system description could convey. We did not try to bloat the text to make it more suitable
-to the category, but tried to layout the task of semantic authoring -- which we see as a
-novel category -- as a general problem to be addressed by the CICM community.
+We concur with the reviewers that the paper is somewhere between a research
+paper and a system description, and we have struggled with this in the paper
+writing process. Indeed we had started out with the former, and then realized
+that there is more to the topic than a system description could convey.
+However, we did not try to bloat the text to make it more suitable to the
+category, but tried to layout the task of semantic authoring -- which we see as
+a novel task category -- as a general problem to be addressed by the CICM
+community.  Given reviewer 1's feedback, we will take hard look at the paper
+and see where we can shorten/streamline it.  While writing, we assumed that not
+all readers would find the ideas as easy as reviewer 1 indicated, considering
+the broad range of backgrounds at CICM.
+
+Reviewer 2 is skeptical regarding the scalability. snify is by now the main
+tool for term annotation in sTeX and has been used for thousands (likely over
+10 000) annotations. Nevertheless, maintaining a catalog across the growing
+ecosystem is a challenge we are currently working on. It is true that the
+'sTeX/ALeA content commons' is mostly shared within our group (though across
+many different lectures), but we hope to grow the community and are in the
+early stages of collaboration with lecturers from IISc and Bath University.
+
+Reviewer 2 also mentions the problem of ambiguity. The annotations disambiguate
+the meaning of terms with the granularity of the domain model, which may be
+insufficient in some cases (e.g. if there is only one entry for natural
+numbers, regardless of whether they include 0 or not).
+
+Reviewer 3 rightfully points out that it would be better to state the
+efficiency gains more precisely. Unfortunately, the efficiency gain cannot
+easily be quantified in a number and have therefore kept the assertions vague.
+It strongly depends on the author's familiarity with the domain model (having
+to search for the system URI of a concept that "should be in the domain model"
+is very time-consuming). Also, authors tend to skip more technical terms
+during annotation if they are not using snify; so they retain efficiency at the
+cost of lowering the quality of annotation. We measured values ranging from a
+factor of 3 (for an author who knows the domain model by heart), to a factor of
+20 (for an author who has never annotated before). The focus mode of snify
+further improves the efficiency in terms of annotations per time. Given the
+reviewer questions we will elaborate on this in the paper.
+
+We will, of course, incorporate the minor corrections suggested by the
+reviewers; these are very helpful to us.
+
+To address some of the minor questions/concerns:
+
+> Reviewer 1 asks why symbolic technologies offer explainability "out of the
+> box"
+
+This statement has to be read in the context of comparing with machine-learning
+or neural AI technologies, where the "context and inference" only exist in the
+form of a weight/parameter distribution over the network. In symbolic
+technologies, the inferences (and/or transformation steps) can be traced and
+turned into argumentations or justifications. While this may not be technically
+be "out of the box" at least the necessary resources are all there. We will
+improve the formulation.
+
+> Reviewer 2 asks where the 150,000 words/10-15% estimate comes from
+
+This is an estimate based on our own lecture materials (a combination of slides
+and notes), specifically our (admittedly rather large) lecture on artificial
+intelligence. The concrete numbers will of course vary by course and by
+instructor, but we were trying to give an idea of the scale. We will clarify
+this in the paper.
+
+> Reviewer 2 asks if there is more related work
+
+To our knowledge, we have covered the spectrum of related work, but we would be
+grateful for pointers to any work we may have missed.
+
+> Reviewer 2 suggests that our use of "verbalization" should be explained
+
+For us, the verbalization of a symbol is its rendering in (written) natural
+language. For example, "integer" and "whole number" are English verbalizations
+of the symbol for integers. We will clarify this in the paper.

### 2025-07-05 doc/cicm25-paper/rebuttal.txt
final version

@@
 
-We concur with the reviewers that the paper is somewhere between a research
-paper and a system description, and we have struggled with this in the paper
-writing process. Indeed we had started out with the former, and then realized
-that there is more to the topic than a system description could convey.
-However, we did not try to bloat the text to make it more suitable to the
-category, but tried to layout the task of semantic authoring -- which we see as
-a novel task category -- as a general problem to be addressed by the CICM
-community.  Given reviewer 1's feedback, we will take hard look at the paper
-and see where we can shorten/streamline it.  While writing, we assumed that not
-all readers would find the ideas as easy as reviewer 1 indicated, considering
-the broad range of backgrounds at CICM.
+We admit that the paper is somewhere between a research paper and a system
+description, and we have struggled with this in the paper writing process.
+Indeed we had started out with the former, and then realized that there is more
+to the topic than a system description could convey. However, we did not try to
+bloat the text to make it more suitable to the category, but tried to layout
+the task of semantic authoring -- which we see as a novel task category -- as a
+general problem to be addressed by the CICM community.  Given reviewer 1's
+feedback, we will take hard look at the paper and see where we can
+shorten/streamline it.  While writing, we assumed that not all readers would
+find the ideas as easy as reviewer 1 indicated, considering the broad range of
+backgrounds at CICM.
 
@@
 
-Reviewer 2 also mentions the problem of ambiguity. The annotations disambiguate
-the meaning of terms with the granularity of the domain model, which may be
-insufficient in some cases (e.g. if there is only one entry for natural
-numbers, regardless of whether they include 0 or not).
-
 Reviewer 3 rightfully points out that it would be better to state the
 efficiency gains more precisely. Unfortunately, the efficiency gain cannot
-easily be quantified in a number and have therefore kept the assertions vague.
-It strongly depends on the author's familiarity with the domain model (having
-to search for the system URI of a concept that "should be in the domain model"
-is very time-consuming). Also, authors tend to skip more technical terms
+easily be quantified in a number and we have therefore kept the assertions
+vague. It strongly depends on the author's familiarity with the domain model
+(having to search for the system URI of a concept that "should be in the domain
+model" is very time-consuming). Also, authors tend to skip more technical terms
 during annotation if they are not using snify; so they retain efficiency at the
 cost of lowering the quality of annotation. We measured values ranging from a
-factor of 3 (for an author who knows the domain model by heart), to a factor of
-20 (for an author who has never annotated before). The focus mode of snify
-further improves the efficiency in terms of annotations per time. Given the
-reviewer questions we will elaborate on this in the paper.
+factor of 3 (for an IDE-supported expert who knows the domain model by heart),
+to a factor of 13 (for an inexperienced annotator with IDE support). The focus
+mode of snify further improves the efficiency in terms of annotations per time.
+Given the reviewer questions we will elaborate on this in the paper.
 
@@
 be "out of the box" at least the necessary resources are all there. We will
-improve the formulation.
+improve the formulation in the paper.
 

### 2025-07-23 doc/cicm25-paper/paper.tex
reworked paper

@@
 \usetikzlibrary{docicon}
+\usepackage{soul}
+\def\edited#1{\hl{#1}}
 
@@
 \def\rrangle{\rangle\kern-.2em\rangle}
+\def\flams{\textsf{FLAMS}\xspace}
+
+\lstset{literate=--1}
 
@@
     % \ednote{Find better title. Was ``Bulk Semantic Annotation with a Partially Known Knowledge Base''}
-Semantic authoring in a flexiformal context --- Bulk annotation of rigorous documents
+        Semantic Authoring in a Flexiformal Context --- Bulk Annotation of Rigorous Documents
 }
 \titlerunning{
-    Semantic authoring in a flexiformal context \textellipsis
+    Semantic Authoring in a Flexiformal Context \textellipsis
 }
@@
   evaluate a simple utility that supports the process of annotating technical terms with
-  semantic references to a flexiformal content commons that enhances annotation
+    semantic references to a flexiformal content commons \edited{and increases} annotation
   productivity by almost an order of magnitude.
@@
 \section{Introduction}
-Arguably, dealing with large document collections is one of the key factors in the knowledge-driven society and economy.
-There are currently two main contenders for machine support in this area.
-Symbolic/logic-based technologies and sub-symbolic/machine-learning-based AI, e.g.\ via LLMs or chatbots.
+Arguably, handling large document collections is a key factor in our knowledge-driven society and economy.
+There are two main contenders for machine support in this area:
+symbolic/logic-based technologies and \edited{statistical}/machine-learning-based AI, e.g.\ via LLMs or chatbots.
 They have complementary strengths and challenges:
 ML-based approaches can be trained on all data of the Internet, but face challenges in precision and explainability. 
-Symbolic technologies, on the other hand, offer precision and explainability out of the box, but face scalability challenges
+Symbolic technologies, on the other hand, \edited{offer precision and explainability%
+%out of the box%
+, but face scalability challenges}
 -- in particular because of the cost of semantic authoring, which we discuss in this paper.
@@
 
-\paragraph{Overview \& Contribution}
+\paragraph{Overview and Contribution}
 In this paper, we argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
@@
 
-Before we discuss semantic authoring in general, let us introduce a concrete example from
-the \sTeX/\ALeA world. We will use it to sharpen our intuitions of the challenges
-involved and to ground the solution we propose.
+Before we discuss semantic authoring in general, let us introduce a concrete example from the \sTeX/\ALeA world.
+    % We will use it to sharpen our intuitions of the challenges involved and to ground the solution we propose.
 
@@
 
-\begin{figure}[ht]\centering
+\begin{figure}[h]\centering
   \fbox{\includegraphics[width=10cm]{../img/minimax-remarks.pdf}}
@@
 
-Take for instance a learning object like the paragraph in \Cref{fig:lo}, where we want to
-annotate the term ``terminal'' with the (semantic) concept of ``a goal state of a search
-problem''. The latter can either be defined earlier in the course or in the domain model
-(or both). Concretely this is about converting the {\LaTeX} string
-
-\begin{lstlisting}[numbers=left,firstnumber=3,
-caption=The unannotated {\LaTeX} sources of \Cref{fig:lo},label=lst:los]
-\item \blue{Note:} Depth-limited minimax requires an evaluation for every
-cut-off state $s$. If $s$ is terminal, we use its utility,
-and otherwise an estimate.
-\end{lstlisting}
-into the partially annotated item:
-\begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
-caption=Annotating ``terminal'' in \Cref{lst:los},label=lst:losa]
-\importmodule[smglom/search]{mod?search-problem}
-[...]
-\item \blue{Note:} Depth-limited minimax requires an evaluation for every
-cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its utility,
-and otherwise an estimate.
-\end{lstlisting}
-In the latter, the author has annotated the word ``terminal'' with the concept it refers
-to with via the \sTeX macro
-\lstinline[mathescape]|\sr{$\llangle URI \rrangle$}{$\llangle verbalization\rrangle$}|.
-
-In \sTeX, objects and concepts are represented by \textbf{symbols}, which are identified
-globally and unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module
-  name}, and a \textbf{symbol name}. Every symbol can have one or more
-\textbf{verbalizations}: words of the domain jargon used to refer to the concept or
-object.  Verbalizations can vary between between subdomains and communities of
-practice. In our example, the verbalization -- i.e. the natural language phrase used to
-express the symbol in English -- ``terminal [state]'' was annotated with the symbol with
-name \lstinline|goal state|. It comes from the module with name \lstinline|search-problem|
-in the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
-macro in line 1. As the symbol name \lstinline|goal state| is unique in the imports here
-it suffices as a relative URI in the \lstinline|\sr| annotation. If the symbol name and
-the desired verbalization coincide (as they often do in English), the short-hand macro
-\lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| for ``utility'' in
-\Cref{lst:losa}).
-
-The annotation refers to a definition in the domain model module shown in
-\Cref{fig:state-space}.  The \lstinline|\definame| and \lstinline|\definiendum| in lines
-5--6 introduce the verbalizations ``goal state'' and ``terminal state'' for the symbol
-\lstinline|search-problem?goal state| introduced by \lstinline|\symdecl*| declaration in
-line 3.
-
-\begin{figure}[ht]\centering
+\begin{figure}[h]\centering
   \fbox{\includegraphics[width=12cm]{../img/search-problem.en.pdf}}
@@
 
+Take for instance a learning object like the paragraph in \Cref{fig:lo}, where we want to
+annotate the term ``terminal'' with the (semantic) concept of ``a goal state of a search
+problem''. The latter can either be defined earlier in the course or in the domain model
+(or both). Concretely this is about converting the unannotated {\LaTeX} string
+
+\begin{lstlisting}[numbers=left,firstnumber=3,
+% caption=The unannotated {\LaTeX} sources of \Cref{fig:lo},label=lst:los
+]
+\item \blue{Note:} Depth-limited minimax requires an evaluation for every
+cut-off state $s$. If $s$ is terminal, we use its utility,
+and otherwise an estimate.
+\end{lstlisting}
+into the partially annotated item
+\begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
+% caption=Annotating ``terminal'' in \Cref{lst:los},label=lst:losa
+]
+\importmodule[smglom/search]{mod?search-problem}
+[...]
+\item \blue{Note:} Depth-limited minimax requires an evaluation for every
+cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its utility,
+and otherwise an estimate.
+\end{lstlisting}
+\edited{
+where the author has annotated the word ``terminal'' with the concept it refers
+to via} the \sTeX macro
+\lstinline[mathescape]|\sr{$\llangle \mathit{symbol} \rrangle$}{$\llangle \mathit{verbalization}\rrangle$}|.
+In \sTeX, \edited{concepts} are represented by \textbf{symbols}, which are identified
+globally and unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module name}, and a \textbf{symbol name}.
+ % Every symbol can have one or more verbalizations. %: words of the domain jargon used to refer to the concept or
+\edited{
+A \textbf{verbalization} is the natural language rendering of a concept.
+% object.
+Verbalizations are, of course, language-dependent and can vary between subdomains and communities of
+practice.
+}
+In our example, the verbalization ``terminal [state]'' was annotated with the symbol with
+name \lstinline|goal state|. It comes from the module with name \lstinline|search-problem|
+in the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
+macro in line 1. As the symbol name \lstinline|goal state| is unique in the imports here
+it suffices as a relative URI in the \lstinline|\sr| annotation. If the symbol name and
+the desired verbalization coincide (as they often do in English), the short-hand macro
+\lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| for ``utility'').
+% in \Cref{lst:losa}).
+
+The annotation refers to a definition in the domain model module shown in
+\Cref{fig:state-space}.  The \lstinline|\definame| and \lstinline|\definiendum| in lines
+5--6 introduce the verbalizations ``goal state'' and ``terminal state'' for the symbol
+\lstinline|search-problem?goal state| introduced by the \lstinline|\symdecl*| declaration in
+line 3.
+
 \subsection{Annotation Workflows and Requirements}\label{sec:workflows}
@@
 \end{inparaenum}
-At ca 150,000 words in a semester's worth of lecture notes and ca. 10--15\% of technical
-terms a rather daunting task.
+\edited{A rather daunting task for a lecture with 50,000 -- 100,000 words in notes and slides and ca. 10--15\,\% of them technical terms.\footnote{
+    \edited{
+        Estimates based on some of our own lectures materials
+        ranging from a small introductory lecture (ca. 30,000 words)
+        to a large two-semester lecture on AI (ca. 140,000 words).
+        The 10--15\,\% estimate appears to be relatively consistent across our lectures
+        but depends on the author's intuition about what a technical term is.
+        Other lecturers or lectures on other topics may have very different values.
+    }
+}}
 
@@
 
-For documents in a language $L$,
-where $L$ is not English, the annotation practices are a bit
-more complex still, in particular because the annotation catalog is usually incomplete.
-
-Note that for $L$-annotation, we need an $L$-specific annotation catalog. 
+\edited{The annotation catalog is language-specific.}
+% For documents in a language $L$,
+% where $L$ is not English, the annotation practices are a bit
+% more complex still, in particular because the annotation catalog is usually incomplete \edited{for $L$}.
+% Note that for $L$-annotation, we need an $L$-specific annotation catalog. 
 \sTeX already supports multilinguality:
@@
   ISO-639 language identifier).
-\item Modules in the domain model allow $L$-translations that share the formal
-  declarations with another module. In particular all definienda in $L$-modules are
-  $L$-specific.
+\item \edited{Modules in the domain model allow translations that share the formal
+    declarations with another module.}
+        % In particular all definienda in $L$-modules are $L$-specific.
 \end{itemize}
-
-But still, $L$-annotation catalogs are usually inadequate for $L$-annotation tasks in
-practice, as
+% But still, $L$-annotation catalogs are usually inadequate for $L$-annotation tasks in
+\edited{In practice, the annotation catalog for languages other than English is often inadequate as}
 \begin{enumerate}[\em i\rm)]
-\item $L$-verbalizations and symbol names do not often coincide and
-\item the domain model may not even have $L$-modules -- especially for the ``smaller''
-  languages.
+    \item \edited{the domain model may not have translations in that language -- especially for the ``smaller'' languages, and}
+    \item \edited{symbol names, usually English words, cannot be used as a verbalizations.}
 \end{enumerate}
- In this situation, we usually have to invest in
-establishing a suitable catalog e.g. by providing $L$-translations (or at least dictionary
-information) in the domain model or delineating modules and annotating definienda that the
+\edited{In this situation, we usually have to invest in
+establishing a suitable catalog -- e.g. by providing translations (or at least dictionary
+information) in the domain model or
+delineating modules and annotating definienda that the
 domain model does not cover in the documents to be annotated.
+}
 
@@
 In programming, the content commons is given formally in the form of the libraries of the
-respective programming language, correspondingly the authored content -- a program -- is
+respective programming language. Correspondingly the authored content -- a program -- is
 also formal and intended for processing by a computer, for which the formality is a
@@
 massively shared math libraries in formal systems and package/library hubs for programming
-libraries. As a consequence, semantic authoring support systems should facilitate the
+libraries. Semantic authoring support systems should therefore facilitate the
 logistics of using and curating shared libraries.
@@
 only support existing annotations and are, therefore, of limited use when creating new
-annotations, which arguably is the most time consuming aspects of semantic authoring.
+annotations, which arguably is the most time consuming aspect of semantic authoring.
 % \ednote{also: SPARQL API}
@@
 \cite{stextools:git} is a simple command-line tool that creates a catalog of
-symbol-verbalization pairs, analyzes \sTeX source files, and then steps through all word
+symbol-verbalization pairs (see also \Cref{sec:workflows}), analyzes \sTeX source files, and then steps through all word
 occurrences in the document that match a verbalization in the catalog.  Matching is based
@@
 top-level lecture notes; then the session consists of all (transitively) included
-files. Other productivity features include a focus mode that can be used to only annotate
+files. Another productivity feature is a focus mode that can be used to only annotate
 a particular word in the rest of the file, session, or the whole \sTeX/\ALeA corpus before
@@
 
-\paragraph{\defianno: Definition Handling as a Pre-process for Multilingual Documents}
+\paragraph{\defianno: Definition Handling as a Pre-process for Multilingual Documents.}
 As a companion to \snify, we have \defianno, a prototype tool for annotating definienda in
@@
 \begin{wrapfigure}{r}{0.4\textwidth}
+    \vskip-1.2cm
   \centering
   \includegraphics[width=0.4\textwidth]{../img/annocounts.pdf}
+    \vskip-0.2cm
   \caption{Number of annotated terms in the \sTeX/\ALeA commons over time.}\label{fig:annocounts}
+  \vskip-0.7cm
 \end{wrapfigure}
-In our experience, \snify's step-through workflow for annotating term references for symbols
-from the domain model is almost an order of magnitude more efficient than writing the
-annotations and imports by hand, even for annotators who are familiar with the domain
-model.
+In our experience, \snify's step-through workflow for annotating term references 
+% for symbols from the domain model 
+is almost an order of magnitude more efficient than writing the
+annotations and imports by hand.\footnote{
+    \edited{
+        The precise productivity gain depends on the annotator's
+        familiarity with the domain model (for the right symbol
+        is time-consuming) but also on other factors like
+        the typing speed.
+        Measurements are complicated by the fact that
+        authors tend to annotate more terms when using \snify.
+        We have observed speed-ups over IDE-based annotation
+        ranging from 3 (for an expert knowing the domain model by heart)
+        to 13 (for a beginner).
+        The focus mode further increases productivity.
+    }
+}
+% , even for annotators who are familiar with the domain model.
 Given the sheer number of term annotations (see \Cref{fig:annocounts}),
@@
 from the domain model) can be integrated into any IDE. In fact we plan to do this for the
-next version of the sTeX plugin for VSCode \cite{sTeX-IDE:git}.
+next version of the \sTeX plugin for VSCode \cite{sTeX-IDE:git}.
 For now, the typical workflow of early adopters is to first write the document (e.g.\ a set of quiz questions) without annotations, and then run \snify to annotate the new content.
@@
 \section{Scaling \snify to the Flexiformal Content Commons }\label{sec:scaling}
-One practical problem with \snify system as presented so far is that it only harvests
+One practical problem with the \snify system as presented so far is that it only harvests
 symbol/verbalization pairs from \sTeX files on the local file system. This means that, for
@@
 The general situation is that on the one hand there is a constantly growing, shared
-\textbf{content commons} of public math archives -- currently 200 with about 14.000 \sTeX
+\textbf{content commons} of public math archives -- currently 200 with about 14\;000 \sTeX
 files -- that are regularly converted to \FTML (\underline{F}lexiformal \underline{T}ext
@@
 
-To cope with this situation, we are working on extending the brand-new \flams system
-\cite{flams:on}\footnote{A paper on \flams is submitted to CICM 2025 and will be cited
-  here if accepted.} with functionality for handling definienda in the FTML2RDF harvester,
-so that all symbol/verbalizations are available in the \flams triple-store. Thus a harvest
-from the flexiformal content commons would be just a simple SPARQL query and some
-postprocessing of the result (de-duplication, stemming, and pairing) away. Note that
+% To cope with this situation, we are working on extending the brand-new \flams system
+% \cite{flams:on}\footnote{A paper on \flams is submitted to CICM 2025 and will be cited
+%   here if accepted.} with functionality for handling definienda in the FTML2RDF harvester,
+% so that all symbol/verbalizations are available in the \flams triple-store. Thus a harvest
+% from the flexiformal content commons would be just a simple SPARQL query and some
+% postprocessing of the result (de-duplication, stemming, and pairing) away.
+\edited{To cope with this situation, we are planning to extend the brand-new} \flams \edited{system, which among other things powers} \url{https://MathHub.info},
+\edited{so that it provides the data necessary to create a
+symbol/verbalization catalog.}
+Note that
 \flams works directly on \FTML, which is generated from \sTeX input by \textsf{rus\TeX}
@@
 Markdown we are developing as a lightweight alternative to \sTeX; it also transforms into
-FTML instead of plain HTML.
+\FTML instead of plain HTML.
 
@@
     \draw[draw,fill=black!10] (-1.3,.5) rectangle ++ (6.4,1); 
-    \node (f) at (-.5,.9) {\flams};
-    \draw[include]  (cc) -- node[left] {\tiny F2RDF} (db);
+    \node (f) at (-0.3,1.0) {\flams};
+      \draw[include]  (cc) -- node[left] {\tiny \begin{tabular}{c}RDF\\exp.\end{tabular}} (db);
     \draw[includeleft] (pl) -- node[right] {\tiny linter} (db);
@@
 Currently, \snify uses a custom Python implementation for harvesting the verbalization cache,
-but we are working on a implementation based on the new \flams \textbf{linter} -- a simple \sTeX parser that
-drives the LSP functionality \sTeX IDE plugins for \textsf{VSCode} and \textsf{emacs}.
-% Here we have extended the \flams \textbf{linter} -- a simple \sTeX parser that
-% drives the LSP functionality the VSCode and \texttt{emacs} \sTeX IDE plugins use to deal
-% with definienda. This is much more efficient (and robust) than the initial regex-based
-% python parser).
-The analogon for this on the \WOIDE side is still under development; it
+but we are working on a implementation based on the new \textbf{\flams linter} -- a simple \sTeX parser that
+drives the LSP functionality of the \sTeX IDE plugins for \textsf{VSCode} and \textsf{emacs}.
+The analog for this on the \WOIDE side is still under development; it
 could be based on \FTML generated by \WOIDE, or \WOIDE could directly export symbol/verbalization
@@
 
-\snify's interface is inspired traditional spell checkers
+\snify's interface is inspired by traditional spell checkers
 like \lstinline|ispell|~\cite{ispellman},
@@
 
-\snify is open source; the source and documentation are available from
+The \snify code is open source; the source and documentation are available from
 \cite{stextools:git}.

### 2025-07-23 scripts/annocounthistory2.py
copied over snify2 code

@@ -1,136 +1,0 @@
-import dataclasses
-import json
-import logging
-import re
-import subprocess
-from datetime import datetime
-from pathlib import Path
-from typing import Optional
-
-from stextools.core.cache import Cache
-from stextools.core.mathhub import get_mathhub_path
-from stextools.core.simple_api import get_linker, get_repos
-
-
-def run_sh(command: str, cwd: Optional[Path]) -> tuple[str, int]:
-    result = subprocess.run(['sh', '-c', command], capture_output=True, text=True, cwd=cwd)
-
-    return result.stdout, result.returncode
-
-
-def prepare():
-    Cache.clear = lambda: None  # type: ignore
-    logging.getLogger('pylatexenc.latexwalker').setLevel(logging.WARNING)
-    # TODO: the linker indicates both real sTeX issues and missing features – we should not suppress them in general
-    logging.getLogger('stextools.core.linker').setLevel(logging.FATAL)
-    logging.basicConfig(level=logging.INFO)
-
-
-current_names = ['sn', 'sns', 'Sn', 'Sns', 'sr']
-intermediate_names = ['symname', 'Symname', 'symref']
-old_names = [prefix + core + suff + suffsuff for core in ['tref', 'Tref'] for suff in ['i', 'ii', 'iii', 'iv'] for
-             suffsuff in ['', 's'] for prefix in ['', 'm']]
-
-
-def anno_count(repo_path: Path) -> int:
-    regex = re.compile(r'\\(' + '|'.join(current_names + intermediate_names + old_names) + r')[{[]')
-    count = 0
-    for texfile in repo_path.glob('**/*.tex'):
-        if any(x in texfile.name for x in ['all.', 'glossary', 'dictionary']):  # those are generated files
-            continue
-        try:
-            count += len(regex.findall(texfile.read_text()))
-        except Exception as e:
-            print(e)
-            print()
-    return count
-
-
-def process_repo(path: Path):
-    branch = 'private' if path.name == 'problems' and 'courses/FAU' in path.as_uri() else 'main'
-    if 'courses/FAU/EiDA/course' in path.as_uri():
-        branch = 'master'
-    if 'courses/FAU/GDP/problems' in path.as_uri():
-        branch = 'main'
-    _, code = run_sh(f'git checkout stex4', path)
-    if code == 0:
-        branch = 'stex4'
-    else:
-        _, code = run_sh(f'git checkout {branch}', path)
[…]

### 2025-07-23 scripts/annocountplot.py

@@ -1,101 +1,0 @@
-from collections import defaultdict
-from datetime import datetime
-from fnmatch import fnmatch
-
-from matplotlib import pyplot as plt
-
-from annocounthistory2 import CommitInfo, load_history
-
-HISTORY_DATA = load_history()
-
-
-SINCE = datetime(2015, 1, 1)
-
-
-FILTERED_DATA = [
-    record for record in HISTORY_DATA.values()
-    if record.date_as_datetime().timestamp() > SINCE.timestamp()
-]
-
-
-def accumulate(data: list[CommitInfo]) -> tuple[list[float], list[int]]:
-    dates = []
-    counts = []
-    counts_by_repo = defaultdict(int)
-    for record in sorted(data, key=lambda x: x.date_as_datetime().timestamp()):
-        dates.append(record.date_as_datetime().timestamp())
-        counts_by_repo[record.repo] = record.anno_count
-        counts.append(sum(counts_by_repo.values()))
-
-    return dates, counts
-
-
-def plot_cumulative(data: list[CommitInfo], label: str, lines: str):
-    dates, counts = accumulate(data)
-    plt.plot(dates, counts, label=label, linestyle=lines)
-
-
-def filtered(*pattern: str) -> list[CommitInfo]:
-    return [record for record in FILTERED_DATA if any(fnmatch(record.repo, p) for p in pattern)]
-
-
-def plot_diff(data: list[CommitInfo], label: str):
-    dates, counts = accumulate(data)
-
-    diff_dates = []
-    diff_counts = []
-
-    for i in range(int(SINCE.timestamp()), int(datetime.now().timestamp()), 1 * 24 * 60 * 60):
-        a = i - 28 * 24 * 60 * 60
-        b = i
-
-        count_before_week = 0
-        count_end_of_week = 0
-        for d, c in zip(dates, counts):
-            if d < a:
-                count_before_week = c
-            if d < b:
-                count_end_of_week = c
-
-        diff_dates.append((a + b) / 2)
-        diff_counts.append(count_end_of_week - count_before_week)
-
-    plt.plot(diff_dates, diff_counts, label=label)
-
-
-def cumulative_plots():
-    plot_cumulative(filtered('smglom/*'), 'smglom', lines='dotted')
-    plot_cumulative(filtered('courses/*/course'), 'course', lines='dashed')
-    plot_cumulative(filtered('*/problems'), 'problems', lines='dashdot')
[…]

### 2025-07-23 stextools/__main__.py

@@ -1,78 +1,60 @@
-import importlib.metadata
 import logging
 from pathlib import Path
-from platform import python_version
 
 import click
 
-from stextools.core.macros import STEX_CONTEXT_DB
-from stextools.utils import ui
-from stextools.core.cache import Cache
-from stextools.core.config import get_config
+from stextools.config import get_config, CACHE_DIR
+from stextools.lexicon.lexgen import lexgen
+from stextools.snify.snify import snify
+from stextools.stepper.interface import set_interface, DEFAULT_INTERFACES
 
 logger = logging.getLogger(__name__)
 
 
-filter_option = click.option('--filter', default=None,
-                             help='Filter pattern to only include some archives (e.g. \'smglom/*,courses/*\')')
+interface_option = click.option(
+    '--interface',
+    default=lambda: get_config().get('stextools.general', 'interface', fallback='console-light'),
+    type=click.Choice(DEFAULT_INTERFACES.keys(), case_sensitive=False),
+    help='Sets the interface (true refers to true color, which is not supported by all terminals).'
+)
 
 
 @click.group()
-@click.option('--keep-cache', is_flag=True, default=False,
-              help='Keep the cache despite changes to the stextools package.'
-                   'This can be useful for stextools development, but may lead to errors.')
-@click.option('--use-true-color',
-              default=lambda: get_config().getboolean('stextools', 'use_true_color', fallback=True),
-              type=bool,
-              help='Use 24-bit ("true") colors. Not all terminals support this.')
-def cli(keep_cache, use_true_color):
-    ui.USE_24_BIT_COLORS = use_true_color
-    if keep_cache:
-        Cache.clear = lambda: None  # type: ignore
+@click.option('--log-file', default=None, type=click.Path(),
+              help='Log file path. If not set, logs to stdout.')
+def cli(log_file):
     logging.getLogger('pylatexenc.latexwalker').setLevel(logging.WARNING)
-    # TODO: the linker indicates both real sTeX issues and missing features – we should not suppress them in general
-    logging.getLogger('stextools.core.linker').setLevel(logging.FATAL)
-    logging.basicConfig(level=logging.INFO)
+    logging.basicConfig(level=logging.INFO, filename=log_file)
 
 
-@cli.command(help='Clear the cache. You should not have to do this '
-                  'as the cache is automatically cleared whenever stextools is updated.')
+@cli.command(name='snify', help='\\sn-ify sTeX documents')
[…]

### 2025-07-23 stextools/core/cache.py

@@ -1,89 +1,0 @@
-import functools
-import logging
-import math
-import pickle
-from pathlib import Path
-from typing import Optional
-
-from stextools.core.mathhub import MathHub
-
-
-logger = logging.getLogger(__name__)
-
-
-CACHE_DIR = Path('~/.cache/stextools').expanduser()
-
-MATHHUB_PICKLE_FILE = CACHE_DIR / 'mathhub.pickle'
-
-
-class Cache:
-    _mh: Optional[MathHub] = None
-    _mh_uptodate: bool = False
-
-    @classmethod
-    @functools.cache
-    def get_stextools_version(cls) -> str:
-        """For now, we will use the last modification time of the stextools package.
-
-        We could also use the git hash, but that would require that the package is installed from git.
-        In the future, a version number should be used."""
-        last_modified = -math.inf
-        for file in Path(__file__).parent.rglob('**/*.py'):
-            last_modified = max(last_modified, file.stat().st_mtime)
-        return str(last_modified)
-
-    @classmethod
-    def get_mathhub(cls, update_all: bool = False) -> MathHub:
-        """Returns the MathHub instance."""
-        if cls._mh is not None:
-            if update_all and not cls._mh_uptodate:
-                cls._mh.load_all_doc_infos()
-                cls._mh_uptodate = True
-            return cls._mh
-
-        cls.ensure_uptodate()
-        if MATHHUB_PICKLE_FILE.exists():
-            logger.info('Loading MathHub info from cache...')
-            with open(MATHHUB_PICKLE_FILE, 'rb') as fp:
-                cls._mh = pickle.load(fp)
-                cls._mh.update()
-        else:
-            logger.info('No MathHub info found in cache - I will start from scratch.')
-            cls._mh = MathHub()
-        if update_all:
-            if cls._mh.load_all_doc_infos() > 20:   # only update cache if enough documents have changed
-                Cache.store_mathhub(cls._mh)
-        return cls._mh
-
-    @classmethod
-    def store_mathhub(cls, mh: MathHub):
-        cls.ensure_uptodate()
-        logger.info('Writing MathHub info to cache...')
-        with open(MATHHUB_PICKLE_FILE, 'wb') as fp:
-            pickle.dump(mh, fp)
-
-    @classmethod
-    def clear(cls):
-        MATHHUB_PICKLE_FILE.unlink(missing_ok=True)
-
-    @classmethod
-    def ensure_uptodate(cls):
-        """Clears outdated data and ensures that the current stextools version is stored."""
[…]

### 2025-07-23 stextools/core/config.py

@@ -1,19 +1,0 @@
-import configparser
-import functools
-import logging
-from pathlib import Path
-
-
-logger = logging.getLogger(__name__)
-
-
-@functools.cache
-def get_config() -> configparser.ConfigParser:
-    config_path = Path('~/.config/stextools/config.ini').expanduser()
-    config = configparser.ConfigParser()
-    if config_path.exists():
-        logger.info(f'Loading config from {config_path}')
-        config.read(config_path)
-    else:
-        logger.info(f'No config file found at {config_path}. Using defaults.')
-    return config

### 2025-07-23 stextools/core/linker.py

@@ -1,305 +1,0 @@
-import logging
-from collections import defaultdict
-from collections.abc import Iterable, Callable
-from typing import Optional, Union, Mapping
-
-from stextools.core.mathhub import MathHub
-from stextools.core.stexdoc import STeXDocument, Symbol, Verbalization, DocInfo, ModuleInfo, Dependency
-from stextools.utils.intifier import Intifier
-
-logger = logging.getLogger(__name__)
-
-
-def top_sort(doc_deps: Mapping[int, Iterable[int]], error_doc_info: Callable[[int], str]) -> Iterable[int]:
-    full_processed: set[int] = set()
-    already_under_consideration: set[int] = set()
-    result: list[int] = []
-
-    def visit(doc: int):
-        if doc in full_processed:
-            return
-        if doc in already_under_consideration:
-            logger.error(f'Circular dependency detected (involving {error_doc_info(doc)})')
-            return
-        already_under_consideration.add(doc)
-        for dep in doc_deps[doc]:
-            visit(dep)
-        full_processed.add(doc)
-        result.append(doc)
-
-    for doc in list(doc_deps):
-        visit(doc)
-
-    return result
-
-
-class Linker:
-    """
-    Heavily optimized:
-        * Objects are converted to integers for faster processing.
-        * Different things are computed in the same pass to avoid multiple iterations over the same data.
-    """
-
-    # intifiers
-    # (conversion to integers makes graph algorithms more efficient)
-    document_ints: Intifier[STeXDocument]
-    module_ints: Intifier[tuple[int, str]]  # (document as int, module name)
-    symbol_ints: Intifier[tuple[int, Symbol]]  # (module as int, symbol)
-    verb_ints: Intifier[tuple[int, Verbalization]]  # (document as int, verbalization)
-
-    # data from _compute_dep_graph
-    file_import_graph: dict[int, set[int]]
-    module_import_graph: dict[int, set[int]]   # module -> set[module]
-    file_to_module: dict[int, set[int]]        # file -> set[module]  (maps to all modules in the file)
-    module_to_file: dict[int, int]             # module -> file
-    # file -> set[(module, range_start, range_end)] - from imports, uses and smodule environments
-    available_module_ranges: dict[int, set[tuple[int, int, int]]]
-    module_to_symbs: dict[int, set[int]]        # module -> set[symbol]
-    symbol_to_module: dict[int, int]            # symbol -> module
[…]

### 2025-07-23 stextools/core/manifest.py

@@ -1,36 +1,0 @@
-"""Code for parsing and editing MANIFEST.MF files."""
-from pathlib import Path
-
-
-class ManifestProcessingException(Exception):
-    pass
-
-
-class Manifest(dict):
-    def __init__(self, path: Path):
-        dict.__init__(self)
-        self.path = path
-
-        # load manifest data
-        if not path.is_file():
-            raise FileNotFoundError(f'{path} is does not exist')
-        with open(self.path) as fp:
-            for line in fp:
-                if not line.strip():
-                    continue
-                before, _, after = line.partition(':')
-                if not before:
-                    raise ManifestProcessingException(f'Failed to process a line in {self.path}: {line!r}')
-                self[before] = after.strip()
-
-    def write(self):
-        """Write the data back to the file."""
-        with open(self.path, 'w') as fp:
-            for key, value in self.items():
-                fp.write(f'{key}: {value}\n')
-
-    def get_dependencies(self) -> list[str]:
-        return [d.strip() for d in self['dependencies'].split(',')]
-
-    def update_dependencies(self, dependencies: list[str]):
-        self['dependencies'] = ','.join(dependencies)

### 2025-07-23 stextools/core/mathhub.py

@@ -1,332 +1,0 @@
-from __future__ import annotations
-
-import fnmatch
-import functools
-import logging
-import multiprocessing
-import os
-import time
-from pathlib import Path
-from typing import Optional, Iterator, Literal, Callable
-
-from stextools.core.manifest import Manifest
-from stextools.core.stexdoc import STeXDocument, DocInfo
-
-logger = logging.getLogger(__name__)
-
-
-class MathHub:
-    def __init__(self, root_path: Optional[Path] = None):
-        self.root_path = root_path or get_mathhub_path()
-        self.archive_lookup: dict[str, Repository] = {}
-        self.update()
-
-    def get_archive(self, archive_name: str) -> Optional[Repository]:
-        return self.archive_lookup.get(archive_name)
-
-    def iter_stex_archives(self) -> Iterator[Repository]:
-        for repo in self.archive_lookup.values():
-            if repo.is_stex_archive():
-                yield repo
-
-    def iter_stex_docs(self) -> Iterator[STeXDocument]:
-        for repo in self.iter_stex_archives():
-            yield from repo.stex_doc_iter()
-
-    def get_archive_from_path(self, path: Path, is_resolved: bool = False) -> Optional[Repository]:
-        if not is_resolved:
-            path = path.absolute().resolve()
-        while not (path / '.git').is_dir():
-            path = path.parent
-            if path == Path('/'):
-                raise FileNotFoundError(f'No parent of {path} has a .git directory')
-
-        return self.get_archive(path.relative_to(self.root_path).as_posix())
-
-    @functools.cache
-    def get_stex_doc(self, path: Path) -> Optional[STeXDocument]:
-        path = path.absolute().resolve()
-        repo = self.get_archive_from_path(path, is_resolved=True)
-        if repo is None:
-            return None
-        return repo.get_stex_doc(path.relative_to(repo.path).as_posix())
-
-    def update(self):
-        """Updates the repo information"""
-        logger.info('Scanning archives...')
-        still_needed: set[str] = set()
-        self.get_stex_doc.cache_clear()
-        for repo in _get_mathhub_repos(self.root_path):
-            repo.get_archive_name.cache_clear()
-            repo.resolve_path_ref.cache_clear()
-            repo.is_stex_archive.cache_clear()
-
-            if repo.get_archive_name() in self.archive_lookup:
-                existing = self.archive_lookup[repo.get_archive_name()]
-                if not existing.path.is_dir():
-                    del self.archive_lookup[repo.get_archive_name()]
[…]

### 2025-07-23 stextools/core/simple_api.py

@@ -1,340 +1,0 @@
-from __future__ import annotations
-
-import dataclasses
-from collections import deque
-from collections.abc import Iterable, Generator
-import itertools
-from pathlib import Path
-from typing import Optional
-
-from stextools.core.cache import Cache
-from stextools.core.linker import Linker
-from stextools.core.mathhub import Repository
-from stextools.core.stexdoc import Symbol, Verbalization, STeXDocument, ModuleInfo
-
-
-class SimpleSymbol:
-    _symbol: Symbol
-    _symbol_int: int
-    _linker: Linker
-
-    def __init__(self, symbol_int: int, linker: Linker):
-        self._symbol = linker.symbol_ints.unintify(symbol_int)[1]
-        self._symbol_int = symbol_int
-        self._linker = linker
-
-    def get_verbalizations(self, lang: Optional[str] = None) -> Generator['SimpleVerbalization']:
-        for verb_int in self._linker.symb_to_verbs[self._symbol_int]:
-            doc_int, verb = self._linker.verb_ints.unintify(verb_int)
-            if lang is not None and verb.lang != lang:
-                continue
-            yield SimpleVerbalization(verb_int, self._linker)
-
-    @property
-    def path_rel_to_archive(self) -> str:
-        return self.declaring_module.path_rel_to_archive + '?' + self._symbol.name
-
-    @property
-    def name(self) -> str:
-        return self._symbol.name
-
-    @property
-    def declaring_file(self) -> 'SimpleFile':
-        return self.declaring_module.file
-
-    @property
-    def macro_range(self) -> tuple[int, int]:
-        return self._symbol.decl_def
-
-    @property
-    def declaring_module(self) -> 'SimpleModule':
-        module = self._linker.symbol_to_module[self._symbol_int]
-        return SimpleModule(module, self._linker)
-
-    def __eq__(self, other) -> bool:
-        return (isinstance(other, SimpleSymbol) and
-                self._symbol_int == other._symbol_int and
-                self._linker is other._linker)
-
-    def __hash__(self) -> int:
-        return hash((self._symbol_int, self._linker))
-
-
-class SimpleModule:
-    _module: ModuleInfo
-    _module_int: int
-    _linker: Linker
-
-    def __init__(self, module_int: int, linker: Linker):
-        doc, modstr = linker.module_ints.unintify(module_int)
-        module = linker.document_ints.unintify(doc).get_doc_info(linker.mh).get_module(modstr)
-        assert module is not None
[…]

### 2025-07-23 stextools/core/stexdoc.py

@@ -1,503 +1,0 @@
-from __future__ import annotations
-
-import dataclasses
-import functools
-import logging
-import re
-import typing
-from pathlib import Path
-from typing import Optional, Iterator, Literal
-
-from pylatexenc.latexwalker import LatexMacroNode, LatexWalker, LatexCommentNode, LatexCharsNode, LatexMathNode, \
-    LatexSpecialsNode, LatexGroupNode, LatexEnvironmentNode
-
-from stextools.core.macros import STEX_CONTEXT_DB
-from stextools.utils.macro_arg_utils import get_first_macro_arg_opt, OptArgKeyVals, get_first_main_arg
-
-if typing.TYPE_CHECKING:
-    from stextools.core.mathhub import MathHub, Repository
-
-
-logger = logging.getLogger(__name__)
-
-
-class DependencyPropFlag:  # note: enum.IntFlag is really slow -- maybe this is better in new Python versions?
-    IS_LIB = 1
-    IS_USE = 2
-    TARGET_NO_TEX = 4
-    IS_INPUT = 8
-    IS_USE_STRUCT = 16
-
-
-# NOTE: According to my benchmark, using `slots=True` noticably slows down pickling/unpickling
-
-@dataclasses.dataclass(frozen=True, eq=True, repr=True)
-class Dependency:
-    archive: str
-    # should always be set, but could be None if the file cannot be determined (still better than no dependency)
-    file_hint: Optional[str]
-    module_name: Optional[str]
-    flags: int     # DependencyPropFlag
-    scope: tuple[int, int]  # range of the document where the dependency is valid
-    intro_range: Optional[tuple[int, int]] = None  # range of the document where the dependency is introduced
-    file_may_be_relative: bool = False
-
-    def get_target_path(self, mh: MathHub, src: STeXDocument) -> Optional[Path]:
-        if self.file_hint is None:
-            return None
-        archive = mh.get_archive(self.archive)
-        if archive is None:
-            return None
-        di = src.get_doc_info(mh)
-        if self.module_name:
-            if len(self.module_name) > 150:
-                logger.warning(f'Ignoring very long module name in {src.path}: {self.module_name[:150]}...')
-                logger.warning(
-                    f'If this is not caused by malformed TeX, please report this issue along with the document.')
-                return None
-            if r := archive.resolve_path_ref(self.file_hint + '/' + self.module_name, 'lib' if self.is_lib else 'source', di.lang):
-                return r
-
[…]

### 2025-07-23 stextools/cycle_finder.py

@@ -1,63 +1,0 @@
-import itertools
-from pathlib import Path
-
-import click
-
-from stextools.core.cache import Cache
-from stextools.core.linker import Linker
-from stextools.utils.ui import print_highlight_selection
-
-
-class CycleFound(Exception):
-    def __init__(self, cycle: list[int]):
-        self.cycle = cycle
-
-
-def cycle_finder(file: str):
-    mh = Cache.get_mathhub(update_all=True)
-    linker = Linker(mh)
-
-    graph = linker.file_import_graph
-
-    covered: set[int] = set()
-
-    def dfs(i: int, stack: list[int]):
-        if i in stack:
-            raise CycleFound(stack + [i])
-        stack.append(i)
-        for child in graph[i]:
-            if child not in covered:
-                dfs(child, stack)
-                covered.add(child)
-        stack.pop()
-
-    doc = mh.get_stex_doc(Path(file))
-    if doc is None:
-        print(f'Failed to find document {file}.')
-        return
-    start = linker.document_ints.intify(doc)
-
-    try:
-        dfs(start, [])
-    except CycleFound as e:
-        print(click.style('I found the following cycle:', bg='bright_cyan'))
-        for i, j in itertools.pairwise(e.cycle):
-            doc = linker.document_ints.unintify(i)
-            print(click.style(doc.path, bold=True))
-            for dep in doc.get_doc_info(mh).flattened_dependencies():
-                if dep.is_use:
-                    continue
-                target_doc = dep.get_target(mh, doc)[0]
-                if target_doc is None:
-                    continue
-                if linker.document_ints.intify(target_doc) == j:
-                    assert dep.intro_range
-                    print_highlight_selection(
-                        doc.path.read_text(), dep.intro_range[0], dep.intro_range[1], 1, bold=False
-                    )
-                    break
-
-        doc = linker.document_ints.unintify(e.cycle[-1])
-        print(click.style(doc.path, bold=True))
-    else:
-        print('No cycle found.')

### 2025-07-23 stextools/dependency_update.py

@@ -1,91 +1,0 @@
-import logging
-from typing import Literal, Optional
-
-import click
-
-from stextools import stexmmtquery
-from stextools.core.cache import Cache
-from stextools.core.mathhub import make_filter_fun
-
-logger = logging.getLogger(__name__)
-
-
-def always_keep_dependency_check(dep: str) -> bool:
-    """Some dependencies should be kept even if we do not detect them"""
-    #     if dep == 'sTeX/meta-inf':
-    #         return True
-    if dep.endswith('/meta-inf'):
-        return True
-    if dep.startswith('MMT/'):
-        return True
-    return False
-
-
-def dependency_check(mode: Literal['test', 'ask', 'write'] = 'test', filter: Optional[str] = None):
-    filter_fun = make_filter_fun(filter)
-    mh = Cache.get_mathhub()
-    if mh.load_all_doc_infos():
-        Cache.store_mathhub(mh)
-
-    logger.info('Getting dependencies from server...')
-    server_dependencies = stexmmtquery.get_dependencies()
-    logger.info(f'Found dependencies for {len(server_dependencies)} archives on the server, '
-                'but I will only update dependencies of locally installed archives.')
-
-    for archive in mh.iter_stex_archives():
-        if not filter_fun(archive.get_archive_name()):
-            continue
-        print('\n\n')
-        print(click.style(archive.get_archive_name(), fg='blue', bold=True))
-        needed_dependencies_my_data = set(
-            dep.archive for doc in archive.stex_doc_iter() for dep in doc.get_doc_info(mh).flattened_dependencies()
-        )
-        # print('Dependencies from my data:', needed_dependencies_my_data)
-        needed_dependencies_server = set(server_dependencies.get(archive.get_archive_name(), []))
-        # TODO: Why are there discrepancies?
-        needed_dependencies = needed_dependencies_my_data | needed_dependencies_server
-        new_dependencies: list[str] = []
-        manifest = archive.get_manifest()
-        for dependency in manifest.get_dependencies():
-            if always_keep_dependency_check(dependency):
-                new_dependencies.append(dependency)
-        for dependency in needed_dependencies:
-            if dependency == archive.get_archive_name():  # no self-dependencies
-                continue
-            if dependency not in new_dependencies:
-                new_dependencies.append(dependency)
-
-        print()
-        # if deps_only_on_server := needed_dependencies_server - needed_dependencies_my_data:
-        #     print(f'WARNING - the following dependencies were not found locally: {deps_only_on_server}')
[…]

### 2025-07-23 stextools/lexicon/lexgenstate.py

@@ -1,0 +1,7 @@
+from stextools.snify.snifystate import SnifyCursor
+from stextools.stepper.document_stepper import DocumentStepperState
+from stextools.stepper.stepper import State
+
+
+class LexGenState(DocumentStepperState, State[SnifyCursor]):
+    pass

### 2025-07-23 stextools/lexicon/lexgenstepper.py

@@ -1,0 +1,158 @@
+import functools
+import os
+from pathlib import Path
+
+from stextools.lexicon.lexgen_commands import SkipCommand
+from stextools.lexicon.lexgenstate import LexGenState
+from stextools.lexicon.ud import sentence_tokenize, word_tokenize, get_word_info
+from stextools.lexicon.wordlex import WordLex
+from stextools.snify.snifystate import SnifyCursor
+from stextools.stepper.command import CommandCollection
+from stextools.stepper.document import STeXDocument
+from stextools.stepper.interface import interface
+from stextools.stepper.stepper import Stepper, StopStepper
+from stextools.stepper.stepper_extensions import QuittableStepper, CursorModifyingStepper, UndoableStepper
+from stextools.stex.flams import FLAMS
+from stextools.stex.local_stex import OpenedStexFLAMSFile
+from stextools.utils.json_iter import json_iter
+
+UPOS_TO_GF_CAT: dict[str, str] = {
+    'ADJ': 'A',
+    'ADP': 'Prep',
+    'ADV': 'Adv',
+    'AUX': '',
+    'CCONJ': 'Conj',
+    'DET': 'Det',
+    'INTJ': '',
+    'NOUN': 'N',
+    'NUM': '',
+    'PART': '',
+    'PRON': 'Pron',
+    'PROPN': 'PN',
+    'PUNCT': 'Punct',
+    'SCONJ': 'Conj',  # is it?
+    'SYM': 'Symb',
+    'VERB': 'V',
+    'X': '',
+}
+
+
+@functools.cache
+def get_wordlex(lang: str, archive: str):
+    parts = archive.split('/')
+    cc = ''.join(
+        part[0].upper() + part[1:] if part else ''
+        for part in parts
+    )
+    directory = Path(os.environ['MATHHUB']) / archive / 'source'
+    return WordLex.load(
+        cc,
+        directory,
+        lang,
+        create_if_nonexistent=True,
+    )
+
+
+class LexGenStepper(QuittableStepper, CursorModifyingStepper, UndoableStepper, Stepper[LexGenState]):
+    def ensure_state_up_to_date(self):
+        cursor = self.state.cursor
+        if not isinstance(cursor.selection, int):
+            return   # already have a selection
+
+        while cursor.document_index < len(self.state.documents):
+            doc = self.state.get_current_document()
+            assert isinstance(doc, STeXDocument), 'currently only STeXDocuments are supported'
+            opened_file = OpenedStexFLAMSFile(str(doc.path))
+
[…]

### 2025-07-23 stextools/pythonapi.py

@@ -1,92 +1,0 @@
-import functools
-import logging
-from typing import Optional
-
-import click
-
-from stextools.core.cache import Cache
-from stextools.core.linker import Linker
-from stextools.core.simple_api import get_linker, SimpleSymbol
-from stextools.snify.selection import VerbTrie
-from stextools.snify.stemming import string_to_stemmed_word_sequence_simplified
-from stextools.utils.ui import latex_format
-
-
-@functools.cache
-def get_linker() -> Linker:
-    mh = Cache.get_mathhub(update_all=True)
-    return Linker(mh)
-
-
-@functools.cache
-def get_verb_trie(lang: str) -> VerbTrie:
-    return VerbTrie(lang, get_linker())
-
-
-def setup_logging():
-    logging.getLogger('pylatexenc.latexwalker').setLevel(logging.WARNING)
-    # TODO: the linker indicates both real sTeX issues and missing features – we should not suppress them in general
-    logging.getLogger('stextools.core.linker').setLevel(logging.FATAL)
-    logging.basicConfig(level=logging.INFO)
-
-
-def serialize_symbol(symbol: SimpleSymbol) -> str:
-    symb_path = symbol.path_rel_to_archive
-    archive = symbol.declaring_file.archive.name
-    return f'[{archive}]{symb_path}'
-
-
-def _get_symbols_for_verbalization(verbalization: str, lang: str) -> list[SimpleSymbol]:
-    vt = get_verb_trie(lang)
-    stemmed = string_to_stemmed_word_sequence_simplified(verbalization, lang)
-
-    subtrie = ([], vt.trie)
-    for word in stemmed:
-        if word not in subtrie[1]:
-            return []
-        subtrie = subtrie[1][word]
-
-    return subtrie[0]
-
-def interactive_symbol_search(verbalization: Optional[str] = None, lang: str = 'en') -> Optional[str]:
-    while True:
-        options = _get_symbols_for_verbalization(verbalization, lang)
-
-        if not options:
-            print('No symbols found for the given verbalization.')
-            print('Enter a new verbalization or "q" to quit.')
-            verbalization = input('verbalization: ').strip()
-            if verbalization.lower() == 'q':
-                return None
-            continue
-
-        for i, symbol in enumerate(options):
-            print(f'[{i}]', serialize_symbol(symbol))
-
-        print()
-        print('Available commands:')
-        print('  [s] - skip')
-        print('  [𝑖] - choose symbol 𝑖')
[…]

### 2025-07-23 stextools/remote_repositories.py

@@ -1,9 +1,20 @@
+import os
 import subprocess
+from pathlib import Path
 
 import gitlab
 from gitlab.v4.objects import Group
 
-from stextools.core.mathhub import get_mathhub_path
+
+def get_mathhub_path() -> Path:
+    path = os.environ.get("MATHHUB")
+    if path is None:
+        raise RuntimeError("MATHHUB environment variable not set")
+    path = Path(path).expanduser().resolve()
+    if not path.exists():
+        raise RuntimeError(f"MATHHUB path {path} does not exist")
+    return path
+
 
 URL = 'https://gl.mathhub.info'
 

### 2025-07-23 stextools/snify/annotate.py

@@ -1,0 +1,499 @@
+import dataclasses
+from copy import deepcopy
+from typing import Optional, Literal, Iterable, Sequence, Any
+
+from pylatexenc.latexwalker import LatexEnvironmentNode
+
+from stextools.snify.catalog import Verbalization
+from stextools.stepper.document import STeXDocument
+from stextools.stex.local_stex import OpenedStexFLAMSFile, get_transitive_imports, FlamsUri, get_transitive_structs
+from stextools.snify.local_stex_catalog import LocalStexSymbol, LocalFlamsCatalog
+from stextools.snify.snify_commands import ImportCommand
+from stextools.stepper.document_stepper import SubstitutionOutcome
+from stextools.snify.snifystate import SnifyState, SnifyCursor
+from stextools.stex.stex_py_parsing import iterate_latex_nodes
+from stextools.stepper.command import Command, CommandInfo, CommandOutcome, CommandCollection
+from stextools.stepper.interface import interface
+from stextools.stepper.stepper_extensions import QuitCommand, QuitOutcome, SetCursorOutcome
+from stextools.stex.flams import FLAMS
+from stextools.utils.json_iter import json_iter
+
+
+class AnnotationAborted(Exception):
+    pass
+
+
+class STeXAnnotateBase(Command):
+    def __init__(
+            self,
+            state: SnifyState,
+            catalog: LocalFlamsCatalog,
+            stepper,
+    ):
+        self.state = state
+        self.catalog = catalog
+        self.stepper = stepper
+        document = state.get_current_document()
+        assert isinstance(document, STeXDocument)
+        self.document: STeXDocument = document
+        self.importinfo = get_modules_in_scope_and_import_locations(
+            self.document,
+            state.cursor.selection[0] if isinstance(state.cursor.selection, tuple) else state.cursor.selection,
+        )
+
+    def annotate_symbol(self, symbol: LocalStexSymbol) -> list[CommandOutcome]:
+        cursor = self.state.cursor
+        outcomes: list[Any] = []
+
+        try:
+            import_thing = self.get_import(symbol)
+        except AnnotationAborted:
+            return []
+
+        if import_thing:
+            outcomes.extend(import_thing)
+
+        sr = self.get_sr(symbol.uri)
+        outcomes.append(
+            SubstitutionOutcome(sr, cursor.selection[0], cursor.selection[1])
+        )
+
+        # at this point, we only have substitutions
+        # -> sort them and update the offsets
+        # TODO: maybe the controller should be responsible for this
+        offset = 0
[…]

### 2025-07-23 stextools/snify/annotate_command.py

@@ -1,454 +1,0 @@
-import dataclasses
-from typing import Optional, Union, Any, Literal, Sequence
-
-import click
-from pylatexenc.latexwalker import LatexMacroNode, LatexSpecialsNode, LatexMathNode, LatexGroupNode, \
-    LatexEnvironmentNode, LatexCommentNode, LatexCharsNode, LatexWalker
-
-from stextools.core.linker import Linker
-from stextools.core.macros import STEX_CONTEXT_DB
-from stextools.core.mathhub import make_filter_fun
-from stextools.core.simple_api import SimpleSymbol, get_symbols, SimpleFile, SimpleModule
-from stextools.snify.commands import ImportCommand
-from stextools.snify.snify_state import SnifyState
-from stextools.stepper.command_outcome import CommandOutcome, Exit, StatisticUpdateOutcome, SubstitutionOutcome, \
-    SetNewCursor
-from stextools.stepper.commands import CommandInfo, Command, QuitSubdialogCommand, show_current_selection, \
-    CommandSectionLabel, CommandCollection
-from stextools.stepper.state import State, SelectionCursor, PositionCursor
-from stextools.utils.fzf import get_symbol_from_fzf
-from stextools.utils.ui import standard_header, pale_color, option_string, color, latex_format
-
-# This stores the keys to fix the order of the symbols within a session
-# (note: cannot cache by SimpleSymbol object, because it changes when resetting the linker)
-_already_determined: dict[tuple[str, str], tuple] = {}
-
-
-def symbol_to_sorting_key(symbol: SimpleSymbol) -> tuple:
-    k = (symbol.declaring_file.archive.name, symbol.path_rel_to_archive)
-    if k in _already_determined:
-        return _already_determined[k]
-    primary = len(list(symbol.get_verbalizations()))
-    secondary = k
-    _already_determined[k] = primary, secondary
-    return primary, secondary
-
-
-@dataclasses.dataclass
-class ImportPositions:
-    use_pos: int
-    top_use_pos: int
-    import_pos: Optional[int]
-    use_env: Optional[str]
-    top_use_env: Optional[str]
-    # scopes
-    top_use_scope: tuple[int, int]
-    use_scope: tuple[int, int]
-    import_scope: Optional[tuple[int, int]]
-
-
-def get_redundant_import_removals(
-        file: SimpleFile, text: str, imported_module: SimpleModule, import_scope: tuple[int, int], is_use: bool,
[…]

### 2025-07-23 stextools/snify/catalog.py

@@ -1,0 +1,149 @@
+import re
+from typing import TypeVar, Generic, Iterable, Optional, Hashable
+
+from stextools.snify.stemming import string_to_stemmed_word_sequence_simplified, string_to_stemmed_word_sequence
+from stextools.utils.linked_str import LinkedStr, string_to_lstr
+
+
+class Verbalization:
+    def __init__(self, verb: str):
+        self.verb = verb
+
+
+Symb = TypeVar('Symb', bound=Hashable)
+Verb = TypeVar('Verb', bound=Verbalization)
+
+
+class Trie(Generic[Symb, Verb]):
+    __slots__ = 'verbs', 'children'
+    def __init__(self):
+        self.children: dict[str, 'Trie[Symb, Verb]'] = {}
+        self.verbs: dict[Symb, list[Verb]] = {}
+
+    def insert(self, key: Iterable[str], symb: Symb, verb: Verb):
+        node = self
+        for k in key:
+            if k not in node.children:
+                node.children[k] = Trie[Symb, Verb]()
+            node = node.children[k]
+        node.verbs.setdefault(symb, []).append(verb)
+
+    def get(self, key: Iterable[str]) -> dict[Symb, list[Verb]]:
+        node = self
+        for k in key:
+            if k not in node.children:
+                return {}
+            node = node.children[k]
+        return node.verbs
+
+    def __contains__(self, item):
+        return item in self.verbs
+
+
+class Catalog(Generic[Symb, Verb]):
+    lang: str
+    trie: Trie[Symb, Verb]
+    symb_to_verb: dict[Symb, list[Verb]]
+    symbols: set[Symb]
+
+    def __init__(self, lang: str, symbverbs: Optional[Iterable[tuple[Symb, Verb]]] = None):
+        self.lang = lang
+        self.trie = Trie[Symb, Verb]()
+        self.symb_to_verb = {}
+        if symbverbs is not None:
+            for symb, verb in symbverbs:
+                self.add_symbverb(symb, verb)
+
+    def symb_iter(self) -> Iterable[Symb]:
+        yield from self.symb_to_verb.keys()
+
+    def add_symbverb(self, symb: Symb, verb: Verb):
+        self.symb_to_verb.setdefault(symb, []).append(verb)
+        key = string_to_stemmed_word_sequence_simplified(verb.verb, self.lang)
+        self.trie.insert(key, symb, verb)
+
+    def add_symb(self, symb: Symb):
[…]

### 2025-07-23 stextools/snify/local_stex_catalog.py

@@ -1,0 +1,197 @@
+"""
+This module contains code for creating a catalog based on local sTeX archives.
+It is significantly optimized for performance (and hence unreadable).
+
+The extraction is based on FLAMS.
+FLAMS is slow enough that caching verbalizations is helpful.
+"""
+
+import dataclasses
+import gzip
+import logging
+import os
+from typing import TypeAlias, Iterable
+
+import orjson
+
+from stextools.config import CACHE_DIR
+from stextools.snify.catalog import Verbalization, Catalog, catalogs_from_stream
+from stextools.stex.local_stex import OpenedStexFLAMSFile, lang_from_path
+from stextools.stex.flams import FLAMS
+from stextools.utils.timer import timelogger
+
+logger = logging.getLogger(__name__)
+
+
+@dataclasses.dataclass
+class LocalStexSymbol:
+    uri: str
+    path: str
+    srefcount: int = 0   # simple heuristic: the more references, the more relevant
+
+    # TODO: symbols have to be hashable... the following is not ideal though
+    def __eq__(self, other):
+        return (
+                isinstance(other, LocalStexSymbol) and
+                self.uri == other.uri and self.path == other.path
+        )
+
+    def __hash__(self):
+        return hash((self.uri, self.path))
+
+
+class LocalStexVerbalization(Verbalization):
+    def __init__(self, verb: str, local_path: str, path_range: tuple[int, int]):
+        super().__init__(verb)
+        self.local_path = local_path
+        self.path_range = path_range
+
+    def __repr__(self):
+        return f"LocalStexVerbalization(verb={self.verb!r}, local_path={self.local_path!r}, path_range={self.path_range!r})"
+
+
+# very rudimentary representation for a verbalization
+# used for efficient serialization and deserialization
+# (language, symbol uri, symbol path, verb, verb start offset, verbend offset)
+# verb path is ommitted as it is assumed to be known from the context
+RawVerbEntry: TypeAlias = tuple[str, str, str, str, int, int]
+
+
+def _verb_and_symb_extraction(j, opened_file: OpenedStexFLAMSFile) -> Iterable[RawVerbEntry | str]:
+    """ recurse through the annotation json to find symrefs and co.
+    It yields verbalizations (RawVerbEntry) and symbols (str) if they are defined/declared.
+    """
+    if isinstance(j, dict):
+        for k, v in j.items():
+            # TODO: add more keys to filter (optimization)
[…]

### 2025-07-23 stextools/snify/skip_and_ignore.py

@@ -1,126 +1,124 @@
 import dataclasses
+import functools
 import itertools
 import re
-from pathlib import Path
 
-from stextools.snify.commands import StateSkipOutcome
-from stextools.stepper.command_outcome import CommandOutcome, TextRewriteOutcome, SetNewCursor
-from stextools.stepper.commands import CommandInfo, Command
-from stextools.stepper.state import State, SelectionCursor, PositionCursor
-from stextools.snify.stemming import string_to_stemmed_word_sequence_simplified
+from stextools.config import CONFIG_DIR
+from stextools.snify.snifystate import SnifyCursor, SnifyState
+from stextools.snify.stemming import string_to_stemmed_word_sequence_simplified, mystem
+from stextools.stepper.command import Command, CommandInfo, CommandOutcome
+from stextools.stepper.document_stepper import TextRewriteOutcome
+from stextools.stepper.stepper import Modification
+from stextools.stepper.stepper_extensions import SetCursorOutcome
 
 
-class SkipOnceCommand(Command):
-    def __init__(self):
+class SkipCommand(Command):
+    def __init__(self, state: SnifyState):
         super().__init__(CommandInfo(
-            pattern_presentation='s',
-            pattern_regex='^s$',
-            description_short='kip once',
-            description_long='Skips to the next possible annotation')
+            pattern_presentation = 's',
+            description_short = 'kip once',
+            description_long = 'Skips to the next possible annotation')
         )
+        self.state = state
 
-    def execute(self, *, state: State, call: str) -> list[CommandOutcome]:
-        assert isinstance(state.cursor, SelectionCursor)
-        return [SetNewCursor(PositionCursor(state.cursor.file_index, state.cursor.selection_start + 1))]
+    def execute(self, call: str) -> list[CommandOutcome]:
+        assert isinstance(self.state.cursor.selection, tuple)
+        return [
+            SetCursorOutcome(
+                new_cursor=SnifyCursor(self.state.cursor.document_index, self.state.cursor.selection[-1] + 1)
+            )
+        ]
+
+
+class StateSkipOutcome(CommandOutcome, Modification[SnifyState]):
+    def __init__(self, word: str, is_stem: bool, session_wide: bool, lang: str, current_document_index: int):
+        self.word = word
+        self.is_stem = is_stem
+        self.session_wide = session_wide
+        self.lang = lang
+        self.current_document_index = current_document_index
+
+    def _get_key_and_dict(self, state: SnifyState):
+        if self.session_wide:
[…]

### 2025-07-23 stextools/snify/snify.py

@@ -1,0 +1,36 @@
+from pathlib import Path
+from typing import Optional
+
+from click.exceptions import Exit
+
+from stextools.stepper.document import documents_from_paths
+from stextools.snify.snifystate import SnifyState, SnifyCursor
+from stextools.snify.snifystepper import SnifyStepper
+from stextools.stepper.session_storage import SessionStorage, IgnoreSessions
+
+
+def snify(files: list[Path]):
+    session_storage = SessionStorage('snify2')
+    result = session_storage.get_session_dialog()
+    if isinstance(result, Exit):
+        return
+    assert isinstance(result, SnifyState) or isinstance(result, IgnoreSessions)
+    state: Optional[SnifyState] = result if isinstance(result, SnifyState) else None
+
+    if state is None:
+        state = SnifyState(
+            SnifyCursor(
+                document_index=0,
+                selection=0,
+            ),
+            documents=documents_from_paths(files)
+        )
+
+    stepper = SnifyStepper(state)
+
+    stop_reason = stepper.run()
+
+    if stop_reason == 'quit':
+        session_storage.store_session_dialog(state)
+    else:
+        session_storage.delete_session_if_loaded()

### 2025-07-23 stextools/snify/snify_commands.py

@@ -1,0 +1,289 @@
+from copy import deepcopy
+from pathlib import Path
+from typing import Sequence, Any
+
+from stextools.snify.catalog import Verbalization
+from stextools.snify.snifystate import SnifyCursor, SnifyState
+from stextools.snify.stemming import string_to_stemmed_word_sequence_simplified, mystem, string_to_stemmed_word_sequence
+from stextools.stepper.document import Document
+from stextools.snify.local_stex_catalog import LocalStexSymbol
+from stextools.stepper.command import CommandOutcome, Command, CommandInfo
+from stextools.stepper.document_stepper import SubstitutionOutcome
+from stextools.stepper.interface import interface
+from stextools.stepper.stepper import Stepper
+from stextools.stepper.stepper_extensions import SetCursorOutcome, FocusOutcome
+
+
+class ImportCommand(Command):
+    def __init__(self, letter: str, description_short: str, description_long: str, outcome: SubstitutionOutcome,
+                 redundancies: list[SubstitutionOutcome]):
+        super().__init__(CommandInfo(
+            pattern_presentation=letter,
+            description_short=description_short,
+            description_long=description_long)
+        )
+        self.outcome = outcome
+        self.redundancies = redundancies
+
+    def execute(self, call: str) -> Sequence[CommandOutcome]:
+        cmds: list[SubstitutionOutcome] = self.redundancies + [self.outcome]
+        cmds.sort(key=lambda x: x.start_pos, reverse=True)
+        return cmds
+
+
+class View_i_Command(Command):
+    def __init__(self, options: list[tuple[Any, Verbalization]]):
+        super().__init__(CommandInfo(
+            show=False,
+            pattern_presentation='v𝑖',
+            pattern_regex='^v[0-9]+$',
+            description_short=' view document for 𝑖',
+            description_long='Displays the document that introduces symbol no. 𝑖')
+        )
+        self.options = options
+
+    def execute(self, call: str) -> Sequence[CommandOutcome]:
+        i = int(call[1:])
+        if i >= len(self.options):
+            interface.admonition('Invalid number', 'error', True)
+            return []
+
+        symbol = self.options[i][0]
+        if not isinstance(symbol, LocalStexSymbol):
+            interface.admonition(f'Unsupported symbol type {type(symbol)}', 'error', True)
+            return []
+
+        with interface.big_infopage():
+            interface.write_header(symbol.path)
+            interface.show_code(
+                Path(symbol.path).read_text(),
[…]

### 2025-07-23 stextools/snify/snify_state.py

@@ -1,42 +1,0 @@
-import dataclasses
-from pathlib import Path
-from typing import Optional
-
-from stextools.core.linker import Linker
-from stextools.stepper.state import State, Cursor
-
-
-@dataclasses.dataclass
-class SnifyFocusInfo:
-    select_only_stem: Optional[str] = None
-
-
-class SnifyState(State[SnifyFocusInfo]):
-    filter_pattern: str
-    ignore_pattern: str
-
-    skip_literal_by_file: dict[Path, set[str]]   # = dataclasses.field(default_factory=dict)
-    skip_literal_all_session: dict[str, set[str]] # = dataclasses.field(default_factory=dict)  # lang -> set of words
-    skip_stem_by_file: dict[Path, set[str]]  # = dataclasses.field(default_factory=dict)
-    skip_stem_all_session: dict[str, set[str]] # = dataclasses.field(default_factory=dict)  # lang -> set of stems
-
-    lang: Optional[str]
-
-    def __init__(self, files: list[Path], cursor: Cursor, filter_pattern: str, ignore_pattern: str,
-                 lang: Optional[str] = None):
-        super().__init__(files, cursor)
-        self.filter_pattern = filter_pattern
-        self.ignore_pattern = ignore_pattern
-
-        self.skip_literal_by_file = {}
-        self.skip_literal_all_session = {}
-        self.skip_stem_by_file = {}
-        self.skip_stem_all_session = {}
-
-        self.lang = lang
-
-    def get_current_lang(self, linker: Linker) -> str:
-        if self.lang is None:
-            self.lang = super().get_current_lang(linker)
-        return self.lang
-

### 2025-07-23 stextools/snify/snifystate.py

@@ -1,0 +1,68 @@
+import dataclasses
+from typing import Optional
+
+from stextools.stepper.document import Document
+from stextools.stepper.document_stepper import DocumentStepperState, DocumentCursor
+from stextools.stepper.stepper import State
+from stextools.stepper.stepper_extensions import FocussableState
+
+
+@dataclasses.dataclass(frozen=True)
+class SnifyCursor(DocumentCursor):
+    selection: int | tuple[int, int]
+
+
+class SnifyState(DocumentStepperState, FocussableState, State[SnifyCursor]):
+    def __init__(self, cursor: SnifyCursor, documents: list[Document]):
+        super().__init__(cursor, documents)
+
+        # (lang, doc_id) -> [word]
+        self.skip_stem_by_docid: dict[tuple[str, int], set[str]] = {}
+        self.skip_by_docid: dict[tuple[str, int], set[str]] = {}
+        # lang -> [word]
+        self.skip_stem: dict[str, set[str]] = {}
+        self.skip: dict[str, set[str]] = {}
+
+        self.stem_focus: Optional[str] = None   # only suggest annotations for this stem (used in focus mode)
+        self.focus_lang: Optional[str] = None  # only annotate documents of this language (used in focus mode)
+
+    def get_skip_words(self, lang: str, doc_index: Optional[int] = None):
+        from stextools.snify.skip_and_ignore import get_srskipped_cached, IgnoreList
+
+        tmp_skip = self.skip.get(lang, set())
+        tmp_doc_skip = self.skip_by_docid.get((lang, doc_index), set()) if doc_index is not None else set()
+        srskipped = get_srskipped_cached(self.get_current_document().get_content()).skipped_literal
+        return (
+            tmp_skip
+            | tmp_doc_skip
+            | srskipped
+            | IgnoreList.get_word_set(lang)
+        )
+
+
+    def get_skip_stems(self, lang: str, doc_index: Optional[int] = None):
+        from stextools.snify.skip_and_ignore import get_srskipped_cached
+
+        tmp_skip = self.skip_stem.get(lang, set())
+        tmp_doc_skip = self.skip_stem_by_docid.get((lang, doc_index), set()) if doc_index is not None else set()
+        srskipped = get_srskipped_cached(self.get_current_document().get_content()).skipped_stems
+        return (
+            tmp_skip
+            | tmp_doc_skip
+            | srskipped
+        )
+
+
+    def get_current_document(self) -> Document:
+        if not self.documents:
+            raise ValueError("No documents available.")
[…]

### 2025-07-23 stextools/snify/snifystepper.py

@@ -1,0 +1,206 @@
+import functools
+from typing import Optional, Any
+
+from stextools.snify.annotate import STeXAnnotateCommand, STeXLookupCommand
+from stextools.snify.catalog import Catalog, Verbalization
+from stextools.snify.local_stex_catalog import local_flams_stex_catalogs, \
+    LocalFlamsCatalog
+from stextools.snify.skip_and_ignore import SkipCommand, SkipUntilFileEnd, SkipForRestOfSession, IgnoreCommand, \
+    AddWordToSrSkip, AddStemToSrSkip
+from stextools.snify.snify_commands import View_i_Command, ViewCommand, ExitFileCommand, RescanOutcome, StemFocusCommand, \
+    StemFocusCommandPlus, PreviousWordShouldBeIncluded, FirstWordShouldntBeIncluded, NextWordShouldBeIncluded, \
+    LastWordShouldntBeIncluded
+from stextools.snify.snifystate import SnifyState, SnifyCursor
+from stextools.stepper.command import CommandCollection, CommandSectionLabel, CommandOutcome
+from stextools.stepper.document import STeXDocument, Document
+from stextools.stepper.document_stepper import DocumentModifyingStepper
+from stextools.stepper.interface import interface
+from stextools.stepper.stepper import Stepper, StopStepper, Modification
+from stextools.stepper.stepper_extensions import QuittableStepper, QuitCommand, CursorModifyingStepper, UndoCommand, \
+    RedoCommand, UndoableStepper
+
+
+class SnifyStepper(DocumentModifyingStepper, QuittableStepper, CursorModifyingStepper, UndoableStepper, Stepper[SnifyState]):
+    def __init__(self, state: SnifyState):
+        super().__init__(state)
+        self.state = state
+        self.current_annotation_choices: Optional[list[tuple[Any, Verbalization]]] = None
+
+    @functools.cache
+    def get_stex_catalogs(self) -> dict[str, LocalFlamsCatalog]:
+        return local_flams_stex_catalogs()
+
+    def get_catalog_for_document(self, doc: Document) -> Optional[LocalFlamsCatalog]:
+        error_message: Optional[str] = None
+        catalog: Optional[Catalog] = None
+
+        if isinstance(doc, STeXDocument):
+            catalogs = self.get_stex_catalogs()
+            if not catalogs:
+                error_message = (
+                    f'Error when processing {doc.identifier}:\n'
+                    'No STeX catalogs available.'
+                )
+            elif doc.language not in catalogs:
+                error_message = (
[…]

### 2025-07-23 stextools/snify/stemming.py

@@ -1,7 +1,13 @@
 import functools
 import re
+from logging import getLogger
 
 from stextools.utils.linked_str import LinkedStr
+from stextools.utils.warnonce import warn_once
+
+logger = getLogger(__name__)
+
+SUPPORTED_LANGUAGES = {'en', 'de', 'fr'}
 
 
 @functools.cache
@@ -12,6 +18,12 @@ def get_stem_fun(lang: str):
     elif lang == 'de':
         from nltk.stem import SnowballStemmer
         return SnowballStemmer('german').stem
+    elif lang == 'fr':
+        from nltk.stem import FrenchStemmer
+        return FrenchStemmer().stem
+    warn_once(logger, f'Unsupported language for stemming: {lang}. No stemming will be applied.')
+    return lambda word: word
+    # raise ValueError(f"Unsupported language: {lang}")
 
 
 @functools.cache  # note: caching results in a huge speedup
@@ -28,7 +40,7 @@ def mystem(word: str, lang: str) -> str:
     elif lang == 'de':
         return ' '.join(stem_fun(w) for w in word.split())
     else:
-        raise ValueError(f"Unsupported language: {lang}")
+        return ' '.join(stem_fun(w) for w in word.split())
 
 
 def string_to_stemmed_word_sequence(lstr: LinkedStr, lang: str) -> list[LinkedStr]:
@@ -47,7 +59,7 @@ def string_to_stemmed_word_sequence(lstr: LinkedStr, lang: str) -> list[LinkedSt
 
 
 def string_to_stemmed_word_sequence_simplified(string: str, lang: str) -> list[str]:
-    # same as above, but without linked strings (more efficient
+    # same as above, but without linked strings (more efficient)
     words: list[str] = []
     for match in re.finditer(r'\b\w+\b', string):
         words.append(mystem(match.group(), lang))

### 2025-07-23 stextools/stepper/base_controller.py

@@ -1,174 +1,0 @@
-from __future__ import annotations
-
-from pathlib import Path
-from typing import TypeVar, Generic, Optional, Sequence
-
-import click
-
-from stextools.core.cache import Cache
-from stextools.core.linker import Linker
-from stextools.stepper.command_outcome import Exit, SubstitutionOutcome, TextRewriteOutcome, SetNewCursor, \
-    StatisticUpdateOutcome, FocusOutcome, CommandOutcome
-from stextools.stepper.commands import show_current_selection, UndoOutcome, RedoOutcome, RescanOutcome, \
-    CommandCollection
-from stextools.stepper.file_management import include_inputs
-from stextools.stepper.modifications import Modification, FileModification, CursorModification, StatisticModification, \
-    PushFocusModification
-from stextools.stepper.state import State, FocusInfoType, PositionCursor, SelectionCursor
-
-S = TypeVar('S', bound=State)
-
-
-class BaseController(Generic[S]):
-    def __init__(self, state: S, new_files: Optional[list[Path]] = None, initial_focus_info: Optional[FocusInfoType] = None):
-        self.state: S = state
-        self.mh = Cache.get_mathhub(update_all=True)
-        self._linker: Optional[Linker] = None
-        self._modification_history: list[list[Modification]] = []
-        self._modification_future: list[list[Modification]] = []   # for re-doing
-
-        if new_files:
-            self.state.push_focus(
-                new_files=include_inputs(self.mh, new_files),
-                new_cursor=None,
-                other_info=initial_focus_info   # type: ignore
-            )
-
-
-    @property
-    def linker(self) -> Linker:
-        if self._linker is None:
-            self._linker = Linker(self.mh)
-        return self._linker
-
-    def reset_linker(self):
-        self._linker = None
-        self.post_reset_linker_hook()
-
-    def post_reset_linker_hook(self):
-        """ called every time after the linker is reset """
-        pass
-
-    def run(self) -> bool:
-        """Returns True iff there are more files to annotate."""
-        while True:
-            if not self.state.focus_stack:
-                return False
-
-            if not self.ensure_cursor_selection():
-                click.clear()
-                self.state.pop_focus()
-                if self.state.focus_stack:
-                    print('Focus mode ended')
-                else:
-                    if self.state.statistic_annotations_added:
-                        print(f'Congratulations! You have added {self.state.statistic_annotations_added} annotations.')
-                    print('There are no more files to annotate.')
-                click.pause()
[…]

### 2025-07-23 stextools/stepper/command.py

@@ -1,0 +1,146 @@
+"""
+This module provides the command infrastructure for the stepper.
+
+Instances of (subclasses of) the `Command` class represent concrete commands
+that can be executed in the current context and their instatiation
+may depend on the current state.
+
+If applied, they return a list of ``CommandOutcome``s that describe the
+changes to be made.
+
+``Command``s should be grouped into a ``CommandCollection``.
+A ``CommandCollection`` can then be "applied", which displays the available commands
+and asks the user for a command that will be executed.
+"""
+
+import dataclasses
+import re
+from typing import Sequence, Iterable
+
+from stextools.stepper.interface import interface
+
+
+@dataclasses.dataclass
+class CommandInfo:
+    """
+    Essentially metadata for a command.
+
+    It describes how the command can be invoked and how the command should be presented to the user.
+    """
+    pattern_presentation: str
+    description_short: str
+    pattern_regex: str = ''         # if not set, will be derived from pattern_presentation
+    description_long: str = ''      # if not set, will be derived from description_short
+
+    show: bool = True   # set to False if the command should only be shown in the help text
+
+    def __post_init__(self):
+        if not self.pattern_regex:
+            self.pattern_regex = '^' + self.pattern_presentation + '$'
+
+        if not self.description_long:
+            self.description_long = self.description_short
+
+
+class CommandOutcome:
+    """Result of executing a command."""
+
+
+@dataclasses.dataclass
+class SimpleCommandOutcome(CommandOutcome):
+    """For simple implementations, it suffices to just return the call."""
+    call: str
+
+
+@dataclasses.dataclass
+class Command:
+    """Instances represent a command that can be executed in the current context.
+
+    Usually, you should create a custom subclass for each command you want to implement.
+    """
+    command_info: CommandInfo
+
+    def execute(self, call: str) -> Sequence[CommandOutcome]:
+        del self   # Intentionally unused in the base class.
+        return [SimpleCommandOutcome(call=call)]
+
+    def standard_display(self):
+        interface.write_command_info(
+            self.command_info.pattern_presentation,
+            self.command_info.description_short,
+        )
+
+    def help_display(self):
+        interface.write_command_info(
+            self.command_info.pattern_presentation,
+            self.command_info.description_long,
+        )
+
+
+class _HelpCommand(Command):
+    """ Should only be instantiated by CommandCollection! """
+    def __init__(self, command_collection: 'CommandCollection'):
+        super().__init__(CommandInfo(
+            pattern_presentation='h',
[…]

### 2025-07-23 stextools/stepper/document.py

@@ -1,0 +1,109 @@
+import abc
+import dataclasses
+from pathlib import Path
+from typing import Iterable, Optional
+
+from pylatexenc.latexwalker import LatexWalker
+
+from stextools.stex.local_stex import lang_from_path
+from stextools.stex.stex_py_parsing import STEX_CONTEXT_DB, get_annotatable_plaintext, get_plaintext_approx
+from stextools.stex.flams import FLAMS
+from stextools.utils.linked_str import LinkedStr
+
+
+@dataclasses.dataclass
+class Document(abc.ABC):
+    identifier: str
+    format: str
+    language: str
+
+    @abc.abstractmethod
+    def get_content(self) -> str:
+        pass
+
+    @abc.abstractmethod
+    def set_content(self, content: str):
+        pass
+
+    def get_annotatable_plaintext(self) -> Iterable[LinkedStr[None]]:
+        raise NotImplementedError()
+
+    def get_plaintext_approximation(self) -> LinkedStr:
+        raise NotImplementedError()
+
+    def get_inputted_documents(self) -> Iterable['Document']:
+        r""" Returns the documents that were inputted to this document.
+        (e.g. \input in LaTeX)
+        """
+        return []
+
+
+class STeXDocument(Document):
+    """ A local stex document. """
+    _content: Optional[str] = None
+    _latex_walker: Optional[LatexWalker] = None
+
+    def __init__(self, path: Path, language: str):
+        self.path = path
+        super().__init__(
+            identifier=str(path.resolve().absolute()),
+            format='sTeX',
+            language=language
+        )
+
+    def get_content(self) -> str:
+        if self._content is None:
+            self._content = self.path.read_text()
+        return self._content
+
+    def set_content(self, content: str):
+        self.write_content(content)
+        self._latex_walker = None
+        FLAMS.load_file(self.identifier)
+
+
+    def get_latex_walker(self) -> LatexWalker:
+        """ Returns a LatexWalker for the document content. """
+        if self._latex_walker is None:
+            content = self.get_content()
+            self._latex_walker = LatexWalker(content, latex_context=STEX_CONTEXT_DB)
+        return self._latex_walker
+
+    def write_content(self, content: str) -> None:
+        """ Writes the content to the file. """
+        self.path.write_text(content)
+        self._content = content
+        self._latex_walker = None
+
+    def get_annotatable_plaintext(self) -> Iterable[LinkedStr[None]]:
+        return get_annotatable_plaintext(
[…]

### 2025-07-23 stextools/stepper/document_stepper.py

@@ -1,0 +1,94 @@
+import dataclasses
+from typing import Optional
+
+from stextools.stepper.command import CommandOutcome
+from stextools.stepper.document import Document
+from stextools.stepper.interface import interface
+from stextools.stepper.stepper import State, Modification, StateType, Stepper
+
+
+@dataclasses.dataclass(frozen=True)
+class DocumentCursor:
+    document_index: int
+
+
+class DocumentStepperState(State[DocumentCursor]):
+    def __init__(self, cursor: DocumentCursor, documents: list[Document]):
+        super().__init__(cursor)
+        self.documents = documents
+
+    def get_current_document(self) -> Document:
+        if not self.documents:
+            raise ValueError("No documents available.")
+        return self.documents[self.cursor.document_index]
+
+
+#######################################################################
+#   MODIFY DOCUMENT
+#######################################################################
+
+class SubstitutionOutcome(CommandOutcome):
+    """Note: command is responsible for ensuring that the index is correct *after* the previous file modification outcomes."""
+    def __init__(self, new_str: str, start_pos: int, end_pos: int):
+        self.new_str = new_str
+        self.start_pos = start_pos
+        self.end_pos = end_pos
+
+
+class TextRewriteOutcome(CommandOutcome):
+    def __init__(self, new_text: str):
+        self.new_text = new_text
+
+
+class DocumentModification(Modification):
+    def __init__(self, document: Document, old_text: str, new_text: str):
+        self.document = document
+        self.old_text = old_text
+        self.new_text = new_text
+
+    def apply(self, state: StateType):
+        current_text = self.document.get_content()
+        if current_text != self.old_text:
+            interface.write_text(
+                (f"\n{self.document.identifier} has been modified since the last time it was read.\n"
+                 f"I will not change the file\n"),
+                style='warning'
+            )
+            interface.await_confirmation()
+            return
+
+        self.document.set_content(self.new_text)
+
+    def unapply(self, state: StateType):
+        current_text = self.document.get_content()
+        if current_text != self.new_text:
+            interface.write_text(
+                (f"\n{self.document.identifier} has been modified since the last time it was written to.\n"
+                 f"I will not change the file\n"),
+                style='warning'
+            )
+            interface.await_confirmation()
+            return
+        self.document.set_content(self.old_text)
+
+
+class DocumentModifyingStepper(Stepper[DocumentStepperState]):
[…]

### 2025-07-23 stextools/stepper/file_management.py

@@ -1,57 +1,0 @@
-from pathlib import Path
-
-import click
-
-from stextools.core.mathhub import MathHub
-from stextools.core.stexdoc import Dependency
-
-
-def include_inputs(mh: MathHub, files: list[Path]) -> list[Path]:
-    """If the files input other files, the user will be asked if those should be included as well.
-    The return value is the list of all files that should be processed.
-    """
-    have_inputs = False
-    for file in files:
-        stexdoc = mh.get_stex_doc(file)
-        if not stexdoc:
-            continue
-        for dep in stexdoc.get_doc_info(mh).dependencies:
-            if dep.is_input:
-                have_inputs = True
-                break
-        if have_inputs:
-            break
-    if have_inputs and click.confirm(
-            'The selected files input other files. Should I include those as well?'
-    ):
-        all_files: list[Path] = []
-        all_files_set: set[Path] = set()
-        todo_list = list(reversed(files))
-        while todo_list:
-            file = todo_list.pop()
-            path = file.absolute().resolve()
-            if path in all_files_set:
-                continue
-            stexdoc = mh.get_stex_doc(path)
-            if stexdoc:
-                all_files.append(path)
-                all_files_set.add(path)
-                dependencies: list[Dependency] = [
-                    dep
-                    for dep in stexdoc.get_doc_info(mh).dependencies
-                    if dep.is_input
-                ]
-                # reverse as todo_list is a stack
-                dependencies.sort(key=lambda dep: dep.intro_range[0] if dep.intro_range else 0, reverse=True)
-                for dep in dependencies:
-                    if not dep.is_input:
-                        continue
-                    target_path = dep.get_target_path(mh, stexdoc)
-                    if target_path:
-                        todo_list.append(target_path)
-            else:
-                print(f'File {path} is not loaded')
-
-        return all_files
-    else:
-        return files

### 2025-07-23 stextools/stepper/interface.py

@@ -1,0 +1,513 @@
+"""
+User interfaces for the stepper module.
+"""
+import dataclasses
+import functools
+import shutil
+import subprocess
+from abc import ABC, abstractmethod
+from contextlib import contextmanager
+from typing import Literal, Optional, TypeAlias, Callable, Any
+
+import click
+from pygments import highlight
+from pygments.formatters.terminal import TerminalFormatter
+from pygments.formatters.terminal256 import TerminalTrueColorFormatter
+from pygments.lexers.markup import TexLexer, MarkdownLexer
+
+from stextools.config import get_config
+
+_Color: TypeAlias = str | tuple[int, int, int]
+
+
+def _get_lines_around(text: str, start: int, end: int, n_lines: int = 7) -> tuple[str, str, str, int]:
+    """
+    returns
+      - the n_lines lines before the start index
+      - the text between the start and end index
+      - the n_lines lines after the end index
+      - the line number of the start index
+    """
+    start_index = start
+    for _ in range(n_lines):
+        if start_index > 0:
+            start_index -= 1
+        while start_index > 0 and text[start_index - 1] != '\n':
+            start_index -= 1
+
+    end_index = end
+    for _ in range(n_lines):
+        if end_index + 1 < len(text):
+            end_index += 1
+        while end_index + 1 < len(text) and text[end_index + 1] != '\n':
+            end_index += 1
+    end_index += 1
+
+    return text[start_index:start], text[start:end], text[end:end_index], text[:start_index].count('\n') + 1
+
+
+class interface:
+    """
+    This is a hack because I messed up the design of this module.
+
+    It has to stay synchronized with the abstract Interface class.
+    An alternative would be to use __getattr__,
+    but that would prevent static type checking.
+    """
+    @staticmethod
+    def clear():
+        actual_interface.clear()
+
+    @staticmethod
+    def big_infopage():
+        return actual_interface.big_infopage()
+
+    @staticmethod
+    def write_text(text: str, style: str = 'default', *, prestyled: bool = False):
+        actual_interface.write_text(text, style=style, prestyled=prestyled)
+
+    @staticmethod
+    def apply_style(text: str, style: str) -> str:
+        return actual_interface.apply_style(text, style=style)
+
+    @staticmethod
+    def get_input() -> str:
+        return actual_interface.get_input()
+
+    @staticmethod
+    def newline():
+        actual_interface.newline()
+
+    @staticmethod
[…]

### 2025-07-23 stextools/stepper/modifications.py

@@ -1,93 +1,0 @@
-import abc
-from pathlib import Path
-from typing import Optional
-
-import click
-
-from stextools.stepper.command_outcome import StatisticUpdateOutcome
-from stextools.stepper.state import State, Cursor
-
-
-class Modification(abc.ABC):
-    files_to_reparse: list[Path]
-
-    @abc.abstractmethod
-    def apply(self, state: State):
-        pass
-
-    @abc.abstractmethod
-    def unapply(self, state: State):
-        pass
-
-
-class FileModification(Modification):
-    def __init__(self, file: Path, old_text: str, new_text: str):
-        self.files_to_reparse = [file]
-        self.file = file
-        self.old_text = old_text
-        self.new_text = new_text
-
-    def apply(self, state: State):
-        current_text = self.file.read_text()
-        if current_text != self.old_text:
-            print(click.style(f"File {self.file} has been modified since the last time it was read", fg='black', bg='bright_yellow'))
-            print(click.style(f"I will not change the file", fg='black', bg='bright_yellow'))
-            click.pause()
-            return
-
-        self.file.write_text(self.new_text)
-
-    def unapply(self, state: State):
-        current_text = self.file.read_text()
-        if current_text != self.new_text:
-            print(click.style(f"File {self.file} has been modified since the last time it was written", fg='black', bg='bright_yellow'))
-            print(click.style(f"I will not change the file", fg='black', bg='bright_yellow'))
-            click.pause()
-            return
-        self.file.write_text(self.old_text)
-
-
-class CursorModification(Modification):
-    def __init__(self, old_cursor: Cursor, new_cursor: Cursor):
-        self.files_to_reparse = []
-        self.old_cursor = old_cursor
-        self.new_cursor = new_cursor
-
-    def apply(self, state: State):
-        state.cursor = self.new_cursor
-
-    def unapply(self, state: State):
-        state.cursor = self.old_cursor
-
-
-class PushFocusModification(Modification):
-    def __init__(self, new_files: Optional[list[Path]], new_cursor: Optional[Cursor], other_info):
-        self.files_to_reparse = []
-        self.new_files = new_files
-        self.new_cursor = new_cursor
-        self.other_info = other_info
-
-    def apply(self, state: State):
-        state.push_focus(new_files=self.new_files, new_cursor=self.new_cursor, other_info=self.other_info)
-
-    def unapply(self, state: State):
-        state.pop_focus()
-
-
[…]

### 2025-07-23 stextools/stepper/session_storage.py

@@ -7,15 +7,14 @@ from datetime import datetime
 from pathlib import Path
 from typing import Optional
 
-import click
-
-from stextools.stepper.command_outcome import CommandOutcome, Exit
-from stextools.stepper.commands import CommandInfo, Command, CommandCollection, QuitSubdialogCommand, QuitProgramCommand
-from stextools.stepper.state import State
-from stextools.utils.ui import option_string, standard_header
+from stextools.config import CONFIG_DIR
+from stextools.stepper.command import CommandOutcome, Command, CommandInfo, CommandCollection
+from stextools.stepper.interface import interface
+from stextools.stepper.stepper import State
+from stextools.stepper.stepper_extensions import QuitOutcome, QuitCommand
 
 # TODO: move this somewhere else?
-PATH = Path('~/.config/stextools/sessions').expanduser()
+PATH = CONFIG_DIR / 'sessions'
 
 
 def format_past_timestamp(time: datetime) -> str:
@@ -75,17 +74,16 @@ class PickSessionCommand(Command):
             description_long='Resume the existing session 𝑖')
         )
 
-    def standard_display(self, *, state: State) -> str:
-        lines: list[str] = ['Resume an existing session:']
+    def standard_display(self):
+        interface.write_text('Resume an existing session:\n')
         for i, session in enumerate(self.sessions):
-            line = option_string(
+            interface.write_command_info(
                 str(i),
-                '  ' + format_past_timestamp(datetime.fromtimestamp(session.metadata['timestamp'])) + ' – ' + session.metadata['description']
+                '  ' + format_past_timestamp(datetime.fromtimestamp(session.metadata['timestamp'])) + \
+                ' – ' + session.metadata['description']
             )
-            lines.append(line)
-        return '\n'.join(lines)
 
-    def execute(self, *, state: State, call: str) -> list[CommandOutcome]:
+    def execute(self, call: str) -> list[CommandOutcome]:
         return [SessionChoiceOutcome(int(call), 'resume')]
 
 
@@ -100,7 +98,7 @@ class DeleteSessionCommand(Command):
             description_long='Delete the existing session 𝑖')
         )
 
-    def execute(self, *, state: State, call: str) -> list[CommandOutcome]:
+    def execute(self, call: str) -> list[CommandOutcome]:
         return [SessionChoiceOutcome(int(call[1:]), 'delete')]
 
 
@@ -110,12 +108,11 @@ class DeleteAllSessionsCommand(Command):
         super().__init__(CommandInfo(
[…]

### 2025-07-23 stextools/stepper/state.py

@@ -1,90 +1,0 @@
-import dataclasses
-from pathlib import Path
-from typing import Optional, TypeVar, Generic
-
-from stextools.core.linker import Linker
-from stextools.core.simple_api import file_from_path, SimpleFile
-
-
-@dataclasses.dataclass(frozen=True)
-class Cursor:
-    file_index: int
-
-
-@dataclasses.dataclass(frozen=True)
-class SelectionCursor(Cursor):
-    file_index: int
-
-    selection_start: int
-    selection_end: int
-
-
-@dataclasses.dataclass(frozen=True)
-class PositionCursor(Cursor):
-    file_index: int
-
-    offset: int
-
-
-
-FocusInfoType = TypeVar('FocusInfoType')
-
-@dataclasses.dataclass
-class Focus(Generic[FocusInfoType]):
-    cursor_on_unfocus: Cursor
-    files_on_unfocus: list[Path]
-    other_info: FocusInfoType
-    # select_only_stem: Optional[str] = None
-
-
-@dataclasses.dataclass
-class State(Generic[FocusInfoType]):
-    """ Editing state for snify. This can be saved to a file and reloaded. """
-    files: list[Path]
-    cursor: Cursor
-
-    statistic_annotations_added: int = 0
-    focus_stack: list[Focus[FocusInfoType]] = dataclasses.field(default_factory=list)
-
-    def push_focus(
-            self,
-            new_files: Optional[list[Path]],
-            new_cursor: Optional[Cursor],
-            other_info: FocusInfoType,
-    ):
-        self.focus_stack.append(
-            Focus(cursor_on_unfocus=self.cursor, files_on_unfocus=self.files, other_info=other_info)
-        )
-        if new_files is not None:
-            self.files = new_files
-        if new_cursor is not None:
-            self.cursor = new_cursor
-
-    def pop_focus(self):
-        focus = self.focus_stack.pop()
-        self.cursor = focus.cursor_on_unfocus
-        self.files = focus.files_on_unfocus
-
-    def get_current_file(self) -> Path:
-        return self.files[self.cursor.file_index]
-
-    def get_current_file_simple_api(self, linker: Linker) -> SimpleFile:
-        file = file_from_path(self.get_current_file(), linker)
-        if file is None:
-            raise RuntimeError(f"File {self.get_current_file()} does not seem to be loaded")
-        return file
-
-    def get_current_file_text(self) -> str:
-        return self.get_current_file().read_text()
-
-    def get_selected_text(self) -> str:
[…]

### 2025-07-23 stextools/stepper/stepper.py

@@ -1,0 +1,98 @@
+from abc import abstractmethod, ABC
+from typing import Optional, TypeVar, Generic, Sequence, Literal, TypeAlias
+
+from stextools.stepper.command import CommandCollection, CommandOutcome
+
+
+CursorType = TypeVar('CursorType')
+
+
+class State(Generic[CursorType]):
+    """
+    The state of the stepper.
+    It may be pickled to restore the stepper's state in a future session.
+    It may therefore be necessary that the stepper keeps an additional state for ephemeral data.
+    """
+    def __init__(self, cursor: CursorType):
+        self.cursor = cursor
+
+
+StateType = TypeVar('StateType', bound=State)
+
+class Modification(ABC, Generic[StateType]):
+    """A change that can be undone. E.g. a file modification."""
+    @abstractmethod
+    def apply(self, state: StateType):
+        pass
+
+    @abstractmethod
+    def unapply(self, state: StateType):
+        pass
+
+
+StopReason: TypeAlias = Literal['done', 'quit', 'fatal_error']
+
+class StopStepper(Exception):
+    """Raised to stop the stepper loop."""
+    def __init__(self, reason: StopReason):
+        super().__init__(f'StopStepper: {reason}')
+        self.reason = reason
+
+
+class Stepper(ABC, Generic[StateType]):
+    """
+    The base class for "ispell-like" functionality.
+    """
+    def __init__(self, state: StateType):
+        self.state = state
+
+        # a single undoing/redoing may undo/redo multiple modifications
+        # (e.g. modify a file and change the cursor position)
+        self.modification_history: list[list[Modification[StateType]]] = []
+        self.modification_future: list[list[Modification[StateType]]] = []
+
+    def run(self) -> StopReason:
+        """Run the stepper until it is stopped."""
+        try:
+            while True:
+                self._single_iteration()
+        except StopStepper as e:
+            return e.reason
+
+    def _single_iteration(self):
+        self.ensure_state_up_to_date()
+        self.show_current_state()
+        outcomes: Sequence[CommandOutcome] = self.get_current_command_collection().apply()
+        new_modifications: list[Modification[StateType]] = []
+        for outcome in outcomes:
+            assert isinstance(outcome, CommandOutcome)
+            modification = self.handle_command_outcome(outcome)
+            if modification:
+                new_modifications.append(modification)
+                modification.apply(self.state)
+                self.reset_after_modification(modification)
+
+        if new_modifications:
+            self.modification_history.append(new_modifications)
[…]

### 2025-07-23 stextools/stepper/stepper_extensions.py

@@ -1,0 +1,181 @@
+from copy import deepcopy
+from typing import Optional, Generic
+
+from stextools.stepper.command import Command, CommandInfo, CommandOutcome
+from stextools.stepper.interface import interface
+from stextools.stepper.stepper import Stepper, Modification, StateType, StopStepper, CursorType
+
+
+#######################################################################
+#   QUIT COMMAND
+#######################################################################
+
+class QuitOutcome(CommandOutcome):
+    pass
+
+
+class QuitCommand(Command):
+    def __init__(self, long_description: Optional[str] = None):
+        super().__init__(CommandInfo(
+            pattern_presentation='q',
+            description_short='uit',
+            description_long=long_description or ''
+        ))
+
+    def execute(self, call: str) -> list[CommandOutcome]:
+        return [QuitOutcome()]
+
+
+class QuittableStepper(Stepper[StateType]):
+    def handle_command_outcome(self, outcome: CommandOutcome) -> Optional[Modification[StateType]]:
+        if isinstance(outcome, QuitOutcome):
+            raise StopStepper('quit')
+
+        return super().handle_command_outcome(outcome)
+
+
+#######################################################################
+#   CURSOR SETTING
+#######################################################################
+
+class SetCursorOutcome(CommandOutcome, Generic[CursorType]):
+    def __init__(self, new_cursor: CursorType):
+        self.new_cursor = new_cursor
+
+
+class CursorModification(Modification[StateType], Generic[StateType, CursorType]):
+    def __init__(self, old_cursor: CursorType, new_cursor: CursorType):
+        self.old_cursor = old_cursor
+        self.new_cursor = new_cursor
+
+    def apply(self, state: StateType):
+        # print(f'CursorModification: changing cursor from {state.cursor} to {self.new_cursor}')
+        state.cursor = self.new_cursor
+
+    def unapply(self, state: StateType):
+        # print(f'CursorModification: changing cursor from {state.cursor} to {self.old_cursor}')
+        state.cursor = self.old_cursor
+
+
+class CursorModifyingStepper(Stepper[StateType]):
+    def handle_command_outcome(self, outcome: CommandOutcome) -> Optional[Modification[StateType]]:
+        if isinstance(outcome, SetCursorOutcome):
+            return CursorModification(deepcopy(self.state.cursor), deepcopy(outcome.new_cursor))
+
+        return super().handle_command_outcome(outcome)
+
+
+#######################################################################
+#   UNDO/REDO
[…]

### 2025-07-23 stextools/stex/flams.py

@@ -1,0 +1,92 @@
+import functools
+from pathlib import Path
+from typing import Any
+
+import orjson
+from cffi import FFI
+
+
+class _Flams:
+    def __init__(self):
+        self._all_files_loaded = False
+        self.ffi = FFI()
+        self.ffi.cdef("""
+void hello_world(size_t arg);
+void initialize();
+void load_all_files();
+char* get_file_annotations(char* path);
+void load_file(char* s);
+char* list_of_loaded_files();
+char* list_of_all_files();
+void free_string(char* s);
+""")
+
+    @functools.cached_property
+    def lib(self):
+        path = '/home/jfs/git/github.com/KWARC/FLAMS/target/release/libflams_ffi.so'
+        lib: Any = self.ffi.dlopen(path)
+        lib.initialize()
+        return lib
+
+    def _cstr_to_json(self, c_str) -> Any:
+        """Convert a C string to a JSON object."""
+        py_str = self.ffi.string(c_str).decode('utf-8')
+        self.lib.free_string(c_str)
+        return orjson.loads(py_str)
+
+    def hello_world(self, arg: int):
+        self.lib.hello_world(arg)
+
+    def load_all_files(self):
+        self.lib.load_all_files()
+        self._all_files_loaded = True
+
+    def require_all_files_loaded(self):
+        if not self._all_files_loaded:
+            self.load_all_files()
+
+    def load_file(self, filepath: str | Path):
+        filepath_c = self.ffi.new('char[]', str(filepath).encode('utf-8'))
+        self.lib.load_file(filepath_c)
+
+    def get_file_annotations(self, filepath: str | Path):
+        filepath_c = self.ffi.new('char[]', str(filepath).encode('utf-8'))
+        c_str = self.lib.get_file_annotations(filepath_c)
+        py_str = self.ffi.string(c_str).decode('utf-8')
+        self.lib.free_string(c_str)
+        if not py_str:
+            self.load_file(filepath)
+            c_str = self.lib.get_file_annotations(filepath_c)
+            py_str = self.ffi.string(c_str).decode('utf-8')
+            self.lib.free_string(c_str)
+        if py_str:
+            return orjson.loads(py_str)
+        return None
+
+    def get_loaded_files(self) -> list[str]:
+        return self._cstr_to_json(self.lib.list_of_loaded_files())
+
[…]

### 2025-07-23 stextools/stex/local_stex.py

@@ -1,0 +1,180 @@
+"""
+Code for working with local sTeX archives.
+
+This is mostly generic code, more specific code is in separate modules (for example for the catalogs).
+"""
+from functools import cached_property
+from pathlib import Path
+from typing import Optional, Iterable
+
+from stextools.stex.flams import FLAMS
+from stextools.utils.json_iter import json_iter
+
+
+class FlamsUri:
+    root: str
+    archive: str = ''
+    path: str = ''
+    module: str = ''
+    symbol: str = ''
+
+    def __init__(self, uri: str):
+        if not isinstance(uri, str):
+            raise TypeError(f'Expected a string, got {type(uri)}')
+        parts = uri.split("?")
+        self.root = parts[0]
+        if len(parts) == 1:
+            return
+        assert len(parts) == 2, f'Unexpected FLAMS URI: {uri}'
+        args = parts[1].split('&')
+        for arg in args:
+            key, value = arg.split('=')
+            if key == 'a':
+                self.archive = value
+            elif key == 'p':
+                self.path = value
+            elif key == 'm':
+                self.module = value
+            elif key == 's':
+                self.symbol = value
+            else:
+                raise ValueError(f'Unexpected FLAMS URI argument: {key}={value}')
+
+    def __str__(self):
+        parts = []
+        if self.archive:
+            parts.append(f'a={self.archive}')
+        if self.path:
+            parts.append(f'p={self.path}')
+        if self.module:
+            parts.append(f'm={self.module}')
+        if self.symbol:
+            parts.append(f's={self.symbol}')
+        return '?'.join([self.root, '&'.join(parts)])
+
+
+
+
+class OpenedStexFLAMSFile:
+    def __init__(self, path: str):
+        self.path = path
+
+    @cached_property
+    def text(self) -> str:
+        with open(self.path, 'r', encoding='utf-8') as f:
+            return f.read()
+
+    @cached_property
+    def _linecharcount(self) -> list[int]:
+        """ returns a list l where l[i] is the number of characters until the beginning of line i
+            In FLAMS, lines are apparently 0-indexed.
+        """
+        result: list[int] = [0]
+        value = 0
+        for line in self.text.splitlines(keepends=True):
+            value += len(line)
+            result.append(value)
+        return result
+
+    def flams_range_to_offsets(self, flams_range) -> tuple[int, int]:
+        lc = self._linecharcount
+        start = lc[flams_range['start']['line']] + flams_range['start']['col']
[…]

### 2025-07-23 stextools/stex/stex_py_parsing.py

@@ -1,0 +1,351 @@
+"""
+Code for parsing sTeX files.
+This is not based on FLAMS (FLAMS only extracts annotations, and we need the informal content as well).
+"""
+from typing import Iterable, Optional
+
+from pylatexenc.latexwalker import get_default_latex_context_db, LatexWalker, LatexMathNode, LatexCommentNode, \
+    LatexSpecialsNode, LatexMacroNode, LatexEnvironmentNode, LatexGroupNode, LatexCharsNode, LatexNode
+from pylatexenc.macrospec import MacroSpec, std_environment, VerbatimArgsParser, ParsedMacroArgs
+
+from stextools.stepper.interface import interface
+from stextools.utils.linked_str import LinkedStr, string_to_lstr, fixed_range_lstr, concatenate_lstrs
+
+STEX_MACRO_SPECS: list = [
+    # from section 3.5 of the stex manual (targets modules)
+    MacroSpec('importmodule', '[{'),
+    MacroSpec('requiremodule', '[{'),
+    MacroSpec('usemodule', '[{'),
+
+    # from section 2.2 of the stex manual (targets tex files)
+    MacroSpec('inputref', '*[{'),
+    MacroSpec('mhinput', '*[{'),
+
+    # targets non-tex files
+    MacroSpec('mhgraphics', '[{'),
+    MacroSpec('cmhgraphics', '[{'),
+    MacroSpec('mhtikzinput', '[{'),
+    MacroSpec('cmhtikzinput', '[{'),
+    MacroSpec('lstinputmhlisting', '[{'),
+    MacroSpec('addmhbibresource', '[{'),
+
+    # others
+    MacroSpec('includeproblem', '[{'),
+    MacroSpec('includeassignment', '[{'),
+
+    # from section 1.3 of the stex manual (targets lib/ directory)
+    MacroSpec('libinput', '[{'),
+    MacroSpec('libusepackage', '[{'),
+    MacroSpec('libusetikzlibrary', '[{'),
+
+    # various others
+    MacroSpec('MSC', '{'),
+    MacroSpec('notation', '{[{'),
+    MacroSpec('symdecl', '*{'),
+    MacroSpec('symdef', '{[{'),
+    MacroSpec('definiendum', '[{{'),
+    MacroSpec('definame', '[{'),
+    MacroSpec('Definame', '[{'),
[…]

### 2025-07-23 stextools/stexmmtquery.py

@@ -1,42 +1,0 @@
-from collections import defaultdict
-
-import requests
-
-
-PREFIX: str = '''
-PREFIX ulo: <http://mathhub.info/ulo#>
-PREFIX archive: <mmt://memory/archive#>
-'''
-
-
-def get_sparql_result(query):
-    r = requests.post('https://stexmmt.mathhub.info/:query/sparql', data=query,
-                      headers={'content-type': 'text/plain'})
-    if not r.ok:
-        r.raise_for_status()
-    return r.json()
-
-
-def archive_id_from_uri(uri: str) -> str:
-    if not uri.startswith('mmt://memory/archive#'):
-        raise ValueError(f'Not an archive URI: {uri}')
-    return uri[len('mmt://memory/archive#'):]
-
-
-def get_dependencies() -> dict[str, list[str]]:
-    query = PREFIX + '''SELECT DISTINCT ?from ?to WHERE {
-    ?from a ulo:library .
-    ?from ulo:contains ?file .
-    ?file ulo:contains ?content .
-    ?content ulo:specifies ?thing .
-    ?thing ulo:include ?included .
-    ?othercontent ulo:specifies ?included .
-    ?otherfile ulo:contains ?othercontent .
-    ?otherfile a ulo:file .
-    ?to ulo:contains ?otherfile .
-}'''
-    results = defaultdict(list)
-    json = get_sparql_result(query)
-    for binding in json['results']['bindings']:
-        results[archive_id_from_uri(binding['from']['value'])].append(archive_id_from_uri(binding['to']['value']))
-    return results

### 2025-07-23 stextools/test/test_catalog.py

@@ -1,0 +1,35 @@
+import unittest
+
+from stextools.snify.catalog import Catalog, catalogs_from_stream, Verbalization
+
+
+def get_test_catalog() -> Catalog:
+    return catalogs_from_stream(
+        [
+            ('en', '?edge', Verbalization('edge')),
+            ('en', '?edgenum', Verbalization('edge number')),
+            ('en', '?int', Verbalization('integer')),
+        ]
+    )['en']
+
+
+class TestCatalog(unittest.TestCase):
+    def test_find_match(self):
+        catalog = get_test_catalog()
+
+        for example, expected_string in [
+            ('\nintegers  are fun...\n', 'integers'),
+            ('\ninteger edge\n', 'integer'),
+            ('The edge   number is', 'edge   number'),
+            ('The edge   degree is', 'edge'),
+        ]:
+            with self.subTest(example=example, expected_string=expected_string):
+                match = catalog.find_first_match(
+                    string=example,
+                    stems_to_ignore=set(),
+                    words_to_ignore=set(),
+                    symbols_to_ignore=set(),
+                )
+                self.assertIsNotNone(match)
+                start, end = match[:2]
+                self.assertEqual(example[start:end], expected_string)

### 2025-07-23 stextools/test/test_macros.py

@@ -1,25 +1,0 @@
-import unittest
-
-from pylatexenc.latexwalker import LatexWalker, LatexMacroNode
-
-from stextools.utils.macro_arg_utils import OptArgKeyVals
-from stextools.core.macros import STEX_CONTEXT_DB
-
-
-def parse_macro(s: str) -> LatexMacroNode:
-    walker = LatexWalker(s, latex_context=STEX_CONTEXT_DB)
-    nodes = walker.get_latex_nodes()[0]
-    assert len(nodes) == 1
-    return nodes[0]
-
-
-class MacroTest(unittest.TestCase):
-    def test_key_val(self):
-        keyvals = OptArgKeyVals.from_first_macro_arg(
-            parse_macro(r'\mhtikzinput[archive=MiKoMH/ComSem,width=10cm]{hou/tikz/ellipsis-ex}').nodeargd
-        )
-        self.assertIsNotNone(keyvals)
-        assert keyvals is not None   # for mypy
-        self.assertEqual(len(keyvals), 2)
-        self.assertEqual(keyvals.get_val('archive'), 'MiKoMH/ComSem')
-        self.assertEqual(keyvals.get_val('width'), '10cm')

### 2025-07-23 stextools/utils/json_iter.py

@@ -1,0 +1,11 @@
+from typing import Iterable
+
+
+def json_iter(j) -> Iterable:
+    yield j
+    if isinstance(j, dict):
+        for value in j.values():
+            yield from json_iter(value)
+    elif isinstance(j, list):
+        for item in j:
+            yield from json_iter(item)

### 2025-07-23 stextools/utils/macro_arg_utils.py

@@ -1,105 +1,0 @@
-from typing import Optional
-
-from pylatexenc.latexwalker import LatexNode, LatexCharsNode, LatexGroupNode
-from pylatexenc.macrospec import ParsedMacroArgs
-
-
-class OptArgKeyVals:
-    """ A class to represent optional arguments with key-value pairs in LaTeX.
-        Note: at the moment it's a relatively hacky implementation that does not support changing values.
-        This might have to be changed in the future.
-    """
-
-    def __init__(self, nodelist: list[LatexNode]):
-        self.nodelist = nodelist
-        self._keyvals: dict[str, str] = {}
-
-        current_key: str = ''
-        looking_for_val: bool = False
-        current_val: str = ''
-        # TODO: the following could also track the source refs to enable edits in the source
-        for node in self.nodelist:
-            if isinstance(node, LatexCharsNode):
-                remainder = node.chars
-                while remainder:
-                    if looking_for_val:
-                        if ',' in remainder:
-                            valrest, _, remainder = remainder.partition(',')
-                            current_val += valrest
-                            self._keyvals[current_key.strip()] = current_val.strip()
-                            current_key = ''
-                            current_val = ''
-                            looking_for_val = False
-                        else:
-                            current_val += remainder
-                            remainder = None
-                    else:
-                        if '=' in remainder:
-                            keyrest, _, remainder = remainder.partition('=')
-                            current_key += keyrest
-                            looking_for_val = True
-                        else:
-                            current_key += remainder
-                            remainder = None
-            else:
-                if looking_for_val:
-                    current_val += node.latex_verbatim()
-                else:
-                    current_key += node.latex_verbatim()
-        if current_key:
-            self._keyvals[current_key.strip()] = current_val.strip()
-
-    def as_dict(self) -> dict[str, str]:
-        return {k: v for k, v in self._keyvals.items()}
-
-    @classmethod
-    def from_first_macro_arg(cls, args: ParsedMacroArgs) -> Optional['OptArgKeyVals']:
-        if not args.argnlist:
-            return None
-        first_arg = args.argnlist[0]
-        if not isinstance(first_arg, LatexGroupNode):  # is this even possible?
-            return None
[…]

### 2025-07-23 stextools/utils/timer.py

@@ -1,0 +1,9 @@
+import contextlib
+import time
+
+@contextlib.contextmanager
+def timelogger(logger, task):
+    start = time.time()
+    yield
+    elapsed = time.time() - start
+    logger.info(f"{task} took {elapsed} seconds")

### 2025-07-23 stextools/utils/ui.py

@@ -1,97 +1,0 @@
-import shutil
-from typing import Optional
-
-import click
-
-
-USE_24_BIT_COLORS: bool = True
-
-
-def width():
-    return shutil.get_terminal_size().columns
-
-
-def print_options(intro: Optional[str], options: list[tuple[str, str]]):
-    if intro:
-        print(intro)
-    for key, description in options:
-        print(option_string(key, description))
-
-
-def option_string(key: str, description: str):
-    return '  ' + click.style(f'[{key}', bold=True) + click.style(']', bold=True) + description
-
-
-def color(simple: str, full: tuple[int, int, int]):
-    if USE_24_BIT_COLORS:
-        return full
-    else:
-        return simple
-
-
-def standard_header(title, bg: str = 'bright_green'):
-    print(standard_header_str(title, bg))
-
-
-def standard_header_str(title, bg: str = 'bright_green') -> str:
-    return click.style(f'{title:^{width()}}', bold=True, fg='black', bg=bg)
-
-
-def pale_color():
-    return color('bright_black', (128, 128, 128))
-
-
-def simple_choice_prompt(options: list[str]):
-    return click.prompt(
-        click.style('>>>', bold=True),
-        type=click.Choice(options),
-        show_choices=False, prompt_suffix=''
-    )
-
-
-def print_highlight_selection(doc_text: str, start: int, end: int, n_lines: int = 7, *, bold: bool = True):
-    a, b, c, line_no_start = get_lines_around(doc_text, start, end, n_lines)
-    doc = latex_format(a) + (
-        '\n'.join(click.style(p, fg='black', bg='bright_yellow', bold=bold) for p in b.split('\n'))
-    ) + latex_format(c)
-
-    for i, line in enumerate(doc.split('\n'), line_no_start):
-        print(click.style(f'{i:4} ', fg=pale_color()) + line)
-
-
-def latex_format(code: str) -> str:
-    from pygments import highlight
-    from pygments.lexers import TexLexer
-
-    if USE_24_BIT_COLORS:
-        from pygments.formatters import TerminalTrueColorFormatter as TerminalFormatter
-    else:
-        from pygments.formatters import TerminalFormatter  # type: ignore
-
-    return highlight(code, TexLexer(stripnl=False, stripall=False, ensurenl=False), TerminalFormatter(style='vs'))
-
-
-def get_lines_around(text: str, start: int, end: int, n_lines: int = 7) -> tuple[str, str, str, int]:
-    """
-    returns
-      - the n_lines lines before the start index
[…]

### 2025-07-23 stextools/__main__.py
reduce mandatory dependencies

@@ -4,8 +4,6 @@ from pathlib import Path
 import click
 
 from stextools.config import get_config, CACHE_DIR
-from stextools.lexicon.lexgen import lexgen
-from stextools.snify.snify import snify
 from stextools.stepper.interface import set_interface, DEFAULT_INTERFACES
 
 logger = logging.getLogger(__name__)
@@ -33,6 +31,7 @@ def cli(log_file):
 )
 @interface_option
 def snify_command(files, interface):
+    from stextools.snify.snify import snify
     set_interface(interface)
     if not files:
         click.echo('No files specified. Please provide paths to files or directories to snify.')
@@ -45,6 +44,7 @@ def snify_command(files, interface):
 )
 @interface_option
 def lexgen_command(files, interface):
+    from stextools.lexicon.lexgen import lexgen
     set_interface(interface)
     if not files:
         click.echo('No files specified. Please provide paths to files or directories to snify.')

### 2025-07-24 doc/cicm25-paper/paper.tex
some edits

@@
   Semantic annotations enable services that make the knowledge in documents more
-  accessible.  In the case of \ALeA, a learning assistant we use as a case study, they
+  accessible.  In the case of \ALeA, an adaptive learning assistant we use as a case study, they
   enable, for example, guided tours and practice problems tailored to the learner's
@@
   In this paper we discuss the nature of the semantic authoring problem and argue that,
-  while sharing aspects of classical (informal) authoring and formal authoring, it is a
-  problem of its own that needs special editing support facilities.  We present and
+  while sharing aspects of both classical (informal) authoring and formal authoring, it is
+  a problem of its own that needs special editing support facilities.  We present and
   evaluate a simple utility that supports the process of annotating technical terms with
-    semantic references to a flexiformal content commons \edited{and increases} annotation
+  semantic references to a flexiformal content commons and increases annotation
   productivity by almost an order of magnitude.
@@
 \section{Introduction}
-Arguably, handling large document collections is a key factor in our knowledge-driven society and economy.
-There are two main contenders for machine support in this area:
-symbolic/logic-based technologies and \edited{statistical}/machine-learning-based AI, e.g.\ via LLMs or chatbots.
-They have complementary strengths and challenges:
-ML-based approaches can be trained on all data of the Internet, but face challenges in precision and explainability. 
-Symbolic technologies, on the other hand, \edited{offer precision and explainability%
-%out of the box%
-, but face scalability challenges}
--- in particular because of the cost of semantic authoring, which we discuss in this paper.
+Arguably, handling large document collections is a key factor in our knowledge-driven
+society and economy.  There are two main contenders for machine support in this area:
+symbolic/logic-based technologies and statistical/machine-learning-based AI, e.g.\ via
+LLMs or chatbots.  They have complementary strengths and challenges: ML-based approaches
+can be trained on all data of the Internet, but face challenges in precision and
+explainability.  Symbolic technologies, on the other hand, offer precision and
+explainability, but face scalability challenges -- in particular because of the cost of
+semantic authoring, which we discuss in this paper.
 
-As a case study, we will use adaptive learning assistants for STEM subjects in higher education, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.
-We contend that machine-learning based technologies are largely unsuitable for this task -- at least until they reach the
-precision and explainability required by the ethics of teaching --
-while the scope of specific domains is just narrow enough that symbolic approaches are viable.
-Conventional (\LaTeX-based) lecture materials can essentially be directly imported into \ALeA,
-but its ability to generate learner-adaptive learning objects -- instrumented with learning support interactions --
-depends on semantic annotations.
-These can be embedded in \LaTeX\ documents using the \sTeX package~\cite{MueKo:sdstex22,sTeX:github:on}.
-In this paper, we will mostly focus on annotating terms with the semantic concepts they refer to,
-but \sTeX supports a much wider range of annotations, such as annotating problems with the learning objectives
-or annotating the defining phrase (definiens) in concept definitions,
-and could, in principle, even be used for full formalization.
-As \sTeX supports the spectrum from informal (un-annotated) to fully formalized content,
-we say that \sTeX content is \textbf{flexiformal}.
+As a case study, we will use adaptive learning assistants for STEM subjects in higher
+education, concretely the \ALeA system \cite{BerBetChu:lssmkm23}.  We contend that
+machine-learning based technologies are largely unsuitable for this task -- at least until
+they reach the precision and explainability required by the ethics of
+teaching\edited{\footnote{Incidentally, the AI Act of the European Union \cite{EUAIAct:on}
+    classifies education as a high-risk area and thus mandates high requirements of
+    explainability, accountability, and human oversight for educational AI applications.}}
+-- while the scope of specific domains is just narrow enough that symbolic approaches are
+viable.  Conventional (\LaTeX-based) lecture materials can essentially be directly
+imported into \ALeA, but its ability to generate learner-adaptive learning objects --
+instrumented with learning support interactions -- depends on semantic annotations.  These
+can be embedded in \LaTeX\ documents using the \sTeX
+package~\cite{MueKo:sdstex22,sTeX:github:on}.  In this paper, we will mostly focus on
+annotating terms with the semantic concepts they refer to, but \sTeX supports a much wider
+range of annotations, such as annotating problems with the learning \edited{prerequisites
+  and} objectives or annotating the defining phrase (definiens) in concept definitions,
+and could, in principle, even be used for full formalization.  As \sTeX supports the
+spectrum from informal (un-annotated) to fully formalized content, we say that \sTeX
+content is \textbf{flexiformal}.
 
@@
 The work reported in this article was conducted as part of the VoLL-KI project (see
-\url{https://voll-ki.de}) funded by the German Research/Education Ministry (BMBF) under
+\url{https://voll-ki.de}) funded by the German Research/Education Ministry (BMFTR) under
 grant 16DHBKI089.
@@
 
-Before we discuss semantic authoring in general, let us introduce a concrete example from the \sTeX/\ALeA world.
-    % We will use it to sharpen our intuitions of the challenges involved and to ground the solution we propose.
+Before we discuss semantic authoring in general, let us introduce a concrete example from
+the \sTeX/\ALeA world.
+% We will use it to sharpen our intuitions of the challenges involved and to ground the
+% solution we propose.
+
+\begin{figure}[h]\centering
+  \fbox{\includegraphics[width=10cm]{../img/minimax-remarks.pdf}}
+  \caption{An annotated learning object from an AI lecture}\label{fig:lo}
+\end{figure}
 
@@
 course materials into semantic documents.
-
-\begin{figure}[h]\centering
-  \fbox{\includegraphics[width=10cm]{../img/minimax-remarks.pdf}}
-  \caption{An annotated learning object from an AI lecture}\label{fig:lo}
-\end{figure}
-
 \begin{figure}[h]\centering
@@
 
-Take for instance a learning object like the paragraph in \Cref{fig:lo}, where we want to
-annotate the term ``terminal'' with the (semantic) concept of ``a goal state of a search
-problem''. The latter can either be defined earlier in the course or in the domain model
-(or both). Concretely this is about converting the unannotated {\LaTeX} string
+Take for instance a learning object like the paragraph in \Cref{fig:lo}\ednote{MK: in
+  diesem Bild sollten keine blauen/annotierten Wörter sein}, where we want to annotate the
+term ``terminal'' with the (semantic) concept of ``a goal state of a search problem''. The
+latter can either be defined earlier in the course or in the domain model (or
+both). Concretely this is about converting the unannotated {\LaTeX} string
 
@@
 ]
-\item \blue{Note:} Depth-limited minimax requires an evaluation for every
+\blue{Note:} Depth-limited minimax requires an evaluation for every
 cut-off state $s$. If $s$ is terminal, we use its utility,
@@
 \end{lstlisting}
-into the partially annotated item
+into the partially annotated 
 \begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
@@
 [...]
-\item \blue{Note:} Depth-limited minimax requires an evaluation for every
+\blue{Note:} Depth-limited minimax requires an evaluation for every
 cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its utility,
@@
 \end{lstlisting}
-\edited{
-where the author has annotated the word ``terminal'' with the concept it refers
-to via} the \sTeX macro
+where the author has annotated the word ``terminal'' with the concept it refers to via the
+\sTeX macro
 \lstinline[mathescape]|\sr{$\llangle \mathit{symbol} \rrangle$}{$\llangle \mathit{verbalization}\rrangle$}|.
-In \sTeX, \edited{concepts} are represented by \textbf{symbols}, which are identified
-globally and unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module name}, and a \textbf{symbol name}.
- % Every symbol can have one or more verbalizations. %: words of the domain jargon used to refer to the concept or
-\edited{
-A \textbf{verbalization} is the natural language rendering of a concept.
-% object.
-Verbalizations are, of course, language-dependent and can vary between subdomains and communities of
-practice.
-}
-In our example, the verbalization ``terminal [state]'' was annotated with the symbol with
-name \lstinline|goal state|. It comes from the module with name \lstinline|search-problem|
-in the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
+In \sTeX, concepts are represented by \textbf{symbols}, which are identified globally and
+unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module name}, and a
+\textbf{symbol name}.  A \textbf{verbalization} is the natural language rendering of a
+concept.  Verbalizations are, of course, language-dependent and can vary between
+subdomains and communities of practice, even if the underlying symbol is the same.  In our
+example, the verbalization ``terminal [state]'' was annotated with the symbol with name
+\lstinline|goal state|. It comes from the module with name \lstinline|search-problem| in
+the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
 macro in line 1. As the symbol name \lstinline|goal state| is unique in the imports here
-it suffices as a relative URI in the \lstinline|\sr| annotation. If the symbol name and
-the desired verbalization coincide (as they often do in English), the short-hand macro
+it suffices to use a relative URI in the \lstinline|\sr| annotation. If the symbol name
+and the desired verbalization coincide (as they often do in English), the short-hand macro
 \lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| for ``utility'').
@@
 % LocalWords:  lang DHBKI FS Todo FTML lexiformal arkup anguage CICM RDF postprocessing
-% LocalWords:  de maths EE CPoint SMD rus WOIDE linter LSP emacs Prot BMBF
+% LocalWords:  de maths EE CPoint SMD rus WOIDE linter LSP emacs Prot BMBF diesem Bild
+% LocalWords:  sollten keine blauen annotierten Wörter sein

### 2025-07-24 doc/cicm25-paper/paper.tex
some more paper editing

@@
 \title{
-    % Semantic Authoring in the \sTeX/\ALeA World
-    % \ednote{Find better title. Was ``Bulk Semantic Annotation with a Partially Known Knowledge Base''}
         Semantic Authoring in a Flexiformal Context --- Bulk Annotation of Rigorous Documents
@@
 
-% \keywords{semantic authoring, \snify, \sTeX, \ALeA, flexiformal content commons}
 \institute{Computer Science, FAU Erlangen N\"urnberg, Germany}
@@
 \begin{abstract}
-%   Semantic annotations can be utilized for learning support services in technical
-%   documents. The resulting active documents support more direct and personalized
-%   interaction with the knowledge conveyed, but require a large investment in the authoring
-%   process.
-% 
-%   In this paper we discuss the nature of the semantic authoring problem and argue that
-%   while sharing aspects of classical, informal authoring and formal authoring, it is a
-%   problem of its own that needs special editing support facilities. We present a simple
-%   utility that supports the process of annotating technical terms with semantic references
-%   to a flexiformal content commons that enhances annotation productivity by almost an
-%   order of magnitude.
-
   Semantic annotations enable services that make the knowledge in documents more
@@
 
-  In this paper we discuss the nature of the semantic authoring problem and argue that,
+  In this paper we discuss the nature of the semantic authoring task and argue that,
   while sharing aspects of both classical (informal) authoring and formal authoring, it is
-  a problem of its own that needs special editing support facilities.  We present and
+  a task of its own that needs special editing support facilities.  We present and
   evaluate a simple utility that supports the process of annotating technical terms with
@@
 \end{abstract}
+
+\keywords{\edited{semantic authoring} \and flexiformality \and annotation \and \snify \and \sTeX \and \ALeA}
 % \setcounter{tocdepth}{3}
@@
 they reach the precision and explainability required by the ethics of
-teaching\edited{\footnote{Incidentally, the AI Act of the European Union \cite{EUAIAct:on}
+teaching\footnote{Incidentally, the AI Act of the European Union \cite{EUAIAct:on}
     classifies education as a high-risk area and thus mandates high requirements of
-    explainability, accountability, and human oversight for educational AI applications.}}
+    explainability, accountability, and human oversight for educational AI applications.}
 -- while the scope of specific domains is just narrow enough that symbolic approaches are
@@
 annotating terms with the semantic concepts they refer to, but \sTeX supports a much wider
-range of annotations, such as annotating problems with the learning \edited{prerequisites
-  and} objectives or annotating the defining phrase (definiens) in concept definitions,
+range of annotations, such as annotating problems with the learning prerequisites
+and objectives or annotating the defining phrase (definiens) in concept definitions,
 and could, in principle, even be used for full formalization.  As \sTeX supports the
@@
 In this paper, we argue that semantic/flexiformal authoring shares aspects with both informal and fully formal authoring,
-but is nevertheless a distinct problem requiring special tool support (\Cref{sec:semauth}).
+but is nevertheless a distinct task requiring special tool support (\Cref{sec:semauth}).
 Using \sTeX/\ALeA as an exemplary case study,
@@
 \begin{figure}[h]\centering
-  \fbox{\includegraphics[width=10cm]{../img/minimax-remarks.pdf}}
-  \caption{An annotated learning object from an AI lecture}\label{fig:lo}
+  \fbox{\includegraphics[width=10cm,trim=0.8cm 0 0 0]{../img/minimax-remarks.pdf}}
+  \caption{An unannotated learning object from an AI lecture}\label{fig:lo}
 \end{figure}
@@
 
-To annotate the word ``terminal'' in the document, the author has to 
+To annotate the word ``terminal'' in the document above, the author has to 
 \begin{inparaenum}[\em i\rm)]
@@
 \end{inparaenum}
-\edited{A rather daunting task for a lecture with 50,000 -- 100,000 words in notes and slides and ca. 10--15\,\% of them technical terms.\footnote{
-    \edited{
+A rather daunting task for a lecture with 50,000 -- 100,000 words in notes and slides and ca. 10--15\,\% of them technical terms.\footnote{
         Estimates based on some of our own lectures materials
@@
 }
-}}
 
@@
 
-\edited{The annotation catalog is language-specific.}
-% For documents in a language $L$,
-% where $L$ is not English, the annotation practices are a bit
-% more complex still, in particular because the annotation catalog is usually incomplete \edited{for $L$}.
-% Note that for $L$-annotation, we need an $L$-specific annotation catalog. 
+The annotation catalog is naturally language-specific.
 \sTeX already supports multilinguality:
@@
   ISO-639 language identifier).
-\item \edited{Modules in the domain model allow translations that share the formal
-    declarations with another module.}
-        % In particular all definienda in $L$-modules are $L$-specific.
+\item Modules in the domain model include translations that share the formal
+    declarations with another module.
+        This induces a ``translation relation'' for verbalizations
+        in different languages with the same symbol URI.
 \end{itemize}
-% But still, $L$-annotation catalogs are usually inadequate for $L$-annotation tasks in
-\edited{In practice, the annotation catalog for languages other than English is often inadequate as}
+In practice, the annotation catalog for languages other than English is often inadequate as
 \begin{enumerate}[\em i\rm)]
-    \item \edited{the domain model may not have translations in that language -- especially for the ``smaller'' languages, and}
-    \item \edited{symbol names, usually English words, cannot be used as a verbalizations.}
+    \item the domain model may not have translations in that language -- especially for the ``smaller'' languages, and
+    \item symbol names, usually English words, cannot be used as a verbalizations.
 \end{enumerate}
-\edited{In this situation, we usually have to invest in
+In this situation, we usually have to invest in
 establishing a suitable catalog -- e.g. by providing translations (or at least dictionary
@@
 domain model does not cover in the documents to be annotated.
-}
 
-\subsection{Semantic Authoring as a Distinct Problem}
-We contend that semantic authoring is a problem that is distinct from traditional
+\subsection{Semantic Authoring as a Distinct Task}
+We contend that semantic authoring is a task that is distinct from traditional
 (informal) authoring and authoring fully formal corpora (e.g. programs or formalizations),
@@
 canonical part of the domain model with the authors. Consequently a large part of the
-authoring problem lies in predicting the state and extent of the domain model of the
+authoring challenge lies in predicting the state and extent of the domain model of the
 reader and creating text that bridges the gap between the reader's domain model and the
@@
 support for now.
+\edited{
+In our experience, LLM-based authoring support is less helpful in technical documents
+than in everyday writing.
+}
 
@@
 In addition to the base text, semantic authoring also has to provide the semantic
-annotations that drive the personalization and interaction. Techniques from formal
+annotations that drive the personalization and interaction downstream. Techniques from formal
 authoring could help here, but in contrast to formal authoring, we cannot just trigger the
@@
 the imaginary unit in formulae is represented e.g. by a semantic macro
-\lstinline|\imaginaryunit|, which can alternatively expand to $i$ (for uses in maths) and
+\lstinline|\imaginaryunit|, which can alternatively \edited{be presented as} $i$ (for uses in maths) or
 $j$ (for EE uses). As a consequence semantic authoring is more geared towards creating
@@
 to their definienda.  A tool for supporting semantic authoring of formulae via notations
-has been presented by Vre{\v{c}}ar et al.\ in~\cite{VreWelKam:tsmmdui24}.  It uses a
-grammar to parse informal, presentation-focused formulae (e.g.\
+has been presented by Vre{\v{c}}ar et al.\ in~\cite{VreWelKam:tsmmdui24},
+\edited{so we disregard this important aspect here.
+The tool} uses a grammar to parse informal, presentation-focused formulae (e.g.\
 \lstinline[keywordstyle={}]|A\times B|) into abstract syntax trees based on semantic
@@
 \lstinline|s!| skips it until the end of the file, and \lstinline|S|
-will skip it in future runs as well by appending a comment to the file.
+will skip it in future runs as well by appending a \snify-specific comment to the file.
 There are numerous other commands, e.g.\ to adjust the selection,
@@
 \lstinline[language={}]|\emph| or \lstinline[language={}]|\textbf|), and then steps
-through them as \snify does.  For each potential definiendum, the user can then use fuzzy
-search to find the corresponding symbol in the domain model.  This populates the catalog
+through them as \snify does.
+For each potential definiendum, the user can then use fuzzy
+search to find the corresponding symbol in the domain model
+\edited{and replace the original macro with an} \sTeX \edited{definiendum annotation}.
+This populates the catalog
 with verbalizations relevant to the document, making subsequent \snify runs more
@@
 annotations and imports by hand.\footnote{
-    \edited{
         The precise productivity gain depends on the annotator's
-        familiarity with the domain model (for the right symbol
-        is time-consuming) but also on other factors like
+        familiarity with the domain model (\edited{searching and identifying}
+        the right symbol is time-consuming) but also on other factors like
         the typing speed.
@@
         to 13 (for a beginner).
-        The focus mode further increases productivity.
-    }
+        The focus mode further increases productivity
+        when considering the overall corpus quality.
 }
@@
 non-English languages in the \sTeX corpus we work with. Coverage in German is only about
-half of that of English and we can already see the practical effects.  The \defianno tool
+half of that of English in the domain model and we can already see the practical effects.  The \defianno tool
 can mitigate this partially by providing verbalizations for
@@
 \snify's interface is inspired by traditional spell checkers
-like \lstinline|ispell|~\cite{ispellman},
-which step through documents and suggest corrections for misspelled words.
+like \lstinline|ispell|~\cite{ispellman}
+that step through documents and suggest corrections for misspelled words,
+\edited{which is of course a very different underlying task.}
 
@@
 browser-based) tools for this task; a prominent example is
-\textsf{brat}~\cite{brat:on}.
+\textsf{brat}~\cite{brat:on} for text annotation.
+\edited{
+    Technical documents with tables, complex formulae, etc. cannot
+    easily be expressed as sequences of words and require different annotation
+    tools like} e.g.\ \textsf{AnnoTize}~\cite{PanzerSchaefer:AnnoTize23}
+\edited{
+    that supports the annotation of HTML documents containing MathML formulae.
+}
 
@@
 facility that allows annotating documents with ontology -- mostly ABox -- references; this
-is the nearest match to our \snify system.
+may be the nearest match to our \snify system.
 
 \section{Conclusion}\label{sec:conclusion}
-In this paper we have presented the semantic authoring problem in a flexiformal context
+In this paper we have presented the semantic authoring task in a flexiformal context
 using the \sTeX corpus and the adaptive learning assistant \ALeA as a concrete example. We
@@
 presented \snify, a command-line tool that bridges the gap between informal text and
-formal symbols in \sTeX.  Given the impact of \snify on the authoring process, we are
+formal/semantic symbols in \sTeX.
+\edited{It is open source; the code and documentation are available at}
+\cite{stextools:git}.
+Given the impact of \snify on the authoring process, we are
 currently working on a deeper integration of \snify into the \sTeX ecosystem and content
 commons as described in \Cref{sec:scaling}.
-
-The \snify code is open source; the source and documentation are available from
-\cite{stextools:git}.
 

### 2025-07-24 doc/cicm25-paper/paper.tex
minor edits

@@
 \addbibresource{extcrossrefs.bib}
+\addbibresource{otherrefs.bib}
 \usepackage[noabbrev]{cleveref}
@@
 % postprocessing of the result (de-duplication, stemming, and pairing) away.
-\edited{To cope with this situation, we are planning to extend the brand-new} \flams \edited{system, which among other things powers} \url{https://MathHub.info},
+\edited{To cope with this situation, we are planning to extend the brand-new} \flams
+system~\cite{flamsongithub}, \edited{which, among other things, powers} \url{https://MathHub.info},
 \edited{so that it provides the data necessary to create a
@@
 Currently, \snify uses a custom Python implementation for harvesting the verbalization cache,
-but we are working on a implementation based on the new \textbf{\flams linter} -- a simple \sTeX parser that
+but \edited{we are in the late stages of developing a new} \snify version using the
+\textbf{\flams linter} -- a simple \sTeX parser that
 drives the LSP functionality of the \sTeX IDE plugins for \textsf{VSCode} and \textsf{emacs}.

### 2025-07-25 doc/cicm25-paper/paper.tex
finalized paper?

@@
 \usepackage[style=alphabetic,hyperref=auto,defernumbers=true,backend=bibtex,firstinits=true,maxbibnames=9,maxcitenames=3,isbn=false]{biblatex}
+\usepackage{xurl}
 \addbibresource{kwarcpubs.bib}
@@
 \usepackage{soul}
-\def\edited#1{\hl{#1}}
+% \def\edited#1{\hl{#1}}
+\def\edited#1{#1}
 
@@
 \def\rrangle{\rangle\kern-.2em\rangle}
-\def\flams{\textsf{FLAMS}\xspace}
+% \def\flams{\textsf{FLAMS}\xspace}
 
@@
 
-% \ednote{
-%     \textbf{Related work: semantic web authoring}
-%     \emph{semantic authoring} already used in context of semantic web (RDFa).
-%     Tools exist and it doesn't have exactly the same challenges as we face, though there is some overlap.
-%     We have a module system and need more dependency management, something else?
-%     This has to be described somewhere.
-% 
-%     Related work: ``User Interfaces for Semantic Authoring of Textual Content: A Systematic Literature Review.''
-% }
-
 \paragraph{Overview and Contribution}
@@
 
-\begin{figure}[h]\centering
+\begin{figure}\centering
   \fbox{\includegraphics[width=10cm,trim=0.8cm 0 0 0]{../img/minimax-remarks.pdf}}
@@
 course materials into semantic documents.
-\begin{figure}[h]\centering
-  \fbox{\includegraphics[width=12cm]{../img/search-problem.en.pdf}}
-\begin{lstlisting}[morekeywords={definame,symdecl,definiendum},numbers=left,
-escapechar=!]
-\begin{smodule}[title=Search Problem]{search-problem}
-[... some imports ...]
-\symdecl*{goal state}
-\begin{sdefinition}
-  [...] Certain \sns{state} are [...] \definame[post=s]{goal state} [...]
-  (also called \definiendum{goal state}{terminal states}).
-\end{sdefinition}
-\end{smodule}
-\end{lstlisting}
-  \caption{Simplified definition of ``goal state'' from the domain model}\label{fig:state-space}
-\end{figure}
 
-Take for instance a learning object like the paragraph in \Cref{fig:lo}\ednote{MK: in
-  diesem Bild sollten keine blauen/annotierten Wörter sein}, where we want to annotate the
+Take for instance a learning object like the paragraph in \Cref{fig:lo},
+    where we want to annotate the
 term ``terminal'' with the (semantic) concept of ``a goal state of a search problem''. The
@@
 both). Concretely this is about converting the unannotated {\LaTeX} string
-
 \begin{lstlisting}[numbers=left,firstnumber=3,
@@
 \Cref{fig:state-space}.  The \lstinline|\definame| and \lstinline|\definiendum| in lines
-5--6 introduce the verbalizations ``goal state'' and ``terminal state'' for the symbol
-\lstinline|search-problem?goal state| introduced by the \lstinline|\symdecl*| declaration in
+5--6 introduce the verbalizations ``goal state'' and ``terminal state'' for the
+\lstinline|goal state| symbol introduced by the \lstinline|\symdecl*| declaration in
 line 3.
+
+\begin{figure}\centering
+  \fbox{\includegraphics[width=12cm]{../img/search-problem.en.pdf}}
+\begin{lstlisting}[morekeywords={definame,symdecl,definiendum},numbers=left,
+escapechar=!]
+\begin{smodule}[title=Search Problem]{search-problem}
+[... some imports ...]
+\symdecl*{goal state}
+\begin{sdefinition}
+  [...] Certain \sns{state} are [...] \definame[post=s]{goal state} [...]
+  (also called \definiendum{goal state}{terminal states}).
+\end{sdefinition}
+\end{smodule}
+\end{lstlisting}
+  \caption{Simplified definition of ``goal state'' from the domain model}\label{fig:state-space}
+\end{figure}
 
@@
 support for now.
-\edited{
 In our experience, LLM-based authoring support is less helpful in technical documents
 than in everyday writing.
-}
 
@@
 informal-formal gap.
+We have experimented with LLM-based generation of annotated (\sTeX) content,
+but the results have been disappointing so far.
 
@@
 Efficient support for the annotation of technical terms in natural language
-has been elusive so far.
+has been elusive so far -- we will address this in the next section.
 
@@
 
+% \newpage
 \paragraph{Evaluation}
 % \section{Practical Evaluation and Future Work}\label{sec:eval}
-\begin{wrapfigure}{r}{0.4\textwidth}
-    \vskip-1.2cm
-  \centering
-  \includegraphics[width=0.4\textwidth]{../img/annocounts.pdf}
-    \vskip-0.2cm
-  \caption{Number of annotated terms in the \sTeX/\ALeA commons over time.}\label{fig:annocounts}
-  \vskip-0.7cm
-\end{wrapfigure}
 In our experience, \snify's step-through workflow for annotating term references 
@@
 For annotators unfamiliar with the domain model, the unassisted annotation task is
+\begin{wrapfigure}{r}{0.4\textwidth}
+    \vskip-0.7cm
+  \centering
+  \includegraphics[width=0.4\textwidth]{../img/annocounts.pdf}
+    % \vskip-0.2cm
+  \caption{Number of annotated terms in the \sTeX/\ALeA commons over time.}\label{fig:annocounts}
+  \vskip-0.7cm
+\end{wrapfigure}
 almost infeasible, and the average unfamiliarity naturally grows with the domain
@@
   \caption{Architecture}\label{fig:arch}\vspace*{-1em}
+  \vskip-0.5cm
 \end{wrapfigure}

### 2025-07-30 scripts/annocountplot.py
final paper version

@@ -72,7 +72,7 @@ def cumulative_plots():
 
 def diff_plots():
     plot_diff(filtered('smglom/*'), 'smglom')
-    plot_diff(filtered('courses/*/course'), 'course')
+    plot_diff(filtered('courses/*/course'), 'courses')
     plot_diff(filtered('*/problems'), 'problems')
 
 

### 2025-07-30 stextools/catalog_export.py
woide catalog export

@@ -1,0 +1,34 @@
+import json
+from datetime import datetime
+from pathlib import Path
+
+from stextools.snify.local_stex_catalog import local_flams_stex_catalogs
+
+
+def catalog_export(path: Path):
+    catalogs = local_flams_stex_catalogs()
+
+    for lang, catalog in catalogs.items():
+        with open(path / f'catalog-{lang}.json', 'w') as fp:
+            catalog_data = {
+                'entries': [
+                    {
+                        'verb': verb.verb,
+                        'reference': symbol.uri,
+                        'origin': {
+                            'value': verb.local_path,
+                            'type': 'localstexfile',
+                        }
+                    }
+                    for symbol in catalog.symb_iter()
+                    for verb in catalog.get_symb_verbs(symbol)
+                ],
+                'language': lang,
+                'date': datetime.now().isoformat(),
+                'creator': 'stextools catalog export',
+            }
+            json.dump(catalog_data, fp, indent=2)
+
+if __name__ == '__main__':
+    import sys
+    catalog_export(Path(sys.argv[1]))

### 2025-07-30 stextools/snify/catalog.py

@@ -57,6 +57,9 @@ class Catalog(Generic[Symb, Verb]):
     def symb_iter(self) -> Iterable[Symb]:
         yield from self.symb_to_verb.keys()
 
+    def get_symb_verbs(self, symb: Symb) -> list[Verb]:
+        return self.symb_to_verb.get(symb, [])
+
     def add_symbverb(self, symb: Symb, verb: Verb):
         self.symb_to_verb.setdefault(symb, []).append(verb)
         key = string_to_stemmed_word_sequence_simplified(verb.verb, self.lang)

### 2025-09-26 doc/aligments2025/doc.tex
alignments abstract draft

\documentclass{ceurart}
\usepackage[show]{ed}
\usepackage{paralist}
\usepackage{amsmath}
\usepackage{cleveref}
\usepackage{listings}
\lstset{
  basicstyle=\ttfamily,
}

\title{Efficiently Linking Documents to Alignment Databases -- A Prototype-Based Exploration}

\author[1]{Michael Kohlhase}
\author[2]{Jan Frederik Schaefer}

\conference{Workshop on Math Datasets: alignments and comparisons (ALIGN2025),
  October 10, 2025, Brasilia, Brazil}

\begin{document}
\maketitle
Both formal and informal mathematical knowledge is fragmented across many different systems and formats.
A first step towards combining these resources is aligning their concepts.

The focus of this paper is the alignment
of concept references in documents -- e.g.\ scientific papers -- with a concept database.
We will use WikiData as the concept database in our examples.
As WikiData links to other resources (e.g.\ nLab, ProofWiki, OEIS, \ldots),
and is also referenced in other databases
(e.g. the MathGloss~\cite{mathgloss:arxiv} alignment database),
% As WikiData is often referenced in other alignment databases,
% e.g.\ MathGloss~\cite{mathgloss:arxiv},
we also get indirect alignments with other systems.

As a running example, we will use the simple declarational sentence
\begin{quote}
Let $n$ be a natural number.
\end{quote}
where we want to align the concept reference ``natural number'' with
the corresponding WikiData identifier.
Such an alignment could, for example,
provide additional information to the reader,
could determine the prerequisite knowledge for understanding the document,
or be a first step towards formalization.
Creating such alignments is difficult.
For our example there are, in fact, three possible alignment targets:
\begin{compactitem}
\item \texttt{Q21199} (natural numbers, possibly including 0)
\item \texttt{Q28920044} (positive integers, i.e.\ natural numbers excluding 0)
\item \texttt{Q28920052} (non-negative integers, i.e.\ natural numbers including 0)
\end{compactitem}
We claim that machine learning techniques are not yet capable of
making the right choice sufficiently reliably,
which puts the burden of creating alignments on human annotators --
ideally the authors.

The most obvious workflow is rather inefficient:
the author has to either recall the correct alignment target from
memory or search for it, and then edit the document to add the alignment annotation.
A more efficient workflow, supported by dedicated tools,
is crucial for large-scale alignment efforts.

The sTeX ecosystem, which uses diverse semantic markup to drive
learning support systems,
faced a similar challenge, which was recently addressed by the snify system.
In this paper, we present a prototypical extension of the snify system
for efficient alignment that is independent of the sTeX ecosystem.
Concretely, we have added the option to use WikiData as the concept database,
rudimentary support for annotating formulae,
the ability to annotate HTML documents,
and a browser-based user interface, which
helps with the annotation of HTML documents.

The snify tool is a simple command line application that
steps through the phrases that could be annotated in a document
and presents the user with a list of potential targets.
\Cref{fig:snify} shows the tool in action.
The user can select the correct entry from the list by typing its number,
which then creates the alignment by inserting a macro into the document
(here e.g.\ \lstinline.\wdalign{Q28920044}{natural number}).
Aside from choosing the correct annotation,
snify has many other commands to skip
phrases, undo annotations, modify the selection, etc.

This workflow transforms the problem from a \emph{recollection}
problem, where the user has to remember the correct alignment target
or otherwise spend time searching for it,
to a \emph{selection} problem, where the user can
pick the correct option from a list of candidates.
In the sTeX ecosystem, the snify tool has significantly increased
the efficiency of annotators, especially for novices who
are not deeply familiar with the domain model (i.e.\ the concept database).

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{img1.png}
    \caption{Screenshot of snify.}\label{fig:snify}
\end{figure}

In the background, snify uses a verbalization catalog
that links concept, here WikiData identifiers,
to their natural language renderings (verbalizations).
WikiData has both language-specific labels and synonyms.
For example, \lstinline.Q28920044. has the
label ``positive integer'' and lists ``natural number''
as a synonym.
This verbalization catalog drives the search for potential targets
and the selection of candidates.
To recognize inflected forms -- e.g.\ ``continuities'' instead of ``continuity''
-- snify uses an off-the-shelf stemmer.

As WikiData also has formula notations
for many concepts, including \LaTeX\ notations like \lstinline.\mathbb{N}.,
we can also use the same workflow to align formula fragments.
Such formula alignment is rather limited:
operators can be aligned, but their arguments are not marked, and
some operators, like the invisible multiplication in ``$ax$'',
are not represented by a glyph at all.
In sTeX, the solution is essentially to write down
a semantic representation of the formula and generate the presentation from it.
A grammar-based approach to generate the semantic representation
for ``conventional'' \LaTeX\ formulae
in collaboration with a human annotator was presented in\ednote{TODO: Luca's work}.
For alignments, this approach could be considered too invasive.

There is no reason to restrict the approach to \LaTeX\ documents.
As a case study, we have added support for HTML documents,
where the alignments are stored in HTML attributes.
Reading plain HTML in the command line interface is not ideal,
especially for complex MathML formulae.
To improve this, we have also created a browser-based interface
that closely imitates the command line interface
but can render HTML instead of showing the raw source (see \cref{fig:snify-browser}).

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{img2.png}
    \caption{Screenshot of the browser-based snify interface.
    Here, a MathML operator is being annotated.
    }\label{fig:snify-browser}
\end{figure}

In summary, we have demonstrated that the snify approach for sTeX annotation
can generalize to alignments with other concept databases
and document formats.
While the formula alignment is rather limited,
it was relatively easy to implement and could be improved in the future.
On a higher level, this shows how relatively simple and naive tools
can enable much more efficient alignment workflows,
and we hope to inspire the community to think about other tools
to help alignment authors -- be it text-to-database or database-to-database alignment.

% Document-to-database alignments are just annotations of specific
% text fragments or formula fragments in the document.
% For example, the text fragment ``natural number'' could be
% with the WikiData concept for natural numbers (Q21199).
% % For formulas, the annotation target can be less clear, for example because
% % some operations, such as the multiplication in ``$ax$'',
% % may not be represented by a glyph.
% % Depending on the document format, there there may still be
% % be an (invisible) symbol that can be annotated -- e.g.\ the unicode character for invisible multiplication.
% % An alternative is to annotate the whole sub-expression, i.e.\ the operation
% % including the operands.\ednote{these remarks break the flow}
% 
% Annotations can be created manually, e.g.\ by the authors,
% or automatically using machine learning techniques.
% Here, we focus on high-quality manual annotation,
% claiming that machine learning models are not sufficiently
% reliable, especially in research-level documents where training data is necessarily scarce.
% Presenting a user with wrong but convincing alignments
% is arguably worse than not providing any alignments at all.
% 
% The key challenge is to convice authors to actually create alignment annotations,
% as it is additional effort with little immediate benefit.
% To change the balance, we could either increase the benefit
% by providing better services based on the annotations,
% or reduce the effort by providing better annotation tools.
% Here we focus on the latter.
% 
% There is already an ecosystem of annotation tools, annotation formats
% and services based on them, developed in the KWARC group:
% \begin{compactitem}
% \item Annotation formats for \LaTeX (sTeX), HTML (FTML),
%     and, prototypically, Microsoft Word (WOIDE).
% \item Services for education (the ALeA platform)
% \item Tool support for manual annotation in sTeX
%     (snify and the sTeX IDE) and, prototypically, in Word (WOIDE).
% \end{compactitem}
% While there have been significant efforts to allow gradual adoption
% and remain as flexible as possible,
% potential users ultimately have to commit to a relatively
% complex ecosystem system of actively developed research software.
% As such, a dedicated teacher may be willing to invest time to
% move their teaching materials into this ecosystem
% to get the benefits of the ALeA learning platform,
% but a researcher writing a paper likely will not.
% 
% In this paper, we present a prototypical extension of the snify system
% for light-weight alignment independent of this ecosystem.
% The snify tool has been developed for the efficient annotation of natural language text
% with concept references and has been used to create an estimated 10,000 annotations.
% 
% Figure\ednote{TODO} shows
% 
% % Example:
% % Let $n < m$ be natural numbers.
% 
% 
% 
% % However, we can adapt some of the ideas and tools
% % for lightweight alignment of documents.
% % For example, the snify system has been designed
% % for large-scale manual creation of sTeX annotations.
% % By adapting it for alignment, we can
% % take advantage of
% 
% 
% % The focus of this paper is the alignment
% % of concept references in documents -- e.g.\ scientific papers -- with a concept database.
% % We will use WikiData as the concept database in our examples.
% % As WikiData is often referenced in other alignment databases,
% % e.g.\ MathGloss~\cite{mathgloss:arxiv},
% % this would also provide indirect alignments with other systems.
% % 
% % These document-to-database alignments are just annotations of specific
% % text fragments or formula fragments in the document.
% % For example, the text fragment ``natural number'' could be
% % with the WikiData concept for natural numbers (Q21199).
% % For formulas, the annotation target can be less clear, for example because
% % some operations, such as the multiplication in ``$ax$'',
% % are not represented by a glyph.
% % Depending on the document format, there there may still be
% % be an (invisible) symbol that can be annotated -- e.g.\ the unicode character for invisible multiplication.
% % An alternative is to annotate the whole sub-expression, i.e.\ the operation
% % including the operands.\ednote{these remarks break the flow}
% % 
% % Annotations can be created manually, e.g.\ by the authors,
% % or automatically using machine learning techniques.
% % Here, we focus on high-quality manual annotation,
% % claiming that machine learning models are not sufficiently
% % reliable, especially in research-level documents where training data is necessarily scarce.
% % Presenting a user with wrong but convincing alignments
% % is arguably worse than not providing any alignments at all.
% % 
% % The key challenge is to convice authors to actually create alignment annotations,
% % as it is additional effort with little immediate benefit.
% % To change the balance, we could either increase the benefit
% % by providing better services based on the annotations,
% % or reduce the effort by providing better annotation tools.
% % Here we focus on the latter.
% % 
% % There is already an ecosystem of annotation tools, annotation formats
% % and services based on them, developed in the KWARC group:
% % \begin{compactitem}
% % \item Annotation formats for \LaTeX (sTeX), HTML (FTML),
% %     and, prototypically, Microsoft Word (WOIDE).
% % \item Services for education (the ALeA platform)
% % \item Tool support for manual annotation in sTeX
% %     (snify and the sTeX IDE) and, prototypically, in Word (WOIDE).
% % \end{compactitem}
% % While there have been significant efforts to allow gradual adoption
% % and remain as flexible as possible,
% % potential users ultimately have to commit to a relatively
% % complex ecosystem system of actively developed research software.
% % As such, a dedicated teacher may be willing to invest time to
% % move their teaching materials into this ecosystem
% % to get the benefits of the ALeA learning platform,
% % but a researcher writing a paper will likely not.
% % 
% % However, we can adapt some of the ideas and tools
% % for lightweight alignment of documents.
% % For example, the snify system has been designed
% % for large-scale manual creation of sTeX annotations.
% % By adapting it for alignment, we can
% % take advantage of
%

\end{document}

### 2025-09-26 doc/aligments2025/doc.tex
more abstract writing

@@
 \usepackage{listings}
+
 \lstset{
@@
 learning support systems,
-faced a similar challenge, which was recently addressed by the snify system.
+faced a similar challenge, which was recently addressed by the snify system\cite{KohSch:safc25}.
 In this paper, we present a prototypical extension of the snify system
@@
 for ``conventional'' \LaTeX\ formulae
-in collaboration with a human annotator was presented in\ednote{TODO: Luca's work}.
+in collaboration with a human annotator was presented in\cite{VreWelKam:tsmmdui24}.
 For alignments, this approach could be considered too invasive.
@@
 to help alignment authors -- be it text-to-database or database-to-database alignment.
+
+% \printbibliography
+
+\bibliography{references}
 

### 2025-09-29 doc/aligments2025/doc.tex
more about alignment task

@@
 
-\title{Efficiently Linking Documents to Alignment Databases -- A Prototype-Based Exploration}
+% \title{Efficiently Linking Documents to Alignment Databases -- A Prototype-Based Exploration}
+\title{Helping Humans Align Efficiently}
 
-\author[1]{Michael Kohlhase}
-\author[2]{Jan Frederik Schaefer}
+\author[1]{Jan Frederik Schaefer}
+\author[2]{Michael Kohlhase}
 
@@
 A first step towards combining these resources is aligning their concepts.
+An alignment, for us, links a concept from one system to a concept in another system,
+indicating that they represent the same object or idea.
+We ignore the added complexity that concepts may only be similar,
+but not a precise match.
+In this paper, we will outline a workflow for supporting
+humans in creating such alignments,
+and present a prototypical extension of the snify tool\cite{KohSch:safc25}
+that implements this workflow.
 
-The focus of this paper is the alignment
-of concept references in documents -- e.g.\ scientific papers -- with a concept database.
-We will use WikiData as the concept database in our examples.
-As WikiData links to other resources (e.g.\ nLab, ProofWiki, OEIS, \ldots),
-and is also referenced in other databases
-(e.g. the MathGloss~\cite{mathgloss:arxiv} alignment database),
-% As WikiData is often referenced in other alignment databases,
-% e.g.\ MathGloss~\cite{mathgloss:arxiv},
-we also get indirect alignments with other systems.
+Concretely, we focus on the workflow where an aligner iterates over
+concepts or concept references in a \emph{source collection}
+to align them, where possible,
+with concepts from a \emph{target collection}.
+While we require that the target collection has unique concept identifiers
+-- for easy referencing --
+this is not necessary for the source collection.
+For example, we can align text fragments in a document
+by annotating them with an identifier from the target collection.
 
-As a running example, we will use the simple declarational sentence
+% The focus of this paper is the alignment
+% of concept references in documents -- e.g.\ scientific papers -- with a concept database.
+% We will use WikiData as the concept database in our examples.
+% As WikiData links to other resources (e.g.\ nLab, ProofWiki, OEIS, \ldots),
+% and is also referenced in other databases
+% (e.g. the MathGloss~\cite{mathgloss:arxiv} alignment database),
+% % As WikiData is often referenced in other alignment databases,
+% % e.g.\ MathGloss~\cite{mathgloss:arxiv},
+% we also get indirect alignments with other systems.
+
+As a running example, assume that we iterate over a scientific document
+and encounter the sentence
 \begin{quote}
@@
 where we want to align the concept reference ``natural number'' with
-the corresponding WikiData identifier.
+the corresponding WikiData concept.
 Such an alignment could, for example,
@@
 or be a first step towards formalization.
-Creating such alignments is difficult.
-For our example there are, in fact, three possible alignment targets:
+% Creating such alignments is difficult.
+In this case there are, in fact, three possible alignment targets:
 \begin{compactitem}
@@
 making the right choice sufficiently reliably,
-which puts the burden of creating alignments on human annotators --
-ideally the authors.
+which puts the burden of creating alignments on humans.
+Even if machine learning techniques are successful,
+we will need human aligners for a gold standard and
+possibly for training data.
 
-The most obvious workflow is rather inefficient:
+Without dedicated tool support,
 the author has to either recall the correct alignment target from
 memory or search for it, and then edit the document to add the alignment annotation.
-A more efficient workflow, supported by dedicated tools,
-is crucial for large-scale alignment efforts.
+The workflow is much more efficient, if a tool
+presents the user with a list of alignment candidates
+(like the three options above),
+from which the user can select the correct one.
 
-The sTeX ecosystem, which uses diverse semantic markup to drive
-learning support systems,
-faced a similar challenge, which was recently addressed by the snify system\cite{KohSch:safc25}.
+To make this work, the target collection has to provide
+an \emph{alignment catalog},
+which links each concept identifier to a human-readable description
+and a set of \emph{verbalizations} (natural language renderings of the concept).
+For example, the alignment catalog from WikiData could have the following entry (extracted from the knowledge graph):
+\begin{compactitem}
+\item \textbf{identifier:} \texttt{https://www.wikidata.org/wiki/Q28920044}
+\item \textbf{description:} \emph{positive integer} (integer greater than zero; natural number explicitly excluding zero)
+\item \textbf{verbalizations:} ``positive integer'', ``integer greater than zero'', ``natural number'', \textellipsis
+\end{compactitem}
+The description is necessary to help the user identify the correct concept
+as the identifiers are not human-readable.
+The (language-specific) verbalizations can be used to find alignment candidates.
+Note that the alignment catalog is independent of the source collection.
+While an alignment support tool requires significant customization
+for the source collection (aligning phrases in a document is
+very different from aligning concepts in an ontology),
+it can easily support different catalogs.
+
+The snify tool\cite{KohSch:safc25} implements the idea of efficient
+annotation by selection of candidates
+specifically for the case where both the source collection
+and the target collection are \LaTeX\ documents
+that use the \texttt{stex} package for semantic markup.
+Using the snify tool has significantly increased
+the efficiency of annotators, especially for novices who
+are not deeply familiar with the domain model (i.e.\ who
+do not know the identifiers by heart).
+
 In this paper, we present a prototypical extension of the snify system
-for efficient alignment that is independent of the sTeX ecosystem.
-Concretely, we have added the option to use WikiData as the concept database,
+to support alignment outside of the sTeX ecosystem.
+Concretely, we have added support for a WikiData-derived alignment catalog,
 rudimentary support for annotating formulae,
@@
 
-The snify tool is a simple command line application that
-steps through the phrases that could be annotated in a document
+% The sTeX ecosystem, which uses diverse semantic markup to drive
+% learning support systems,
+% faced a similar challenge, which was recently addressed by the snify system\cite{KohSch:safc25}.
+% In this paper, we present a prototypical extension of the snify system
+% for efficient alignment that is independent of the sTeX ecosystem.
+% Concretely, we have added the option to use WikiData as the concept database,
+% rudimentary support for annotating formulae,
+% the ability to annotate HTML documents,
+% and a browser-based user interface, which
+% helps with the annotation of HTML documents.
+
+The snify tool is implemented as a command line application.
+It steps through the phrases in a document that match verbalizations
+from the alignment catalog
 and presents the user with a list of potential targets.
+To recognize inflected forms -- e.g.\ ``continuities'' instead of ``continuity''
+-- snify uses an off-the-shelf stemmer.
 \Cref{fig:snify} shows the tool in action.
-The user can select the correct entry from the list by typing its number,
+The user can select the correct entry from the list by entering its number,
 which then creates the alignment by inserting a macro into the document
 (here e.g.\ \lstinline.\wdalign{Q28920044}{natural number}).
-Aside from choosing the correct annotation,
-snify has many other commands to skip
-phrases, undo annotations, modify the selection, etc.
+If the user does not want to create an alignment,
+there are many other commands to skip phrases,
+undo annotations, modify the selection, etc.
 
-This workflow transforms the problem from a \emph{recollection}
-problem, where the user has to remember the correct alignment target
-or otherwise spend time searching for it,
-to a \emph{selection} problem, where the user can
-pick the correct option from a list of candidates.
-In the sTeX ecosystem, the snify tool has significantly increased
-the efficiency of annotators, especially for novices who
-are not deeply familiar with the domain model (i.e.\ the concept database).
+% This workflow transforms the problem from a \emph{recollection}
+% problem, where the user has to remember the correct alignment target
+% or otherwise spend time searching for it,
+% to a \emph{selection} problem, where the user can
+% pick the correct option from a list of candidates.
+% In the sTeX ecosystem, the snify tool has significantly increased
+% the efficiency of annotators, especially for novices who
+% are not deeply familiar with the domain model (i.e.\ the concept database).
 
@@
     \includegraphics[width=0.7\textwidth]{img1.png}
-    \caption{Screenshot of snify.}\label{fig:snify}
+    \caption{Screenshot of aligning
+    \LaTeX\ documents with WikiData concepts using snify.}\label{fig:snify}
 \end{figure}
 
-In the background, snify uses a verbalization catalog
-that links concept, here WikiData identifiers,
-to their natural language renderings (verbalizations).
-WikiData has both language-specific labels and synonyms.
-For example, \lstinline.Q28920044. has the
-label ``positive integer'' and lists ``natural number''
-as a synonym.
-This verbalization catalog drives the search for potential targets
-and the selection of candidates.
-To recognize inflected forms -- e.g.\ ``continuities'' instead of ``continuity''
--- snify uses an off-the-shelf stemmer.
+% In the background, snify uses a verbalization catalog
+% that links concept, here WikiData identifiers,
+% to their natural language renderings (verbalizations).
+% WikiData has both language-specific labels and synonyms.
+% For example, \lstinline.Q28920044. has the
+% label ``positive integer'' and lists ``natural number''
+% as a synonym.
+% This verbalization catalog drives the search for potential targets
+% and the selection of candidates.
+% To recognize inflected forms -- e.g.\ ``continuities'' instead of ``continuity''
+% -- snify uses an off-the-shelf stemmer.
 
-As WikiData also has formula notations
-for many concepts, including \LaTeX\ notations like \lstinline.\mathbb{N}.,
-we can also use the same workflow to align formula fragments.
-Such formula alignment is rather limited:
+Aside from verbalizations,
+WikiData also has formula notations
+for many concepts, including \LaTeX\ notations like \lstinline.\mathbb{N}..
+This allows us to use the same workflow to align formula fragments,
+but there are significant limitations:
 operators can be aligned, but their arguments are not marked, and
@@
 
-There is no reason to restrict the approach to \LaTeX\ documents.
-As a case study, we have added support for HTML documents,
+% There is no reason to restrict the approach to \LaTeX\ documents.
+To explore other formats, we have added support for HTML documents,
 where the alignments are stored in HTML attributes.
@@
 especially for complex MathML formulae.
-To improve this, we have also created a browser-based interface
+To improve this, we have created a browser-based interface
 that closely imitates the command line interface
@@
 
-In summary, we have demonstrated that the snify approach for sTeX annotation
-can generalize to alignments with other concept databases
-and document formats.
+In summary, we have described a workflow for efficient human alignment
+based on an alignment catalog,
+and presented a prototypical extension of the snify tool
+that explores this workflow
+for the alignment of text and formulae in \LaTeX\ and HTML documents
+with WikiData concepts.
+% 
+% In summary, we have demonstrated that the snify approach for sTeX annotation
+% can generalize to alignments with other concept databases
+% and document formats.
 While the formula alignment is rather limited,
@@
 and we hope to inspire the community to think about other tools
-to help alignment authors -- be it text-to-database or database-to-database alignment.
+that could help alignment authors.
+% -- be it text-to-database or database-to-database alignment.
 

### 2025-09-29 stextools/aligner/aligner.py
some wikidata alignment UI improvements

@@ -1,0 +1,131 @@
+import dataclasses
+import textwrap
+from typing import Literal, Optional
+
+from stextools.stepper.command import CommandCollection, Command, CommandInfo, CommandOutcome
+from stextools.stepper.interface import interface
+from stextools.stepper.stepper import Stepper, State, StopStepper, Modification
+
+
+class AlignerState(State[int]):
+    def __init__(self, cursor: int, alignments: dict[str, Optional[str]]):
+        super().__init__(cursor)
+        self.alignments = alignments
+
+
+@dataclasses.dataclass
+class ConceptDbEntry:
+    identifier: str
+    label: str
+    description: str
+    description_format: Literal['tex', 'plain']
+    verbalizations: list[str]
+
+
+class AnnotateModification(Modification[AlignerState], CommandOutcome):
+    def __init__(self, concept_id: str, target_id: Optional[str]):
+        self.concept_id = concept_id
+        self.target_id = target_id
+
+    def apply(self, state: AlignerState):
+        state.alignments[self.concept_id] = self.target_id
+        state.cursor += 1
+
+    def unapply(self, state: AlignerState):
+        if self.concept_id in state.alignments:
+            del state.alignments[self.concept_id]
+        state.cursor -= 1
+
+
+class WdAnnotateCommand(Command):
+    def __init__(
+            self,
+            concept_id: str,
+            options: list[ConceptDbEntry],
+    ):
+        self.concept_id = concept_id
+        self.options = options
+
+        Command.__init__(
+            self,
+            CommandInfo(
+                pattern_presentation='𝑖',
+                pattern_regex='^[0-9]+$',
+                description_short=' annotate with 𝑖',
+                description_long='Annotates the current concept with option number 𝑖'
+            )
+        )
+
+    def standard_display(self):
+        for i, cde in enumerate(self.options):
+            prevlen = len(cde.label) + len(cde.identifier) + 2
+            description = '\n'.join(textwrap.wrap(
+                ' ' * prevlen +
+                cde.description or "no description",
+                width=80, tabsize=6,
+                ))
+            interface.write_command_info(
+                str(i), f'{cde.label} ({cde.identifier}): {description}'
+            )
+
+    def execute(self, call: str) -> list[CommandOutcome]:
+        index = int(call)
+        if index < 0 or index >= len(self.options):
+            interface.write_text(f'Invalid option number: {index}\n', style='error')
+            return []
+        selected = self.options[index]
[…]

### 2025-09-29 stextools/snify/snifystepper.py

@@ -215,7 +215,7 @@ class SnifyStepper(DocumentModifyingStepper, QuittableStepper, CursorModifyingSt
             else:
                 content = doc.get_body_content()
             interface.write_text(
-                '<div style="border: 1px solid black; padding: 5px; margin: 5px; max-height: 400px; overflow: auto;">' +
+                '<div style="border: 1px solid black; padding: 5px; margin: 5px; max-height: 40vh; overflow: auto;">' +
                  content +
                 '</div>' + r'''
                 ''',

### 2025-09-29 stextools/snify/wikidata.py

@@ -5,6 +5,7 @@ import html
 import itertools
 import logging
 import re
+import textwrap
 from typing import Sequence, Optional
 
 import orjson
@@ -55,7 +56,13 @@ class QueryFragments:
         FILTER NOT EXISTS {?item wdt:P31 wd:Q10376408 }.
     } UNION {VALUES ?item { wd:Q204 wd:Q199 wd:Q200 }}. 
     '''
-
+    # item_has_description = r'?item schema:description ?description . FILTER(LANG(?description) = "en")'
+    item_has_description = r'''
+    SERVICE wikibase:label {
+        bd:serviceParam wikibase:language "en".
+        ?item schema:description ?description .
+    }
+    '''
     item_has_unicode_notation = r'?item wdt:P913?/wdt:P487 ?notation'
     item_has_tex_notation = r'?item wdt:P913?/wdt:P1993 ?notation'
 
@@ -91,7 +98,7 @@ def _get_cached_verbs(lang: str):
 
     labels = send_query(f'''
     SELECT DISTINCT ?item ?itemLabel WHERE {{
-      {QueryFragments.item_is_math_concept}
+      {{ {QueryFragments.item_is_math_concept} }} UNION {{ {QueryFragments.item_is_math_concept_small} }} .
       {QueryFragments.item_label_in_lang(lang)}
     }}
     ''')
@@ -135,12 +142,43 @@ def _get_cached_symbols(format: str):
     ''')
     notation_data = {}
     for binding in notations['results']['bindings']:
-        notation_data[binding['item']['value']] = binding['notation']['value']
+        notation_data.setdefault(binding['item']['value'], []).append(binding['notation']['value'])
 
     with gzip.open(cache_file, 'wb') as fp:
         fp.write(orjson.dumps(notation_data))
 
     return notation_data
+
+def _get_descriptions() -> dict[str, str]:
+    CACHE_DIR.mkdir(parents=True, exist_ok=True)
+    cache_file = CACHE_DIR / f'wd_descriptions_en.json.gz'
+    if cache_file.exists():
+        with gzip.open(cache_file, 'rb') as fp:
+            return orjson.loads(fp.read())
+
+    logger.info(f'Downloading wikidata descriptions')
+
+    descriptions = send_query(f'''
+    SELECT DISTINCT ?item ?description WHERE {{
+      {{ {QueryFragments.item_is_math_concept} }} UNION {{ {QueryFragments.item_is_math_concept_small} }} .
[…]

### 2025-09-29 stextools/stepper/resources/browser_interface.js

@@ -10,7 +10,7 @@ async function fetchUpdates() {
             // TODO: snify-specific code shouldn't be here
             const target = document.getElementById("snifyhighlight")
             if (target) {
-                target.scrollIntoView({behavior: "smooth", block: "center"});
+                target.scrollIntoView({behavior: "instant", block: "center"});
             }
         } else if (element.type === 'input') {
             const input = document.createElement('input');

### 2025-09-30 doc/aligments2025/doc.tex
third abstract iteration

@@
 \usepackage{listings}
+\usepackage{tikz}
+\usetikzlibrary{shapes.geometric} 
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows}
+\usetikzlibrary{arrows.meta}
 
@@
 }
+
+\def\emph#1{\textbf{#1}}
 
@@
 Both formal and informal mathematical knowledge is fragmented across many different systems and formats.
-A first step towards combining these resources is aligning their concepts.
+Aligning the concepts across these resources
+is a key contribution towards making the data FAIR
+(Findable, Accessible, Interoperable, Reusable).
+% A first step towards combining these resources
+% and making the data FAIR (Findable, Accessible, Interoperable, Reusable)
+% is aligning their concepts.
 An alignment, for us, links a concept from one system to a concept in another system,
@@
 We ignore the added complexity that concepts may only be similar,
-but not a precise match.
-In this paper, we will outline a workflow for supporting
-humans in creating such alignments,
-and present a prototypical extension of the snify tool\cite{KohSch:safc25}
-that implements this workflow.
+but not a precise match (see \cite{10.1007/978-3-319-62075-6_7}
+for a discussion of this issue).
+Most authors and curators try to generate alignments from existing datasets
+automatically (via rule-based matching, machine learning, or LLM techniques),
+but few parallel corpora exist for mathematics, and unsupervised approaches
+seem difficult for the subtle distinctions crucial in mathematical knowledge.
+We claim that human-created alignments are necessary
+-- to ensure high quality, or at least to create datasets
+for training and evaluation.
 
-Concretely, we focus on the workflow where an aligner iterates over
-concepts or concept references in a \emph{source collection}
-to align them, where possible,
-with concepts from a \emph{target collection}.
-While we require that the target collection has unique concept identifiers
--- for easy referencing --
-this is not necessary for the source collection.
-For example, we can align text fragments in a document
-by annotating them with an identifier from the target collection.
-
-% The focus of this paper is the alignment
-% of concept references in documents -- e.g.\ scientific papers -- with a concept database.
-% We will use WikiData as the concept database in our examples.
-% As WikiData links to other resources (e.g.\ nLab, ProofWiki, OEIS, \ldots),
-% and is also referenced in other databases
-% (e.g. the MathGloss~\cite{mathgloss:arxiv} alignment database),
-% % As WikiData is often referenced in other alignment databases,
-% % e.g.\ MathGloss~\cite{mathgloss:arxiv},
-% we also get indirect alignments with other systems.
-
-As a running example, assume that we iterate over a scientific document
-and encounter the sentence
+\paragraph{Example}
+Alignments can be created between diverse resources,
+including knowledge graphs, libraries for automated theorem provers,
+and documents.
+As a running example, consider the simple sentence
+% and encounter the sentence
 \begin{quote}
@@
 provide additional information to the reader,
-could determine the prerequisite knowledge for understanding the document,
-or be a first step towards formalization.
+or determine the prerequisite knowledge for understanding the document.
+It could also be a first step towards full formalization:
+if we also align e.g.\ the type of natural numbers
+in Lean's mathlib with Wikidata,
+we have an indirect alignment between the document and mathlib.
+
 % Creating such alignments is difficult.
-In this case there are, in fact, three possible alignment targets:
+In our example there are, in fact, three possible alignment targets:
 \begin{compactitem}
@@
 \end{compactitem}
-We claim that machine learning techniques are not yet capable of
-making the right choice sufficiently reliably,
-which puts the burden of creating alignments on humans.
-Even if machine learning techniques are successful,
-we will need human aligners for a gold standard and
-possibly for training data.
+This kind of ambiguity is a major challenge for automated alignment.
+While a human aligner, especially one familiar with the document,
+may well be able to correctly disambiguate,
+they face the additional challenge of remembering what concepts exist
+or searching for them, which takes significant time and effort.
 
-Without dedicated tool support,
-the author has to either recall the correct alignment target from
-memory or search for it, and then edit the document to add the alignment annotation.
-The workflow is much more efficient, if a tool
-presents the user with a list of alignment candidates
-(like the three options above),
-from which the user can select the correct one.
+% For a human aligner,
+% there is the additional challenge of knowing which concepts exist
 
-To make this work, the target collection has to provide
-an \emph{alignment catalog},
-which links each concept identifier to a human-readable description
+% We claim that machine learning techniques are not yet capable of
+% making the right choice sufficiently reliably,
+% which puts the burden of creating alignments on humans.
+% Even if machine learning techniques are successful,
+% we will need human aligners for a gold standard and
+% possibly for training data.
+
+\paragraph{Contribution}
+In this paper,
+we propose a workflow for human alignment supported by a simple tool
+that iterates over the data to be aligned
+and presents the user with a list of alignment candidates to select from
+(see \Cref{fig:workflow}).
+% we will outline how the alignment workflow can
+% be made much more efficient with if (relatively simple) tools
+% suggest alignment candidates that the user can select from.
+For the implementation, we build on the snify architecture for selection-based,
+corpus-based semantic annotation~\cite{KohSch:safc25}
+and extend it, prototypically, to the alignment problem.
+
+\begin{figure}
+    \centering
+    \begin{tikzpicture}
+        \node[draw,rectangle] (show) at (0, 2) {Show concept (reference) + alignment candidates};
+        \node[trapezium left angle=60, trapezium right angle=-60,trapezium,draw] (command) at (0, 1) {User command};
+        \node[draw,rectangle] (create) at (-1, -1) {Create alignment};
+        \draw[-Stealth,thick] (command) to[in=100,out=210] (create);
+        \draw[-Stealth,thick] (command) to[in=50,out=260] (create);
+        \node at (-1.65, 0) {\itshape\footnotesize\begin{tabular}rSelect\;\ \\alignment\quad\ \\candidate\quad\;\ \end{tabular}};
+        \node at (0.5, 0) {\itshape\footnotesize\begin{tabular}l\quad Enter other\\\;\;alignment\\target\end{tabular}};
+        \node[draw,rectangle] (next) at (0, -2) {Go to next concept (reference)};
+        \draw[-Stealth,thick] (show) -- (command);
+        \draw[-Stealth,thick] (create) -- (next);
+        \draw[-Stealth,thick] (command) to[out=0,in=40] (next.north east);
+        \node at (2.9, 0) {\itshape\footnotesize Skip};
+        \draw[-Stealth,thick] (next) to[out=180,in=225] (show.south west);
+    \end{tikzpicture}
+    \caption{Flowchart of tool support for human aligners.}\label{fig:workflow}
+\end{figure}
+
+% and present a extension of the snify tool~\cite{KohSch:safc25}
+% that implements this workflow.
+
+\begin{figure}
+    \centering
+    \begin{tikzpicture}[xscale=3, yscale=0.7]
+        \node[circle,draw,minimum size=0.5cm] (h1) at (-1, 1.5) {};
+        \node[circle,draw,minimum size=0.5cm] (h2) at (-1, 0.5) {};
+        \node at (-1, -0.4) {$\vdots$};
+        \node[circle,draw,minimum size=0.5cm] (h3) at (-1, -1.5) {};
+        \node at (-1, -2.5) {Harvesters};
+        \node[draw, rectangle] (catalog) at (0, 0) {\begin{tabular}cConcept\\Glossary\end{tabular}};
+        \node[circle,draw,minimum size=0.5cm] (f1) at (1, 1.5) {};
+        \node[circle,draw,minimum size=0.5cm] (f2) at (1, 0.5) {};
+        \node at (1, -0.4) {$\vdots$};
+        \node[circle,draw,minimum size=0.5cm] (f3) at (1, -1.5) {};
+        \node at (1, -2.5) {Frontends};
+        \draw[-Stealth] (h1) -- (catalog);
+        \draw[-Stealth] (h2) -- (catalog);
+        \draw[-Stealth] (h3) -- (catalog);
+        \draw[-Stealth] (catalog) -- (f1);
+        \draw[-Stealth] (catalog) -- (f2);
+        \draw[-Stealth] (catalog) -- (f3);
+    \end{tikzpicture}
+    \caption{
+        Proposed architecture:
+        Custom harvesters extract a concept glossary from concept collections.
+        The concept glossary can be used by frontends customized
+        to the data we want to align.
+    }\label{fig:architecture}
+\end{figure}
+
+\paragraph{Proposed Architecture}
+To support the workflow from \Cref{fig:workflow},
+we propose an architecture centered around a \emph{concept glossary}
+(see \Cref{fig:architecture}).
+The concept glossary is extracted from the concept collection
+that we want to align with -- in our example, WikiData --
+using a custom \emph{harvester}. % -- for example, using a SPARQL query.
+The concept glossary contains for each concept
+a unique \emph{identifier}, a human-readable \emph{description},
 and a set of \emph{verbalizations} (natural language renderings of the concept).
-For example, the alignment catalog from WikiData could have the following entry (extracted from the knowledge graph):
+For example, a concept glossary extracted from WikiData could have the following entry:
 \begin{compactitem}
-\item \textbf{identifier:} \texttt{https://www.wikidata.org/wiki/Q28920044}
-\item \textbf{description:} \emph{positive integer} (integer greater than zero; natural number explicitly excluding zero)
-\item \textbf{verbalizations:} ``positive integer'', ``integer greater than zero'', ``natural number'', \textellipsis
+\item \textit{identifier}: \texttt{https://www.wikidata.org/wiki/Q28920044}
+\item \textit{description}: ``\textit{positive integer} (integer greater than zero; natural number explicitly excluding zero)``
+\item \textit{verbalizations}: ``positive integer'', ``integer greater than zero'', ``natural number'', \textellipsis
 \end{compactitem}
@@
 The (language-specific) verbalizations can be used to find alignment candidates.
-Note that the alignment catalog is independent of the source collection.
-While an alignment support tool requires significant customization
-for the source collection (aligning phrases in a document is
-very different from aligning concepts in an ontology),
-it can easily support different catalogs.
+In practice, not every concept collection has verbalizations,
+but alignments can mitigate this:
+if concept $A$ is aligned with concept $B$ and $B$ has the verbalization $v$,
+then we can also use $v$ as a verbalization for $A$.
 
-The snify tool\cite{KohSch:safc25} implements the idea of efficient
-annotation by selection of candidates
-specifically for the case where both the source collection
-and the target collection are \LaTeX\ documents
+The concept glossary is used by \emph{frontends} that enable
+the alignment workflow sketched in \Cref{fig:workflow}.
+The details depend on what we want to align:
+aligning phrases in a document is, for example,
+very different from aligning concepts in an ontology.
+Note that harvesters and frontends are largely independent
+as they are only connected via the concept glossary.
+This facilitates reusing harvesters for new frontends
+or, vice versa, reusing frontends for harvesters
+for new different concept collections.
+
+% The concept glossary is used b
+% 
+% 
+% Note that the alignment catalog is independent of the source collection.
+% While an alignment support tool requires significant customization
+% for the source collection (aligning phrases in a document is
+% very different from aligning concepts in an ontology),
+% it can easily support different catalogs.
+
+% Concretely, we focus on the workflow where a (human) aligner iterates over
+% concepts or concept references in a \emph{source collection}
+% to align them, where possible,
+% with concepts from a \emph{target collection}.
+% While we require that the target collection has unique concept identifiers
+% -- for easy referencing --
+% this is not necessary for the source collection.
+% For example, we can align text fragments in a document
+% by annotating them with an identifier from the target collection.
+
+% Without dedicated tool support,
+% the author has to either recall the correct alignment target from
+% memory or search for it, and then edit the document to add the alignment annotation.
+% The workflow is much more efficient, if a tool
+% presents the user with a list of alignment candidates
+% (like the three options above),
+% from which the user can select the correct one.
+% 
+% To make this work, the target collection has to provide
+% an \emph{alignment catalog},
+% which links each concept identifier to a human-readable description
+% and a set of \emph{verbalizations} (natural language renderings of the concept).
+
+\paragraph{snify}
+The snify tool~\cite{KohSch:safc25} implements
+a a workflow similar to the one shown in \Cref{fig:workflow}
+for semantic annotation.
+Both the harvester and the frontend act on \LaTeX\ documents
 that use the \texttt{stex} package for semantic markup.
@@
 
-In this paper, we present a prototypical extension of the snify system
+In this paper, we build on the snify architecture
 to support alignment outside of the sTeX ecosystem.
-Concretely, we have added support for a WikiData-derived alignment catalog,
-rudimentary support for annotating formulae,
-the ability to annotate HTML documents,
-and a browser-based user interface, which
-helps with the annotation of HTML documents.
+Concretely, we have added a harvester for mathematical concepts from WikiData,
+rudimentary support for aligning formulae,
+a frontend for aligning HTML documents,
+and a browser-based user interface,
+which helps with the annotation of HTML documents.
 
@@
 It steps through the phrases in a document that match verbalizations
-from the alignment catalog
+from the concept glossary
 and presents the user with a list of potential targets.
@@
 WikiData also has formula notations
-for many concepts, including \LaTeX\ notations like \lstinline.\mathbb{N}..
+for many concepts, including \LaTeX\ notations like \lstinline.\mathbb{N}.
+(in the concept glossary, we could treat them as verbalizations in another language).
 This allows us to use the same workflow to align formula fragments,
@@
 for ``conventional'' \LaTeX\ formulae
-in collaboration with a human annotator was presented in\cite{VreWelKam:tsmmdui24}.
+in collaboration with a human annotator was presented in~\cite{VreWelKam:tsmmdui24}.
 For alignments, this approach could be considered too invasive.
@@
 In summary, we have described a workflow for efficient human alignment
-based on an alignment catalog,
-and presented a prototypical extension of the snify tool
+based on an concept glossary
+and presented a prototypical implementation based on the snify tool
 that explores this workflow

### 2025-09-30 doc/aligments2025/doc.tex
minor edits

@@
 
-\paragraph{snify}
+\paragraph{Building on snify}
 The snify tool~\cite{KohSch:safc25} implements
@@
 
+\paragraph{Supporting Formulae and HTML}
 Aside from verbalizations,
@@
 
+\paragraph{Conclusion}
 In summary, we have described a workflow for efficient human alignment
